---
title: "Breeding bird indicators"
subtitle: |
  [NO_BIRD_001] (farmland and open lowland) \
  [NO_BIRD_002] (mountains) \
  [NO_BIRD_003] (wetlands) \
  [NO_BIRD_004] (forests)
indicatorID: 
  - NO_BIRD_001
  - NO_BIRD_002
  - NO_BIRD_003
  - NO_BIRD_004
indicatorName: 
  - Breeding birds
folderName: NO_BIRD_001-004
country: country
continent: continent
ECT: B1 - Compositional State Characteristics
Realm: 
  - Terrestrial (T)
  - Terrestrial (T)
  - Freshwater-Terrestrial (TF)
  - Terrestrial (T)
Biome: 
  - T3 Shrublands and shrubby woodlands biome; T4 Savannas and grasslands biome; T7 Intensive land-use biome
  - T3 Shrublands and shrubby woodlands biome; T6 Polar-alpine (cryogenic) biome
  - TF1 Palustrine wetlands biome
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - ''
yearAdded: 2025 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2025
status: complete # options are complete, incomplete, or deprecated
Version: "000.003" 
VersionComment: Similar to 000.002, but with integrated metadata and open science badges.
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BIRD_001-004
Normalised: No
SpatialAggregationPathway: Aggregate
AuthorList: Nater, C.R., Pavón-Jordán, D., Kolstad, A.
data_availability: gold
code_reproducibility: silver
open_science_badge: silver
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Chloé R. Nater
    email: chloe.nater@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Diego Pavón-Jordán
    affiliations:
      - ref: myID
  - name: Anders L. Kolstad
    email: anders.kolstad@nina.no  # Enter email
    affiliations:
      - ref: myID
date: February 3, 2025
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
bibliography: references.bib
hide: FALSE
---

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(odbc)
library(rtrim)
library(data.table)
knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
source(here::here("_common.R"))
source(here::here("extract_yaml.R"))
```

::: {layout-ncol="5"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]),
  folder_name = folder_name)
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```
::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::


::: {.callout-tip collapse="true"}
## Logg

<!--# Update this logg with short messages for each update -->

-   Oct. 2023 - Complete workflow [described](https://ninanor.github.io/ecosystemCondition/breeding-birds.html)
-   3 Feb. 2025 - Started restructuring documentation for ecRxiv and updating indicator values
-   24 Nov. 2025 - Restrukted to the new template and adding open science badges (ALK)
:::

<hr />

## 1. Summary

Birds are an important component of biodiversity and are well known to both the general public and to scientists. Birds also perform important ecosystem functions and respond to natural and anthropogenic pressures, such as land-use change, climate change, and changes in food supply (esp. insects and seeds). This makes bird population sizes a useful metric for monitoring ecosystem condition. The Norwegian bird monitoring program *Norsk Hekkefulgovervåking* (previously named TOV-E) has been running since 2007 (with broad national coverage since about 2010) and covers the entire country with about 500 sampling locations and covering all ecosystem. We use this data to create four thematic multi-species bird indices, all using the same methodological approach. Using set criteria, bird species are categorised as belonging to one and only one main ecosystem. For each of our four ecosystem we extract monitoring data for these habitat specific species, and as long as there is sufficient data for that species we estimate a temporal trend for the change is population size. This means we are excluding rare and geographically restricted species, and focus on relatively common species. The trends are centered on a baseline year and all results are reported as percentage change relative to this year. We use 2010 as the reference year because at that time the bird monitoring had good national coverage in Norway. After we combine all the species for the same ecosystem into a muli-species index (MSI). The MSI should idealy be interpretted as time series, i.e. by looking at line graphs, but for reporting in standardised ecosystem accounting tables we have chosen to represent each three year accounting period by taking the mean of the MSI for those three years. The value represents a change relative to year 2010.

## 2. About the underlying data {#sec-aboutdata}

These indicators are based on a shared data set originating from the Norwegian Breeding Birds Monitoring Program ([*Hekkefulgovervåkingen*](https://hekkefuglovervakingen.nina.no/), previously TOV-E) [@kålås2022]. This is a national bird census monitoring program, origination gradually from 2005 to about 2011, with 492 sampling locations thought the country, each with 12 - 20 sampling points. About 80% of the sampling locations are visited yearly. The data are gathered mostly by experienced volunteer bird-watchers following a [standardized sampling protocol](https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf). Observers will visit each route once a year between 20. May and 10. June (depending on latitude) and carry out the census in good weather conditions. The sampling locations are selected randomly withing each of five regions: east, south, west, cnetral, and northern Norway. Note that becaus eof the standardised way that the R scripts for the TRIM analyses are made, we have not been able to account for this spatial structure in the data, for example by doing some post-stratification.

The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route (see below for further information on the sampling design). A total of 229 species are typically recorded during this census. After 5 minutes, the observer walks to the next point in a straight line (when possible). During this 'transect' the observer is requested to record also a subset of species that are considered rare or difficult to capture during the 5-minute point counts (n = 121).

The data are entered into a [database](https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx) and quality-checked by NINA. Subsequently, the raw data are processed using a combination of country-specific wrangling routines and internationally standardized scripts distributed by PECBMS, the [Pan European Bird Monitoring Scheme](https://pecbms.info/), which together make up the now integrated workflow below:

TRIM: [TRends and Indices for Monitoring data (TRIM)](https://pecbms.info/methods/software/trim/) is a freeware program developed by Statistics Netherlands in the framework of wildlife statistics

RTRIM: A set of three R script used for doing TRIM analyses. Versions exist for execution in shell and in R. See [here](https://pecbms.info/methods/software/trim/)

RSWAN: A tool for combining TRIM output data.

MSI: Multi-species index

### 2.1 Spatial and temporal resolution and extent

*Norsk Hekkefuglovervåking* is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats along both latitudinal and altitudinal gradients present in Norway. Each route has between 12 and 20 counting points (ca. 9000 across Norway), which typically are 300 m apart, describing a square with side 1.5 km long (the observer walks on a straight line between two consecutive points). Deviations of this typical route shape are normally due to the landscape configuration (e.g. rough terrain, presence of a lake or river, sea, glacier, etc.). Due to limited resources, only a random sample of 370 routes are surveyed every year with an aim to survey at least 80% of the routes yearly.

The bird monitoring program was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010. For many species, data are already of good coverage and high quality starting in 2008. The monitoring scheme is still ongoing as of 2025.

A predecessor monitoring scheme existed prior to the establishment of “hekkefuglovervåkingen” and led to the collection of decent data in the period from 1996 up to and including 2008. Although the predecessor monitoring programmes and “hekkefuglovervåkingen” are not the same, the overlap for the years 2007 and 2008 makes it possible to combine abundance indices from across time using tools like “RSWAN”, which is provided by PECBMS. For a subset of 55 species, longer time-series of abundance indices (1996-2023, as of now) are therefore available. We have, however, not included this data in the current version of this condition variable.

As per now, the breeding bird indicators are calculated at the national scale only and the data are therefore not linked to any geofiles here. We note that some species may be sufficiently common and have enough data to allow for calculating indicator values for two major regions, Northern Norway and Southern Norway, but this requires careful evaluation that needs to be planned and organized as part of a future project.

### 2.2 Original units

The observers in the field record “number of individuals” seen or heard during the census. Those values are then converted into “number of pairs” by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only). In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of change relative to the population in a given baseline year (in Norway this is typically 2008). The abundance indices then form the basis of further analyses and aggregation routines.

### 2.3 Additional comments about the dataset

NINA performs quality-checks of the data delivered by the observers to ensure that the new data entered to the database are reliable. In addition, NINA completes the annual dataset by entering the zeroes (undetected species) at surveyed routes and -1s (for not surveyed routes). This ensures that all species have data (either a positive count, a zero or a -1) for each route and year.

From all species recorded within the *Norsk Hekkefuglovervåking* monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses. Abundance indices of 71 of those species, representing those with wide distributions and high continental-scale relevance, are reported annually to PECBMS. Longer (combined) time series can be calculated for 55 species, and 54 of those serve as indicators in the Norwegian Nature Index [@jakobsson2020]. The established breeding bird indicators (condition variables in SEEA EA terminology) contain varying numbers of species: the farmland bird index has 7 species, the mountain bird index has 8 species, and the forest bird index has 24 species. For the new index for wetland birds, we used 7 out of the 8 species suggested for this purpose [@husby2011]. The one species we dropped was the Common Crane *(Grus grus)* as it has seen strong recent population increases that are linked to anthropogenic impacts unrelated to the status of their wetland breeding habitats (Bård Stokke, pers. communication).

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

B1 – Compositional state characteristics.

### 3.2 Ecosystem condition characteristic

The metric is aimed at describing a characteristic of ecosystems in good condition as having high and stable bird populations.

### 3.3 Other standards

The metric is categorised under *biologisk mangfold* in the Norwegian classification system [@nybø2017]. It is important to note though that the indicators represent species abundance (trends) in this context and are **not** indicators of species richness or species distribution.

### 3.4 Collinearities with other indicators

None that we consider very obvious or problematic, excep if one chose to produce a parallel bird index based on individual bird indicators from the Norwegian Nature Index (see @sec-rl).

### 3.5 Impact factors

Birds are affected by a large range of both natural and anthropogenic stressors, and what stressors are most important will vary from species to species. In general, across all species, and of the human caused impact factors, we can mention, climate change, insect declines, and land-use change (e.g. habitat loss, reduced connectivity, edge effects).

## 4. Reference condition and levels

### 4.1 Reference condition

In Norway, the reference condition for natural ecosystems have been one of largely intact ecosystems without large-scale human modifications [@nybø2017]. For semi-natural ecosystems, human influence is allowed, and indeed a prerequisite, but then in the sense of appropriate, stable and traditional land-use practices. Fro agricultural ecosystems we lack a precise definition for a reference condition. The metrics are not normalised, and therefore there is no direct way to compare the values to a reference condition.

### 4.2 Reference levels {#sec-rl}

None. Note that the use of a baseline year to center the trend analyses is not equivalent to the use of reference levels, and do not allow quantitative statement of the type *this value is very high/low.* Note also that reference levels for individual bird species has been set in the Norwegian Nature Index, and it is possible to compile similar normalised bird indices by combining these single bird indicators [see @nybø2023 page 55] .

## 5. Uncertainties

One source of uncertainty comes from the assumption that multi-species abundance indices use a representative selection of species sampled in a representative set of habitats. However, due to resource- and logistic constraints, that is often not the case and non-random sampling along, for example, rural-urban gradients may bias breeding bird indices derived from data from "hekkefuglovervåkingen" [@husby2020].

## 6. References

::: {#refs}
:::

## 7. Datasets

### 7.1 Dataset A - Norsk Hekkefulgovervåking {#sec-data1}

Data is downloaded from the [internal NINA database](https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx?ReturnUrl=/Fugl/). See @sec-aboutdata for details on the dataset.

## 8. Spatial units

The sampling locations are selected with a regional stratification, but we do not try to account for this spatial structure here (see @sec-aboutdata for discussion). Therefore the only spatial unit we rely on the the national scale (the ecosystem accounting area).

## 9. Analyses

### Workflow setup {#sec-workflow}

The code for the following analyses has been developed in a separate GitHub repository: https://github.com/NINAnor/birdIndicators As a first step, we therefore need to read in all the functions from that repository. The repository has not been converted into a package yet, so we need to download the entire repository and source all function files in the “R” folder manually:

```{r}
#| code-summary: "Download functions from GitHub repository"
#| output: false
temppath <- here::here("indicators/NO_BIRD_001-004/temp")
destpath <- paste0(temppath, "/birdIndicators_Repo.zip")

if(!file.exists(destpath)) {
  download.file(url = "https://github.com/NINAnor/birdIndicators/archive/refs/heads/main.zip",
               destfile = destpath)
  unzip(zipfile = destpath,
       exdir = temppath)
}

sourceDir <- function(path, trace = TRUE, ...) {
  for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
    if(trace) cat(nm,":")
    source(file.path(path, nm), ...)
    if(trace) cat("\n")
  }
  invisible()
}
sourceDir(paste0(temppath, "/birdIndicators-main/R"))
```

Next, we need to define a range of relative and absolute directories. The main reason for that is that many of the scripts delivered by PECBMS function by iteratively writing and reading files in a specific folder structure. We use `here::here()` to go from relative to absolute paths. While all code would ideally be rewritten to work with relative paths and without toggling working directories, PECBMS tends to distribute new versions of their scripts periodically and with no formal version control in place. This means that it is imperative to leave their core code as untouched as possible to facilitate including new versions (by copy paste) as they come.

```{r}
#| code-summary: "Create local directories"
## Absolute path for root folder:
wd <- here::here("indicators/NO_BIRD_001-004")

# Storage of raw PECBMS Trim outputs
folder <- paste0(wd, "/PECBMS_Files")

# Storage of processed PECBMS Trim outputs
subFolderName <- "Species_files"

# general and working folders
general_folder_rel <- paste0(wd, "/temp/birdIndicators-main/data")
working_folder_rel <- paste0(folder, "/", subFolderName)

# Storage of MSI results
MSI_results_folder <- paste0(wd, "/MSI_Results")

# NINAs local storage of legacy Trim output files
legacyFile_folder <- "P:/41201612_naturindeks_2021_2023_database_og_innsynslosning/Hekkefugl_Dataflyt/LegacyFiles_PECBMS_Trim"

# Absolute data and output paths for RSWAN (not needed)
# general_folder <- general_folder_rel
# working_folder <- working_folder_rel
# output_folder <- paste0(working_folder, "/output/") 
# output_folder2 <- paste0(working_folder, "/output") 


## Create directories if they are not available
dir.list <- c(
  folder, 
  general_folder_rel, 
  working_folder_rel
  )

for(i in 1:length(dir.list)){
  if(!file.exists(dir.list[i])){ 
  dir.create(dir.list[i])
  }
}

```

### Data sets

The workflow uses the up-to-date raw data from the Norwegian bird monitoring. One can also combine this with the calculated abundance indices from the predecessor monitoring scheme in Norway, and that way get a longer time series, but we will not do that here. Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow as the different outcomes require different information.

### Monitoring data {#sec-download}

The route-based raw data from “hekkefuglovervåking” are compiled after each breeding season and curated and stored in NINA’s database. Before storing new data in the database, a thorough quality-check is performed. From that database we download the data relevant for the [PECBMS Trim analyses](https://pecbms.info/methods/software/trim/) and for the specified range of years (here from 2008 up to and including 2024). The function `downloadData_TRIM` does that for a specified year range (from minYear up to and including maxYear), and requires specifying which driver is used for accessing the database (via argument DriverName). When running from the NINA RStudio server, we need to use “FreeTDS”. When running locally from NINA computers, we need to select “SQL Server”. Note that downloading these data requires user-specific access privileges which can be requested from NINA IT department (‘datahjelp’).

```{r}
#| eval: false
#| code-summary: "Download Trim data from NINA database"
minYear <- 2008
maxYear <- 2024

Trim_data <- downloadData_TRIM(
  minYear = minYear, 
  maxYear = maxYear,
  drop_negativeSpp = TRUE, 
  # DriverName = "FreeTDS"
  DriverName = "SQL Server"
  )
# save temp file
saveRDS(Trim_data, file = paste0(wd, "/temp/Trim_data.rds"))
```

```{r}
#| include: false
Trim_data <- readRDS(paste0(wd, "/temp/Trim_data.rds"))
```

### Lists of species information

The species list files should have been downloaded together with the rest of the birdIndicators repository (@sec-workflow) and we can use `makeSpeciesLists()` to assemble the information.

```{r}
#| eval: false	
#| code-summary: "Assemble species lists"
sppLists <- makeSpeciesLists(
  Trim_data = Trim_data, 
  spp_DataPath = paste0(wd, "/temp/birdIndicators-main/data"))
Spp_selection <- sppLists$sppData

# save data
saveRDS(sppLists, file = paste0(wd, "/data/sppLists.rds"))
saveRDS(Spp_selection, file = paste0(wd, "/data/Spp_selection.rds"))
```

```{r}
#| echo: false
sppLists <- readRDS(paste0(wd, "/data/sppLists.rds"))
Spp_selection <- readRDS(paste0(wd, "/data/Spp_selection.rds"))
```

The table `Spp_selection` contains information for each species on which reporting routes should be included in. The multi-species indices are grouped into a single category (“MSI”) here, but we can extract the species that go into the MSI for each ecosystem via the additionally provided lists.

::: {layout-ncol="2"}
Farmland MSI species:

```{r}
#| echo: false
#| results: asis
temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_farmland)])
cat(paste0("- ", temp), sep = "\n")
```

Forest MSI species:

```{r}
#| echo: false
#| results: asis
temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_forest)])
cat(paste0("- ", temp), sep = "\n")
```

Mountain MSI species:

```{r}
#| echo: false
#| results: asis
temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_mountain)])
cat(paste0("- ", temp), sep = "\n")
```

Wetlands MSI species:

```{r}
#| echo: false
#| results: asis
temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_wetlands)])
cat(paste0("- ", temp), sep = "\n")
```
:::

### PECBMS RTRIM analyses

In the next step, the downloaded raw data (see @sec-download) is analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data" [@R-rtrim2020]). The RTRIM analyses are run for a subset of 85 species which have sufficient data for it using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/). For the application here, we have written wrapper function around the PECBMS code for each step of the RTRIM workflow.

```{r}
#| eval: false
#| code-summary: "TRIM analyses"

# Write PECBMS arguments input files for each species
# These are called *_arg_input_stratum under PECBMS_Files
argument_file <- 
  setupInputFiles_PECBMS_trimShell(
    Spp_selection = Spp_selection,
    folderPath = folder)

## Subset data to contain only relevant species
PECBMS_data <- 
  makeInputData_PECBMS(
    Trim_data = Trim_data,
    Spp_selection = Spp_selection,
    convertNA = TRUE, 
    save_allSppData = TRUE, 
    returnData = TRUE,
    wd = wd)

## Run analyses using the PECBMS Rtrim shell
runRtrimShell_PECBMS(folder = folder)

## Process results
# Returns file called all_Indices_All_Trends in PECBMS_Files
trimResults_PECBMS <- processRtrimOutput_PECBMS(folder = folder)
```

The results of the the RTRIM workflow are a series of .csv and .txt files. Here we copied hem into a designated folder:

```{r}
#| eval: false
#| code-summary: "Copy and rename files files"
## Collect (and rename) species and summary files
collectSpeciesFiles_PECBMS(
  folder = folder, 
  subFolderName = subFolderName)
```

### Aggregation to multi-species indices (MSIs)

The breeding bird indicators for ecosystem condition accounts / assessments are multi-species indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.

The function for calculating MSIs requires specifying whether or not to make use of combined time series generated by RSWAN (argument `useCombTS`), which year to use as a base/reference (argument `baseYear`), and which year to use as a changepoint (argument `changepointYear`).

We are not including data from the bird monitoring that was in effect before the current *Hekkefulgovervåkinga*, and therefore we do not need to use the RSWAM method to combine time series.

We set the base year to be 2010 to coincide with the time when the bird monitoring became truly [national in coverage](https://ninanor.github.io/ecosystemCondition/breeding-birds.html#temp-coverage-bb). The baseline year only works to center the time series around a fixed time. The relative change in the time series remains unchanged. The use of arbitrary baseline years (arbitrary in the sense that they do not represent a reference state in the SEEA EA use of the term, or any other necessarily desirable value) means that point estimates should really be interpreted with caution when seen in isolation. It is the development of the trends in the time series that is important to understand and convey. But having a spatially representative baseline means there is less chance of a spatial bias in the sampling to affect the absolute level of the time time series, but as we have explained, the temporal trends are unaffected.

The choice of change point year does not affect index values, but specifies until and from which year additional, shorter term trends should be calculated (i.e. if `changepointYear = 2008`, the procedure will additionally provide trend estimates from starting year to 2008, and from 2008 to the last year in the data). We set this to 2022 below because we have to set something, but we will not investigate this part of the analyses any further.

```{r}
#| code-fold: false
useCombTS <- FALSE
baseYear <- 2010
changepointYear <- 2022
```

The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument). We list these for all four ecosystem indices:

```{r, eval = FALSE}
IndexNames <- c("FarmlandBirds", "ForestBirds", "MountainBirds", "WetlandBirds")
Index_sppLists <- list(sppLists$sppLists$MSI_farmland,
                       sppLists$sppLists$MSI_forest,
                       sppLists$sppLists$MSI_mountain,
                       sppLists$sppLists$MSI_wetlands)
```

We will calculate all four indices starting from whenever the data starts (2008) and using 2010 as the reference year.

```{r}
#| eval: false
#| code-summary: "Calculate MSI values"
MSI_longTerm <- data.frame()

for(i in 1:4){
  message(paste0(crayon::bold("Calculating ", IndexNames[i], " Index...")))
  results <- calculateIndex_MultiSpecies(working_folder = working_folder_rel, 
                                         Spp_subset = Index_sppLists[[i]], 
                                         IndexName = IndexNames[i], 
                                         results_folder = MSI_results_folder,
                                         useCombTS = useCombTS,
                                         baseYear = baseYear,
                                         changepointYear = changepointYear)
  MSI_longTerm <- rbind(MSI_longTerm, results)
  message("")
}
# save output
saveRDS(MSI_longTerm, file = paste0(wd, "/data/MSI_results.rds"))
```

```{r, echo = FALSE}
MSI_results <- readRDS(paste0(wd,"/data/MSI_results.rds"))
```

The MSI is provided both as point estimates (`MSI`) with associated standard error (`sd_MSI`) and confidence interval (`lower_CL_MSI` and `upper_CL_MSI`), and as a smoothed trend model fit (characterised by an average trend, `Trend`, and an associated confidence interval, `lower_CL_trend` and `upper_CL_trend`):

Additionally, the results are saved as a standardized set of files used for reporting to PECBMS.

We can now visualize the results both as point estimates and trend fit:

```{r fig-MSI2015}
#| fig-cap: 'Time series of the multispecies indices (MSIs) for the all ecosystems. The trend is centered to equal 100% in year 2015.'

myplotTimeSeries_MSI <- function(MSI, results_folder){
  
  ## Define custom colors
  plotCols <- colorspace::divergingx_hcl(palette = "Temps", n = length(unique(MSI$IndexName)))
  
  p2 <- ggplot(MSI, aes(x = year, y = MSI)) + 
    geom_point(aes(colour = IndexName), size = 3) + 
    geom_pointrange(aes(ymin = MSI - sd_MSI, ymax = MSI + sd_MSI, colour = IndexName)) + 
    geom_line(aes(y = Trend, colour = IndexName), linetype = "dashed") +
    geom_ribbon(aes(ymin = lower_CL_trend, ymax = upper_CL_trend, fill = IndexName), alpha = 0.2) + 
    scale_x_continuous(name = "Year", breaks = min(MSI$year):max(MSI$year)) + 
    scale_color_manual(name = "Index", values = plotCols) +
    scale_fill_manual(name = "Index", values = plotCols) + 
    ylab("MSI Value (% of baseline)") + 
    theme_bw() + 
    theme(panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1)) 
  
    print(p2)
  }
  


myplotTimeSeries_MSI(
  MSI = MSI_results,
  results_folder = MSI_results_folder
                   )
```

### Choise of metric to represent accounting period

The MSI should ideally be interpreted as time series (e.g. from looking at @fig-MSI2015). However, ecosystem accounting tables are limited in how they can convey temporal trends. Usually this is limited to a starting and closing value. The value for a gives year (point estimates) can fluctuate due to natural variation (e.g. weather events) and may not reflect any long or medium term trend in ecosystem condition. The MSI trend is better at smoothing out these interannual trends. However, it may be too good at smoothing out these trends, and we want to avoid data from previous accounting periods to affect subsequent periods. Our solution is to use the mean of the MSI point estimates from the years inside each accounting period (three year periods). Errors are aggregated by simulating values for each time step. assuming a normal distribution, and sampling equally from each. We must stress that the absolute variable value for an accounting period is only interprettable as a short term trend in the data, and to not indicate whether this value is especial high or low compared to historical values or any desirable state.

> We must stress that the absolute variable value for an accounting period is only interprettable as a short term trend in the data, and to not indicate whether this value is especial high or low compared to historical values or any desirable state.

These bird indices are variables in the SEEA EA terminology and should be reported in the variable account.

## 10. Results

```{r}
#| message: false
output <- MSI_results |>
  
# errors are 0 in 2015 so we will not use that year, or any year before, at least not for this excercise
  filter(year > 2015) |> 

# creating three accounting periods
  mutate(accounting_period_final_year = case_when(
    year <= 2018 ~ 2018,
    year <= 2021 ~ 2021,
    year <= 2024 ~ 2024,
    .default = NA
      )) |>
  mutate(accounting_period = case_when(
    year <= 2018 ~ "2016-2018",
    year <= 2021 ~ "2019-2021",
    year <= 2024 ~ "2022-2024",
    .default = NA
      )) |>

# create probability distributions for each point estimate, 
# assuming a normal distribution
  rowwise() |>
  mutate(
    dist = list(rnorm(10000, MSI, sd_MSI))) |>

# sample 9000 observations randomly across the three years 
# inside each accountin period to get a shared distribution
  group_by(IndexName, accounting_period_final_year, accounting_period) |>
    reframe(
      variableValue = mean(MSI),
      dist = list(sample(
        unlist(dist),
        size = 9000,
        replace = TRUE))) |>

# get the 25th and 75th percentile from this shared distribution
  ungroup() |>
  rowwise() |>
  mutate(
    "75th percentile" = quantile(unlist(dist), 0.75),    # ERROR 
    "25th percentile" = quantile(unlist(dist), 0.25), 
  
 # Add some variables for the final table 
    ID = case_when(
      IndexName == "FarmlandBirds" ~ "NO_BIRD_001",
      IndexName == "ForestBirds" ~ "NO_BIRD_004",
      IndexName == "MountainBirds" ~ "NO_BIRD_002",
      IndexName == "WetlandBirds" ~ "NO_BIRD_003"
    ),
    ecosystem = case_when(
      IndexName == "FarmlandBirds" ~ "Farmland & Open lowland",
      IndexName == "ForestBirds" ~ "Forests",
      IndexName == "MountainBirds" ~ "Alpine",
      IndexName == "WetlandBirds" ~ "Wetlands"
    )) |>
  add_column(
    ECT = "B1",
    "Accounting area" = "Norway") |>

# Select which columns to show, and set their order
  select(
    Indicator = IndexName, 
    ID, 
    ECT, 
    "Accounting area",
    "Accounting period" = accounting_period,
    "Variable value" = variableValue,  
    "25th percentile",  
    "75th percentile")

# print table
output |>
    kbl()

```

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::