---
title: "ABI - land-use intensity index"
subtitle: |
  [NO_ABIN_001] (Mountains)
indicatorID: 
  - NO_ABIN_001 
indicatorName: 
  - Land-use intensity
verbatimName: 
  - Arealbruksintensitet
folderName: NO_ABIN_001
country: Norway
continent: Europe
ECT: C1
Realm: 
  - Terrestrial (T)
Biome: 
  - T3 Shrublands and shrubby woodlands biome
  - T6 Polar-alpine (cryogenic) biome
Ecosystem: 
  - NA
verbatimEcosystem: Fjell
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: This is a pre-released version of this workflow and it has not yet been code-reviewed. 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_ABIN_001
Normalised: No
SpatialAggregationPathway: Aggregate (geographic regions)
Precision: No precision estimates are reported
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: false
author:
  - name: Anders Kolstad
    email: anders.kolstad@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research
AuthorList: Kolstad, A.
date: September 4, 2026
data_availability: bronze
code_reproducibility: silver
open_science_badge: bronze
callout-icon: false
lightbox: true
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
bibliography: references.bib
codeReviewers:
  - name: Pre-published version - No code review performed
    email: jane.smith@example.no
    affiliations:
      - id: reviewerInst
        name: Example Institute
    reviewType: code review
    github: janesmith
---

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(terra)
library(stars)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 04 September 2026 - Started working
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
Ecosystems and their species communities are dependent on movement of individuals and genes in order to maintain the ecological integity and functioning. Fragmentation reduces the ability of some species to move through the landscape, and this fragmentation can take the form of both physical barriers as well as any disturbance that alterns species behavours i was that makes some areas less accessible. How sensitive the species are to fragmentation varies widely. Generally, large mammals are the most susceptible group. Ecosystem condition indicators should ideally not represent pressures or drivers directly, but in the absence of good and religable data on landskape connectivity, fragmentation, asa condition variable, is used here as a proxy.

2. **Methods**:
The underlying dataset is called land-use intensity index, or ABI in the Norwegian abbreviation, and is an index representing the amount of technical infrastructur within an area. The index values are calcualted for all 25 m grid cells inside the Norwegian mainland. The types of infrastructur that affect the index are: physical construction, arificial surfaces and 'other' human impact. This last category include regulated water bodies, tractor road and agricultural lands. The data is available for the year 2017 and 2025. The data is masked to include only mountain areas, and variable is then taking the mean value for each of the five geographic regions in Norway, as well as for all of Norway

3. **Interpretation**:
High values represenet higher presence of human infrastructure and land-use related disturbances, than smaller values do. The absolute value of the variable can be compared between regions. Within a region, the development over time tells us if human-cause fragmentation is increasing or decreasing in that area.

4. **Results and development needs**:
The land use intensity index in Norwegian mountains has risen slightly between 2017 and 2025, but it is difficult to asses how severe this trend is. The index values are highest in western Norway, and lowest in central Norway. The workflow and cpondition variable would benefit from better temporal resolution, and a copmparison against other ecosystems.
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->



## 2. About the underlying data

This workflow used the land use intensity index for Norway [@erikstad_index_2023]. In Norwegian it is called *arealbruksintensitet*; hance the abbreviation ABI. The data is slightly different than the version reported in Erikstad et al. [-@erikstad_index_2023]. It is range-normalised to values between 0 and 1. It also includes in its calculations the presence of agriculture and regulated lakes. 


<!--# Give a general introduction the datasets you have used, and provide more specific details in the sub-chapters.  -->

### 2.1 Spatial and temporal resolution and extent

The data is available from 2017 and 2025 but there is considerable time lag in how the raw, underlying data is updates.

### 2.2 Sampling design

The sampling is complete, wall-to-wall, but to speed up calculations I have resampled a certain number of raster cells to calculate area statistics from.

### 2.3 Original units

The units are arbitrary.

### 2.4 Instructions for citing, using and accessing data

the method and data can be sited as Erikstad et al. [-@erikstad_index_2023]. The data will be made available upon request.

### 2.5 Additional comments about the dataset

Many version exist of this dataset, with different resolutions and different human interventions being included in the index calculations.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)
This is an ecosystem condition variable or a proxy of such, and belong in  the C1 - landscape characteristics class.

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->



### 3.2 Ecosystem condition characteristic

The variable represents landscape fragmentation, or i works as a coarse proxy for landscape connectivity.
<!--# 

A short (one to three word) tag or name for the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.

Examples: 'intact hydrology'; 'near-natural species composition'; 'soil stability'; sustainable trophic interactions'. 

The term 'characteristic' is used in a similar way  to the term 'criteria' in Multiple Criteria Decision Making.  

-->

### 3.3 Collinearities with other indicators

None.

### 3.4 Impact factors

The variable (index) is sensitive to any human infrastructure or activity that is captured in the index, such as roads, buildings and agriculture.

### 3.5 Spatial aggregation
The data is aggregated with an arithmetic mean for each region.

## 4. Reference condition and levels

### 4.1 Reference condition
Not relevant
<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels
Not relevant
<!--# 

If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ refers to the reference level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.

 -->


#### 4.2.1 Spatial resolution and validity
Iinterprettable at the scale of geographic regions (as is reported).
<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->

## 5. Uncertainties
### 5.1 General uncertainties
Not assessed.
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

### 5.2 Reported uncertainties
No uncertainties are reported for this variable.
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->



## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

## 7. Import and prepare datasets
Importing the ABI data, an ecosystem delineatoiun mask, and a deliminatior file for the five geographical regions.

```{r paths}
if (grepl("/data/", getwd())){
  root <- "/data/P-Prosjekter2/172520_utvikling_av_forstyrrelsesindeks/ABI/Jenny/ABI/output/abi_exports/"
}else{
  root <-  "P:/172520_utvikling_av_forstyrrelsesindeks/ABI/Jenny/ABI/output/abi_exports/"
}

pathReg <- here::here("data/regions.rds")
path17 <- paste0(root, "2017/ABI_2017_enkel_v1_norm.tif")
path25 <- paste0(root, "2025/ABI_2025_enkel_v1_norm.tif")
pathMask <- here::here("data/fjellmasken.tif")

```

```{r readData}
#| eval: false
reg <- readRDS(pathReg)
data17t <- terra::rast(path17)
data25t <- terra::rast(path25)
theMask <- terra::rast(pathMask)
```


## 8. Spatial units

The data is a wall-to-wall raster og 25m resolution, but i resample it to make it manageable, but still representative.

## 9. Analyses

Workflow:

1. Aligning rasters and masking the ABI data against an ecosystem mask
2. Sampling the masked ABI data and retaining 1 mill cells
3. Vectorize sampled data
4. Intersect with geographic region dataset
5. Summarise
6. Repeat for the other time step and combine 

```{r mainAnalyses}
#| eval: false

# align the data
theMask2 <- resample(
  theMask,
  data17t,
  method = "near"
)
writeRaster(theMask2, here::here("data/theMask2.tif"), overwrite = TRUE)


# mask against ecosystem 
data17t_masked <- mask(
  data17t,
  theMask2,
  maskvalues = 0
)

writeRaster(data17t_masked, here::here("data/data17t_masked.tif"), overwrite = TRUE)

data17t_masked <- terra::rast(here::here("data/data17t_masked.tif"))

# draw 500k random samples
# This takes 20 ish minutes
samp <- spatSample(
  data17t_masked,
  size = 5e5,
  method = "random",
  na.rm = TRUE,
  xy = TRUE,
  as.df = TRUE
)

# make into points
samp_sf <- sf::st_as_sf(
  samp,
  coords = c("x", "y"),
  crs = crs(data17t_masked)
)

# join with region data
samp_sf <- sf::st_join(
  samp_sf,
  reg["region"],
  left = TRUE
)

# drop geometry to speed up calculations
samp_df <- samp_sf |>
  sf::st_drop_geometry() |>
  tidyr::drop_na()

# check sample size
samp_df |>
  count(region)

saveRDS(samp_df, here::here("data/samp_df.rds"))

# It's maybe a bit low for southern Norway with < 20k points
names(samp_df)
samp_tbl <- samp_df |>
  rename(ABI = file127fc4846f8db) |>
  group_by(region) |>
  summarise(mean_2017 = mean(ABI))

samp_tblNat <- samp_df |>
  rename(ABI = file127fc4846f8db) |>
  summarise(mean_2017 = mean(ABI)) |>
  add_column(region = "Norge") |>
  bind_rows(samp_tbl)

saveRDS(samp_tblNat, here::here("data/samp_tbl.rds"))

# same for 2025
theMask2 <- resample(
  theMask,
  data25t,
  method = "near"
)
data25t_masked <- mask(
  data25t,
  theMask2,
  maskvalues = 0
)
writeRaster(data25t_masked, here::here("data/data25t_masked.tif"), overwrite = TRUE)
# data25t_masked <- terra::rast(here::here("data/data25t_masked.tif"))

# draw random sample
# This takes 20 ish minutes
samp2 <- spatSample(
  data25t_masked,
  size = 5e5,
  method = "random",
  na.rm = TRUE,
  xy = TRUE,
  as.df = TRUE
)

# make into points
samp2_sf <- sf::st_as_sf(
  samp2,
  coords = c("x", "y"),
  crs = crs(data25t_masked)
)

# join with region data
samp2_sf <- sf::st_join(
  samp2_sf,
  reg["region"],
  left = TRUE
)

# drop geometry to speed up calculations
samp2_df <- samp2_sf |>
  sf::st_drop_geometry()

# check sample size
samp2_df |>
  dplyr::count(region)

names(samp2_df)[1] <- "ABI"

samp2_tbl <- samp2_df |>
  tidyr::drop_na() |>
  group_by(region) |>
  summarise(mean_2025 = mean(ABI))

samp2_tblNat <- samp2_df |>
  tidyr::drop_na() |>
  summarise(mean_2025 = mean(ABI)) |>
  add_column(region = "Norge") |>
  bind_rows(samp2_tbl)


saveRDS(samp2_tblNat, here::here("data/samp2_tbl.rds"))


```

```{r rasterizeRegions}
#| eval: true

abi17 <- readRDS(here::here("data/samp_tbl.rds"))
abi25 <- readRDS(here::here("data/samp2_tbl.rds"))

abi17 <- abi17 |>
  select(area = region, value = mean_2017) |>
  add_column(year = 2017)


abi25 <- abi25 |>
  select(area = region, value = mean_2025) |>
  add_column(year = 2025)

abi <- abi17 |>
  bind_rows(abi25)
```

## 10. Results



```{r fig-plot}
#| fig-cap: 'Time series of spatially aggregeed human infrastructure or land use intensity index.'
region_labels <- c(
  "1" = "Østlandet", "2" = "Sørlandet", "3" = "Vestlandet",
  "4" = "Midt-Norge", "5" = "Nord-Norge", "Norge" = "Norge"
)
region_levels <- c("Norge", "Midt-Norge", "Nord-Norge", "Østlandet", "Sørlandet", "Vestlandet")

region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(region_levels)),
  region_levels
)

plot_data <- abi |>
  dplyr::mutate(
    region_name = dplyr::recode(area, !!!region_labels),
    series_type = dplyr::if_else(area == "Norge", "Norge", "Region")
  ) |>
  dplyr::group_by(year, area, series_type) |>
  dplyr::mutate(region_name = factor(area, levels = region_levels))

ggplot2::ggplot(
  plot_data,
  ggplot2::aes(x = year, y = value, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type)
  ) +
  ggplot2::scale_colour_manual(values = region_cols, name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(Norge = 3, Region = 1), guide = "none") +
  ggplot2::scale_x_continuous(breaks = c(2017, 2025))+
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 2)
    )) +
  ggplot2::labs(y = "Gj. ABI", x = "")

```

## 11. Export file

```{r exportFigure}
#| eval: false
ggsave(here::here("img/ABI.png"))
```



## 12. Author contributions

This workflow is the work of a single author, but I would like to acknowledge the help from Erik Stange in providing and explaining the datasets.
<!-- Fill in CReDIT and MeRIT statements to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

<!-- **CRediT statement**

* Conceptualization: [Initials]
* Methodology: [Initials]
* Data curation: [Initials]
* Formal analysis: [Initials]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [Initials]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD]. -->



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
