---
title: "Clear-cutting"
subtitle: |
  NO_AIFH_001
indicatorID: 
  - NO_AIFH_001
indicatorName: 
  - Clear-cutting
verbatimName: 
  - Flatehogst
folderName: NO_AIFH_001
country: Norway
continent: Europe
ECT: C1 Landscape and Seascape Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T2.2 Deciduous temperate forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: ecTools workflow; national refs X0 = 100% clear-cutting, X100 = 0% clear-cutting
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_AIFH_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county) - [Scale] - Aggregate (geographic region) - Aggregate (national level)
Precision: Approximate 95% bootstrap interval (2.5th–97.5th percentile of 1000 bootstrap draws) as inferential uncertainty around the point estimate (median). At county level, bootstrapping is applied directly to NFI plot data; at regional and national level, uncertainty is propagated via Monte Carlo resampling of county draws.
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
  docx:
    embed-resources: true
execute: 
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Ken Olaf Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Mari Jönsson
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Ulrika Jansson Asplund
    affiliations:
      - name: Norwegian Institute for Nature Research
AuthorList: Halbritter, A. H., Storaunet, K. O., Jönsson, M., Snoksrud, O., and Jansson Asplund, U.
date: June 9, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
csl: ambio.csl
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
codeReviewers:
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    affiliations:
      - id: reviewerInst
        name: Norwegian Institute for Nature Resaech (NINA)
    reviewType: code review
    github: DrMattG
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->
```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```
::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 09 Jun. 2026 - NO_AIFH_001, ecTools workflow
- 18 July 2026 - Accepted
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:

The widespread adoption of clear-cut forestry has altered the structure and dynamics of Norwegian forests. Clear-cut harvesting resets stand development and interrupts the continuity of canopy cover, reducing the occurrence of late-successional structural features such as old trees, large trees, dead wood, and vertically complex forest stands.
As a result, forests that have not been clear-cut are more likely to retain some components of structural complexity and continuity than forests managed under clear-cut regimes.   

These structural attributes are important for many forest species, particularly those associated with mature and old forests, stable microclimates, and continuous tree cover.
The indicator therefore reflects the proportion of forest that has not been clear-cut, providing a measure of the extent to which these ecological characteristics have been maintained.  

2. **Methods**:

Data comes from the National Forest Inventory (NFI) LSK extract.
At the plot level, clear-cutting status is recorded as a categorial varaible where 1 identifies clear-cut plots,  2 identifies plots without clear-cutting, and 3 indicates undetermined status in the NFI variable FLATEHOGD.
This variable is available from 2018 onwards, so the indicator is estimated for one reporting period only: 2020–2024.
Non-scaled indicator values are estimated by bootstrapping county-level data, and the resulting draws are scaled and aggregated to regional and national levels. 

3. **Interpretation**:

Current proportion of clear-cut forest is compared to a reference value with no clear-cut harvesting.
An indicator value of 1 means that 0% of the forest area has been clear-cut (_X~100~_).
An indicator value of 0 means that 100% of the forest area has been clear-cut (_X~0~_). 

4. **Results and development needs**:

The national indicator value for the current reporting period (2020–2024) is 0.7.
The value varies between the regions, with the highest value in Nord (0.93) and the lowest value in Øst (0.51).


### 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: Åpne hogstformer (flatehogst, frøstillingshogst) har endret stukturen og dynamikken i norske skoger. Flatehogst bryter kontinuiteten i både strukturell variasjon som kjennetegner seine suksesjonsstadier (som gamle trær, død ved og flersjiktede bestand) og i krondekke.
Disse strukturelle egenskapene er viktige for mange skoglevende arter, særlig arter knyttet til eldre skog, stabile mikroklimatiske forhold og kontinuerlig tresjikt.
Indikatoren viser derfor andelen skog som ikke har vært flatehogd, og gir dermed, sammen med andre indikatorer, et mål på i hvilken grad disse økologiske egenskapene og den strukturelle kontinuiteten er opprettholdt.   

2. **Metoder**: Dataene kommer fra Landsskogtakseringens LSK-uttrekk.
På flate-nivå registreres flatehogst i variabelen FLATEHOGD.
Variabelen finnes fra og med 2018, så indikatoren estimeres for én rapporteringsperiode: 2020–2024.
Ikke-skalerte verdier estimeres ved bootstrap på fylkesnivå, og trekningene skaleres og aggregeres til region- og riksnivå. 

3. **Tolkning**: Dagens andel av ikke flatehogd skog sammenlignes referansetilstanden hvor ingen skog er avvirket med flatehogst.
En indikatorverdi på 1 tilsvarer 0 % arealandel med flatehogst (_X~100~_).
En indikatorverdi på 0 tilsvarer 100 % arealandel med flatehogst (_X~0~_).

4. **Resultater og utviklingsbehov**: Den nasjonale indikatoren for den nåværende rapporteringsperioden (2020–2024) er 0.7.
Verdien varierer mellom regionene, med høyest verdi i Nord (0.93) og lavest verdi i Øst (0.51).

## 2. About the underlying data

### 2.1 Spatial and temporal extent, resolution, and representativity

The indicator is based on plot-level data from the Norwegian National Forest Inventory (NFI), covering Norwegian forest land.
The underlying data is available from 2018 onwards.
The indicator is therefore reported for a single five-year period, 2020–2024.
Earlier periods might be added if FLATEHOGD becomes available for previous years. 

### 2.2 Sampling design

Estimates of clear-cut forest area were based on permanent sample plots from the Norwegian National Forest Inventory (NFI).
For all plots measured during the 2018–2022 inventory cycle, previous field records and historical aerial photographs were interpreted to determine whether the forest had been subject to clear-cutting or not, back to ca. 1940.
Each plot was classified as either clear-cut, not clear-cut or not possible to determine [@viken2026_flatehogst].
8% of the investigated plots could not be determined because of insufficient historical aerial-photo coverage [@viken2026_flatehogst].  

### 2.3 Original units

At the plot level, the NFI variable `FLATEHOGD` uses code 1 for clear-cut plots, code 2 for no clear-cutting, and code 3 for undetermined status.
Plots with codes 1, 2, or 3 are retained; only missing values are excluded.
Aggregated values are the area-weighted mean proportion of forest area that has not been clear-cut. 

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

C1 Landscape and Seascape Characteristics

### 3.2 Ecosystem condition characteristic

The indicator represents the proportion of not clear-cut forest of the total forest landscape, since such harvesting practices became dominant in Norway after the 1940s.
Not clear-cut forests may occur as small and scattered patches in regions with intensive forest management, and as more continuous areas in regions with less intensive forest management.
The indicator is assessed as a landscape characteristic (C1) in the ecosystem condition typology.  

### 3.3 Collinearities with other indicators

Correlation with other indicators of old forest is expected.
Some correlation with habitat indicators (e.g. dead wood, old/large trees) might also be expected. 

### 3.4 Impact factors

Clear-cut harvesting is the impact factor for this indicator. 

### 3.5 Spatial aggregation

The spatial aggregation pathway is: aggregate (NFI plots to county) – [scale] – aggregate (geographic region) – aggregate (national level).
Non-scaled values are area-weighted means of the plot-level `FLATEHOGD` variable.
Scaling uses national reference levels X0 = 100% clear-cut area share and X100 = 0% clear-cut area share (proportions 1 and 0 on the original scale).

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances.

### 4.2 Reference levels

X0 is set to 100% clear-cut area share (worst condition) and X100 to 0% clear-cut area share (best condition).
Values are scaled linearly between X0 and X100 (@sec-aifh-reference-values).
Because the non-scaled indicator is an area proportion bounded between 0 and 1 (0%–100% clear-cut), values can not fall outside the reference range and no truncation is needed.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined as one national value for the share of the assessed forest area with no clear-cutting.
Although indicator values from NFI plots technically may be aggregated at finer spatial units than counties, scaling at such finer levels may result in biases. 

## 5. Uncertainties

### 5.1 General uncertainties

For 8 % of the NFI plots it was not possible to determine if the area had been subjected to clear-cut harvesting or not. These plots were classified as not clear-cut, resulting in a somewhat conservative estimate of the extent of clear-cut forest.
Regional and national values are area-weighted summaries of county estimates, and therefore inherit both model uncertainty at county level and any mismatch between county boundaries and ecological variation within regions.
The indicator does not account for the intensity and timing of clear-cutting, nor for the size and fragmentation of forests that have not been clear-cut. 

### 5.2 Reported uncertainties

County non-scaled values are bootstrapped with 1000 draws per county per period.
County draws are scaled using the national _X~0~_ and _X~100~_.
Uncertainty is propagated when aggregating to regional and national levels.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Share of forest area with clear-cut

The NFI records clear-cutting status in the variable `FLATEHOGD` (column name after `clean_names()`: `flatehogd`).
Code 1 identifies clear-cut plots, code 2 identifies plots without clear-cutting, and code 3 indicates undetermined status.
Plots with codes 1, 2, or 3 are retained; only missing values are excluded.
Codes 2 and 3 are both treated as no clear-cutting in the indicator calculation.
`flatehogd` is available from 2020 onward, and analyses use the reporting period 2020–2024.
Both productive and unproductive forest plots (`arealtype` 1 and 12) are included.

```{r load-data}
#| code-summary: "Load data"

lsk_data_file <- here("data/LSK_data_okologisk_tilstand_2026-06-03.csv")

lsk_data_raw <- read_csv(
  lsk_data_file,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties, regions and nationally.
Where counties are split into parts, those parts are merged to a single county name before aggregation.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-aifh-derived-vars}

This section prepares the NFI plot data for indicator calculation.
We define one five-year reporting period (2020–2024).

```{r reporting-periods}
#| code-summary: "Define reporting periods"

min_indicator_year <- 2020L

reporting_periods <- list(
  # "2010-2014" = 2010:2014,  # enable when data is available
  # "2015-2019" = 2015:2019,  # enable when data is available
  "2020-2024" = 2020:2024
)

```

We filter the raw dataset to inventory years from 2020 onward (`sesong >= min_indicator_year`) and include productive and unproductive forest plots (`arealtype` 1 and 12).
Plots with non-missing `flatehogd` (codes 1, 2, or 3) are retained, and split NFI county labels are standardised by merging county parts to a single county name.
Code 1 is recoded to 1 (clear-cut); codes 2 and 3 are recoded to 0 (no clear-cutting).
For each included plot, the area-weighted contribution to later aggregation is `indicator_continuous * au_areal`.

```{r aifh-create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(sesong >= min_indicator_year) |>
  filter(arealtype %in% c(1, 12)) |> 
  filter(!is.na(flatehogd)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    fylke_name = normalize_county_parts(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = dplyr::if_else(flatehogd == 1, 1, 0),
    indicator_continuous_weighted = indicator_continuous * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r aifh-select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    fylke_name,
    fylke_name_clean,
    flatehogd,
    indicator_continuous,
    indicator_continuous_weighted
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values {#sec-aifh-non-scaled}

For the reporting period (2020–2024), we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted mean of all included plots: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
Plots with code 3 contribute their area to the denominator and are treated as no clear-cutting (0) in the numerator.

```{r calculate-aifh-all-periods}
#| code-summary: "Calculate non-scaled values"

 aifh_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 aifh_results_county_all_periods   <- purrr::map_dfr(aifh_results, "county")
 aifh_results_region_all_periods   <- purrr::map_dfr(aifh_results, "region")
 aifh_results_national_all_periods <- purrr::map_dfr(aifh_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each bootstrap replicate computes the area-weighted share of clear-cut forest area on the non-scaled scale, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(lsk_data_file), file.mtime(here::here("R/NO_AIFH_001.qmd")))

county_draws_long <- purrr::imap_dfr(aifh_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-aifh-county-nonscaled) shows the direct area-weighted mean share of clear-cut forest area per county, with approximate 95% bootstrap confidence intervals (2.5th–97.5th percentile of 1000 draws).

```{r tbl-aifh-county-nonscaled}
#| tbl-cap: "Non-scaled share of clear-cut forest area (%) by county. Point estimate is the direct area-weighted mean; 95% CI is derived from 1000 bootstrap draws."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, region_code, total_area) |>
  dplyr::summarise(
    ci_low  = quantile(non_scaled_value, 0.025) * 100,
    ci_high = quantile(non_scaled_value, 0.975) * 100,
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aifh_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots, indicator_value),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    mean_pct = round(indicator_value * 100, 1),
    ci_low   = round(ci_low, 1),
    ci_high  = round(ci_high, 1),
    `95% CI` = paste0(ci_low, " – ", ci_high)
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Clear-cut area (%)` = mean_pct,
    `95% CI`,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Clear-cut area (%)`,
    decimals = 1
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-aifh-reference-values}

National reference levels are _X~0~_ = 100% clear-cut area share, and _X~100~_ = 0% clear-cut area share (proportions 1 and 0 on the original scale).

```{r aifh-reference-values}
#| code-summary: "National reference levels"

x0 <- 1
x100 <- 0


```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 1, X100 = 0).
Because the non-scaled indicator is an area proportion bounded between 0 and 1 (0%–100% clear-cut), values can not fall outside the reference range and no truncation is needed.
@fig-aifh-norm shows the mapping.

```{r county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aifh_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-aifh-norm}
#| fig-cap: "Linear normalisation of clear-cut area share (proportion) to indicator values (0–1), using national reference levels X0 = 100% and X100 = 0%. Points are 500 sampled bootstrap draws for the reporting period 2020–2024."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 5
#| fig-width: 6

norm_plot_data <- county_draws_scaled |>
  dplyr::slice_sample(n = min(500, nrow(county_draws_scaled)))

aifh_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::labs(
    x = "Clear-cut area share (proportion)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
aifh_norm_plot$layers[[1]]$aes_params$size <- 3

suppressMessages(
  ggsave(
    filename = here::here("img/aifh_norm_plot.png"),
    plot = aifh_norm_plot,
    scale = 1.5
  )
)

aifh_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r aifh-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

aifh_mc_cache_deps <- c(
  lsk_data_file,
  here::here("R/functions.R"),
  here::here("R/NO_AIFH_001.qmd")
)
aifh_mc_rds_regional <- here::here("data/aifh_mc_regional.rds")

if (
  file.exists(aifh_mc_rds_regional) &&
  file.mtime(aifh_mc_rds_regional) >= max(file.mtime(aifh_mc_cache_deps), na.rm = TRUE)
) {
  aifh_mc_regional <- readRDS(aifh_mc_rds_regional)
} else {
  aifh_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(aifh_mc_regional, aifh_mc_rds_regional)
}
```

```{r aifh-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

aifh_mc_regional_for_national <- aifh_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

aifh_mc_rds_national <- here::here("data/aifh_mc_national.rds")

if (
  file.exists(aifh_mc_rds_national) &&
  file.mtime(aifh_mc_rds_national) >= max(
    c(file.mtime(aifh_mc_cache_deps), file.mtime(aifh_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  aifh_mc_national <- readRDS(aifh_mc_rds_national)
} else {
  aifh_mc_national <- ecTools::ec_upscale(
    data        = aifh_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(aifh_mc_national, aifh_mc_rds_national)
}
```

```{r aifh-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

aifh_mc_combined <- dplyr::bind_rows(
  aifh_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  aifh_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- aifh_results_county_all_periods |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- aifh_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- aifh_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- aifh_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    aifh_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- aifh_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    aifh_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

```

The figure below (@fig-aifh-timeSeries) summarises regional and national indicator values for the reporting period 2020–2024.
When additional periods are enabled in `reporting_periods`, the same plot will show values across periods.

```{r fig-aifh-timeSeries}
#| fig-cap: "Indicator values for share of forest area without clear-cutting by region for the reporting period 2020–2024. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

aifh_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
aifh_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
aifh_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(aifh_region_levels)),
  aifh_region_levels
)

aifh_plot_data <- aifh_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!aifh_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = aifh_region_levels))

ggplot2::ggplot(
  aifh_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = aifh_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county, regional, and national scales for the reporting period 2020–2024.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to X0 (100% clear-cut area share) and 1 corresponds to X100 (0% clear-cut area share).

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-aifh-county-results to browse county rows for the reporting period 2020–2024.

```{r tbl-aifh-county-results}
#| tbl-cap: 'County indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Values are scaled to 0–1 using national X0 and X100. Uncertainty is derived from 1000 bootstrap draws. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-aifh-region-results}
#| tbl-cap: 'Regional indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-aifh-national-results}
#| tbl-cap: 'National indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`aifh_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  aifh_mc_combined,
  here::here("data/results_NO_AIFH_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [UJA, KOS]
* Methodology: [UJA, KOS, AHH]
* Data curation: [OS]
* Formal analysis: [AHH]
* Validation: [AHH, OS, KOS]
* Visualization: [AHH]
* Writing – original draft: [AHH, UJA]
* Writing – review & editing: [All]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info
This workflow uses `renv`. Run `renv::restore()` to recreate the working environment. 
:::
