---
title: "Nature Index product"
subtitle: |
  [NO_NIPR_001] (waterbirds) \
  [NO_NIPR_002] (<i>Lutra lutra</i>)
indicatorID: 
  - NO_NIPR_001
  - NO_NIPR_002
indicatorName: 
  - Waterbirds
  - <i>Lutra lutra</i> (Eurasian otter)
folderName: NO_NIPR_001_002
country: Norway
continent: Europe
ECT: B1 - Compositional State Characteristics
Realm: 
  - Freshwater (F)
Biome: 
  - F2 Lakes biome
  - F1 Rivers and streams biome
Ecosystem: 
  - F2.4 Freeze-thaw lakes
  - F1.3 Freeze-thaw streams
yearAdded: 2025
yearLastUpdate: 2025
status: complete
Version: "001.000"
VersionComment: First complete version
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_NIPR_001_002
Normalised: Yes
SpatialAggregationPathway: Scale - Aggregate (area-weighted arithmetic mean) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research
AuthorList: Sandvik, H.  
date: November 24, 2025
data_availability: gold
code_reproducibility: silver
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(ggridges)
library(magrittr)
library(paletteer)
library(pals)
library(readr)
library(svglite)
library(usethis)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here::here("extract_yaml.R"))
```

<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="5"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]),
  folder_name = folder_name)
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```
::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID: ", id, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 24 Oct 2025 - Version 001.000 started: first complete version
- 28 Oct 2025 - Version 001.000 submitted to review
- 24 Nov 2025 - Otter added to version 001.000
:::
  

<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**: 
The indicators consist of abundance indices for (1) eleven waterbird species and (2) Eurasian otters that are included in the [Nature Index for Norway](https://www.naturindeks.no/) (**NI**, hereafter). 
Waterbirds breed in lakes, and otters live in rivers. 
Both can be regarded as indicators of intact ecological condition, as they are sensitive to anthropogenic pressures such as habitat loss and pollution.
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->
2. **Methods**: 
The indicator is a "Nature Index product", meaning that it is not identical to, but derived from, the NI and is estimated from data that are read directly from the NI database. 
The averaging across species involves weighting by functional group, so that the three trophic levels represented among waterbids (herbivores, mesopredators and top predators) receive the same weights.
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->
3. **Interpretation**:
For the waterbird indicator, a value of 1 signifies that waterbird abundances are equal to those in an intact lake ecosystem in the period 1961--1990. 
For the otter indicator, a value of 1 signifies that otters are present in all river systems with suitable habitat.  
Values decreasing toward zero reflect proportionally lower abundances or presences.
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. -->
4. **Results and development needs**:
The indicators are ready for use. 
Its values for 2024 mean that (1) the average waterbird abundance is approximately 77% of the one expected in an intact lake ecosystem, and that (2) otters are present in 78% of suitable habitat.
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->


## 2. About the underlying data

This indicator is a "Nature Index product" ([Nater 2023b](#nat23b)), i.e. it is estimated using data from the [Nature Index for Norway](https://www.naturindeks.no/) or **NI** ([Miljødirektoratet 2025](#mil25)). 

This documentation focuses on the waterbird indicator (NO_NIPR_001). 
The otter indicator (NO_NIPR_002) is calculated along the same lines but is not documented step-by-step. 
For more information about the latter, please refer to the [Nature Index](https://www.naturindeks.no/Indicators/oter_ferskvannsbestand).

The original source of the waterbird abundance data making up this indicator is the Norwegian Breeding Birds Monitoring Program ([*Hekkefulgovervåkingen*](https://hekkefuglovervakingen.nina.no/)).

<!--# Describe the data you have used in more detail, it's origin, biases, availability ect.-->

### 2.1 Spatial and temporal resolution and extent

For most bird species included in this indicator, separate NI values are provided for Southern and Northern Norway. The **spatial** resolution of this indicator is thus roughly one half of Norway.

The NI is updated every *five years*, which thus is the **temporal** resolution of the indicator.

<!--# Describe the temporal and spatial resolution and extent of the data used -->

### 2.2 Sampling design

The species included in this indicator are sampled in connection with the Norwegian Breeding Birds Monitoring Program. 
The sampling design is the same as described for the indicators [NO_BIRD_001--004]((https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BIRD_001-004)) ([Nater et al. 2025](#nat25)). 
Please refer to that description for more detail.

<!-- How was the data sampled? Describe potential sampling biases -->

### 2.3 Original units

The observers in the field record “number of individuals” seen or heard during the census. 
These raw counts are converted into species-specific abundance indices. 
See  Nater et al. ([2025](#nat25)) for more detail.

<!--# What are the original units for the most relevant  variables in the data-->

### 2.4 Instructions for citing, using and accessing data

Upon publication of NI 2025 (later this year), there will be reading access to the NI database. 
The data should be cited by referring to the [NI webpage](https://www.naturindeks.no/) or the NI 2025 summary report (Chloé Nater et al., currently in prep.).

<!--# Is the data openly available? If not, how can one access it? What are the key references to the datasets?   -->


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The ECT of waterbirds is **B1** (compositional state) because the indicator is a measure of the abundances of selected species.

<!--# 
Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. 
-->

#### 3.1.1 Other condition typologies

In the Norwegian classification of ecological characteristics (*økologiske egenskaper*) according to Nybø & Evju ([2017](#nyb17), p. 35), the indicator represents **biodiversity** (*biologisk mangfold*).

<!--# Optional: Add text about other spesific standards, e.g. national standards, and how the indicator relates to these -->

### 3.2 Ecosystem condition characteristic

"The metric is aimed at describing a characteristic of ecosystems in good condition as having high and stable bird populations" ([Nater et al. 2025](#nat25)).

<!--# 
Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decition Making.  
-->

### 3.3 Collinearities with other indicators

There are no known collinearities with other Ecosystem Condition indicators. 

<!--# Describe known collinearities with other metrices (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

### 3.4 Impact factors

"Birds are affected by a large range of both natural and anthropogenic stressors, and what stressors are most important will vary from species to species. In general, across all species, and of the human caused impact factors, we can mention, climate change [...] and land-use change (e.g. habitat loss, reduced connectivity, edge effects)" ([Nater et al. 2025](#nat25)).

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->

### 3.5 Spatial aggregation

The variable is scaled at the initial polygon level. 
Spatial aggregation is done using are-weighted arithmetic means. 
Truncation is done, if necessary, of the spatially aggregated indicator values. 
Aggregation across species involved weighting by trophic level, ensuring that the trophic levels have equal weight, irrespective of the number of species by which any trophic level is represented. 
There is no additional transformation.

<!-- Describe in word the same aggregation pathway presented in the metadata.xlsx. If you have scaled against the reference value, it is especially important to know if you did any avergaing or summations etc. beforehand. An example text for a rather simple case could be: "The variable is scaled at the initial polygon level and the point estimate is a area-weighted arithmetic mean of all polygins. There is no transformation, and truncation is not needed as the indicator becomes naturally bound between 0 and 1". -->


## 4. Reference condition and levels

### 4.1 Reference condition

For the NI, the reference condition is defined as "an ecosystem which is little affected by human activity", and it "consists of ecosystems with species assemblages and abundances which would have been present in an intact ecosystem in the period 1961--1990" ([Miljødirektoratet 2020/2025](#mil25) /About).

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels

Bird abundance indices are scaled linearly between the reference levels for X~100~ and X~0~. 
X~100~ has been defined by expert judgement, separately for each species and (normally) regions of Norway. 
X~0~ is a natural zero.

<!--# 
If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ refers to the reference level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.
-->

#### 4.2.1 Spatial resolution and validity

The spatial resolution of the reference levels is the same as the spatial resolution of the variable values. 
In other words, for most bird species included here, there are two reference levels for Norway.

<!--# 
Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 
-->


## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties in NI data have several sources, the most import of which is sampling error, implying that the indicator value presented for a given geographic area may not be fully representative of the entire area. 
For indicator values that are based on expert judgements, there is also the obvious possibility of errors of judgement.

<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

### 5.2 Reported uncertainties

The uncertainties reported are directly based on the ones in the NI database. 
They are inferential and refer to the uncertainty around the point estimates.

<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->


## 6. References

<span id="ber21">Berge SE & Pedersen B.</span> 2021. Nature Index system documentation: mathematical framework, database, web-portals, scripts and API. *NINA Rep* **1990**: 1--66. [[hdl:11250/2739962](https://hdl.handle.net/11250/2739962)]

<span id="mil25">Miljødirektoratet.</span> 2020/2025. Naturindeks. <https://www.naturindeks.no>

<span id="nat23a">Nater CR.</span> 2023a. NIflex: flexible and customizable implementations of the Norwegian Nature Index. <https://github.com/NINAnor/NIflex>

<span id="nat23b">Nater CR.</span> 2023b. Nature index product. <https://ninanor.github.io/ecosystemCondition/nature-index-product.html> in: Kolstad AL et al. (eds) *Indicators for ecosystem condition in Norway*. [[doi:10.5281/zenodo.7854835](https://doi.org/10.5281/zenodo.7854835)]

<span id="nat25">Nater CR, Pavón-Jordán D & Kolstad AL.</span> 2025. Breeding bird indicators, v. 000.002. [[ecRxiv:NO_BIRD_001-004](https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BIRD_001-004)]

<span id="nyb17">Nybø S & Evju M (eds).</span> 2017. *Fagsystem for fastsetting av god økologisk tilstand. Forslag fra et ekspertråd.* Klima- og miljødepartementet, Oslo. <https://www.regjeringen.no/id2558481/>

<span id="ped20">Pedersen B et al.</span> 2020. NIcalc: R-package for calculation of the Nature Index. <https://github.com/NINAnor/NIcalc>

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


## 7. Import and prepare datasets

<!--# Describe the unique datasets seperately under seperate headers (Dataset A, Dataset B, etc.-->

### 7.1 Nature Index

The Nature Index data are read from the NI database (<https://naturindeks.nina.no>). 
To be able to do so, we first need to install the **R** packages `NIcalc` ([Pedersen et al. 2020](#ped20), [Berge & Pedersen 2021](#ber21)) and `NIflex` ([Nater 2023a](#nat23a)).

```{r}
#| output: false

if(system.file(package = 'NIcalc') == "") {
  devtools::install_github("NINAnor/NIcalc", build_vignettes = TRUE)
}

if(system.file(package = 'NIflex') == "") {
  devtools::install_github("NINAnor/NIflex", build_vignettes = TRUE)
}

library(NIcalc)
library(NIflex)
```

Reading access is needed to download data from the NI database. 
For the time being, this requires a username and password, which can be requested by contacting the NI core team (<naturindeks@nina.no>). 
The following code assumes that you have obtained these credentials and stored them as local environmental variables in the project's .Renviron file.

```{r}
#| eval: false
# Because you may not have obtained the credentials necessary to run this code,
# it is only displayed but not executed

NIdb_username <- Sys.getenv("NIdb_username")
NIdb_password <- Sys.getenv("NIdb_password")
```

<!--# Describe the main dataset, typicaly the one containing the variable of (most) interest. Change the header from Dataset A to the name of the actual dataset. -->


## 8. Spatial units

For most bird species included in this indicator, the spatial units are Southern and Northern Norway. 
The results are reported with mainland Norway as the only spatial unit.
The Ecosystem Accounting Area covered by the data is mainland Norway. 

<!--# 
Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpretted at. Potential spatial delineation data should eb introduced under 7.1. Datasets. We recomend using the SEEA EA terminology opf Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). 
-->


## 9. Analyses

The steps and choices involved in estimating a Nature Index product have been documented by Nater ([2023b](#nat23b)). 
The current analyses follow this template.

The data for waterbirds are read using the convenience function `calculateCustomNI` from the NIflex package. 
As a default, this function reads all indicators of a specified ecosystem. 
One can then choose to omit certain indicators from the calculation.

We thus need to start by defining the NI year and ecosystem (by its Norwegian name).

```{r}
year <- 2024
EcoSystem <- "Ferskvann" # = freshwater
```

For the freshwater ecosystem, the following indicators are available:

```{r}
load("../data/indicato.RData")
rownames(ferskvann) <- NULL
print(ferskvann[, -(5:7)])
```

We thus need to exclude all indicators except waterbirds.

```{r}
omit <- 
  c(1, 50, 11, 233, 31, 228, 231, 36, 38, 43, 268, 270,                104, 134, 135, 146,                                         213)
# c(1, 50, 11, 233, 31, 228, 231, 36, 38, 43, 268, 270, 287, 385, 386, 104, 134, 135, 146, 298, 275, 333, 277, 278, 296, 312, 279, 213),
#   ål AUR pit aip BUNN aspt radd DPL krp EMU fmås FOSS have krik kvin laks OTER oter pti  sila smål stok stor stra svar svøm topp TIc
```

The species we keep are waterbirds, here shown with scientific and English names added.

```{r}
wb <- ferskvann[-which(ferskvann$id %in% omit), c(1, 2, 4)]
wb <- wb[order(wb$name), ]
wb$functionalGroup <- sapply(strsplit(wb$functionalGroup, " "), "[[", 1)
wb$scientificName <- c("Clangula hyemalis",
                       "Anas crecca",
                       "Bucephala clangula",
                       "Mergus serrator",
                       "Gavia stellata",
                       "Anas platyrhynchos",
                       "Gavia arctica",
                       "Actitis hypoleucos",
                       "Melanitta nigra",
                       "Phalaropus lobatus",
                       "Aythya fuligula")
wb$EnglishName    <- c("Long-tailed duck",
                       "Eurasian teal",
                       "Common goldeneye",
                       "Red-breasted merganser",
                       "Red-throated diver",
                       "Mallard",
                       "Black-throated diver",
                       "Common sandpiper",
                       "Common scoter",
                       "Red-necked phalarope",
                       "Tufted duck")
wb <- wb[, c(1, 2, 4, 5, 3)]
print(wb)
```

Furthermore, we would like to use trophic weighting, which means that trophic levels (or "functional groups") are assigned equal weight, irrespective of the number of species by which any trophic level is represented. 
Among the waterbird indicators, there are two herbivores, six mesopredators and three top predators.

```{r}
TrophicWeights <- TRUE
```

Now we are ready to calculate the waterbird indicator.

```{r}
#| eval: false
#| output: false

# This code chunk requires reading access to the NI database.
# It is only displayed but not executed.
# Instead, a cached version of the output is read (silently).

set.seed(12)

waterbirds <- calculateCustomNI(
  ecosystem = EcoSystem,
  indicators = NULL,
  theme = "none",
  dropInd = omit,
  username = NIdb_username,
  password = NIdb_password,
  KeyIndicators = FALSE,
  KeyWeight = 0.5,
  AreaWeights = TRUE,
  TrophicWeights = TrophicWeights,
  NAImputation = FALSE,
  years = year,
  OutputType = "EcologicalCondition",
  funArguments = NULL,
  Diagnostics = TRUE,
  TestRun = FALSE,
  norwegianNames = TRUE,
  saveSteps = FALSE
)
```
```{r}
#| echo: false
# This is the cached output from a previous call. 
load("../data/watrbird.RData")
```

Let's double check the species included.

```{r}
print(waterbirds$CustomIndex$wholeArea$"2024"$indicatorData)
```

That is as intended. 
In the final step, we calculate the quantiles of the output.

```{r}
print(round(quantile(waterbirds$CustomIndex$wholeArea$"2024"$index,
            c(0.025, 0.25, 0.5, 0.75, 0.975)), 6))
```

<!--# 
Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.
-->

### 9.1 Otter

The otter indicator (NO_NIPR_002) is calculated along the same lines. 
The main difference consists of the indicators to be omitted:

```{r}
omit <- 
  c(1, 50, 11, 233, 31, 228, 231, 36, 38, 43, 268, 270, 287, 385, 386, 104,           146, 298, 275, 333, 277, 278, 296, 312, 279, 213)
# c(1, 50, 11, 233, 31, 228, 231, 36, 38, 43, 268, 270, 287, 385, 386, 104, 134, 135, 146, 298, 275, 333, 277, 278, 296, 312, 279, 213),
#   ål AUR pit aip BUNN aspt radd DPL krp EMU fmås FOSS have krik kvin laks OTER oter pti  sila smål stok stor stra svar svøm topp TIc
```

Trophic weighting is not needed. 
The indicator is thus calculated as the average condition of the freshwater population and the coastal population of otters 
(where "coastal population" refers to all otter populations in municipalities with a coastline, irrespective of the actual habitat use of the species).


## 10. Results

The indicators are reported in terms of *scaled* indicator values.


| Indicator   | Name | Ecosystem | Best estimate | Lower quartile | Upper quartile |
|:------------|:-----------|:-------|:-------------:|:--------------:|:--------------:|
| NO_NIPR_001 | Waterbirds | Lakes  | 0.774         | 0.742          | 0.804          |
| NO_NIPR_002 | Otter      | Rivers | 0.782         | 0.781          | 0.783          |

: Waterbird and otter indicator values for Norway in 2024 {.striped .hover}


<!--# 
Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.
-->

<!--# 
## 11. Export file

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a seperate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 
-->

```{r export}
#| eval: false
```
