---
title: "Temperate deciduous trees by share of forest area in the boreonemoral zone"
subtitle: |
  [NO_TDTA_001]
indicatorID: 
  - NO_TDTA_001
indicatorName: 
  - NO_TDTA
country: Norway
continent: Europe
ECT: B1 Biotic compositional characteristic
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2024 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2025
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_TDTA_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: No # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Aggregate # e.g. Scale - Transform - Aggregate (length-weighted arithmetic mean) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Framstad, E., Storaunet, K. O. and Halbritter, A. H.
date: November 8, 2025
data_availability: gold
code_reproducibility: bronze
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("indicators/NO_TDTA_001/R/functions.R"))
```

```{r extractFromYML}
#| echo: false
#| cache: false
this_file <- get_file_info()

# Read the YAML front matter
lines <- readLines(this_file$path, warn = FALSE)
dash_lines <- which(trimws(lines) == "---")
yaml_block <- lines[(dash_lines[1] + 1):(dash_lines[2] - 1)]
yaml_data <- yaml::yaml.load(paste(yaml_block, collapse = "\n"))

# Robust flatten function
flatten_to_string <- function(x) {
  if (is.list(x)) {
    # recursively flatten any nested lists
    x <- unlist(x, recursive = TRUE)
  }
  # convert to single string
  paste(x, collapse = "; ")
}

# Apply to all YAML entries
yaml_data_flat <- lapply(yaml_data, flatten_to_string)

# Create tibble
meta <- tibble::enframe(yaml_data_flat, name = "Variable", value = "Value")

# Check structure
#str(meta)

st <- meta |>
  dplyr::filter(Variable == "status") |>
  pull(Value)
version <- meta |>
  dplyr::filter(Variable == "Version") |>
  pull(Value)
auth <- meta |>
  dplyr::filter(Variable == "AuthorList") |>
  pull(Value)
year <- meta |>
  dplyr::filter(Variable == "yearAdded") |>
  pull(Value)
id <- meta |>
  dplyr::filter(Variable == "indicatorID") |>
  pull(Value)
name <- meta |>
  dplyr::filter(Variable == "indicatorName") |>
  pull(Value)
url <- meta |>
  dplyr::filter(Variable == "url") |>
  pull(Value)
badges<-meta |> 
  dplyr::filter(Variable %in% c("data_availability",
         "code_reproducibility", "open_science_badge")) |> 
  pull(Value)

meta <- meta |>
  dplyr::mutate(Variable = dplyr::case_match(Variable,
    "indicatorID" ~ "Indicator ID" ,
    "indicatorName" ~ "Indicator Name",
    "continent" ~ "Continent",
    "country" ~ "Country",
    "ECT" ~ "Ecosystem Condition Typology Class",
    "yearAdded" ~ "Year added",
    "yearLastUpdate" ~ "Last update",
    "VersionComment" ~ "Version comment",
    "SpatialAggregationPathway" ~ "Spatial aggregation pathway",
    .default = Variable
  )) |>
  dplyr::filter(Variable != "authors")

```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="5"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```
::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID: ", id, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 08 Nov. 2025 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
This indicator represents the share of forest area in the boreonemoral (hemi-boreal) bioclimate zone that has a substantial share (≥30%) of temperate deciduous tree species (of the genera Acer, Fagus, Fraxinus, Quercus, Tilia, Ulmus, as well as *Alnus glutinosa*). Temperate deciduous trees are important habitat or substrate for a wide range of fungi and insects, as well as other species. The main current distribution of temperate deciduous trees is in the boreonemoral zone with a mild climate, largely reflecting the general climate of the last 50 or more years. In forests shaped by natural disturbances and succession, temperate deciduous trees should constitute an increasing proportion of the forest area in response to climate change. Forestry with an emphasis on spruce (*Picea abies*) and land use change have reduced the proportion of forests dominated by temperate deciduous trees.

2. **Methods**:
Data for this indicator come from the National Forest Inventory (NFI). The frequency of temperate deciduous trees has been recorded for NFI plots since 2000 according to current methods (@viken2021-feltinstruks). Here we include only NFI plots in the boreonemoral zone.

3. **Interpretation**:
The proportion of forest dominated by temperate deciduous trees may be compared to a reference value for all forest in the boreonemoral zone, based on expert assessments supported by empirical data for near-natural forests and knowledge of ecosystem dynamics in intact forest ecosystems. Such reference values have not yet been determined for this indicator.

4. **Results and development needs**:
Otherwise, the indicator is ready for use as a non-scaled variable. The indicator should not be used at finer spatial resolution than regions within the boreonemoral zone.



## 2. About the underlying data

The frequency and volume of temperate deciduous trees are recorded for each 250 m2 NFI inventory plot (@viken2021-feltinstruks). Here we only include NFI plots in the boreonemoral zone, where temperate deciduous trees have their main distribution. There are about 1760 NFI plots in forest ecosystems in the boreonemoral zone, distributed on a regular 3x3 km grid. The indicator value is an unbiased area-weighted estimate of the proportion of forest dominated by temperate deciduous trees (quantified as the NFI variables Eik_pr and Edellauv_pr) for included NFI plots in any selected region. All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year. For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024. The data are available upon request from NFI. 

### 2.1 Spatial and temporal resolution and extent

Data are available from 2000, when current recording methods for the variables Eik_pr and Edellauv_pr were established. Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. The data are representative for the forest land within the boreonemoral zone of Norway. Spatial resolution should be limited to regions within the boreonemoral zone.

### 2.2 Sampling design

...

### 2.3 Original units

The indicator is specified as the share (%) of forest area dominated by temperate deciduous trees.


### 2.3 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2025-10-20, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

### 2.4 Additional comments about the dataset

None

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The indicator represents the share of forest dominated by temperate deciduous trees, i.e., tree species associated with particularly high species richness. Hence, this indicator represents a B2 Biotic structural characteristic.

### 3.1.1 Other condition typologies

None


### 3.2 Ecosystem condition characteristic

The indicator represents the share of forest dominated by temperate deciduous trees, i.e., tree species associated with particularly high species richness. The indicator is characterised as Functionally important species and biophysical structures.

### 3.3 Collinearities with other indicators

There are no apparent correlations with other indicators.


### 3.4 Impact factors

Natural disturbances at various temporal and spatial scales will contribute to the establishment and dominance of some of the temperate deciduous trees, whereas others depend on more stable environments. High populations of wild ungulates will limit recruitment of most temperate deciduous trees. Natural climate variability will affect the viability of the various species. Forestry practices like clear-cutting and dense planting of spruce will reduce the proportion of old temperate deciduous trees. As deciduous trees are often located in the southern, more densely populated lowlands of Norway, other human activities are a clear threat. Human-induced climate change is likely to expand the distribution and dominance of temperate deciduous trees.

### 3.5 Spatial aggregation

...

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems in the boreonemoral zone, where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

An upper reference value (X100) has not yet been determined. It would likely be based on expert assessments supported by information on (1) the distribution and frequency of temperate deciduous trees in NFI plots assessed to be little affected by recent human activities, (2) the distribution of other dominating tree species in various regions/counties, and (3) general assessments of natural forest dynamics and the effects on temperate deciduous trees. Reference values would be expert assessments without estimated uncertainty. The lower reference value (X0) is set to 0%. For lack of better knowledge, the relationship between indicator values and ecosystem condition is assumed to be linear for values between X100 and X0. Hence, indicator values are scaled between 0 and 1 based on a linear scaling function between X0 and X100; in principle, scaled indicator values above 1 are truncated to 1.


#### 4.2.1 Spatial resolution and validity

The reference value will be given as an overall aggregated value for the share of forest dominated by temperate deciduous trees for larger areas, such as counties, within the boreonemoral zone. Although indicator values from NFI plots may technically be aggregated at finer spatial units than counties, scaling indicator values at such finer spatial levels may result in biases.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability in the values of the variables Eik_pr and Edellauv_pr among NFI plots and its quantification for each NFI plot. This uncertainty is quantified by bootstrapping. There will also be uncertainty related to the setting of the upper reference value and to the shape of the relationship between indicator values and ecosystem condition. These uncertainties are not quantified.

### 5.2 Reported uncertainties

Uncertainty is quantified by bootstrapping.
The uncertainties reported here are inferential and refer to the uncertainty around the point estimates, reflecting spatial variation among NFI plots and recording errors.


## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

### 7.1 Dataset A

<!--# Import and prepare the first dataset for analysis. Change the header from Dataset A to the name of the actual dataset. -->

## 8. Spatial units

<!--# Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpreted at. Potential spatial delineation data (maksing layers etc.) should be introduced under chapert 7. We recommend using the SEEA EA terminology of Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). -->

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->


```{r export}
#| eval: false
```

<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
