---
title: "Dead wood share"
subtitle: |
  NO_DTVS_001
indicatorID: 
  - NO_DTVS_001
indicatorName: 
  - Dead wood share
verbatimName: 
  - Død ved andel
folderName: NO_DTVS_001
country: Norway
continent: Europe
ECT: B2 Structural state characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2024 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_DTVS_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Ken Olaf Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
AuthorList: Halbritter, A. H., Framstad, E., Snoksrud, O., and Storaunet, K. O.
date: November 8, 2025
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
csl: ambio.csl
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
codeReviewers:
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    affiliations:
      - id: reviewerInst
        name: Norwegian Institute for Nature Resaech (NINA)
    reviewType: code review
    github: DrMattG
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->
```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```
::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 11 Nov. 2025 - Original PR
- 12 May. 2026 - Updated to use new template
- 18 July 2026 - Accepted
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
The volume of dead wood as a share of the volume of all living, dead and harvested wood is a robust measure of the naturalness of forests.
Dead wood is a key resource as substrate or food for a wide range of forest organisms, especially fungi and invertebrates.
However, the expected volumes of both dead and living wood in intact forest ecosystems vary between forest types depending on site productivity, dominating tree species and climate conditions.
By comparing the volume of dead wood to the volume of all wood for the same geographical area, the effects of such site factors may partly be cancelled out.  

2. **Methods**:
Data for this indicator come from the National Forest Inventory (NFI), where the volumes of living and standing dead trees, as well as the cumulative volume of harvested trees have been recorded for each NFI plot since the mid-1990s and lying dead wood since 2010 (by current methods).
The indicator is the share of dead wood volume to the volume of all living, harvested and dead wood on each plot.
For each reporting period, area-weighted means of this share are calculated for counties, regions and nationally using au_areal. The volume of harvested trees does not cover trees harvested before approx. 1986. Hence, the total harvested volume is underestimated, resulting in a small overestimation of the indicator value.

3. **Interpretation**:
Recorded dead wood volume as a share of the volume of all wood is compared to a national reference level for productive forest, based on expert assessments supported by empirical data for near-natural forests and knowledge of ecosystem dynamics in intact forest ecosystems [@storaunet-skarpaas-2022].
The lower reference level _X~0~_ is 0% and the upper reference level _X~100~_ is 33% for the share of dead wood of all wood.
The empirical data in @storaunet-skarpaas-2022 is generally from productive forest. Since the NFI-data of the dead wood share within old or near-natural forests categories seems similar between unproductive and low-productive forest plots, we assume the same _X~100~_ reference value of 33% for unproductive forest.
Thus, the indicator represents all forest land. 

4. **Results and development needs**:
For the current reporting period (2020-2024) the national indicator value is 0.29, with similar values for most regions, except for region North with a value of 0.45. The indicator values are rather stable over time, except for small increases for regions West and North. There are no specific development needs for this indicator, except that a thorough review of reference values and the scaling function is needed.


### 1.2 Native language summary - Norwegian

**Bakgrunn og begrunnelse**:
Volumet av død ved som andel av volumet av all levende, død og avvirket ved er et robust mål på naturnærhet i skog.
Død ved er en nøkkelressurs som substrat eller næring for et bredt spekter av skogorganismer, særlig sopp og virvelløse dyr. Forventet volum av både død og levende ved i intakt skog varierer med bonitet, dominerende treslag og klimaforhold.
Ved å sammenligne volumet av død ved med volumet av all ved for samme geografiske område, kan virkningen av produktivitet, treslag og klima delvis utlignes. 

**Metoder**:
Data for indikatoren kommer fra Landsskogtakseringen (LSK), der volum av levende og stående døde trær samt kumulativt volum av hogd virke er registrert for hver prøveflate siden midten av 1990-tallet, og liggende død ved siden 2010 (etter gjeldende metode).
Indikatoren er volumandelen død ved utgjør av all levende, avvirket og død ved på hver prøveflate.
For hver rapporteringsperiode beregnes arealvektede gjennomsnitt for fylker, regioner og nasjonalt nivå med au_areal.
Avvirket ved dekker ikke trær hogd før ca. 1986, så indikatorverdien kan være noe for høy. 

**Tolkning**:
Registrert volum av død ved som andel av volumet av all ved sammenlignes med en nasjonal referanseverdi for produktiv skog: nedre referansenivå _X~0~_ er 0 % og øvre referansenivå _X~100~_ er 33 %.
De empiriske bakgrunnsdataene for å fastsette referanse-verdier [@storaunet-skarpaas-2022] er hovedsakelig fra studier i produktiv skog.
Siden LSK-data av dødved-andel i ulike arealkategorier av lite påvirket skog viser omtrent samme nivåer i uproduktiv og lavproduktiv skog, antas samme referanseverdier også i den uproduktive skogen. Indikatoren er derfor beregnet for alt skogareal. 

**Resultater og utviklingsbehov**:
For den nåværende rapporteringsperioden (2020–2024) er den nasjonale indikatorverdien 0.29, med lignende verdier for de fleste regionene, bortsett fra region Nord som har en høyere indikatorverdi (0.45).
Indikatorverdiene er nokså stabile over tid, bortsett fra en svak økning for regionene Vest og Nord.
Det er ingen spesifikke utviklingsbehov for denne indikatoren, bortsett fra at en grundig gjennomgang av referanseverdier og skaleringsfunksjonen er nødvendig. 

## 2. About the underlying data

The volume of living trees and standing and lying dead wood > 10 cm in diameter is recorded for each NFI inventory plot [@viken2021-feltinstruks].
The NFI plot data also include a measure of the cumulative volume of trees harvested since the establishment of permanent inventory plots around 1990.
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an area-weighted mean of the plot-level share of dead wood volume to the volume of all living, harvested and dead wood.
Only plots with complete volume data are included in these means; plots with missing dead wood, living wood or harvested wood volume are excluded before aggregation. Results are reported for counties, regions and nationally.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI. 

### 2.1 Spatial and temporal extent, resolution, and representativity

Data are available from 2010, when the volume of lying dead wood was first recorded based on current methods, and the NFI grid was extended to cover all forest land.
A different method for quantifying the volume of dead wood was applied during the 1994-1998 forest inventory.
Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010.
The data are representative for all productive forest land of Norway. Aggregated reporting uses counties, regions and the whole country. 

### 2.2 Sampling design
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
For each NFI plot, all living and dead standing trees with a diameter at breast height of at least 5 cm are recorded, with additional variables for each tree allowing estimates of its volume [@viken2021-feltinstruks].
For this indicator only the volumes of living and standing dead trees of at least 10 cm diameter at breast height are included.
Within each NFI plot, items of lying dead wood at least 10 cm in diameter and 1 m long are recorded along four transects in the cardinal directions, allowing estimates of lying dead wood volume.
For each NFI plot, the estimated volumes for standing and lying dead wood at least 10 cm in diameter are combined.
At each inventory, the removal of living and standing dead trees since the previous inventory is recorded, and their volumes calculated.
The cumulative volume of the harvested trees is aggregated since permanent plots were established ca. 1990. 

### 2.3 Original units

The indicator is specified as the share of the volume of dead wood to volume of all living, harvested and dead trees.
At plot level this is a proportion; aggregated values are area-weighted means of that proportion. 

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The indicator represents particularly important structures for species dependent on dead wood and other characteristics of intact forests.
Hence, this indicator represents a B2 Biotic structural characteristic. 

### 3.2 Ecosystem condition characteristic


### 3.3 Collinearities with other indicators

Some correlation with other indicators for old or intact forest, such as biologically old forest, may be expected.
The inherent correlation with the indicator dead wood volume seems far lower than might be expected. 

### 3.4 Impact factors

Natural disturbances at different temporal and spatial scales, such as fires, floods, windthrows, snow breaks, attacks by pathogens, or tree senescence, will contribute to the production of dead wood.
Forestry may produce dead wood of small dimensions and coarser dead wood to a limited degree. 
Empirical data show a negative relationship between the share of dead wood to all living, harvested and dead wood and the number of stumps per hectare, as a measure of logging activity [@storaunet-skarpaas-2022].
Forest fire prevention and salvage logging after fires or windthrows reduce the availability of dead wood.
Human climate change may increase the decomposition rate of dead wood, whereas increased frequency and severity of storms will increase production of dead wood.

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each reporting period, non-scaled values for county (fylke), region, and Norway are obtained by pooling all included forest plots in the unit: the area-weighted mean of the plot-level dead wood share.
Where NFI county labels are split into parts, those labels are merged to a single county name before aggregation.
Only plots with complete dead wood, living wood and harvested wood volume enter the aggregation.
Scaling and truncation are applied after non-scaled county aggregates have been calculated: each county is scaled against the national lower reference level _X~0~_ (0% share) and upper reference level _X~100~_ (33% share) using a linear scaling function, and values above 1 are truncated to 1 at that step.
Regional and national scaled point estimates are area-weighted means of the already scaled county values, using each county's represented area as weight.
Non-scaled regional and national point estimates are calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

The reference value is based on expert assessments of the relationship between the indicator and measures of logging impact, based on reported dead wood data for a range of site conditions and dominating tree species [@storaunet-skarpaas-2022; and references therein].
A proportion of 1/3 dead wood to all living, harvested and dead wood (by volume) is judged to be a robust measure for forests with only minor impact from forestry over the last 100 years for a range of site conditions and dominating tree species.
Reference values are expert assessments without estimated uncertainty.
Hence, the lower reference level _X~0~_ is set to 0% and the upper reference level _X~100~_ is set to 33% for the share of dead wood to all wood.
The national reference levels used in the analyses are listed in @sec-dtvs-reference-values.
The same _X~0~_ and _X~100~_ are applied when scaling each county before regional and national aggregation.
For lack of better knowledge, the relationship between indicator values and ecosystem condition is assumed to be linear for values between _X~0~_ and _X~100~_.
Hence, indicator values are scaled between 0 and 1 based on a linear scaling function between _X~0~_ and _X~100~_; scaled indicator values above 1 are truncated to 1.

#### 4.2.1 Spatial resolution and validity

The reference value is given as an overall national value for the share by volume of dead wood to all living, harvested and dead wood.
Although indicator values from NFI plots may technically be aggregated at finer spatial units than counties, scaling indicator values at such finer spatial levels may result in biases.
County, regional and national reporting use the merged county geography described in section 3.5.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability of the volume of living, harvested and dead wood among NFI plots and to the measurements of this volume for each NFI plot.
There is also uncertainty related to the setting of the upper reference value and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties are not quantified.

### 5.2 Reported uncertainties

The uncertainties reported in  @sec-results are inferential and refer to uncertainty around the point estimates.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within each county and recalculating the area-weighted mean dead wood share.
For scaled county values, bootstrap draws are transformed using the same national reference levels _X~0~_ and _X~100~_ as the point estimates (truncation at scaling).
For scaled regional and national values, uncertainty is propagated from the scaled county bootstrap draws.
The resulting distributions are summarised with median and quantile intervals.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Living, harvested and dead wood volume per hectare

The volumes of living and dead wood per hectare are quantified for each NFI plot according to methods described in [@viken2021-feltinstruks].
Harvested volume within each plot is quantified as the cumulative volume of harvested trees since the establishment of the permanent NFI-plots in the late 1980s.
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used.
The plot-level indicator is the share of dead wood volume to the volume of all living, harvested and dead wood.


```{r load-data}
#| code-summary: "Load data"

# Load data file
lsk_data_file <- here("data/LSK_data_okologisk_tilstand_2026-06-03.csv")
aats_cache_deps <- c(
  lsk_data_file,
  here::here("R/functions.R"),
  here::here("R/NO_DWTO_001.qmd")
)

lsk_data_raw <- read_csv(
  lsk_data_file,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties, regions and nationally.
Where counties are split into parts, those parts are merged to a single county name before aggregation.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-dtvs-derived-vars}

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

We filter the raw dataset to include both productive and unproductive forest plots (`arealtype` 1 and 12), remove records missing the required volume variables, standardise county names by merging split NFI labels to county level.
We compute the share of dead wood volume relative to all wood volume.
The indicator is calculated as `dw_prha / (dw_prha + vm10_prha + kum_avvirka)`, where `dw_prha` is dead wood volume per hectare, `vm10_prha` is living wood volume per hectare, and `kum_avvirka` is cumulative harvested wood volume per hectare.
For each included plot, the area-weighted contribution to later aggregation is `dead_wood_share * au_areal`.
Only plots with complete volume data remain in the dataset used for county, regional and national aggregation.

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    arealtype %in% c(1, 12)
  ) |>
  filter(!is.na(dw_prha), !is.na(vm10_prha), !is.na(kum_avvirka)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    fylke_name = normalize_county_parts(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    
    # Calculate total wood volume (dead + living + harvested)
    total_wood_volume = dw_prha + vm10_prha + kum_avvirka,
    # Indicator value: proportion of dead wood volume to total wood volume
    dead_wood_share = if_else(total_wood_volume > 0, dw_prha / total_wood_volume, 0),
    dead_wood_share_weighted = dead_wood_share * au_areal,
    indicator_continuous = dead_wood_share,
    indicator_continuous_weighted = dead_wood_share_weighted
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    fylke_name,
    fylke_name_clean,
    dw_prha,
    vm10_prha,
    kum_avvirka,
    total_wood_volume,
    dead_wood_share,
    dead_wood_share_weighted,
    indicator_continuous,
    indicator_continuous_weighted
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values {#sec-dtvs-non-scaled}

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.

```{r calculate-dtvs-all-periods}
#| code-summary: "Calculate non-scaled values"

 dtvs_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 dtvs_results_county_all_periods   <- purrr::map_dfr(dtvs_results, "county")
 dtvs_results_region_all_periods   <- purrr::map_dfr(dtvs_results, "region")
 dtvs_results_national_all_periods <- purrr::map_dfr(dtvs_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted non-scaled dead wood share, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(here::here("data/LSK_data_okologisk_tilstand_2026.csv")), file.mtime(here::here("R/NO_DTVS_001.qmd")))

county_draws_long <- purrr::imap_dfr(dtvs_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-dtvs-county-nonscaled) shows the direct area-weighted mean dead wood share per county for all reporting periods, with approximate 95% bootstrap confidence intervals (2.5th–97.5th percentile of 1000 draws).

```{r tbl-dtvs-county-nonscaled}
#| tbl-cap: "Non-scaled dead wood volume share (%) by county for all reporting periods. Point estimate is the direct area-weighted mean; 95% CI is derived from 1000 bootstrap draws."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, region_code, total_area) |>
  dplyr::summarise(
    ci_low  = round(quantile(non_scaled_value, 0.025) * 100, 2),
    ci_high = round(quantile(non_scaled_value, 0.975) * 100, 2),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dtvs_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots, indicator_value),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    mean_pct = round(indicator_value * 100, 2),
    `95% CI` = paste0(ci_low, " – ", ci_high)
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Dead wood share (%)` = mean_pct,
    `95% CI`,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Dead wood share (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-dtvs-reference-values}

National reference levels are _X~0~_ = 0% and _X~100~_ = 33% for the share of dead wood volume to all wood volume.

```{r dtvs-reference-values}
#| code-summary: "National reference levels"

x0 <- 0
x100 <- 0.33


```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, _X~0~_ = 0, _X~100~_ = 33%; truncation at scaling).
@fig-dtvs-norm shows the mapping.

```{r county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dtvs_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-dtvs-norm}
#| fig-cap: "Linear normalisation of dead wood volume share (proportion of all wood volume) to indicator values (0–1), using national reference levels _X~0~_ = 0% and _X~100~_ = 33%. Points are 500 sampled bootstrap draws, pooled across all reporting periods."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 5
#| fig-width: 6

norm_plot_data <- county_draws_scaled |>
  dplyr::slice_sample(n = min(500, nrow(county_draws_scaled)))

dtvs_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::labs(
    x = "Dead wood share (proportion of all wood volume)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
dtvs_norm_plot$layers[[1]]$aes_params$size <- 3

suppressMessages(
  ggsave(
    filename = here::here("img/dtvs_norm_plot.png"),
    plot = dtvs_norm_plot,
    scale = 1.5
  )
)

dtvs_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r dtvs-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

dtvs_mc_cache_deps <- c(
  here::here("data/LSK_data_okologisk_tilstand_2026.csv"),
  here::here("R/functions.R"),
  here::here("R/NO_DTVS_001.qmd")
)
dtvs_mc_rds_regional <- here::here("data/dtvs_mc_regional.rds")

if (
  file.exists(dtvs_mc_rds_regional) &&
  file.mtime(dtvs_mc_rds_regional) >= max(file.mtime(dtvs_mc_cache_deps), na.rm = TRUE)
) {
  dtvs_mc_regional <- readRDS(dtvs_mc_rds_regional)
} else {
  dtvs_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(dtvs_mc_regional, dtvs_mc_rds_regional)
}
```

```{r dtvs-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

dtvs_mc_regional_for_national <- dtvs_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

dtvs_mc_rds_national <- here::here("data/dtvs_mc_national.rds")

if (
  file.exists(dtvs_mc_rds_national) &&
  file.mtime(dtvs_mc_rds_national) >= max(
    c(file.mtime(dtvs_mc_cache_deps), file.mtime(dtvs_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  dtvs_mc_national <- readRDS(dtvs_mc_rds_national)
} else {
  dtvs_mc_national <- ecTools::ec_upscale(
    data        = dtvs_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(dtvs_mc_national, dtvs_mc_rds_national)
}
```

```{r dtvs-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

dtvs_mc_combined <- dplyr::bind_rows(
  dtvs_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  dtvs_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- dtvs_results_county_all_periods |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- dtvs_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- dtvs_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- dtvs_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    dtvs_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- dtvs_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    dtvs_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  dtvs_results_county_all_periods |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = 33,
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      dtvs_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      dtvs_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county level, truncation is evaluated by comparing the non-scaled point estimate to _X~100~_ (33%).
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r dtvs-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_pct, original_value_pct) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded _X~100~_ (33%) and required truncation.\n")
}

```

The time series below (@fig-dtvs-timeSeries) summarises regional and national values.

```{r fig-dtvs-timeSeries}
#| fig-cap: "Indicator values for dead wood volume share by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

dtvs_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
dtvs_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
dtvs_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(dtvs_region_levels)),
  dtvs_region_levels
)

dtvs_plot_data <- dtvs_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!dtvs_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = dtvs_region_levels))

ggplot2::ggplot(
  dtvs_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = dtvs_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to _X~0~_ (0% dead wood share) and 1 corresponds to _X~100~_ (33% dead wood share).

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-dtvs-county-results to browse county rows across all reporting periods.

```{r tbl-dtvs-county-results}
#| tbl-cap: 'County indicator values for dead wood volume share by reporting period. Values are scaled to 0–1 using national _X~0~_ and _X~100~_. Uncertainty is derived from 1000 bootstrap draws. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Dead wood volume share",
    ID = "NO_DTVS_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-dtvs-region-results}
#| tbl-cap: 'Regional indicator values for dead wood volume share by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Dead wood volume share",
    ID = "NO_DTVS_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-dtvs-national-results}
#| tbl-cap: 'National indicator values for dead wood volume share by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Dead wood volume share",
    ID = "NO_DTVS_001",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`dtvs_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  dtvs_mc_combined,
  here::here("data/results_NO_DTVS_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [KOS, EF]
* Methodology: [KOS, EF, AHH]
* Data curation: [OS]
* Formal analysis: [AHH]
* Validation: [AHH, EF, KOS]
* Visualization: [AHH]
* Writing – original draft: [AHH, EF]
* Writing – review & editing: [All]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info
This workflow uses `renv`. Run `renv::restore()` to recreate the working environment. 
:::