---
title: "Functional plant community indicators"
subtitle: |
  [NO_FUNC_001] (farmland and open lowland) \
  [NO_FUNC_002] (mountains) \
  [NO_FUNC_003] (wetlands) \
  [NO_FUNC_004] (forests)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Joachim Töpper
    email: joachim.topper@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research
  - name: Erik Framstad
    affiliations:
      - ref: myID               # To reuse affiliations referecen the id like this
date: March 4, 2025 # Enter date 
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
source(here::here("_common.R"))
```

```{r}
#| echo: false
meta <- readxl::read_xlsx("../metadata.xlsx")
st <- meta |>
  filter(Variable == "status") |>
  pull(Value)
version <- meta |>
  filter(Variable == "Version") |>
  pull(Value)
auth <- meta |>
  filter(Variable == "authors") |>
  pull(Value)
year <- meta |>
  filter(Variable == "yearAdded") |>
  pull(Value)
id <- meta |>
  filter(Variable == "indicatorID") |>
  pull(Value)
name <- meta |>
  filter(Variable == "indicatorName") |>
  pull(Value)
url <- meta |>
  filter(Variable == "url") |>
  pull(Value)

meta <- meta |>
  mutate(Variable = case_match(Variable,
    "indicatorID" ~ "Indicator ID" ,
    "indicatorName" ~ "Indicator Name",
    "country" ~ "Country",
    "continent" ~ "Continent",
    "ECT" ~ "Ecosystem Condition Typology Class",
    "yearAdded" ~ "Year added",
    "yearLastUpdate" ~ "Last update",
    .default = Variable
   )
  ) |>
  filter(Variable != "authors")

```

<!--# The following parts are autogenerated. Do not edit. -->

```{r}
#| echo: false
#| results: asis
status(st)
```

::: {layout-ncol="2"}



> **Recomended citation**: `r paste(auth, " ", year, ". ", name, " (ID: ", id, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

> **Version**: `r version`

:::

```{=html}
<details>
<summary>Show metadata</summary>
```

```{r tbl-meta}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false

meta |>
  select(Variable, Value) |>
  kbl(col.names = NULL) 

```

```{=html}
</details>
```

::: {.callout-tip collapse="true"}

## Logg

<!--# Update this logg with short messages for each update -->
- 01 Jan. 1901 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
Functional plant community indicators represent the response of the vascular plant species composition to environmental variables such as available temperature, moisture, nitrogen, light, pH, soil disturbance, and others and can thus be used to describe the functional signature of plant communities by calculating community-weighted means of functional trait values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. For different ecosystems different environmental variables are subject to change and thus relevant for condition accounting. In principal, both too much and too little of any environmental resource can represent a challenge leading to shifts in plant species composition. Thus, functional plant community indicators are two-sided indicators that can show deviations of a plant community towards both excess or limiting amounts of a resource.

Here, we build functional plant community indicators by combining field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) with functional plant indicator values from Tyler et al. (2021). We calculate the functional signatures of plant communities in the monitored sites with respect to various environmental variables depending on ecosystem type (not all variables are relevant for all ecosystem types). These functional signatures are then compared to reference distributions of functional signature, separately for each ecosystem type, calculated from ‘generalized species lists’ underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These functional plant community indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

For farmland and open lowland ecosystems, the main challenges today are abandonment, climatic changes, pollution, and recreational overuse. Therefore, we use functional plant indicator values for light, moisture, pH, nitrogen, phosphorus, grazing/mowing, and soil disturbance. 
For mountain ecosystems, the main challenges today are climatic changes, species range shifts, and pollution. Therefore, we use functional plant indicator values for nitrogen, light, and heat requirement.
For wetland ecosystems, the main challenges today are drainage, shrub/tree encroachment and pollution. Therefore, we use functional plant indicator values for moisture, light, pH, and nitrogen
For forest ecosystems, the main challenges today are climatic changes and pollution. Therefore, we use functional plant indicator values for moisture and nitrogen.
Note, that these indicators do not address the issue of loss of area, which is the overarching and dominating reason for nature loss both globally and in Norway. Thus, the functional plant community indicators do not cover this pressure.

## 2. About the underlying data
We use three sets of data for building functional plant community indicators for ecological condition:

- as test data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- Swedish functional plant indicator data from Tyler et al. (2021)

(1) **ANO monitoring data**: ANO stands for areal-representativ naturovervåking, i.e. area representative nature monitoring. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN). In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) **NiN reference data**: The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the ‘expected combination of frequency and cover’ of occurrence

1 = < 1/32

2 = 1/32 - 1/8

3 = 1/8 - 3/8

4 = 3/8 - 4/5

5 = 3/8 - 4/5 + dominance

6 = > 4/5

For the purpose of this project, these steps are simplified to maximum expected combination of frequency and cover, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(3) **The Swedish plant indicator values** published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we use indicator data for light, heat requirement, moisture, pH, nitrogen, phosphorus, grazing/mowing, and soil disturbance as these are thought to be subject to potential change due to ongoing pressures in the respective ecosystems (see details above under 'Summary').

### 2.1 Spatial and temporal resolution and extent

The ANO monitoring data available now represent one cycle of sampling from 2020-2024 comprising ca. 1000 sites across Norway.
ANO has no defined end date. 2025 is the first year of repeated sampling for about 20% of the approximately 1000 ANO sites for all of mainland Norway (i.e. the Ecosystem Accounting Area). Annual sampling is spatially representative, allowing representative annual estimates. 
Thus, at this point there is no actual time series to these data yet, and the indicator evaluation does therefore not include any temporal analyses.

### 2.2 Original units

Abundance data for vascular plants in ANO is given as % cover within a 1 m2 sample quadrat. This is transformed to relative abundance data per quadrat.  Each ANO quadrat is classified to its respective ecosystem type according to ‘Nature in Norway’ (NiN; https://artsdatabanken.no/NiN). Environmental indicator values from Tyler et al. (2021) are given on an ordinal scale from 1 (lowest affinity to the respective variable) to n (highest affinity to the respective variable), the number of levels varying between environmental indicators.

### 2.3 Additional comments about the dataset

none

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

All functional plant community indicators reflect the plant species composition and thus represent ECT class 'B1 Biotic compositional characteristics'.
Through the coupling of compositional data to environmental indicator values, the functional plant community indicators in addition represent the following ECT classes:
Functional plant community indicators for nitrogen, phosphorus and pH reflect the availability of nutrients and soil chemical properties and thus represent the ECT class 'A2 Chemical characteristics'.
Functional plant community indicators for heat requirement, moisture and light reflect the availability of abiotic, physical environmental resources and thus represent the ECT class 'A1 Physical characteristics'.
Functional plant community indicators for grazing/mowing and soil disturbance reflect the availability of physical structuring conditions  and thus represent the ECT class 'A1 Physical characteristics'.
 
### 3.2 Ecosystem condition characteristic

Functional plant community indicators in general are aimed at describing to which degree ecosystems fulfill the criterion 'Intact plant community', i.e. to that degree they are in good condition with respect to a plant community composition reflecting the functional properties of the respective ecosystem in the reference condition.

The single, specific functional plant community indicators represent concrete functional properties of an ecosystem and deviations from the reference thus shed light on which drivers affect the plant community.

For instance, the functional plant community indicator on nitrogen represents the influence of nitrogen on the plant community composition in the given ecosystem. Thus, deviations from the reference community towards e.g. higher nitrogen values reflect an environment characterized by higher availability of nitrogen which has led to shifts in plant species composition towards higher abundances of species with higher affinity to nitrogen. This functional plant indicator for nitrogen reflects both the chemical properties of the ecosystem but also the ecosystem’s productivity. Hence, the indicator represents the ecosystem characteristics of both Abiotic factors and Productivity.

### 3.3 Other standards

None

### 3.4 Collinearities with other indicators

Different functional plant community indicators for different environmental variables (light, moisture, ...) might show a certain degree of co-linearity with one another. Therefore, any aggregation of scaled functional plant community indicators to one single functional plant community indicator index should rely on using the indicator with worst scaled value and not on for instance an average of all indicators.

Functional plant community indicators can potentially exhibit co-linearities with other vegetation oriented indicators, such as NDVI, though this depends greatly on which functional plant community indicator ends up as the leading one (i.e. the one with the lowest scaled score).

### 3.5 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->
Plant community composition is sensitive to consistent directional changes in environmental drivers, but the main pressures may vary between ecosystems.

Farmlands and open lowland ecosystems have been kept open by historic human activities and land use. With land-use change, intensivation of certain agricultural areas and abandonment of many other, encroachment of shrubs and trees is one of the main challenges today for anthropogenically open systems which change the plant community composition in a directional way. In addition, climatic changes, pollution and wear and tear (for instance from recreational overuse) alter the environmental conditions for both human influenced and naturally open ecosystems directionally, for instance through increased precipitation, drought spells and high critical loads of nitrogen deposition. 

Mountain ecosystems are characterized as limited by cold temperatures and usually low availability of nutrients. The main challenge today are thus climate warming and climate-warming related species range shifts (which introduce strong changes in biotic interactions between species) and pollution (e.g. nitrogen deposition) which affect the plant community composition in a directional way.

Wetland ecosystems are primarily characterized by high moisture and are thus threatened by drainage and shrub/tree encroachment, which both change the plant community composition dramatically. In addition, pollution can cause directional shifts in plant communities (e.g. nitrogen deposition).

The plant community composition in forest ecosystems is subject to changes with climate warming and precipitation changes as well as pollution.

Also alien species represent a major threat to all ecosystems, and this pressure may cause shifts in functional plant community signatures as found for naturally open systems around the Oslofjord (Evju et al. 2024).
## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be a plant community composition representing natural and undisturbed conditions (for natural ecosystems) or moderately disturbed, human influenced conditions in e.g. semi-natural grasslands under moderate long-term traditional grazing or mowing land management (but note that the plant species composition will vary between grazed and mowed grasslands). The reference condition for functional plant community indicators is represented by the generalized species lists for ecosystems as described in the Norwegian ecosystem categorization system NiN (Halvorsen et al. 2020).

### 4.2 Reference levels

<!--# 

If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ referes to the referece level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.

 -->


#### 4.2.1 Spatial resolution and validity

<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->

## 5. Uncertainties

<!--# Describe the main uncertainties or sources of error in the indicator or the underlying data. -->

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

## 7. Datasets

<!--# Describe the unique datasets seperately under seperate headers (Dataset A, Dataset B, etc.-->

### 7.1 Dataset A

<!--# Describe the main dataset, typicaly the one containing the variable of (most) interest. Change the header from Dataset A to the name of the actuall dataset. -->

### 7.2. Dataset B

<!--# Describe additional datasets in a similar was as above. Deleteor add ned subheaders as needed. -->

## 8. Spatial units

<!--# 

Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpretted at. Potential spatial delineation data should eb introduced under 7.1. Datasets. We recomend using the SEEA EA terminology opf Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). 

-->

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in seperate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a seperate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```
