---
title: "PIT (periphyton index of trophic status)"
subtitle: "[NO_WPIT_001]"
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research
date: April 25, 2025 
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
knitr::opts_chunk$set(echo = TRUE)

library(foreign)
library(sf)
library(readxl)
library(httr)
library(jsonlite)
```

```{r source}
#| echo: false
source(here::here("_common.R"))
```

```{r}
#| echo: false
meta <- readxl::read_xlsx("../metadata.xlsx")
st <- meta |>
  filter(Variable == "status") |>
  pull(Value)
version <- meta |>
  filter(Variable == "Version") |>
  pull(Value)
auth <- meta |>
  filter(Variable == "authors") |>
  pull(Value)
year <- meta |>
  filter(Variable == "yearAdded") |>
  pull(Value)
id <- meta |>
  filter(Variable == "indicatorID") |>
  pull(Value)
name <- meta |>
  filter(Variable == "indicatorName") |>
  pull(Value)
url <- meta |>
  filter(Variable == "url") |>
  pull(Value)

meta <- meta |>
  mutate(Variable = case_match(Variable,
    "indicatorID" ~ "Indicator ID" ,
    "indicatorName" ~ "Indicator Name",
    "country" ~ "Country",
    "continent" ~ "Continent",
    "ECT" ~ "Ecosystem Condition Typology Class",
    "yearAdded" ~ "Year added",
    "yearLastUpdate" ~ "Last update",
    .default = Variable
   )
  ) |>
  filter(Variable != "authors")

```

<!--# The following parts are autogenerated. Do not edit. -->

```{r}
#| echo: false
#| results: asis
status(st)
```

::: {layout-ncol="2"}



> **Recomended citation**: `r paste(auth, " ", year, ". ", name, " (ID: ", id, ") ", "v. ", version, ". ecRxiv: https://view.nina.no/ecRxiv/", sep="")`

> **Version**: `r version`

:::

```{=html}
<details>
<summary>Show metadata</summary>
```

```{r tbl-meta}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false

meta |>
  select(Variable, Value) |>
  kbl(col.names = NULL) 

```

```{=html}
</details>
```

::: {.callout-tip collapse="true"}

## Logg

<!--# Update this logg with short messages for each update -->
- 22 Apr 2025 - Version 000.001 started
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

Freshwater ecosystems are negatively affected by anthropogenic activities that change the chemical composition of the water, such as pollution or run-off causing eutrophication. 
PIT ("periphyton index of trophic status") is a measure of the sensitivity of *periphyton* to the availability of phosphates, where "periphyton" is an umbrella term for algae and microorganisms growing on submerged surfaces. 
PIT is estimated by calculating the average sensitivity scores of the periphyton species sampled in a given location ([Schneider & Lindstrøm 2011](#sch11)). 
Rivers characterised by periphyton communities with low sensitivity to high phosphate concentration indicate a functional shift towards more disturbed ecosystems.
In this way, PIT detects eutrophication. 

PIT is one of the indicators that is measured and reported in connection with the [*Water Framework Directive*](#wfd) (hereafter **WFD**). 
In Norway, the WFD is implemented by means of [*Vannforskriften*](#vannf), and relevant indicators ("parameters", in WFD terminology) are described in national guidelines ([DGVF 2025](#veil2)). 
The data collected are archived and made available in the [*Vannmiljø*](https://vannmiljo.miljodirektoratet.no/) database. 

These WFD data can be used in other contexts, including ecological condition assessments. 
The procedure implemented here follows a dataflow that has been described and documented earlier ([Sandvik 2019](#san19), [Sandvik 2024](#san24)) and that consists of the following steps:

* Selecting measurements from the relevant time period.
* Removing measurements that do not fulfil the requirements for the specific indicator.
* Identifying the water body for each measurement.
* Scaling and transforming measurements to "modified Ecological Quality Ratios" according to the normative definitions for "ecological status" categories provided by the WFD.
* Fitting a linear model to the measurements, with explanatory variables being typology factors of the water bodies, reporting periods, and surveillance activities.
* Averaging measurements taken in the same water body.
* Extrapolating mEQR values to water bodies that lack data, based on the model fitted.
* Using simulations to estimate confidence and prediction intervals.
* Aggregating estimates geographically using area-weighted averages.
* Truncating the values obtained to the interval from 0 to 1.

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of commmon language summary. It is a good idea to include a bullet point list of the spesific steps in the workflow. Include a mention of the following aspects:

What does the metric represent?
Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenig impact factors?
What kind of data is used? 
Shortly, how is the data customized (modified, estimated, integarted) to fit its purpuse as an indicator?
What is the current status of  the metric (can it be used or is it still in development)?
How should the metric be used and interpretted, and how should it not be used/interpretted?
-->


## 2. About the underlying data

WFD data in Norway are collected in a common database called [*Vannmiljø*](https://vannmiljo.miljodirektoratet.no/), which is maintained by the [Norwegian Environment Agency](https://www.miljodirektoratet.no/). 
This database contains information on all *measurements* taken in connection with the WFD ("registrations") as well as all *locations* at which measurements are taken ("water localities"). 
Each water locality is assigned to a *water body*. 
Information on water bodies can be retrieved from two other databases: the typology of water bodies is available from [*Vann-nett*](https://vann-nett.no/), whereas a geospatial dataset can be [downloaded](https://karteksport.miljodirektoratet.no/) from the Norwegian Environment Agency.

All data used are publicly available, but not all of them are machine-readable (for details, see [Datasets](#datasets) below).

<!--# Describe the data you have used in more detail, it's origin, biases, availability ect.-->

### 2.1 Spatial and temporal resolution and extent

The **spatial** resolution of WFD indicators is on the level of *water bodies*. 
The WFD defines a water body (specifically a "body of surface water") as "a discrete and significant element of surface water such as a lake, a reservoir, a stream, river or canal, [or] part of a stream, river or canal" (WFD, Article 2, paragraph 10; cf. *Vannforskriften* §&nbsp;3, letter a; for implementation, see [DGVF 2018](#veil1), ch. 2).

There is no specific **temporal** resolution, as measurements of WFD indicators are reported sporadically. 
Some water bodies are sampled regularly (e.g., every fifth year). 
For most water bodies, however, no such fixed surveillance schemes exist. 
Still, the total amount of measurements for many WFD indicators is sufficient to estimate annual values that are spatially aggregated to the national and regional level.

<!--# Describe the temporal and spatial resolution and extent of the data used -->

### 2.2 Original units

PIT is calculated as a dimensionless score with possible values between 1.92 and 68.91, where sensitivity to eutrophication increases with increasing scores.
The score of a species represents *ten times the natural logarithm of the average total phosphorus*, measured in µg/L, at the sites where the species occurs ([Schneider & Lindstrøm 2011](#sch11)). 
However, in the connection with the WFD, values are normally reported as "normalised ecological quality ratios" (nEQR) between 0 and 1, which corresponds to *indicator* values sensu [SEEA EA](#seeaea).

<!--# What are the original units for the most relevant  variables in the data-->

<!--# ### 2.3 Additional comments about the dataset

Text here -->


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The ECT of PIT is **A2** (chemical state) because the indicator measures the effect of organic water pollution on the periphyton community.

**B1** (compositional state) may be regarded as a *secondary* ECT class of PIT, as PIT is estimated based on the set of periphyton species present in the locality sampled, which is a measure of *biodiversity*.

<!--# 
Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. 
-->

### 3.2 Ecosystem condition characteristic

Intact ecosystems are characterised by having no organic pollution, and PIT represents the organic pollution of rivers.

<!--# 
Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decition Making.  
-->

### 3.3 Other standards

In the Norwegian classification of ecological characteristics (*økologiske egenskaper*) according to Nybø & Evju ([2017](#nyb17), p. 35), the indicator represents **abiotic conditions** (*abiotiske forhold*).

In the WFD's distinction between **quality elements**, the indicator represents the **aquatic flora** (WFD, Annex V).

The WFD enforces the environmental **objectives** that water bodies shall not have an indicator value below X~60~, and that water bodies already obtaining an indicator value above X~80~ shall not drop below this threshold (WFD, Article 4; cf. *Vannforskriften* §&nbsp;4).

<!--# Optional: Add text about other spesific standards, e.g. national standards, and how the indicator relates to these -->

### 3.4 Collinearities with other indicators

PIT is correlated to total phosphorus (Ptot) and may be broadly correlated to other measures of river eutrophication, specifically ASPT.
<!--# these indicator abbreviations can be linked to the appropriate fact sheets at a later stage -->

<!--# Describe known collinearities with other metrices (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

### 3.5 Impact factors

PIT is affected by eutrophication and organic pollution, which can be the result from local or up-stream run-off (e.g. from fertilisers), direct pollution, or long distance pollution.

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->


## 4. Reference condition and levels

### 4.1 Reference condition

The WFD defines the reference condition of a water body as "the condition of a quality element where there is little or no human impact on the water body. The reference condition is therefore also referred to as the natural condition" ([DGVF 2025](#veil2)).

Reference conditions may be specific to certain water-body types. 
Water bodies judged to be in the reference condition are referred to as *reference water bodies*.

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels

According to the WFD, reference levels are defined for X~0~, X~20~, X~40~, X~60~, X~80~, and X~100~. 
These are used to classify the "ecological status" of water bodies as "bad" (X~0~ <= X <= X~20~), "poor" (X~20~ < X <= X~40~), "moderate" (X~40~ < X <= X~60~), "good" (X~60~ < X <= X~80~), and "high" (X~80~ < X <= X~100~). 
The WFD (Annex V, section 1.2) provides the following normative definition of "high status": 
"The values of the biological quality elements for the surface water body reflect those normally associated with that type under undisturbed conditions, and show no, or only very minor, evidence of distortion."
An approach used for many WFD indicators, including this one, is to define X~100~ as the median of the variable values measured in reference water bodies.

Reference levels for PIT are provided in Table 5.1a of DGVF ([2025](#veil2)). 
A more detailed description of the calculation of AIP can be found in Table V5.1.1. 

According to these tables, possible variable values range from 1.92 to 68.91, corresponding to a continuum from maximum eutrophication tolerance to maximum eutrophication sensitivity of the periphyton sampled. 

<!--# 
If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ refers to the reference level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.
-->

#### 4.2.1 Spatial resolution and validity

The spatial resolution of WFD reference levels is the water body. 
Reference levels of AIP differ between river water types. 
Depending on the natural alkalinity (calcium concentration), river water bodies are grouped in two categories, each with their own set of reference levels ([DGVF 2025](#veil2), Table 5.1a).

<!--# 
Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 
-->


## 5. Uncertainties

The main uncertainty for national and regional values of the indicator is *sampling error*. 
Measurements of variable values are only sampled in a small fraction of all relevant water bodies. 
The approach below extrapolates indicator values to water bodies that have not been sampled, but this extrapolation is necessarily uncertain. 
This uncertainty is quantified using prediction intervals.

The underlying problem of the mentioned uncertainty is that the *representativeness* of the sampled water bodies for the non-sampled water bodies is unknown. 
A specific issue regarding representativeness is that quite some WFD measurements are taken *precisely because the sampled water body is known or assumed to be unrepresentative*. 
The main reasons are the following:

* Some water bodies are sampled because it is known or suspected that they do not fulfil the normative objective that X > X~60~ (in order to justify and design measures to improve ecological condition and, later, to evaluate the effect of the measures taken).
* Some water bodies are sampled because they are regarded as reference water bodies, i.e. assumed to be in, or very close to, the reference condition.

The approach below attempts to correct for these biases by weighting measurements by the "surveillance activity" (*overvåkingsaktivitet*) during which they have been sampled. 
Scores between &minus;3 and +3 have been assigned to each surveillance activity, where the sign indicates the direction of the deviation (poorer or better than a randomly chosen water body), and the absolute value is an estimate of the magnitude of the bias. 
The approach has been described (in Norwegian) by Sandvik ([2019](#san19): pp. 17--19) and extended by Framstad et al. ([2023](#fra23): pp. 121--122).

<!--# Describe the main uncertainties or sources of error in the indicator or the underlying data. -->


## 6. References

<span id="veil1">DGVF [Direktoratsgruppa for vannforvaltning].</span> 2018. *Veileder 1:2018 Karakterisering. Metodikk for å karakterisere og vurdere miljømålsoppnåelse etter vannforskriftens § 15.* Miljødirektoratet, Trondheim. <https://www.vannportalen.no/veiledere/veileder-12018-karakterisering-metodikk-fora-karakterisere-og-vurdere-miljooppnaelse-etter-vannforskriften--15/>

<span id="veil2">DGVF [Direktoratsgruppa for vannforvaltning].</span> 2025. *Veileder for klassifisering av miljøtilstand i kyst- og ferskvann* [version of 28 Jan 2025]. Vannportalen. <https://www.vannportalen.no/veiledere/klassifiseringsveileder/>

<span id="fra23">Framstad E, Czúcz B, Schartau AK, Simensen T, Nybø S & Sandvik H.</span> 2023. Naturregnskap og økologisk tilstand: Samsvar mellom fagsystemet for økologisk tilstand, vannforskriften, FNs rammeverk og EUs forslag til naturregnskap. *NINA Rapp* **2327**: 1&ndash;122. [[hdl:11250/3104185](https://hdl.handle.net/11250/3104185)]

<span id="nyb17">Nybø S & Evju M (eds).</span> 2017. *Fagsystem for fastsetting av god økologisk tilstand. Forslag fra et ekspertråd.* Klima- og miljødepartementet. <https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/>

<span id="san19">Sandvik H.</span> 2019. Ferskvann i vannforskriften og naturindeksen. Forslag til samordning og dataflyt. *NINA Rapp* **1723**: 1&ndash;38. [[hdl:11250/2631056](http://hdl.handle.net/11250/2631056)]

<span id="san24">Sandvik H.</span> 2024. NI_vannf: dataflyt fra vannforskriften til naturindeks. <https://github.com/NINAnor/NI_vannf> [[doi:10.5281/zenodo.10278000](https://doi.org/10.5281/zenodo.10278000)]

<span id="sch11">Schneider S & Lindstrøm E-A.</span> 2011. The periphyton index of trophic status PIT: a new eutrophication metric based on non-diatomaceous benthic algae in Nordic rivers. *Hydrobiologia* **665**: 143&ndash;155. [[doi:10.1007/s10750-011-0614-7](https://doi.org/10.1007/s10750-011-0614-7)]

<span id="seeaea">**SEEA EA**:</span> United Nations et al. 2021. *System of Environmental-Economic Accounting--Ecosystem Accounting (SEEA EA).* <https://seea.un.org/ecosystem-accounting>

<span id="vannf">**Vannforskriften**.</span> 2006. Forskrift om rammer for vannforvaltningen. *Nor Lovtid I* **2006**: 2004&ndash;2034. [[FOR-2006-12-15-1446](https://lovdata.no/dokument/SF/forskrift/2006-12-15-1446)]

<span id="wfd">**WFD**:</span> EU. 2000. Directive 2000/60/EC of the European Parliament and of the Council of 23 October 2000 establishing a framework for Community action in the field of water policy. *Off J EU L* **43** (327): 1&ndash;72. [[CELEX:32000L0060](http://eur-lex.europa.eu/LexUriServ/LexUriServ.do?uri=CELEX:32000L0060:en:NOT)]
  
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


## 7. Datasets

<!--# Describe the unique datasets seperately under seperate headers (Dataset A, Dataset B, etc.-->

### 7.1 Vannmiljø

*Vannmiljø* is a database maintained by the [Norwegian Environment Agency](https://www.miljodirektoratet.no/) that is used to collect registrations related to ecological condition of water according to the WFD. 
This database contains information on all *measurements* taken in connection with the WFD ("registrations") as well as all *locations* at which measurements are taken ("water localities"). 
Other information available from *Vannmiljø* includes lists of WFD variables and indicators ("parameters"), of surveillance activities, of measurement units used, of species recorded etc. 

The database has long been available for manual download via [https://vannmiljo.miljodirektoratet.no](https://vannmiljo.miljodirektoratet.no/). 
Recently, the data have also been made available via an [API](https://vannmiljowebapi.miljodirektoratet.no/swagger/ui/index#/), which is utilised here. 

<!--# Describe the main dataset, typicaly the one containing the variable of (most) interest. Change the header from Dataset A to the name of the actual dataset. -->

### 7.2. Vannforekomster

In order to carry out the subsequent analyses, all WFD registrations need to be assigned to the correct water body (*vannforekomst*). 
Information on water bodies has currently to be downloaded from separate sources. 
A geospatial dataset of all water bodies is publicly available, but it needs to be downloaded manually:

* Navigate to [https://karteksport.miljodirektoratet.no](https://karteksport.miljodirektoratet.no/).
* In the menu to the left of the map, tick "Vannforekomster" in the "Produkt" section...
* ...and "Nasjonalt" in the "Definer område" section.
* Click "Neste".
* Provide your e-mail in the "Epost" box.
* Choose "ESRI Filgeodatabase (ESPG:4326)" in the "Format" box.
* Upon pressing "Send bestilling", a compressed geodatabase (gdb) will be sent to your e-mail.

The uncompressed geodatabase is needed for the analyses. 
It has to have the name "VF.gdb" in order to be recognised. 
(A cached version called "VF.RData" is used in the demo below.)

<!--# Describe additional datasets in a similar was as above. Delete or add ned subheaders as needed. -->

### 7.3. Vann-nett

In addition to the location of water bodies, the subsequent analyses need information on the water body *typology*. 
In WFD context, typology refers to a set of factors describing abiotic characteristics of the water bodies, such as altitude, size and alkalinity. 
This information is publicly available from a separate database called *Vann-nett* via the link [https://vann-nett.no/waterbodies/report](https://vann-nett.no/waterbodies/report). 

A spreadsheet workbook exported from *Vann-nett* is needed for the analyses. 
It has to have the name "V-L.csv" (for lakes) or "V-R.csv" (for rivers) in order to recognised.

*(Note that Vann-nett has recently been upgraded. In the new version, typology factors are not included in the list of variables about water bodies that are available for export. For the time being, water body typology thus has to be based on an export from the previous version of Vann-nett, which is no longer online. It is to be hoped that the relevant information will be made available in the new version soon.)* <!--# 
I have requested this information to be added. I hope that it is only a matter of time before it is available again. -->

<!--# Describe additional datasets in a similar was as above. Delete or add ned subheaders as needed. -->

<!--# ### 7.4. Innsjødatabasen

Describe additional datasets in a similar was as above. Delete or add ned subheaders as needed. -->


## 8. Spatial units

Data (measurements) are assumed to be representative for their respective water bodies, which correspond to the Ecosystem Assets of freshwater ecosystems. 
The Ecosystem Accounting Area covered by the data is mainland Norway. 

<!--# 
Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpretted at. Potential spatial delineation data should eb introduced under 7.1. Datasets. We recomend using the SEEA EA terminology opf Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). 
-->

## 9. Analyses

### 9.1. Preliminaries

At the outset, we define two variables containing the official abbreviations (i.e., according to *Vannmiljø*) of the indicator and of the relevant water category, respectively. 
("Water categories" are **R**iver, **L**ake and **C**oast.)

```{r}
indikator <- "PIT"
vannkategori <- "R"
```

The dataflow used here has been described in the GitHub repository [NI_vannf](https://github.com/NINAnor/NI_vannf/). 
We start by downloading the functions and data files from archived version 1.5 of that repo.

```{r}
if(!exists("../data/VF.RData")) {
  download.file("https://zenodo.org/records/15276139/files/NINAnor/NI_vannf-v1.5.zip", 
                "../data/NI_vannf.zip")
  folder <- "../data/NI_vannf"
  unzip("../data/NI_vannf.zip", exdir = folder)
  while (length(list.files(folder)) == 1) {
    folder <- paste0(folder, "/", list.files(folder))
  }
  whichDataFile <- c(
    "fnr.xlsx",
    "knr.xlsx",
    "navnNVEl.csv",
    "navnVL.csv",
    "navnVM.csv",
    "navnVN.csv",
    #"NVE.RData",
    paste0("V-", vannkategori, ".csv"),
    "VF.RData",
    "VM-aktiv.xlsx",
    "VM-param.xlsx"
  )
  dummy <- file.copy(from = paste0(folder, "/data/", whichDataFile),
                     to   = "../data/",
                     overwrite = FALSE)
  dummy <- file.copy(from = paste0(folder,
                                   "/klassegr/klassegrenser_",
                                   indikator, 
                                   ".xlsx"),
                     to   = "../data/",
                     overwrite = FALSE)
  dummy <- file.copy(from = list.files(paste0(folder, "/R/"), full.names = TRUE),
                     to   = ".",
                     overwrite = FALSE)
  dummy <- file.remove("../data/NI_vannf.zip")
  unlink("../data/NI_vannf", recursive = TRUE)
}
```

Next, we load the functions contained in that repo.

```{r}
source("Funksjon.R")
for (filnavn in list.files(".", full.names = TRUE)) {
  if (substr(filnavn, nchar(filnavn), nchar(filnavn)) %=% "R") {
    source(filnavn)
  }
}
```

### 9.2. Reading registrations and other information from Vannmiljø

Now that the preliminaries are in place, we first read the measurements of the specified indicator from the *Vannmiljø* API.

```{r}
DATA <- lesMaalinger(indikator, filsti = "../data")
```

The output informs us that the operation was successful and how many registrations have been read.

A list of water localities (*vannlokaliteter*) is likewise read from the *Vannmiljø* API. 
The main function argument is the water category.

```{r}
VL <- lesVannlokaliteter(vannkategori, filsti = "../data")
```

The output informs us that the operation was successful and how many water localities there are.

The function `lesMer` creates three more variables that are needed for analyses:

* `Parametere`, a data frame containing names, abbreviations and allowed values for WFD parameters (variables and indicators), read from "VM-param.xlsx"
* `Aktiviteter`, a data frame containing surveillance activities and their bias scores, read from "VM-aktiv.xlsx"
* `Enheter`, a data frame containing measurement units, read from the *Vannmiljø* API

```{r}
lesMer()
```

### 9.3. Reading data on water bodies

The preparation of the list of water bodies (*vannforekomster*) requires two files: 
one containing the water body typology, which needs to be downloaded from *Vann-nett* and named "V-R.csv" (for **R**iver water bodies); 
and another containing a geodatabase of all water bodies, named "VF.gdb".

```{r, eval=FALSE, echo=TRUE}
V <- lesVannforekomster(vannkategori, filsti = "../data")
```
```{r, eval=TRUE, echo=FALSE}
# The geodatabase is too huge for GitHub, so a cached version is used instead
V <- lesVannforekomster(vannkategori, filsti = "../data",
                        CACHE = "VF.RData")
```

The output informs us of several adjustments made to the data, but that the operation was successful and how many water bodies (of the relevant water category) there are.

### 9.4. Calculating mEQRs using reference levels

According to the WFD and DGVF ([2025](#veil2)), measurements of WFD variables are transformed into so-called nEQR values ("normalised Ecological Quality Ratios") between 0 and 1, which correspond to indicator values sensu [SEEA EA](#seeaea). 
This is achieved in the following steps:

1. *Truncation.* Variable values are truncated, where measurements that are poorer than X~0~ are set to X~0~, and measurements that are better than X~100~ are set to X~100~.
2. *Scaling.* The truncated variable values are linearly scaled so that X~0~ becomes 0 and X~100~ becomes 1. The resulting indicator values are referred to as **EQR**s ("Ecological Quality Ratios").
3. *Transformation.* EQR values are transformed in a stepwise-linear manner, so that the indicator values 0.2, 0.4, 0.6 and 0.8 correspond to the thresholds between the normative definitions of "bad", "poor", "moderate", "good" and "high" ecological status according to the WFD. The resulting indicator values are referred to as **nEQR**s ("normalised EQRs").

Truncation, however, destroys information about variation in ecological condition that is important for estimating (modelling and extrapolating) the indicator value and its uncertainty. 
Therefore, the dataflow from WFD to an Ecological Condition Account should be based on values that are *scaled and transformed but not truncated*.
This is achieved by modifying the order of the steps from truncation&ndash;scaling&ndash;transformation to scaling&ndash;transformation&ndash;aggregation&ndash;truncation (following [Sandvik 2019](#san19): pp. 13--15, [Sandvik 2024](#san24)):

1. *Scaling.* The *un*truncated variable values are linearly scaled so that X~0~ becomes 0 and X~100~ becomes 1. Note that the resulting values may be negative (if X is poorer than X~0~) or larger than 1 (if X is better than X~100~).
2. *Transformation.* The scaled values are transformed in a stepwise-linear manner in accordance with the threshold values defined by the WFD. The resulting indicator values are here referred to as **mEQR**s ("modified EQRs"). This step requires an adjustment in cases were the minimum and/or maximum values of X can be much lower or higher than X~0~ or X~100~. The approach chosen here ensures that no indicator value can be below &minus;0.2 or above 1.2.
3. *Aggregration.* Indicator values for the specific water bodies need to be spatially aggregated (see below). 
4. *Truncation.* When mEQR values are truncated, they become identical to nEQR values. However, truncation is here done only after the spatial aggregation of the modelled and simulated values.

Let us have closer look at the reference levels of the indicator (which are given in [DGVF 2025](#veil2)).

```{r}
klassegrenser <- hentKlassegrenser("../data/klassegrenser_" %+% indikator %+% ".xlsx")
if (nrow(klassegrenser) > 1) {
  m <- matrix(0, 8, 2, dimnames = list(
    c("Minimum value:", "Range of X_" %+% seq(0, 100, 20) %+% ":", "Maximum value:"), 
    c("from", "to")))
  m[, 1] <- apply(klassegrenser, 2, min, na.rm = TRUE)
  m[, 2] <- apply(klassegrenser, 2, max, na.rm = TRUE)
  klassegrenser <- klassegrenser[which(!is.na(klassegrenser[, 1]))[1], ]
} else {
  m <- t(klassegrenser)
  dimnames(m) <- list(c("Minimum", "X_0   =", "X_20  =", "X_40  =", "X_60  =",
                        "X_80  =", "X_100 =", "Maximum"), "")
}
print(m)
```

Since reference values differ between water types, their ranges are shown above. 

The following graph shows the scaling and transformation of the indicator (using the WFD colour coding for ecological status classes).

```{r}
#| lab: meqr-fig
#| fig-cap: "Scaling and transformation from variable measurements to mEQR values"

xlim <- c(0, 70)
xaxis <- seq(0, 70, 10)
yaxis <- seq(-0.2, 1.2, 0.2)
x <- seq(min(klassegrenser), max(klassegrenser), le = 1000)
y <- mEQR(x, klassegrenser)
forklar <- "Original variable values"
if (unique(DATA$enhet) %!=% "3" & exists("Enheter")) {
  forklar <- forklar %+% " (" %+% 
    Enheter[which(Enheter[, 1] == unique(DATA$enhet)), 2] %+% ")"
}
plot(x, y, ty = "l", main = indikator,
     xlim = xlim,         xaxs = "r", xaxt = "n", xlab = forklar,
     ylim = c(-0.2, 1.2), yaxs = "r", yaxt = "n", ylab = "mEQR values",
     cex.main = 1.5, cex.lab = 1.2)
rect(xlim[1]-144,-0.4, xlim[2]+144, 0.0, col = rgb(0.0, 0.0, 0.0, 0.1), border = NA)
rect(xlim[1]-144, 0.0, xlim[2]+144, 0.2, col = rgb(1.0, 0.0, 0.0, 0.2), border = NA)
rect(xlim[1]-144, 0.2, xlim[2]+144, 0.4, col = rgb(1.0, 0.5, 0.0, 0.2), border = NA)
rect(xlim[1]-144, 0.4, xlim[2]+144, 0.6, col = rgb(1.0, 1.0, 0.0, 0.2), border = NA)
rect(xlim[1]-144, 0.6, xlim[2]+144, 0.8, col = rgb(0.0, 1.0, 0.0, 0.2), border = NA)
rect(xlim[1]-144, 0.8, xlim[2]+144, 1.0, col = rgb(0.0, 0.0, 1.0, 0.2), border = NA)
rect(xlim[1]-144, 1.0, xlim[2]+144, 1.4, col = rgb(0.5, 0.0, 1.0, 0.1), border = NA)
lines(c(xlim[1] - 144,    klassegrenser[7]), c(+1, 1), lwd = 1.5, col = "blue")
lines(c(klassegrenser[7], klassegrenser[7]), c(-1, 1), lwd = 1.5, col = "blue")
lines(c(xlim[1] - 144,    klassegrenser[2]), c(+0, 0), lwd = 1.5, col = "red")
lines(c(klassegrenser[2], klassegrenser[2]), c(-1, 0), lwd = 1.5, col = "red")
lines(x, y, lwd = 3)
axis(1, xaxis, erstatt(xaxis, "-", "−"), cex.axis = 0.9)
axis(2, yaxis, c("−0.2", "0.0", 0.2, 0.4, 0.6, 0.8, "1.0", 1.2), cex.axis = 0.9)
text(xlim[2], seq(0.1, 0.9, 0.2), c("bad", "poor", "moderate", "good", "high"), 
     pos = 2, offset = 0)
rf <- klassegrenser[7] > klassegrenser[2]
text(klassegrenser[2], -0.15, expression(X[0]),   pos=ifelse(rf, 4, 2), offset=0.3)
text(klassegrenser[7], -0.15, expression(X[100]), pos=ifelse(rf, 2, 4), offset=0.3)
```

Note that variable values may be substantially smaller than X~100~ (blue line). 
Therefore, an asymptotic transformation is used for X < X~100~ in order to keep mEQR values below 1.2.
Variables values may also be somewhat greater than X~0~ (red line).

### 9.5. Estimating the indicator values and their uncertainty

With the necessary files in place, the calculations can start. 
This is done using the function WFD2ECA ("Water Framework Directive to Ecological Condition Account"), which carries out the following steps (cf. [Sandvik 2019](#san19), [Sandvik 2024](#san24)):

* *Removing measurements that have been taken outside the specified accounting periods.* Accounting times are assumed to be every third year. The value for 2024 is based on measurements taken in 2022, 2023 and 2024. In order to increase the sample available for modelling, earlier time periods are included as well. Accounting years chosen are 2012, 2015, 2018 and 2021, each of them representing the three years up to and including the accounting year. Measurements taken before 2010 are thus excluded.
* *Removing measurements with values that lie outside the interval for which the indicator is defined.* The *Vannmiljø* database may contain values that are incompatible with the definition of the indicator. These must have been mismeasured or misreported and need to be excluded. In addition, other measurements taken by the same operator at the same date are excluded.
* *Identifying the water locality for each measurement.* Measurements taken in an unknown water locality are excluded.
* *Identifying the water body for each water locality.* Measurements taken in an unknown water body are excluded.
* *Removing measurements that are taken in a water category for which the indicator is not defined.* For example, if a WFD indicator that is defined for rivers is measured in a lake, this measurement is excluded.
* *Removing measurements that are taken in a water body type for which the indicator is not defined.* For example, if a WFD indicator that is defined for lime-poor lakes is measured in a lime-rich lake, this measurement is excluded.
* *Removing measurements that do not fulfil specific requirements for the indicator.* Some WFD indicators need to be measured in certain months or to be derived from a certain minimum number of measurements per season. These requirements are described in DGVF ([2025](#veil2)). Where it is possible to test whether these requirements are met based on data from *Vannmiljø*, this is done by the function.
* *Removing measurements for reporting periods with too little data.* A certain minimum sample size is needed to estimate the status for a given reporting period. If the actual sample size is smaller than this threshold, the respective reporting period is omitted, and the measurements taken during this period are excluded. The default threshold used here is 100 measurements per reporting period.
* *Scaling measurements to untruncated nEQR values ("mEQR" values).* This step is explained [above](#calculating-meqrs-using-reference-levels).
* *Fitting a linear model to the measurements.* A linear model is fitted to the mEQR values, with explanatory variables being reporting periods, typology factors and surveillance activities. *Typology factors* of water bodies are described in DGVF ([2025](#veil2), especially Table 3.4) and include altitudinal zone, size class, lime content and organic matter content, among others. *Surveillance activities* describe the management target of each measurement (see [Vannmiljø](https://vannmiljokoder.miljodirektoratet.no/activity)). Measurements taken under the WFD can be expected to be biased towards poorer status than the average. Therefore, each measurement is here weighted by the representativeness of its surveillance activity, so as to partially correct for the bias in the data available (see [Sandvik 2019](#san19): pp. 17--19, [Framstad et al. 2023](#fra23): pp. 121--122). Model selection is based on AIC ([Akaikes Information Criterion](https://en.wikipedia.org/wiki/Akaike_information_criterion)).
* *Averaging measurements taken in the same water body.* If more than one measurement has been taken in the same water body within the same reporting period, the unweighted mean of these measurements is used to represent the water body.
* *Extrapolating mEQR values to water bodies that lack data.* If no measurement has been taken in a water body, the most likely indicator value is predicted for this water body, based on the model fitted previously.
* *Using simulations to estimate confidence and prediction intervals.* Based on the model fitted and a re-sampling procedure, the uncertainty surrounding the estimates obtained in the previous steps is quantified using confidence intervals for water bodies that have been sampled, and prediction intervals for water bodies that have not been sampled. This is a time-consuming step. For illustration purposes, the number of iterations is here limited to 1,000. With so few iterations, the estimates of uncertainty are unreliable, however. For reporting one should, therefore, use more iterations, at least 10,000 (and preferably 100,000).
* *Aggregating estimates geographically.* The values obtained for water bodies are aggregated for administrative units (municipalities, counties, regions and/or the whole country). Aggregation takes the form of averages weighted by area (for lake water bodies) or length (for river water bodies). Note that this step deviates from the WFD, which does not practice aggregation. (Rather, summary statistics following the WFD are provided as the number or proportion of water bodies that fulfil the objectives of the WFD.)

The function produces summaries of the steps performed (for the time being, the information is provided in Norwegian only).

```{r, eval=FALSE, echo=TRUE}
utmating <- WFD2ECA(
  DATA = DATA,
  vannforekomster = V,
  vannlokaliteter = VL,
  parameter = indikator,
  vannkategori = vannkategori,
  iterasjoner = 1000,
  SEED = 479001600,
  vis = FALSE
)
```
```{r, eval=TRUE, echo=FALSE}
# To avoid messy output, the arguments "bredde" and "tell" are added
# This affects nothing but the layout
utmating <- WFD2ECA(
  DATA = DATA,
  vannforekomster = V,
  vannlokaliteter = VL,
  parameter = indikator,
  vannkategori = vannkategori,
  iterasjoner = 1000,
  SEED = 479001600,
  bredde = 90,
  vis = FALSE, # this simplifies/shortens the information about model fitting
  tell = FALSE
)
```

### 9.6. Summary of the output

The output is a list with one list element per geographic reporting level (i.e., in this case, one for the national and one for the regional level).
For each geographical unit on this level and each reporting year, the predicted value and the simulated values are saved.

```{r}
ut <- utmating
attributes(ut) <- attributes(ut)[1:3]
# to simplify the output, most attributes are removed
str(ut)
rm(ut)
```

Note that the output consists of mEQR values, i.e. it may contain values below 0 or above 1. 
The values thus need to be truncated in a final step. 
They then correspond to nEQR values sensu WFD and indicator values sensu [SEEA EA](#seeaea).

```{r}
{
cat("Range of values prior to truncation (mEQR): " %+% 
    format(round(min(utmating$landsdel, na.rm=TRUE), 4), nsm=4, sci=FALSE) %+%
    " to " %+%
    format(round(max(utmating$landsdel, na.rm=TRUE), 4), nsm=4, sci=FALSE) %+% "\n")
tomrom <- ifelse(nchar(format(round(min(utmating$landsdel, na.rm=TRUE), 4), 
                              nsm=4, sci=FALSE)) %=% 7, " ", "")
trunker <- function(x) {
  if (is.list(x) & length(x) > 0) {
    for (i in 1:length(x)) {
      x[[i]] <- trunker(x[[i]])
    }
  } else {
    x[] <- ifelse(x[] > 1, 1, x[])
    x[] <- ifelse(x[] < 0, 0, x[])
  }
  return(x)
}
utmating <- trunker(utmating)
cat("Range of values _after_  truncation (nEQR): " %+% tomrom %+%
    format(round(min(utmating$landsdel, na.rm=TRUE), 4), nsm=4, sci=FALSE) %+%
    " to " %+%
    format(round(max(utmating$landsdel, na.rm=TRUE), 4), nsm=4, sci=FALSE) %+% "\n")
}
```

The estimates and their confidence intervals may be summarised as tables.

```{r}
{
  kvantiler <- c(0.025, 0.25, 0.5, 0.75, 0.975)
  print(round(t(apply(utmating$Norge[  "Norge", , -1], 1, quantile, kvantiler)), 8)) 
  print(round(t(apply(utmating$landsdel[, "2024", -1], 1, quantile, kvantiler)), 4))
}
```

<!--# 
Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.
-->


## 10. Results

The results are here tabulated (a) for the whole country and different years, and (b) for regions in 2024. 
These estimates are based on 10,000 iterations<!-- not yet! (¤) --> and therefore differ from the ones above.

*NB: Please note that these results are not the final ones for 2024. The deadline for reporting WFD measurements taken in 2024 is September 2025, meaning that the values for 2024 are so far based mainly on measurements taken in 2022 and 2023.* 


| Year | Best estimate | Lower quartile | Upper quartile |
|:-----|:-------------:|:--------------:|:--------------:|
| 2024 | 0.936         | 0.934          | 0.937          |
| 2021 | 0.911         | 0.909          | 0.913          |
| 2018 | 0.869         | 0.867          | 0.871          |
| 2015 | 0.606         | 0.604          | 0.609          |

: PIT for Norway in different years {.striped .hover}

&nbsp;

| Region     | Best estimate | Lower quartile | Upper quartile |
|:-----------|:-------------:|:--------------:|:--------------:|
| Østlandet  | 0.957         | 0.955          | 0.960          |
| Sørlandet  | 0.937         | 0.934          | 0.940          |
| Vestlandet | 0.899         | 0.895          | 0.903          |
| Midt-Norge | 0.955         | 0.952          | 0.959          |
| Nord-Norge | 0.922         | 0.919          | 0.925          |

: PIT for regions of Norway in 2024 {.striped .hover}

<!--# 
Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.
-->

<!--# 
## 11. Export file

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a seperate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 
-->

```{r export}
#| eval: false
```
