---
title: "Tree Canopy Cover Density"
subtitle: |
  [NO_TCCD_001] - Tree Canopy Cover Density indicator in forests  \
  [NO_TCCD_100 to NO_TCCD_104] - Tree Canopy Cover Density in urban areas 
indicatorID: 
  - NO_TCCD_001
  - NO_TCCD_100
  - NO_TCCD_101
  - NO_TCCD_102
  - NO_TCCD_103
  - NO_TCCD_104
indicatorName: 
  - Tree Canopy Cover Density
verbatimName: 
  - Kronedekke
folderName: NO_TCCD_001_104
country: Norway
continent: Europe
ECT: B2 - Structural State Characteristics
Realm: 
  - Terrestrial
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
  - T7 Intensive land-use biome
  - T7 Intensive land-use biome
  - T7 Intensive land-use biome
  - T7 Intensive land-use biome
  - T7 Intensive land-use biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T7.4 Urban and industrial
  - T7.4 Urban and industrial
  - T7.4 Urban and industrial
  - T7.4 Urban and industrial
  - T7.4 Urban and industrial
verbatimEcosystem: 
- Skog
- Bebygd og opparbeidet areal
yearAdded: 2025 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "000.002" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Version 000.002 reviewed with new available data.
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_TCCD_001_104
Normalised: No
SpatialAggregationPathway: Aggregation at small polygon scale within target area (TCCD area weighted mean) - Aggregation at focal area scale (mean and standard error; city, region, national scales)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: false
author:
  - name: Sylvie Clappe # Enter name
    email: sylvie.clappe@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Bálint Czúcz            #  Enter subsequent authors like this, or remove if not relevant
    email: balint.czucz@nina.no
    affiliations:
      - ref: myID               # To reuse affiliations referecen the id like this
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026 # Enter date 
data_availability: silver
code_reproducibility: silver
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
bibliography: references.bib
hide: FALSE
codeReviewers:
  - name: XXXXX
    email: XXXXXX
    affiliations:
      - id: reviewerInst
        name: XXXXXX
    reviewType: code review
    github: XXXXXXX
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(terra)
library(sf)
library(dplyr)
library(readxl)
library(httr2)
library(glue)
library(tictoc)
library(duckdb)
library(duckdbfs)
library(duckspatial)
library(mirai)
library(purrr)
library(stringr)
library(exactextractr)
library(carrier)
library(hdar)
library(DT)
library(numform)
library(giscoR)
library(conflicted)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```

::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log
<!--# Update this logg with short messages for each update -->
- 05 August 2025 - First draft produced
- 01 September 2025 - Original PR
- 09 October 2025 - First review
- 23 October 2025 - Second review
- 25 November 2025 - Restructue metadata and add open science badges (ALK)
- 13 July 2026 - Reviewed code to include new available data
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
Tree canopy cover, or briefly *tree cover*, is the area of the earth that lies underneath the canopy/foliage of high (typically >5 m) woody vegetation when viewed vertically from above. This is typically expressed as the proportion/share of area of the land covered by tree crowns within a broader area of interest (**i.e.,** as the *density* of tree canopies/tree cover). Accordingly, *tree canopy cover density* (TCCD), or *tree cover density* in short, is typically measured from above by using some airborne or space-borne remote sensing technique. 

In principle, TCCD can be quantified for all terrestrial ecosystem types (ETs). From an EU perspective there are, however, two major ETs for which TCCD is considered particularly relevant: 

* The "*tree cover density*" of *forests* (ET04 in the EU ecosystem typology; @et-typo-eurostat) is one of the eight *mandatory* condition variables prescribed by the EU Regulation on Environmental Economic Accounting ([EU Reg. 2024/3024](https://eur-lex.europa.eu/eli/reg/2024/3024/oj/eng)), 

* The average *tree canopy cover*	of *urban ecosystems* is listed as a voluntary condition variable in Eurostat's condition accounting guidelines (@handbook2026) for the ecosystem type *Settlements and Other Artificial Areas* (ET01 in the EU ecosystem typology) **#BALINT: THIS VOLUNTARY INDICATOR IS NOT IN THE OFFICIAL HANDBOOK**

2. **Methods**:
We used three main data sources: (i) High Resolution Layer Tree Cover and Forests produced by Copernicus; (ii) the official Norwegian ecosystem map (Grunnkart) produced by NIBIO; and (iii) Local Administrative Units produced by Eurostat (municipalities boundaries).

For ET01 (Settlements and Other Artificial Areas) and ET04 (Forests and woodlands), the calculation of TCCD for a specific reference period and ecosystem accounting area (EAA) consists of two two main steps :

1. Identify the '*target area*' for which the indicator should be calculated (cf. @Czucz2025a);

2. Calculate the mean TCCD within this area (for the selected accounting period).

There are often several ways of defining the *target area* at Step 1, especially when it comes to urban ecosystems. Accordingly, in this document, we will present four different variants of the urban indicator, based on different definitions of the *target area* (see point 4 below). Besides offering a flexible approach for performing the "right" calculation for various reporting context, these variants are also aligned across other *urban* ecosystem condition variables (URGR, URAQ), thus making it easier to establish consistency across the different variables and their policy use. 


3. **Interpretation**:
A value of 1 signifies that the ecosystem type is fully covered by tree canopies. For some forests this might be the ideal condition, for other forests and urban ecosystems lower tree covers can still indicate very good conditions. Low values (TCCD close to zero) reflect a bad condition in both forest and urban ecosystems. 


4. **Results and development needs**:
For forests we only calculate one version of this indicator: 

* **TCCD_001**:  Average Tree Cover Density in the forest areas of Norway. Here the *target area* is defined as the total area of ET04 (*Forest and woodlands*) in the Norwegian national ecosystem type map (Grunnkart), which complies with both Eurostat's guidance (@handbook2026) and the SEEA EA principles. 

For urban ecosystems we present five variants:

* **TCCD_100**: The *target area* is the *entire area* of the municipalities considered as "cities" according to Eurostat DEGURBA classification (`DEGURBA = 1`). As the total area of Norwegian municipalities is typically dominated by natural ecosystems (forests, wetlands, mountains, etc.), this version of the indicator will overestimate TCCD in urban ecosystems within municipalities.

* **TCCD_101**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as *"cities"* according to Eurostat's degree of urbanisation (DEGURBA) classification (`DEGURBA = 1`). 

* **TCCD_102**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as "towns and suburbs" according to Eurostat's DEGURBA classification (`DEGURBA = 2`).

* **TCCD_103**: The *target area* is the area of the ecosystem type ET01 (*Settlements and other artificial areas*) within the municipalities considered as "rural areas" according to Eurostat DEGURBA classification (`DEGURBA = 3`). TCCD_101-TCCD_103 are complementary to each other, they present an overview of the tree cover across all three main types of municipalities in Norway. As the three DEGURBA categories comprise all Norwegian municipalities, the area-weighted mean (for ET01) of these three indicators will be equal to TCCD_104 for any reporting period.  

* **TCCD_104**: The *target area* is the total area of the ecosystem type ET01 (*Settlements and other artificial areas*) in Norway. This is the option which is most aligned with the principles of SEEA EA (@seea-ea), where an ecosystem's condition indicators are expected to be representative of their "parent" ecosystem type (see also @Czucz2025a). 


**TCCD_100** contradicts key principles of ecosystem accounting (*i.e.,* the target area covers vast areas of forests, wetlands, grasslands, heathlands, lakes, etc. in addition to a typically small settlement area), so this option should probably not be used for reporting, we just present to maintain consistency with the other urban ecosystem condition indicators (URGR, URAQ) presented in ecRxiv. The most SEEA-EA-consistent variant is TCCD_104, which also seems to be aligned with the relatively brief description available for the voluntary variable ("average across all Settlements and other artificial areas"). The calculation of TCCD_101 to TCCD_103 for diverse ecosystem accounting areas (EAAs) might also be useful for the additional analytical insight offered by distinguishing the different types of settlements according to their degree of urbanisation.   


## 2. About the underlying data
We use the following two main datasets for our calculations: 

* **Copernicus High Resilution Layer Tree Cover Density (HRL TCD)**: We used the *Tree Cover Density* product of the Copernicus Land Monitoring Service to calculate mean tree canopy cover density over different ecosystem types. 

* **Grunnkart**: Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). This map aims at implementing Eurostat's EU ecosystem typology (@et-typo-eurostat) for Norway. Here we use this map for delineating the area of ecosystem type ET01 (*Settlements and other artificial areas*) and ET04 (*Forest and woodlands*).

In addition we used administrative datasets to delineate the individual municipalities (i.e., Local Administrative Units, LAU), and to determine their "degree of urbanisation" (DEGURBA).


### 2.1 Spatial and temporal resolution and extent
*Copernicus HRL TCD* is available for the whole Europe, including mainland Norway (excluding Svalbard) as a raster file with a high resolution (10m x 10m). 

All other spatial data files, including the *Grunnkart* and the *Local Administrative Units* (LAU) are vector data (polygons). Unlike traditional land cover maps (e.g., Corine Land Cover), the Grunnkart does not have a single, universal Minimum Mapping Unit (MMU). Instead, the spatial resolution of the Grunnkart is directly inherited from the underlying source datasets used in its creation. This is because the new map is a synthesis and reclassification of existing data rather than a new, independent mapping effort. As the majority of spatial information available in public datasets focusses on human settlements and infrastructure, urban ecosystems are relatively finely represented in the Grunnkart, compared to other, less anthropologically modified ecosystem types. This makes the contours of ET01 overly-detailed, often delineating individual buildings or roads passing through a forest as "urban ecosystems". This also makes the dataset somewhat unfit for its purpose (@Rendon2025) and cumbersome to process. The delineation of ET04 is of comparable detail and quality as forests is the main ecosystem type in Norway.


### 2.2 Original units
The pixel values of Copernicus HRL TCD express the estimated share of tree cover in percentages (*i.e.*, the % of the pixel covered by tree canopies). As each pixel covers ~100 m^2^, this actually corresponds to the estimated tree cover in 1m^2^. 


### 2.3 Additional comments about the dataset
The DEGURBA dataset was used to classify the Norwegian municipalities into three categories (1: "cities", 2: "towns and suburbs", 3: "rural areas") based on their degree of urbanisation. This data are produced by Eurostat.


#### 2.3.1 Instructions for citing, using and accessing data
||Publicly accessible to download| Data and metadata available Available at|
|----|--|--------|
| **Copernicus Tree Cover Density**  | Yes | Data and description downloadable from the Copernicus [website](^https://land.copernicus.eu/en/products/high-resolution-layer-forests-and-tree-cover?tab=overview$) and through the Copernicus [WEkEO API](^https://wekeo.copernicus.eu/$) | 
**Grunnkart**  | No, only upon request to [NIBIO](^https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d$) | We used test version 0.2, description available in @gk2 and @strand2024 |
| **EU data about the municipalities**  | Yes | Data and description available from Eurostat: [LAU](^https://ec.europa.eu/eurostat/web/gisco/geodata/statistical-units/local-administrative-units$) and [DEGURBA](^https://ec.europa.eu/eurostat/web/nuts/local-administrative-units$) |


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)
Tree Canopy Cover Density can be considered under ECT class B2 (structural state characteristics) as it describes the physical structure of the parent ecosystem (ET04 forests and ET01 urban ecosystems, respectively).


### 3.2 Ecosystem condition characteristic
Trees are key structural components of almost all terrestrial ETs. The variable Tree Canopy Cover Density is a simple yet powerful metric describing the abundance of trees, thus structurally characterising the ETs. 

In principle, as trees can act as microhabitats and they also moderate microclimate, an increased TCCD should indicate an increased support for species and genetic diversity (a positive directionality, sensu @Czucz2021). Depending on the type and subtype of the parent ecosystem this relationship is, however, not always true. Some forest subtypes have a naturally open structure, in such forest types an additional increase in TCCD cannot be seen as an improvement of condition. Similarly, in clear-cut areas sapling stands emerge as the forest regenerates. These stands can be quite dense and offer a high Tree Canopy Cover Density even though the overall condition of the forest is poor as still regenerating from human intervention.


### 3.3 Other standards
In this documentation, we present different variations of the Tree Canopy Cover Density indicator. For forest ecosystems TCCD_001 aligns well both with SEEA EA principles and the requirements for national reporting under EU Regulation 2024/3024 (@eu-2024-3024). For urban ecosystems we recommend the use of TCCD_104 for all reporting purposes, while TCCD_101--103 can provide additional insights by distinguishing the TCCD of rural regions, urbanised regions, and major cities from each other.


### 3.4 Collinearities with other indicators
The TCCD of urban ecosystems will definitely be highly correlated with the abundance of *urban green areas* (URGR), which is one of the mandatory EU condition variables for ET01. As trees provide the air filtration service, which allows to decrease the concentration of pollutants in that air (e.g., particulate matter 2.5) there might also be a correlation with the other mandatory urban indicator, the concentration of *PM25* in urban ecosystems (URAQ).

In more general, the abundance of trees in terrestrial ecosystems is connected to (and possibly more or less correlated with) several further characteristics of these ETs, e.g.:

* other complementary structural characteristics (e.g. the abundance of "gaps" or the share of open areas is expected  to be closely (and inversely) correlated to TCCD)

* near surface microclimate (wind, moisture, temperature), and in particular the availability of light

* habitat availability for several species groups (connected to trees / forests)

* landscape connectivity for several species groups (connected to trees / forests)

* the successional age of communities, the frequency of major disturbance events

### 3.5 Impact factors
There are a number of natural and anthropogenic processes that affect the number and size of trees in terrestrial ecosystems. In forests the most important such process is probably human management (forestry activities), the type and intensity of which is decisive for the closedness of tree canopies. Other relevant factors include a number of environmental variables (which often also determine the (sub)type of the forest) as well as disturbance regimes (e.g. fires). 

TCCD in urban ecosystems is also overwhelmingly determined by anthropogenic processes, including conscious urban planning.


## 4. Reference condition and levels

### 4.1 Reference condition
Following SEEA EA (@seea-ea), the most suitable *reference condition* for forests is their natural condition, where the structure and composition of the forest is dominated by natural processes, including dynamic equilibria with natural disturbance regimes (SEEA EA 5.70-71). This actually implies slightly different reference conditions for the different forest types. 

For urban ecosystems SEEA EA suggests the identification of an "anthropogenically derived" reference condition (SEEA EA 5.72, a.k.a. "best attainable condition, Vallecillo et al. 2022), which is also an option for highly modified forest types, without any natural counterparts. As this type of reference condition cannot be "sampled" (there is no clear reference set of "cities" in an "optimal/ideal/natural condition"), the best way to operationalise this type of reference condition is to determine optimal reference levels (X~100~) for each variable independently (@Czucz2025b), which is also discussed below. 


### 4.2 Reference levels
In this exercise we have not determined *reference levels* and have not applied a proper rescaling for any of the TCCD variables. Nevertheless, we have identified several considerations related to the future selection of these reference levels. 

* **X~100~**: For forests the optimal reference level (X~100~) probably needs to be different for each relevant forest subtype, reflecting the canopy structure of the natural state of the given forest type (*e.g.,* boreal rainforest *vs* a steep ravine forest). For forest types with a naturally closed canopy $X~100~ = 1$ (100% tree cover) can probably also be a suitable reference level. For urban ecosystems and highly anthropogenic forest types the X~100~ can probably be anchored best using linear extrapolation after the good/bad threshold (X~60~) and the pessimal scale endpoint have already been set (cf. @Czucz2025b; Paracchini et al. in prep).

* **X~0~**: The natural zero (0% tree cover) is probably a suitable pessimal reference level (null value) for all TCCD variables, both in forests and in urban ecosystems.

* **X~60~**: The threshold value separating "good" from "not good" condition is probably the most challenging reference level for TCCD. For anthropogenic ecosystem types (e.g. urban ecosystems) this value can probably be best set through a careful analysis of relevant environmental (e.g. planning) norms or targets, reflecting diverse societal perceptions about what can be seen as "good enough" in diverse contexts. For natural forests where both X~100~ and X~0~ are known, X~60~might also be identified using linear interpolation.



#### 4.2.1 Spatial resolution and validity
X~0~ (the null value of TCCD) can be seen as a universal reference level (RL), which is valid for any ET, including any forest subtype. For all other RLs (X~60~, X~100~) the value of the RL will probably need to depend on the ecosystem subtype, and/or possibly also on bioclimatic zone (i.e. "homogeneous ecosystem areas" sensu Vallecillo et al. 2022). 

As in the case of forests each pixel of the map can belong to a different forest type, TCCD should ideally be rescaled at the local scale (ecosystem asset by ecosystem asset), before any aggregation would happen. Accordingly, a proper rescaling of TCCD into proper "condition indicators" (sensu SEEA EA) will ideally demand a detailed map of the spatial distribution of the different forest subtypes in Norway. The identification of the relevant spatial datasets, ecosystem subtypes, and reference levels needs to be aligned with each other, and this will be a key task for making this ecosystem condition variable really meaningful for policy. In the case of this variable, where the same coverage value may mean a good condition in one forest type, while a very bad condition in another one, aggregating unscaled pixel values across broader areas may lead to confusing and contradictory results.


## 5. Uncertainties
### 5.1 General uncertainties
There are several sources of uncertainties affecting the values of this indicator and its subsequent use:

- <u>data uncertainties in Copernicus TCD</u>: The main input data source is itself the output of a complicated machine learning algorithm. While on a pan-European scale the accuracy (including both producer and user accuracy) of the HRL TCD product exceeds 90%, in Norway this accuracy seems to be significantly lower (@Dostalova2021). This can largely be attributed to the complex and uneven topography of the country, which is known to create challenges for satellite-based classification algorithms, leading to lower performance. More information availbale on the metadata documents [here](^https://land.copernicus.eu/en/products/high-resolution-layer-forests-and-tree-cover/tree-cover-density-2018-present-raster-10-m-europe-yearly$)

- <u>spatial, temporal and semantic uncertainties of the Grunnkart</u>: the current Grunnkart combines many different static data sources with various temporal reference dates, which does not seem to present an easily reproducible and updatable data source. This means that the outlines of the urban ecosystems (ET01) contain an unknown amount of uncertainty ("ad-hoc-ness"). The Grunnkart is currently a static dataset whose timeline has not been clearly documented as relying on temporally heterogeneous sources. Neverthless, we used it for the entire study period.

- <u>uncertainty of defining the target area</u>: the indicators TCCD_100 to TCCD_104 have been compiled in accordance with different interpretations of the urban areas, see more in Section 1.1. The instructions outlined in @handbook2026 do not align well with SEEA EA principles. All these indicators present a different definition of the EEA, which greatly impacts the results and is thus an important source of uncertainty.

### 5.2 Reported uncertainties
We do not quantify uncertainties around our indicators in this document due to lack of resources.


## 6. References
::: {#refs}
:::


## 7. Import and prepare datasets
All the datasets used to compile the different versions of the URAQ indicators have been thoroughly described in the previous sections, links to them are also provided for more information.

We used the following four datasets: 

* **D-1**: *Copernicus HRL TCD* (copernicus.eu): a series of raster datasets used to calculate mean PM2.5 concentration over different spatial scales. 

* **D-2**: *Grunnkart* (NIBIO): a high-resolution vector dataset used to delineate urban areas (ET01: settlements and other artificial areas).

* **D-3** *LAU data* (Local Administrative Units, Eurostat GISCO): a vector dataset was used as the primary data source to delineate the  municipalities.

* **D-4** *DEGURBA*  (Degree of urbanisation) A tabular dataset assigning one of three DEGURBA classes (1: "cities"; 2: "towns and suburbs"; 3: "rural areas") to each municipality. These classes were used to breakdown the results for URAQ_001-003.


We downloaded, cleaned and formatted the data prior to compiling URGR indicators. Section 7 only presents how to load the cleaned data. In the data analyses, we will use a mix of `duckspatial` and `terra` packages. The first one is faster at handling big vector data than `sf` but raster operations are not part of it, so we use `sf` and `terra` to handle vector-raster operations. Hence vector datasets are written in a duckdb database whose connection is set below:

```{r}
#| eval: true
#| message: false
#| code-summary: "Create a DuckDB connection"

# # For users who don't have a set up duckdb database
# conn_tcd <- ddbs_create_conn(dbdir = "tempdir", threads = 10, memory_limit_gb = 30)

# If users already have a duckdb database they want to use
conn_tcd <- ddbs_create_conn(dbdir = paste0(str_remove(tempdir(), "Rtmp\\w*"), "duckdb_conn/tccd.duckdb"), threads = 10, memory_limit_gb = 30)

```

### 7.1 Copernicus HRL TCD (D-1)
We just load the virtual rasters we created for the years 2018 to 2021. These virtual rasters were previously built from different TCD tiles for each year. The only cleaning steps were to apply a mask for Norway (to select the relevant tiles) and re-classify the value 255 as NA (outside area).

```{r}
#| eval: true
#| code-summary: "Load Copernicus IMD"

# Path to the Copernicus IMD data
path_tcd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/TCD_Copernicus"
path_tcd_list <- dir(path_tcd, recursive = FALSE, full.names = TRUE)
yrs_tcd <- str_sub(path_tcd_list, start = -4)
file_tcd <- map_vec(seq_along(path_tcd_list), function(i){paste0(path_tcd_list[i], "/TCD_", yrs_tcd[i], ".tif")})

# Set target CRS
tcd_crs <- st_crs(rast(file_tcd[1]))

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs")))

```



### 7.2 Grunnkart (D-2)
D-2 is stored in the temporary duckdb database. We create one table for ET01 (Settlements and other artificial areas), and one for ET04 (Forests and woodlands). The latter will be used for TCCD_001, and the former for TCCD_101 to 104.

The cleaning process of the Grunnkart included several steps: (i) correction of geometry; (ii) union of the polygons per ecosystem type at level 1 of the EU typology; (iii) BALINT.

To facilitate analyses in block processes afterwards, we create one table per fylke in the duckdb database. We add a step of re-projection to the CRS from Copernicus TCD to facilitate the data analyses later on. We make the geometry valid after transformation and clean the duckdb database of the original data.

```{r}
#| eval: false
#| code-summary: "Load Grunnkart - ET01 and ET04"

# Path
gk_path <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/GK_processed/fylke_level"
gk_list <- dir(gk_path, pattern = "UTM33",recursive = FALSE, full.names = TRUE)


# Store in duckdb database
walk(gk_list, function(x){
  
  dbExecute(conn_tcd, 
      sprintf("
        CREATE OR REPLACE TABLE '%s' AS
        SELECT okosystemtype1_name, geom
        FROM ST_Read('%s')
        WHERE okosystemtype1_name = 'Settlements and other artificial areas' 
          OR okosystemtype1_name = 'Forest and woodlands' 
      ", 
paste0("gk_et014_", str_extract(x, "(?<=UTM33_f).{2}")),
x))
  
})

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_tcd)

# Transform into IMD CRS
tables <- dbListTables(conn_tcd)
tbl_list <- keep(tables, startsWith(tables, "gk_et014"))


walk(tbl_list, function(tbl){
  
# transform  
ddbs_transform(tbl, tcd_crs, conn = conn_tcd) %>% 

    # make valid
    ddbs_make_valid(conn = conn_tcd, name = paste0("cln_", tbl), overwrite = TRUE)
  
})

# Check validity
tables <- dbListTables(conn_tcd) %>%
            keep(~ startsWith(.x, "cln_gk_et014"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_tcd) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_tcd) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_tcd) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# # Write the intermediate product in case of crash
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCCD_tmp_layers/GK_et014_3035_")
# tables <- grep('^cln_gk_et014', dbListTables(conn_tcd), value = TRUE)
# 
# walk(tables, function(x){
# 
#   as_duckspatial_df(x, conn = conn_tcd) %>%
#   ddbs_collect() %>%
#   st_write(paste0(path_gpkg, str_extract(x, "(?<=et014_).{2}"), ".gpkg"))
# })

# # Read intermediate layer (if necessary)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCCD_tmp_layers")
# gk_list <- dir(path_gpkg, recursive = FALSE, pattern = "GK_et01", full.names = TRUE)
# 
# walk(gk_list, function(tbl) {
# 
#   ddbs_data <- ddbs_open_dataset(tbl, conn = conn_tcd)
# 
#   ddbs_write_table(data = ddbs_data,
#                    conn = conn_tcd,
#                    name = paste0("cln_gk_et014_", str_extract(tbl, "(?<=3035_).{2}")),
#                    overwrite = TRUE)
# 
#   message("Wrote layer: ", tbl)
# })


# Clean duckdb database of the raw tables
tables <- dbListTables(conn_tcd) 
tables <- keep(tables, startsWith(tables, "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_tcd, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }else{

    dbExecute(conn_tcd, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }

}

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_tcd)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs")))
```


### 7.3 LAU (D-3) and DEGURBA (D-4)
LAU data can be directly downloaded from Eurostat using the `giscoR` package. We download the data in temporary directory.

```{r}
#| eval: false
#| code-summary: "Load LAU data"
# Load administrative boundaries
# Create a temporary directory to store the cache
temp_dir <- tempdir()

# Download the dataset
lau_no <- gisco_get_lau(year = "2024",
                        epsg = "3035",
                        update_cache = TRUE,
                        cache_dir = temp_dir,
                        verbose = FALSE,
                        country = "NO",
                        gisco_id = NULL)
# Clean LAU codes
lau_no$GISCO_ID <- str_remove(lau_no$GISCO_ID, "NO_")

```

Now we load DEBURBA data and add it to `lau_no`.

```{r}
#| eval: false
#| code-summary: "Add DEGURBA to LAU data"

# Source function to add fylke number
source(paste0(here::here(),"/indicators/NO_URAQ_001-005/R/functions/add_fylke.R"))

# Path to DEGURBA data
path_degurba <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_DEGURBA_Eurostat/EU-27-LAU-2024-NUTS-2024.xlsx"

# Read excel file with Degurba variable and select Norway
degurba_lau_no <- read_excel(path_degurba, sheet = "NO") %>%
                    mutate(Fylke_nummer = add_fylke(., kommune_nummer = "LAU CODE")[[1]],
                           Fylke_navn = add_fylke(., kommune_nummer = "LAU CODE")[[2]],)

# Join lau_no with DEGURBA
lau_dgurb <- inner_join(lau_no,
                        degurba_lau_no,
                        by = join_by("GISCO_ID" == "LAU CODE")) %>%
              select(Fylke_nummer, Fylke_navn, GISCO_ID, LAU_NAME, DEGURBA, geometry) %>%
              st_as_sf()

# Re-project in the CRS of TCD data and store in duckdb database
st_transform(lau_dgurb, crs = tcd_crs) %>%
      ddbs_make_valid(conn = conn_tcd,
                      "cln_lau",
                      overwrite = TRUE)
 
# Check validity
valid_check <- ddbs_is_valid("cln_lau",
                              conn = conn_tcd) %>%
                 filter(!is_valid) %>%
                 ddbs_collect() %>%
                 dim()

empty_check <- ddbs_is_empty("cln_lau",
                             conn = conn_tcd) %>%
                  filter(is_empty == TRUE) %>%
                  ddbs_collect() %>%
                 dim()

simple_check <- ddbs_is_simple("cln_lau",
                               conn = conn_tcd) %>%
                  filter(!is_simple) %>% 
                  ddbs_collect() %>%
                 dim()

print(valid_check)
print(empty_check)
print(simple_check)

# Verify table is in my duckdb conn_ug database
ddbs_list_tables(conn_tcd)

# Clean duckdb database of the raw tables
tables <- dbListTables(conn_tcd) 
tables <- keep(tables, startsWith(tables, "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_tcd, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }else{

    dbExecute(conn_tcd, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }

}

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs")))

```


## 8. Spatial units
The indicator values of the TCCD variable can be interpreted at any spatial scale, even for the pixels (10x10 m) of the original TCD raster maps (basic spatial unit, BSU). The smallest policy-relevant spatial units for all variables characterising ET01 are probably the contiguous forest patches and settlement areas (ecosystem assets, EA). Nevertheless, coarser ecosystem accounting areas (EAA), like the *fylke* (county) or the *entire country* can also be highly relevant as reporting areas.  


## 9. Analyses

### 9.1 Spatial intersections
All of the indicators are based on the same process so we group all the spatial intersections. 

**Target area** = the area of ET04 for TCCD_001; the area of LAU classified as "cities" (`DEGURBA = 1`) for TCCD_100; the area of ET01 within "cities" LAUs (`DEGURBA = 1`) for TCCD_101 or "towns and subburbs" LAUs (`DEGURBA = 2`) for TCCD_102 or "rural" LAUs (`DEGURBA = 3`) for TCCD_103, or all LAUs for TCCD_104.  

**Target feature** = the tree cover density of Copernicus TCD.

**Compilation process** = we start by a spatial intersection between ET04/ET01 and LAUs. Then we do a spatial intersection between ET04/ET01 and TCD. This results in the tree cover density percentage for each ET04/ET01 polygon. We then just need to do an area-weighted mean at national or LAU level. For TCCD_100, we only do the intersection between TCD and LAU.


We start by doing the spatial intersection between ET01/ET04 and LAUs.

*Note 1: we add an extensive cleaning steps to ensure that the resultsing first intersection only has polygons or multipolygons gemoetries as it is a requirement from `exact_extract`.*

*Note 2: we modified the function ddbs_intersecion from the `duckspatial` package so it could return the columns of all the input datasets. This required us to take the code of ddbs_intersection() directly from its Github repo: https://github.com/Cidree/duckspatial/tree/main. It also required us to download a bunch of functions, from this same repo, necessary to run ddbs:intersection().*

```{r}
#| eval: false
#| code-summary: "Spatial intersection ET01/ET04 and LAU"

## Intersect ET01 with LAUs - block processing to avoid memory issues
# Source function to do ddbs_intersection and return all columns
source(paste0(here::here(),"/indicators/NO_TCCD_001_104/R/functions/utils_fun_ddbs.R"))
source(paste0(here::here(),"/indicators/NO_TCCD_001_104/R/functions/ddbs_intersection_tcd.R"))

# Block intersections
fylke_nb <- dbGetQuery(conn_tcd, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

walk(seq_along(fylke_nb), function(i){

  # Input data  
  et <- as_duckspatial_df(paste0("cln_gk_et014_", fylke_nb[i]), conn = conn_tcd)
  
  lau <- as_duckspatial_df("cln_lau", conn = conn_tcd) %>%
          filter(Fylke_nummer == !!fylke_nb[i])

  
  # Intersection
  ddbs_intersection_tcd(lau,
                        et,
                        conn = conn_tcd,
                        conn_x = conn_tcd,
                        conn_y = conn_tcd,
                        keep_cols = "both") %>% 
    
    ddbs_make_valid() %>%
    
    # Need to remove linestrings and geometrycollection for extact_extract
    mutate(geometry = ST_CollectionExtract(geometry, 3L)) %>% # clean up geometrycollections
    filter(!ST_IsEmpty(geometry)) %>% # remove empty geometry resulting from the step above
    filter(ddbs_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>% # keep only polygons and multipolygons
    
    ddbs_area(new_column = "area_m2", name = paste0("lau_et014_", fylke_nb[i]), conn = conn_tcd, overwrite = TRUE)
  
})

# Verify table is in my duckdb conn_pm database
ddbs_list_tables(conn_tcd)

## Check valididty
tables <- dbListTables(conn_tcd) %>%
            keep(~ startsWith(.x, "lau_et014_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_tcd) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_tcd) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_tcd) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# ## Export the temporary layers, just in case of crash
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCCD_tmp_layers/ET014_LAU_3035.gpkg")
# 
# for (lyr in fylke_nb){
# 
# as_duckspatial_df(paste0("lau_et014_", lyr), conn = conn_tcd) %>%
#                 ddbs_collect() %>%
#     select(!fid) %>% # if intermediate layer used
# 
#     st_write(path_gpkg,
#              layer = lyr,
#              append = TRUE)
# }

# ## Read intermediate layer (if needed)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "TCCD_tmp_layers/ET014_LAU_3035.gpkg")
# layers <- st_layers(path_gpkg)$name
# 
# walk(layers, function(lyr) {
# 
#   ddbs_data <- ddbs_open_dataset(path_gpkg, layer = lyr, conn = conn_tcd)
# 
#   ddbs_write_table(data = ddbs_data,
#                    conn = conn_tcd,
#                    name = paste0("lau_et014_", lyr),
#                    overwrite = TRUE)
# 
#   message("Wrote layer: ", lyr)
# })


## Clean duckdb database
tables <- dbListTables(conn_tcd)
tables <- tables[!tables %in% c(keep(tables, startsWith(tables, "cln_lau")),
                               keep(tables, startsWith(tables,"lau_et014")))]

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_tcd, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }else{

    dbExecute(conn_tcd, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_tcd, tbl)))

  }

}


## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs")))

```

Now we can use `exact_extract()` to retrieve the tree cover density overlapping the ET04 polygons.

```{r}
#| eval: true
#| results: false
#| message: false
#| code-summary: "Spatial intersection ET04-LAU and TCD data"
#| 

## Merge then back together and store them in R environment for parallel processing at the next step
tables <- dbListTables(conn_tcd)
tables <- keep(tables, startsWith(tables, "lau_et014_"))

lau_et04_all <- map(tables, function(tbl) {

  as_duckspatial_df(tbl, conn = conn_tcd) %>%
    ddbs_collect()}) %>%
  bind_rows() %>%
  filter(okosystemtype1_name == "Forest and woodlands")


## Fylke number for parallelisation
fylke_nb <- dbGetQuery(conn_tcd, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

## Path to TCD virtual rasters 
path_tcd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/TCD_Copernicus"
path_tcd_list <- dir(path_tcd, recursive = FALSE, full.names = TRUE)
yrs_tcd <- str_sub(path_tcd_list, start = -4)
file_tcd <- map_vec(seq_along(path_tcd_list), function(i){paste0(path_tcd_list[i], "/TCD_", yrs_tcd[i], ".tif")})


## Retrieve IMD per year and fylke
daemons(5)

lau_et04_tcd <- map(seq_along(file_tcd), function(yr){
  
  # Year
  year <- yrs_tcd[yr]
  
  message("Processing: ", year)
  
    # Parallelise
  tcd_yr <- map(seq_along(fylke_nb), in_parallel(function(fyk){
    
    # Libraries needed
    library(duckspatial)
    library(exactextractr)
    library(dplyr)
    library(terra)
    library(duckdb)
    library(DBI)

    # et01 for filke fyk
    lau_et04 <- filter(lau_et04_all,
                       Fylke_nummer == fylke_nb[fyk])
    
    # IMD input
    tcd <- rast(file_tcd[yr])
    
    # Run exact_extract
    exact_extract(tcd, lau_et04, fun = c("mean"), append_cols = TRUE) %>%
      mutate(`Accounting period` = year)
    
  },
  file_tcd = file_tcd,
  year = year,
  fylke_nb = fylke_nb,
  lau_et04_all = lau_et04_all,
  yr = yr))
  
})

daemons(0)


# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs",
                    "lau_et04_tcd")))

```


We do the same with ET01 polygons.
```{r}
#| eval: true
#| results: false
#| message: false
#| code-summary: "Spatial intersection ET01-LAU and TCD data"
#| 

## Merge then back together and store them in R environment for parallel processing at the next step
tables <- dbListTables(conn_tcd)
tables <- keep(tables, startsWith(tables, "lau_et014_"))

lau_et01_all <- map(tables, function(tbl) {

  as_duckspatial_df(tbl, conn = conn_tcd) %>%
    ddbs_collect()}) %>%
  bind_rows() %>%
  filter(okosystemtype1_name == "Settlements and other artificial areas")


## Fylke number for parallelisation
fylke_nb <- dbGetQuery(conn_tcd, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

## Path to TCD virtual rasters 
path_tcd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/TCD_Copernicus"
path_tcd_list <- dir(path_tcd, recursive = FALSE, full.names = TRUE)
yrs_tcd <- str_sub(path_tcd_list, start = -4)
file_tcd <- map_vec(seq_along(path_tcd_list), function(i){paste0(path_tcd_list[i], "/TCD_", yrs_tcd[i], ".tif")})


## Retrieve IMD per year and fylke
daemons(5)

lau_et01_tcd <- map(seq_along(file_tcd), function(yr){
  
  # Year
  year <- yrs_tcd[yr]
  
  message("Processing: ", year)
  
    # Parallelise
  tcd_yr <- map(seq_along(fylke_nb), in_parallel(function(fyk){
    
    # Libraries needed
    library(duckspatial)
    library(exactextractr)
    library(dplyr)
    library(terra)
    library(duckdb)
    library(DBI)

    # et01 for filke fyk
    lau_et01 <- filter(lau_et01_all,
                       Fylke_nummer == fylke_nb[fyk])
    
    # IMD input
    tcd <- rast(file_tcd[yr])
    
    # Run exact_extract
    exact_extract(tcd, lau_et01, fun = c("mean"), append_cols = TRUE) %>%
      mutate(`Accounting period` = year)
    
  },
  file_tcd = file_tcd,
  year = year,
  fylke_nb = fylke_nb,
  lau_et01_all = lau_et01_all,
  yr = yr))
  
})

daemons(0)


# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs",
                    "lau_et04_tcd",
                    "lau_et01_tcd")))

```

Finally, we do the extract the tree cover density with the LAU classified as "cities" (`DEGURBA = 1`) 
```{r}
#| eval: true
#| message: false
#| code-summary: "Spatial intersection LAU and TCD data"
#| 

## Merge then back together and store them in R environment for parallel processing at the next step
lau_1 <- as_duckspatial_df("cln_lau", conn = conn_tcd) %>%
              ddbs_collect() %>%
              filter(DEGURBA == 1)


## Fylke number for parallelisation
fylke_nb <- dbGetQuery(conn_tcd, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

## Path to TCD virtual rasters 
path_tcd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/TCD_Copernicus"
path_tcd_list <- dir(path_tcd, recursive = FALSE, full.names = TRUE)
yrs_tcd <- str_sub(path_tcd_list, start = -4)
file_tcd <- map_vec(seq_along(path_tcd_list), function(i){paste0(path_tcd_list[i], "/TCD_", yrs_tcd[i], ".tif")})


## Retrieve IMD per year and fylke
daemons(5)

lau_1_tcd <- map(seq_along(file_tcd), function(yr){
  
  # Year
  year <- yrs_tcd[yr]
  
  message("Processing: ", year)
  
    # Parallelise
  tcd_yr <- map(seq_along(fylke_nb), in_parallel(function(fyk){
    
    # Libraries needed
    library(duckspatial)
    library(exactextractr)
    library(dplyr)
    library(terra)
    library(duckdb)
    library(DBI)

    # et01 for filke fyk
    lau <- filter(lau_1,
                    Fylke_nummer == fylke_nb[fyk])
    
    # IMD input
    tcd <- rast(file_tcd[yr])
    
    # Run exact_extract
    exact_extract(tcd, lau, fun = c("mean"), append_cols = TRUE) %>%
      mutate(`Accounting period` = year)
    
  },
  file_tcd = file_tcd,
  year = year,
  fylke_nb = fylke_nb,
  lau_1 = lau_1,
  yr = yr))
  
})

daemons(0)


# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd")))

```


### 9.2 Compile TCCD indicators

#### TCCD_001
```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_001"

## TCCD national level - area-weighted mean
tccd_001_nat <- map(seq_along(lau_et04_tcd), function(x){
                      lau_et04_tcd[[x]] %>%
                        bind_rows() %>%
                        select(!fid) %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_001_nat <- tccd_001_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in forests",
                         ID = "TCCD_001",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)


## TCCD municipal level
tccd_001_mun <- map(seq_along(lau_et04_tcd), function(x){
                        lau_et04_tcd[[x]] %>%
                          bind_rows() %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                        select(!fid) %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`)) %>%
                          relocate(LAU_NAME, .after = GISCO_ID) %>%
                          rename("Accounting area (municipality name)"= LAU_NAME,
                                 "Accounting area (municipality number)"= GISCO_ID) }) 
  
tccd_001_mun <- tccd_001_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in forests",
                         ID = "TCCD_001",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)


                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "tcd_crs",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun")))
```



#### TCCD_100
```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_100"

# Compile LAUs area
lau_area <- as_duckspatial_df("cln_lau", conn = conn_tcd) %>%
              ddbs_area(new_column = "area_m2_lau") %>%
              ddbs_collect() %>%
              st_drop_geometry() %>%
              select(GISCO_ID, area_m2_lau)

## TCCD national level - area-weighted mean
tccd_100_nat <- map(seq_along(lau_1_tcd), function(x){
                      lau_1_tcd[[x]] %>%
                        bind_rows() %>%
                        left_join(., lau_area, by = "GISCO_ID") %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2_lau, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_100_nat <- tccd_100_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_001",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## TCCD municipal level
tccd_100_mun <- map(seq_along(lau_1_tcd), function(x){
                        lau_1_tcd[[x]] %>%
                          bind_rows() %>%
                          left_join(., lau_area, by = "GISCO_ID") %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2_lau, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                            relocate(LAU_NAME, .after = GISCO_ID) %>%
                            rename("Accounting area (municipality name)"= LAU_NAME,
                                   "Accounting area (municipality number)"= GISCO_ID) }) 
  
tccd_100_mun <- tccd_001_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_100",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun",
                    "tccd_100_nat",
                    "tccd_100_mun")))
```


#### TCCD_101-104
```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_101"

## TCCD national level - area-weighted mean
tccd_101_nat <- map(seq_along(lau_et01_tcd), function(x){
                      lau_et01_tcd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 1) %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_101_nat <- tccd_101_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_101",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## TCCD municipal level
tccd_101_mun <- map(seq_along(lau_et01_tcd), function(x){
                        lau_et01_tcd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 1) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                            relocate(LAU_NAME, .after = GISCO_ID) %>%
                            rename("Accounting area (municipality name)"= LAU_NAME,
                                   "Accounting area (municipality number)"= GISCO_ID) }) 
  
tccd_101_mun <- tccd_101_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_101",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun",
                    "tccd_100_nat",
                    "tccd_100_mun",
                    "tccd_101_nat",
                    "tccd_101_mun")))
```

```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_102"

## TCCD national level - area-weighted mean
tccd_102_nat <- map(seq_along(lau_et01_tcd), function(x){
                      lau_et01_tcd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 2) %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_102_nat <- tccd_102_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_102",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## TCCD municipal level
tccd_102_mun <- map(seq_along(lau_et01_tcd), function(x){
                        lau_et01_tcd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 2) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                            relocate(LAU_NAME, .after = GISCO_ID) %>%
                            rename("Accounting area (municipality name)"= LAU_NAME,
                                   "Accounting area (municipality number)"= GISCO_ID) }) 
  
tccd_102_mun <- tccd_102_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_102",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun",
                    "tccd_100_nat",
                    "tccd_100_mun",
                    "tccd_101_nat",
                    "tccd_101_mun",
                    "tccd_102_nat",
                    "tccd_102_mun")))
```

```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_103"

## TCCD national level - area-weighted mean
tccd_103_nat <- map(seq_along(lau_et01_tcd), function(x){
                      lau_et01_tcd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 3) %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_103_nat <- tccd_103_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_103",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## TCCD municipal level
tccd_103_mun <- map(seq_along(lau_et01_tcd), function(x){
                        lau_et01_tcd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 3) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                            relocate(LAU_NAME, .after = GISCO_ID) %>%
                            rename("Accounting area (municipality name)"= LAU_NAME,
                                   "Accounting area (municipality number)"= GISCO_ID) }) 
  
tccd_103_mun <- tccd_103_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_103",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun",
                    "tccd_100_nat",
                    "tccd_100_mun",
                    "tccd_101_nat",
                    "tccd_101_mun",
                    "tccd_102_nat",
                    "tccd_102_mun",
                    "tccd_103_nat",
                    "tccd_103_mun")))
```


```{r}
#| eval: true
#| message: false
#| code-summary: "Compile TCCD_104"

## TCCD national level - area-weighted mean
tccd_104_nat <- map(seq_along(lau_et01_tcd), function(x){
                      lau_et01_tcd[[x]] %>%
                        bind_rows() %>%
                        summarise("Average tree canopy cover density (%)" = round2(stats:::weighted.mean(mean, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

tccd_104_nat <- tccd_104_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Tree canopy cover density in urban areas",
                         ID = "TCCD_104",
                         `Accounting area` = "Norway",
                         .before = `Average tree canopy cover density (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "file_tcd",
                    "lau_et04_tcd",
                    "lau_et01_tcd",
                    "lau_1_tcd",
                    "tccd_001_nat",
                    "tccd_001_mun",
                    "tccd_100_nat",
                    "tccd_100_mun",
                    "tccd_101_nat",
                    "tccd_101_mun",
                    "tccd_102_nat",
                    "tccd_102_mun",
                    "tccd_103_nat",
                    "tccd_103_mun",
                    "tccd_104_nat")))
```



## 10. Results

From the summary table that we exported previously, we create a summary table with the following indicators:

* **TCCD_001**: Average TCCD in the forest areas of Norway (class Forest and woodlands of the EU Ecosystem Typology)

* **TCCD_100**: Average TCCD in the *whole area of* the municipalities considered as "cities" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 1`). 

* **TCCD_101**: Average TCCD in the urban areas (class Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "cities" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 1`)

* **TCCD_102**: Average TCCD in the urban areas (class Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "towns and suburbs" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 2`)

* **TCCD_103**: Average TCCD in the urban areas (class Settlements and other artificial areas of the EU Ecosystem Typology) within the municipalities considered as "rural" according to Eurostat classification: degree of urbanisation equal to 1 (`DEGURBA = 3`)

* **TCCD_104**: Average TCCD in the urban areas of Norway (class Settlements and other artificial areas of the EU Ecosystem Typology) 

Below, we present two kinds of summary table: one national, and one presenting a breakdown by municipalities. Note that due to methodological issues, results for Røst and Træna are not available for TCCD_001. Also, due to methodological limitations, no standard error is associated with TCCD_100 which are indicated as "Not available". Further improvements will be implemented with future revisions of the TCCD indicators.

### National Results
```{r natRes}
#| eval: true
#| code-summary: "National summary table"

# Merge all tables
res_national <- bind_rows(tccd_001_nat,
                          tccd_100_nat,
                          tccd_101_nat,
                          tccd_102_nat,
                          tccd_103_nat,
                          tccd_104_nat)


# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_tcd",
                    "res_national",
                    "tccd_001_mun",
                    "tccd_100_mun",
                    "tccd_101_mun",
                    "tccd_102_mun",
                    "tccd_103_mun")))

# Show table
datatable(res_national, filter = "top") %>%
  formatRound(columns = "Average tree canopy cover density (%)", digits = 2)
```

### Municipality level
Træna and Røst are missing from TCCD_001.
```{r}
#| eval: true
#| message: false
#| code-summary: "Municipal summary table"


# Merge all tables
res_municipal <- bind_rows(tccd_001_mun,
                           tccd_100_mun,
                           tccd_101_mun,
                           tccd_102_mun,
                           tccd_103_mun)

# Close duckdb connection
ddbs_stop_conn(conn_tcd)

# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "res_municipal",
                    "res_national")))

# Show table
datatable(res_municipal, filter = "top") %>%
  formatRound(columns = "Average tree canopy cover density (%)", digits = 2)
```


## 11. Export file
Finally we present a sample code which can export the results into csv or xlsx files: 

```{r export}
#| eval: false
#| code-summary: "Code to export the result tables above"

#CSV
write.table(res_national, "TCCD_NO_national_2018_2024.csv")
write.table(res_municipal, "TCCD_NO_regional_2018_2024.csv")

#Excel
write_xlsx(res_national, "TCCD_NO_national_2018_2024.xlsx")
write_xlsx(res_municipal, "TCCD_NO_regional_2018_2024.xlsx")

```

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
