---
title: "ASPT (invertebrate index of trophic status)"
subtitle: |
  [NO_ASPT_001]
indicatorID: 
  - NO_ASPT_001
indicatorName: 
  - ASPT (invertebrate index of trophic status)
folderName: NO_ASPT_001
country: Norway
continent: Europe
ECT: A2 - Chemical State characteristics
Realm: 
  - Freshwater (F)
Biome: 
  - F1 Rivers and streams biome
Ecosystem: 
  - F1.3 Freeze-thaw streams
yearAdded: 2025
yearLastUpdate: 2025
status: deprecated
Version: "001.900"
VersionComment: Superseded by NO_WSPT_001
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_ASPT_001
Normalised: Yes
SpatialAggregationPathway: Scale - Transform - Aggregate (length-weighted arithmetic mean) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research
AuthorList: Sandvik, H.  
date: December 4, 2025
data_availability: none
code_reproducibility: none
open_science_badge: none
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here::here("extract_yaml.R"))
```





<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="5"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]),
  folder_name = folder_name)
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```
::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 24 Jan 2025 - Version 000.001 (should have been version 001.000) started: first complete version
- 18 Feb 2025 - Version 000.001 (should have been version 001.000) submitted to review
- 06 Mar 2025 - Version 000.001 (should have been version 001.000) revised and ready for publication
- 04 Dec 2025 - Version 001.900: superseded by NO_WSPT_001, version 002.000
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

The indicator NO_ASPT_001 (**ASPT**) has from version 002.000 been renamed as NO_WSPT_001
(so that all indicators derived from the Water Framewotk Directive start with "W").

* Please refer to [NO_WSPT_001](https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_WSPT_001) for documentation.


<!--# 
Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.
-->

<!--# 
## 11. Export file

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```

<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::