---
title: "Urban Green Areas"
subtitle: "[NO_URGR_001 to NO_URGR_005] - Green areas in settlements"
indicatorID: 
  - NO_URGR_001
  - NO_URGR_002
  - NO_URGR_003
  - NO_URGR_004
  - NO_URGR_005
indicatorName: 
  - Share of Urban Green Areas
verbatimName: 
  - Andel grønt areal 
  - Part des espaces verts urbains
folderName: NO_URGR_001-005
country: Norway
continent: Europe
ECT: B2 - Structural State Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T7 Intensive land-use biome
Ecosystem: 
  - T7.4 Urban and industrial
verbatimEcosystem: Bebygd og opparbeidet areal
yearAdded: 2025 
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "000.003" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Version 000.002 reviewed with new available data
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_URGR_001-005
Normalised: No
SpatialAggregationPathway: Aggregate (target area; area weighted mean) - Aggregate at focal area (area weighted mean)
Precision: No metric available #Describe the precition metric you have used. Precision may be reported either as variation among observed values (e.g. standard deviation, variance, range, interquartile range) or as uncertainty around an estimated value (e.g. standard error, confidence interval, credible interval). 
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Sylvie Clappe             
    email: sylvie.clappe@nina.no 
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research 
  - name: Bálint Czúcz              
    email: balint.czucz@nina.no
    affiliations:
      - ref: myID
AuthorList: Clappe, S., Czúcz, B.
date: July 13, 2026
data_availability: bronze
code_reproducibility: silver
open_science_badge: bronze
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
bibliography: references.bib
codeReviewers:
  - name: XXXXX
    email: XXXXXX
    affiliations:
      - id: reviewerInst
        name: XXXXXX
    reviewType: code review
    github: XXXXXXX
---


```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(terra)
library(sf)
library(dplyr)
library(readxl)
library(httr2)
library(glue)
library(duckdb)
library(duckspatial)
library(mirai)
library(purrr)
library(stringr)
library(exactextractr)
library(carrier)
library(DT)
library(numform)
library(conflicted)
library(giscoR)

# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```

::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this log with short messages for each update -->
- 30 Jun. 2025 - Version used for NINA Report 2604 
- 01 Sep. 2025 - Original version for ecRxiv
- 08 Oct. 2025 - Cleanup, simplifications
- 17 Oct. 2025 - Last cleanup
- 24 Oct. 2025 - Second revision
- 27 Oct. 2025 - Minor edits by ecRxiv (ALK)
- 26 Nov. 2025 - Embed metadata and include open science badges (ALK)
- 13 July 2026 - Reviewed code to include new available data
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary
1. **Background and rationale**:
Green areas in settlements, commonly also called _urban green areas_, provide several critical services for urban residents, such as improving air quality, reducing noise levels, regulating air temperature, providing recreation opportunities, and creating an aesthetically pleasing living environment. The amount of urban green areas available in a settlement has significant impacts on the mental and physical health of its inhabitants. Accordingly, the _share of urban green areas_ is a useful condition indicator for urban ecosystems. This is in line with the EU Regulation on Environmental Economic Accounting (EU Regulation 2024/3024; @eu-2024-3024), which lists _"green areas in cities and adjacent towns and suburbs"_ among the mandatory ecosystem condition indicators to characterise the condition of the ecosystem type _settlements and other artificial areas_ (ET01, henceforward “urban ecosystems” for short;  @handbook2026).[^1] The diverse definitions of _urban green areas_ often also include small water bodies in an urban context ("_urban blue_"), which are thus considered as a part of _urban green_. In line with this, the indicator presented here also integrates urban blue areas. 

2. **Methods**:
There are 3 major steps for calculating this indicator in an arbitrary _ecosystem accounting area_ (e.g. a specific municipality, metropolitan region, county, or the entire country):

1. Identify the _‘target area’_ for which the indicator should be calculated (i.e., the total "urban" area, cf. @Czucz2025a).
2. Within this target area, identify the _'focal features'_ (i.e. the "green areas", cf. @Czucz2025a).
3. Divide the area of the _focal features_ with that of the _target area_.

There are several ways of defining the _target area_ ("urban") and the _focal features_ ("green areas"), and there are several different datasets that can be used. In this document, we present five variants of the indicator based on different interpretations and datasets -- further possible variants are discussed in @Czucz2025a. 

3. **Interpretation**:
A value of 1 would mean that a settlement is fully covered in green areas: a situation that is both impossible, and probably also undesirable. Nevertheless, higher values are generally better, because they mean a greener, more "liveable" city.   

4. **Results and development needs**:
Overview of the five indicators that we calculated in this document. For a national use we recommend URGR_004, as this is variant combines an ecologically meaningful definition of the focal features ("green areas") with a calculation pathway compatible with the Systems of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA; @seea-ea), whereby the target area corresponds to the parent ecosystem[^2]. Nevertheless, this indicator does not follow the official guidance from the European Commission. If users need to report this indicator in an official EU reporting, we recommend to use URGR_005 (possibly with URGR_001 or URGR_004 as complementary indicators). The calculation of URGR_001--URGR_003 for diverse ecosystem accounting areas (EAAs) might also be useful for the additional analytical insight offered by distinguishing the different types of settlements according to their degree of urbanisation. 

* **URGR_001**: The *target area* is the area of the ecosystem type ET01 (urban ecosystems) within the municipalities considered as *"cities"* according to Eurostat's degree of urbanisation (DEGURBA) classification (`DEGURBA = 1`). The definition of the *focal features* are defined as the non-impervious areas in the Copernicus Imperviousness Density layer (IMD). This option is close to the Eurostat voluntary indicator (2) (@handbook2026)[^2].

* **URGR_002**: The *target area* is the area of the ecosystem type ET01 (urban ecosystems) within the municipalities considered as "*towns and suburbs*" according to Eurostat's DEGURBA classification (`DEGURBA = 2`). The definition of the *focal features* is the same than for URGR_001 (non-built-up pixels in Copernicus IMD). 

* **URGR_003**: The *target area* is the area of the ecosystem type ET01 (urban ecosystems) within the municipalities considered as "*rural areas*" according to Eurostat DEGURBA classification (`DEGURBA = 3`). The definition of the *focal features* is the same than for URGR_001 (non-built-up pixels in Copernicus IMD). URGR_002 and URGR_003 are complementary to URGR_001, they present an overview of the share of green areas in the three main types of settlements in Norway. The area-weighted mean of these three indicators (URGR_001-URGR_003, weigthed by the total area of settlements (ET01) in the respective categories) will be equal to URGR_004 for any reporting period.

* **URGR_004**: The *target area* is the total area of the ecosystem type ET01 (urban ecosystems) in Norway, while the *focal features* are defined as for URGR_001 (non-built-up pixels in Copernicus IMD). This is the option which is most aligned with the principles of SEEA EA (@seea-ea), where an ecosystem condition indicators are expected to be representative of their "parent" ecosystem type (see also @Czucz2025a). 

* **URGR_005**: The *target area* consists of "metropolitan zones" defined as the area of the municipalities considered as *cities* by Eurostat (`DEGURBA = 1`) combined with the *towns and suburbs* (`DEGURBA = 2`) that are adjacent to them.[^3] The *total area* (whole municipalities) of these "metropolitan zones" is considered as target area. The *focal features* are defined as any ecosystem type that have a green (or blue) cover during (the majority of) the vegetation season. As the total area of Norwegian municipalities is typically dominated by natural ecosystems (forests, wetlands, mountains, etc.) this option actually contradicts the "parent ecosystem principle" of ecosystem accounting (@Czucz2025a), and the resulting indicator is more similar to a little ecosystem extent account than to a condition variable. Nevertheless this option is the mandatory condition variable in the Accounting Regulation (EU Regulation 2024/3024; @eu-2024-3024), and it is presented as the main approach for EU reporting in Eurostat's guidelines @handbook2026. 


<!-- Summary footenotes -->
[^1]: The KM-GBF headline indicator [12.1](^https://www.gbf-indicators.org/metadata/headline/12-1$) ("average share of the built-up area of cities that is green/blue space for public use for all") serves a seemingly similar purposes, but unlike the indicator presented here, it only covers urban green and blue areas that are _accessibile_ to the public. Similarly, two of the Norwegian national environmental indicators (miljøindikator [3.1.2](^https://miljostatus.miljodirektoratet.no/miljomal/friluftsliv/miljomal-3.1/miljoindikator-3.1.2$) and [3.1.3](^https://miljostatus.miljodirektoratet.no/miljomal/friluftsliv/miljomal-3.1/miljoindikator-3.1.3$)). The share of _all green_ areas, and the share of _accessible green_ areas can both be considered as relevant characteristics of the urban ecosystems, with slightly different functionalities: while the total amount of _all_ green surfaces is critically important for most regulating ecosystem services, the share of _accessible_ greenspaces is vital for cultural services, in particular recreation. The variable that is presented here addresses all green areas. Accessible green areas, if needed, can be implemented in the future as a separate condition indicator. 

[^2]: The difference lay in the definition of the *target area* which does not include adjacent towns and suburbs; and  the use of Copernicus IMD to define the *target features* instead of any ecosystem types having a green (or blue) cover.

[^3]: Currently there are four such metropolitan areas in Norway: Oslo (consisting of Oslo\*, Ås, Asker, Bærum\*, Lier, Lillestrøm, Lørenskog\*, Nittedal, Nordre Follo\*, Rælingen\*), Bergen (consisting of Bergen\*, Bjørnafjorden), Trondheim (consisting of Trondheim\*, Malvik), and Stavanger (consisting of Stavanger\*, Gjesdal, Klepp, Randaberg, Sandnes\*, Sola, Strand, Time). Here we marked the municipalities of `DEGURBA = 1` ("cities") with an asterisk (\*).  




## 2. About the underlying data
We use the following three main datasets in our calculations:

* **Copernicus HRL IMD 2018**: High resolution raster layers from the Copernicus Land Monitoring Service used to define the *focal features* (green areas) in the variants URGR_001-004. More concretely, we define "green areas" as the total area of non-impervious surfaces within the target area, which thus covers all green and blue surfaces, including any parks, gardens, lawns, hedges, green roofs, ponds, rivers, forests, etc. that are small enough to belong to the *target area*.[^4]

* **Grunnkart**: Norway's official *ecosystem type map* designed to support ecosystem accounting (@strand2024; @gk2). This map is used to define the *focal features* in the variant aligned with the EU mandatory indicator (URGR_005) and define the *target area* (URGR_001-004). In URGR_005 we define *urban green areas* as any occurrences of the following ecosystem types within the *target area* (the entire area of the metropolitan municipalities): 1.4 Urban greenspace; 1.5.2 Cemeteries; 2. Cropland; 3. Grassland; 4. Forest and woodlands; 5. Heathlands and shrub; 6. Sparsely vegetated ecosystems; 7. Inland wetlands; 8. Rivers and canals; 9. Lakes and reservoirs; 11. Coastal beaches, dunes and wetlands.
This list includes all ecosystems, within the EU Ecosystem Typology (@et-typo-eurostat), that have a green (or blue) cover during (the majority of) the vegetation season (including large natural ecosystems outside an urban context). 

In addition we use several further administrative datasets to delineate the individual municipalities (i.e. Local Administrative Units; LAU), as well as to determine their "degree of urbanisation" (DEGURBA), and to assign them to regions (counties, "fylke").

[^4]: Eurostat recommends the use of the CLC+ Backbone product by default (@handbook2026). CLC+ Backbone could have been an alternative data source for delineating the "focal features" instead of IMD. Nevertheless, IMD allowed us to use a much higher (1x1 m, subpixel-level) spatial resolution for detecting "green areas" (see more details in section 2.1), which is why we chose this product over CLC+ backbone.


### 2.1 Spatial and temporal resolution and extent
The nominal resolution of Copernicus HRL IMD is 100 m^2^ (10 x 10 m), but each pixel contains information about the density (%) of (non-)impervious surfaces in the grid cell, making it possible to determine the density of focal features (urban green) with a useful resolution of 1 m^2^ (variants URGR_001-004). This layer is normally updated in every third year, and the updates usually become available with a delay of 2-3 years. 

All other spatial data files, including the Grunnkart and the Local Administrative Units (LAU) are vector data (polygons). Unlike traditional land cover maps, the Grunnkart does not have a single, universal Minimum Mapping Unit (MMU). Instead, the spatial resolution of the Grunnkart is directly inherited from the underlying source datasets used in its creation. This is because the new map is a synthesis and reclassification of existing data rather than a new, independent mapping effort. As the majority of spatial information available in public datasets focuses on human settlements and infrastructure, urban ecosystems are relatively finely represented in the Grunnkart, especially compared to seminatural and natural ecosystem types. This makes the contours of ET01 overly-detailed, often delineating individual buildings or roads passing through a forest as “urban ecosystems”. This also makes the dataset somewhat unfit for its purpose (@Rendon2025) and cumbersome to process.


### 2.2 Original units
The original unit of Copernicus HRL imperviousness is percentages (the share of impervious areas in the pixel), which is the complement of our quantity of interest (1-imperviousness, interpreted as the share of "urban green" surfaces in the pixel, also in %). This is congruent with the units of the final variable (the share of urban green in the "target area"), which can be expressed both as a number between 0 and 1, as well as in percentages (%). 

### 2.3 Additional comments about the dataset
#### 2.3.1 Instructions for citing, using and accessing data

|| Publicly accessible to download | Data and metadata available at |
|----|--|--------|
| **Copernicus HRL IMD** | Yes | Available from the [EEA](https://land.copernicus.eu/en/products/high-resolution-layer-imperviousness)|
| **Grunnkart**  | No, only upon request to [NIBIO](https://kartkatalog.geonorge.no/metadata/nasjonalt-grunnkart-for-arealanalyse-testversjon-2/28c28e3a-d88f-4a34-8c60-5efe6d56a44d) | We used testversion 0.2, description available in @strand2024 and @gk2 |
|**Local Administrative Units**| Yes | Available from Eurostat plateform [Gisco](https://ec.europa.eu/eurostat/web/gisco/geodata/statistical-units/local-administrative-units) |
|**DEGURBA**| Yes | Available from [Eurostat](https://ec.europa.eu/eurostat/web/nuts/local-administrative-units)


## 3. Indicator properties
### 3.1 Ecosystem Condition Typology Class (ECT)
The condition variable *urban green areas* (i.e. the share of green areas in urban ecosystems) belongs to ECT class B2 (structural state characteristics).


### 3.2 Ecosystem condition characteristic
The share of *urban green* areas characterises the amount of biologically active surfaces that is present in an urban environment. Urban green spaces provide important ecosystem services that are essential to the wellbeing of the human communities living in urban settlements. The presence of green spaces also ensures habitats for wildlife with urban areas.

### 3.3 Other standards
In this documentation, we present different variations of the urban green indicator indicator. We recommend to use **URGR_004** as it relies upon the most ecologically correct definition, in line with the SEEA EA (@seea-ea). It however does not follow the official guidance from the European Commission for national reporting under EU Regulation 2024/3024 (@eu-2024-3024). If users need to report this indicator in an official EU reporting, we recommend to use **URGR_005**.


### 3.4 Collinearities with other indicators
This indicator might present a high correlation with the urban Tree Cover Density (TCCD) indicator. There may also be significant correlations with the Urban Air Quality (URAQ) indicator, as well as any future indicator reflecting the *accessibility* of urban green areas.[^1] 


### 3.5 Impact factors
The share of urban green spaces is overwhelmingly determined by anthropogenic processes, including conscious urban planning.


## 4. Reference condition and levels

### 4.1 Reference condition
For urban ecosystems, the SEEA EA suggests the identification of an “anthropogenically derived” reference condition (SEEA EA 5.72, a.k.a. “best attainable condition, @valecillo2022). This type of reference condition cannot be”sampled” (there is no clear reference set of “cities” in an “optimal/ideal/natural condition”), but an optimal reference level (X~100~) can possibly still be established (e.g. via extrapolation from an X~60~, see @Czucz2025b).


### 4.2 Reference levels
For this variable, the zero value can be considered as a universally meaningful "pessimal level" (X~0~): a situation with no green areas represents the worst possible condition of an urban area from the perspective of this variable. 
In principle, it could also be possible to assign further reference levels (e.g. a "good-bad" threshold, X~60~) based on elements from planning norms and thresholds (e.g. the "Blue-Green Factor", NS 3845), or from related national/international targets. Nevertheless, the meaning and relevance of these thresholds and targets in the different types of settlements must still be carefully studied before they are operatively used for rescaling this condition variable. And if both X~0~ and X~60~ are identified, then it will also be possible to set a meaningful X~100~ by means of extrapolation (@Czucz2025b).

In any case, the identification of the reference levels for this variable is not a purely ecologist task, and in the context of the current project the necessary interdisciplinary understanding could not be achieved.


#### 4.2.1 Spatial resolution and validity
X~0~ (no urban green) can be seen as a universal reference level, which is valid for any occurrences and subtypes of ET01 (in any type of settlement) at any spatial resolution over all Norway. Ideally, further reference levels (determined in the future) should also be universal reference levels – even if planning norms are different in the different municipalities. This would lead to a cleaner and more transparent indicator, and then the individual municipalities could express their differences by setting different targets for this variable. As each settlement consists of denser and greener parts, the spatial scale of applying the reference levels will also be critical.

## 5. Uncertainties
### 3.1 General uncertainties
There are several sources of uncertainties affecting the values of this indicator and its subsequent use:

- <u>data uncertainties in Copernicus IMD</u>: uncertainties have several sources all taking roots in the modelling used to process the Sentinel raw data. They can come from the pre-processing of the raw data necessary to correct the images (e.g., radiometrical/sensor calibration; geophysical corrections; cloud cover corrections;creation of the mosaic). But they also arise when the pre-processed data are classified through machine learning to produce the final layer of imperviousness. Classification is never definitive and produces probabilities for a pixel to belong to a specific class. Although, according to the official validation results, this dataset has a thematic accuracy above 90%, so this source of uncertainty is relatively minor. Finally, another source of uncertainty rely in the significant methodological changes between the three years downloaded: 2018, 2021, and 2024. Care should be taken when comparing imperviousness across years.[^5]

- <u>spatial, temporal and semantic uncertainties of the Grunnkart</u>: the current Grunnkart combines many different static data sources with various temporal reference dates, which does not seem to present an easily reproducible and updatable data source. This means that the outlines of the urban ecosystems (ET01) contain an unknown amount of uncertainty ("ad-hoc-ness"). The Grunnkart is currently a static dataset whose timeline has not been clearly documented as relying on temporally heterogeneous sources. Neverthless, we used it for the entire study period.

- <u>semantic uncertainties in the definition of *green areas* (focal features)</u>: for URGR_001-004 we defined *green* as the lack of imperviousness. Nevertheless, there may be some cases when an area in a city is neither impervious nor green (e.g. unpaved parking lots). No definition and data source can be completely free from such "semantic mismatches", and ideally their assessment should be part of a validation study for each indicator. Nevertheless, in the frame of the present study, we could not assess the magnitude of these semantic mismatches. 

- <u>thematic bias for URGR_005</u>: in this variant the green areas are dominated by a huge mass of natural ETs, and the value of indicator is ultimately determined by the shape & size of the municipalities (LAUs), i.e. distant administrative designations that have nothing to do with the condition of the respective settlement areas. From the perspective of the indicator this is a huge bias, which makes it completely unfit for purpose. 


<!-- Footnote for uncertainty section -->
[^5]: There has been significant changes is data processing between 2018, 2021 and 2024. First, 2021 and 2024 layers have been merged with CLC+ BackBone sealed layer. This ensures comparability between the two Copernicus products. Copernicus IMD 2018 however has not be merged with CLC+ BackBone sealed layer. Years 2021 and 2024 have been produced with the same convolutional neural network called U-Net, while the model used for 2018 is not specified in the metadata but is stated as being different. Finally, the threshold of prediction for an imperviousness density to be reported in a pixel was 90% in 2021 but 70% in 2024. This means that care should be taken when using these data for temporal series analyses. To ensure consistency across time, it would have been preferable to use the Imperviousness change layers that are designed to be homogeneous over time, correcting for the different changes in methodology. However, these layers are only available at 20m resolution. Downscaling theses layers to 10m or upscaling the original layers of HRL Imperviousness to 20m would results in additional uncertainties.


### 3.2 Reported uncertainties
Due to resources limitations, we could not quantify uncertainties linked to UURGR indicators. Future version of this documentation will look into exploiting the Copernicus reported "Confidence layers" for each of the IMD datasets. 


## 6. References
<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->


::: {#refs}
:::


<!-- Aune-Lundberg, L., Steinnes, M., Lund, M.O., Arneberg, E. & Munsterhjelm, N. 2025. Testversjon 2 – Nasjonalt grunnkart for bruk i arealregnskap. NIBIO Rapport 11/50/2025. (in Norwegian) -->

<!-- Czúcz, B., Framstad, E., Clappe, S. & Nowell, M. 2025a. Exploring the use of Eurostat’s mandatory and voluntary indicators in Norway. NINA Report 2604. Norwegian Institute for Nature Research. -->

<!-- Czúcz, B., Lappalainen, M., Jernberg, S., Korpinen, S. 2025b. Report on definition of reference conditions that describe good ecosystem condition. SELINA Deliverable 3.3. https://project-selina.eu/ -->

<!-- Eurostat 2023. Ecosystem condition accounts – Guidance Note, Fourth Draft. Eurostat Unit E2, Task Force on Ecosystem Accounting, 05 October 2023. Doc. ENV/EA/TF/2023_3/4v2. -->

<!-- Eurostat 2024. EU Ecosystem typology. Technical Note. Version: September 2024. Eurostat. Supplementary document to Doc. ENV/EA/TF/2024_3/2a. --> 

<!-- Eurostat 2025. Ecosystem Condition Accounts – Guidance Note, 6th Draft. Eurostat Unit E2, Task Force on Ecosystem Accounting, 26-27 February 2025. Doc. ENV/EA/TF/2025_1/2. https://circabc.europa.eu/ui/group/922b4700-1c83-4099-b550-763badab3ec0/library/5074c555-99ea-4931-ace8-9351fa174f8b -->

<!-- Strand, G.-H., Steinnes, M., Arneberg, E., Lund, M.O., Munsterhjelm, N., Aune-Lundberg, L. & Rørholt, A. 2024. Grunnkart for bruk i arealregnskap. NIBIO Rapport 10/28/2024. (in Norwegian) -->

<!-- UN et al. 2024. System of Environmental Economic Accounting – Ecosystem Accounting. United Nations, Department of Economic and Social Affairs, Statistics Division. Statistical Papers, Series F No. 124. -->

## 7. Import and prepare datasets
All the datasets used to compile the different versions of the URGR indicators have been thoroughly described in the previous sections, links to them are also provided for more information.

We used the following four datasets: 

* **D-1**: *IMD Copernicus data* (EEA): a series of raster datasets used to calculate share of urban green/blue spaces at different spatial scales. The rationale is that any non-impervious pixels are urban green/blue.

* **D-2**: *Grunnkart* (NIBIO): a high-resolution vector dataset used to delineate urban areas (ET01: settlements and other artificial areas) and other ecosystem types for URGR_005.

* **D-3** *LAU data* (Local Administrative Units, Eurostat GISCO): a vector dataset was used as the primary data source to delineate the  municipalities.

* **D-4** *DEGURBA*  (Degree of urbanisation): a tabular dataset assigning one of three DEGURBA classes (1: "cities"; 2: "towns and suburbs"; 3: "rural areas") to each municipality. These classes were used to breakdown the results for URGR_001-004.

We downloaded, cleaned and formatted the data prior to compiling URGR indicators. Section 7 only presents how to load the cleaned data. In the data analyses, we will use a mix of `duckspatial` and `terra` packages. The first one is faster at handling big vector data than `sf` but raster operations are not part of it, so we use `sf` and `terra` to handle vector-raster operations. Hence vector datasets are written in a duckdb database whose connection is set below:

```{r}
#| eval: true
#| code-summary: "Create a DuckDB connection"

# # For users who don't have a set up duckdb database
# conn_ug <- ddbs_create_conn(dbdir = "tempdir", threads = 10, memory_limit_gb = 30)

# If users already have a duckdb database they want to use
conn_ug <- ddbs_create_conn(dbdir = paste0(str_remove(tempdir(), "Rtmp\\w*"), "duckdb_conn/urgr.duckdb"), threads = 10, memory_limit_gb = 30)


```

### 7.1 Copernicus IMD (D-1)
We have three years and 159 rasters to load, we first create a list of files before reading them and storing them in a list object. The only cleaning step was to create a virtual raster. We did not apply any mask for Norway on the IMD data as the use of the `exact_extract()` function will restrict the urban green indicators to the polygons terrestrial urban ecosystems or terretsrial LAUs boundaries.

```{r}
#| eval: true
#| code-summary: "Load Copernicus IMD"

# Path to the Copernicus IMD data
path_imd <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/IMD_Copernicus"
path_imd_list <- dir(path_imd, recursive = FALSE, full.names = TRUE)
yrs_imd <- str_sub(path_imd_list, start = -4)
file_imd <- map_vec(seq_along(path_imd_list), function(i){paste0(path_imd_list[i], "/IMD_", yrs_imd[i], ".tif")})

# Set target CRS
imd_crs <- st_crs(rast(file_imd[1]))

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "imd_crs")))

```


### 7.2 Grunnkart (D-2)
D-2 is stored in the temporary duckdb database. We create one table for ET01 (Settlements and other artificial areas), and one for the other ecosystem types (ETs), excluding ET12 (Marine ecosystems) as they are excluded from Eurostat's definition of URGR_005.

The cleaning process of the Grunnkart included several steps: (i) correction of geometry; (ii) union of the polygons per ecosystem type at level 1 of the EU typology; (iii) BALINT.

To facilitate analyses in block processes afterwards, we create one table per fylke in the duckdb database. We add a step of re-projection to the CRS from Copernicus IMD to facilitate the data analyses later on. We make the geometry valid after transformation and clean the duckdb database of the original data.

```{r}
#| eval: false
#| code-summary: "Load Grunnkart - not ET01"

# Path to GK
gk_path <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/GK_processed/fylke_level"
gk_list <- dir(gk_path, pattern = "UTM33",recursive = FALSE, full.names = TRUE)

# Store in duckdb database
walk(gk_list, function(x){
  
  dbExecute(conn_ug, 
      sprintf("
        CREATE TABLE '%s' AS
        SELECT okosystemtype1_name, geom
        FROM ST_Read('%s')
        WHERE okosystemtype1_name IS DISTINCT FROM 'Settlements and other artificial areas'
      ", 
paste0("gk_", str_extract(x, "(?<=UTM33_f).{2}")),
x))
  
})

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

# Transform into IMD CRS
tbl_list <- dbListTables(conn_ug)

walk(tbl_list, function(tbl){
  
# transform  
ddbs_transform(tbl, imd_crs, conn = conn_ug) %>% 

    # make valid
    ddbs_make_valid(conn = conn_ug, name = paste0("cln_gk_not_et01_", str_extract(tbl, "(?<=gk_).{2}")), overwrite = TRUE)
  
})

# Check validity
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "cln_gk_not"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# # Write the intermidaite product in case of crash
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers/GK_not_et01_3035_")
tables <- grep('^cln_gk_', dbListTables(conn_ug), value = TRUE)

walk(tables, function(x){

  as_duckspatial_df(x, conn = conn_ug) %>%
  ddbs_collect() %>%
  st_write(paste0(path_gpkg, str_extract(x, "(?<=cln_gk_not_et01_).{2}"), ".gpkg"))
})

# Read intermediate layer (if necessary)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers")
# gk_list <- dir(path_gpkg, recursive = FALSE, pattern = "GK_not_et01", full.names = TRUE)
# 
# walk(gk_list, function(tbl) {
# 
#   ddbs_data <- ddbs_open_dataset(tbl, conn = conn_ug)
# 
#   ddbs_write_table(data = ddbs_data,
#                    conn = conn_ug,
#                    name = paste0("cln_gk_not_et01_", str_extract(tbl, "(?<=3035_).{2}")),
#                    overwrite = TRUE)
# 
#   message("Wrote layer: ", tbl)
# })

# Clean duckdb database
tables <- dbListTables(conn_ug)

tbl_list <- keep(tables, startsWith(tables, "cln_gk_not"))

tables <- tables[!tables %in% tbl_list]

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("file_imd",
                    "imd_crs",
                    "pull",
                    "conn_ug")))
```

```{r}
#| eval: false
#| code-summary: "Load Grunnkart - ET01"

# Path
gk_path <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/GK_processed/fylke_level"
gk_list <- dir(gk_path, pattern = "UTM33",recursive = FALSE, full.names = TRUE)


# Store in duckdb database
walk(gk_list, function(x){
  
  dbExecute(conn_ug, 
      sprintf("
        CREATE TABLE '%s' AS
        SELECT okosystemtype1_name, geom
        FROM ST_Read('%s')
        WHERE okosystemtype1_name = 'Settlements and other artificial areas'
      ", 
paste0("gk_et01_", str_extract(x, "(?<=UTM33_f).{2}")),
x))
  
})

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

# Transform into IMD CRS
tables <- dbListTables(conn_ug)
tbl_list <- keep(tables, startsWith(tables, "gk_et01"))

walk(tbl_list, function(tbl){
  
# transform  
ddbs_transform(tbl, imd_crs, conn = conn_ug) %>% 

    # make valid
    ddbs_make_valid(conn = conn_ug, name = paste0("cln_", tbl), overwrite = TRUE)
  
})

# Check validity
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "cln_gk_et01"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# # Write the intermediate product in case of crash
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers/GK_et01_3035_")
tables <- grep('^cln_gk_et', dbListTables(conn_ug), value = TRUE)

walk(tables, function(x){

  as_duckspatial_df(x, conn = conn_ug) %>%
  ddbs_collect() %>%
  st_write(paste0(path_gpkg, str_extract(x, "(?<=et01_).{2}"), ".gpkg"))
})

# Read intermediate layer (if necessary)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers")
# gk_list <- dir(path_gpkg, recursive = FALSE, pattern = "GK_et01", full.names = TRUE)
# 
# walk(gk_list, function(tbl) {
# 
#   ddbs_data <- ddbs_open_dataset(tbl, conn = conn_ug)
# 
#   ddbs_write_table(data = ddbs_data,
#                    conn = conn_ug,
#                    name = paste0("cln_gk_et01_", str_extract(tbl, "(?<=3035_).{2}")),
#                    overwrite = TRUE)
# 
#   message("Wrote layer: ", tbl)
# })


# Clean duckdb database of the raw tables
tables <- dbListTables(conn_ug) 
              
tables <- tables[!tables %in% keep(tables, startsWith(tables, "cln_"))]

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("file_imd",
                    "imd_crs",
                    "pull",
                    "conn_ug")))
```

The last thing we need to do is to load subtypes 1.4 Urban greenspace and 1.5.2 Cemeteries from the Grunnkart. For this, we kept an intermediary product from the cleaning process we did on the Grunnkart. This intermediary layer is the Grunnkart whose geometry has been corrected but the ETs has not yet been merged at level 1.

```{r}
#| eval: false
#| code-summary: "Load Grunnkart - Cemeteries and Urban greenspaces"

# list the files
gdb_folder <- "/data/R/GeoSpatialData/LandUse/Norway_Arealregnskap/Original/GrunnkartArealregnskap FGDB-format"

gdb_files <- list.files(gdb_folder, pattern = "_gdb\\.gdb$", 
                        full.names = TRUE)

# SQL query to read and select only urban cemeteries and greenspaces
walk(gdb_files, function(x){
  
  dbExecute(conn_ug, 
      sprintf("
        CREATE TABLE '%s' AS
        WITH src_data AS (
        SELECT okosystemtype_2, okosystemtype_3, geo
        FROM ST_Read('%s')
      )
        SELECT okosystemtype_2 AS okosystemtype_1, geo
        FROM src_data
        WHERE okosystemtype_2 = 'Urban greenspace' 
        
        UNION ALL
        
        SELECT okosystemtype_3 AS okosystemtype_1, geo
        FROM src_data
        WHERE okosystemtype_3 = 'Cemeteries'
      ", 
      
paste0("gk_urb_green_features_", str_extract(x, "(?<=format/).{2}")),
x))
  
})

# Project to IMD CRS, Clean slivers, make valid, union
tbl_list <- paste0("gk_urb_green_features_", str_extract(gdb_files, "(?<=format/).{2}"))

walk(tbl_list, function(tbl){
  
# transform  
ddbs_transform(tbl, imd_crs, conn = conn_ug) %>% 

    # remove slivers
    ddbs_area(new_column = "area_m2_proj") %>% 
        filter(area_m2_proj > 0.05) %>%
    
    # make valid
    ddbs_make_valid() %>% 
    
    # decompose complex geometry types into individual types
    ddbs_dump() %>% 
    
    # union
    ddbs_union_agg(by = "okosystemtype_1") %>% 
    
    # make valid
    ddbs_make_valid() %>% 
    
    # decompose and write in duckdb database
    ddbs_dump(conn = conn_ug, name = paste0("cln_", tbl), overwrite = TRUE) 
  
})

# Check validity
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "cln_gk_urb"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

# Clean duckdb database of the raw tables
tables <- dbListTables(conn_ug)

tables <- tables[!tables %in% keep(tables, startsWith(tables, "cln_"))]

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}

# Verify tables are in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

# Clean the R environment
rm(list = setdiff(ls(), 
                  c("file_imd",
                    "imd_crs",
                    "pull",
                    "conn_ug")))

```

### 7.3 LAU (D-3) and DEGURBA (D-4)
LAU data can be directly downloaded from Eurostat using the `giscoR` package. We download the data in temporary directory.

```{r}
#| eval: false
#| code-summary: "Load LAU data"
# Load administrative boundaries
# Create a temporary directory to store the cache
temp_dir <- tempdir()

# Download the dataset
lau_no <- gisco_get_lau(year = "2024",
                        epsg = "3035",
                        update_cache = TRUE,
                        cache_dir = temp_dir,
                        verbose = FALSE,
                        country = "NO",
                        gisco_id = NULL)
# Clean LAU codes
lau_no$GISCO_ID <- str_remove(lau_no$GISCO_ID, "NO_")


```

Now we load DEBURBA data and add it to `lau_no`.

```{r}
#| eval: false
#| code-summary: "Add DEGURBA to LAU data"

# Source function to add fylke number
source(paste0(here::here(),"/indicators/NO_URGR_001-005/R/functions/add_fylke.R"))

# Path to DEGURBA data
path_degurba <- "/data/P-Prosjekter2/412413_2023_no_egd/git_data/Adminstrative_boundaries_NO/LAU_DEGURBA_Eurostat/EU-27-LAU-2024-NUTS-2024.xlsx"

# Read excel file with Degurba variable and select Norway
degurba_lau_no <- read_excel(path_degurba, sheet = "NO") %>%
                    mutate(Fylke_nummer = add_fylke(., kommune_nummer = "LAU CODE")[[1]],
                           Fylke_navn = add_fylke(., kommune_nummer = "LAU CODE")[[2]],)

# Join lau_no with DEGURBA
lau_dgurb <- inner_join(lau_no,
                        degurba_lau_no,
                        by = join_by("GISCO_ID" == "LAU CODE")) %>%
              select(Fylke_nummer, Fylke_navn, GISCO_ID, LAU_NAME, DEGURBA, geometry) %>%
              st_as_sf()

# Re-project in the CRS of IMD data and store in duckdb database
st_transform(lau_dgurb, crs = imd_crs) %>%
      ddbs_make_valid(conn = conn_ug,
                      "cln_lau",
                      overwrite = TRUE)
 
# Check validity
valid_check <- ddbs_is_valid("cln_lau",
                              conn = conn_ug) %>%
                 filter(!is_valid) %>%
                 ddbs_collect() %>%
                 dim()

empty_check <- ddbs_is_empty("cln_lau",
                             conn = conn_ug) %>%
                  filter(is_empty == TRUE) %>%
                  ddbs_collect() %>%
                  dim()

simple_check <- ddbs_is_simple("cln_lau",
                               conn = conn_ug) %>%
                  filter(!is_simple) %>% 
                  ddbs_collect() %>%
                  dim()


# Verify table is in my duckdb conn_ug database
ddbs_list_tables(conn_ug)

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("file_imd",
                    "imd_crs",
                    "pull",
                    "conn_ug")))

```

## 8. Spatial units
The indicator values of URGR-001-004 can be interpreted at any spatial scale above the _characteristic scale_ of urban green areas which largely corresponds to the MMU for natural ecosystem types (any forests, lakes, etc below the MMU will be considered as parts of the urban ET, while the larger ones will be seen as non-urban natural ETs). So calculating URGR_001-004 for any planning and administrative unit above the quatier (bydel) level can generally be seen as meaningful. Nevertheless, coarser ecosystem accounting areas (EAA), like the *fylke*s or the *entire country* can also be highly relevant as reporting areas.  

For the variant URGR-005 the smallest interpretable reporting units are the entire metropolitan areas (the entire administrative area of specific groups of "cities and adjacent towns and suburbs").


## 9. Analyses

### 9.1 URGR_001-2-3-4
**Target area** = the area of ET01 in the LAU whose `DEGURBA` is `1` (URGR_001), `2` (URGR_002), `3` (URGR_003). The sum of all gives ET01 in all LAUs (URGR_004)
**Target feature** = the non-impervious pixels of Copernicus IMD.
**Compilation process** = spatial intersection between LAUs data and ET01. Then spatial intersection with IMD data, for all years. We compile a "non-impervious" percentage for each polygon and do an area-weighted mean at LAU level. These four indicators rely on the same compilation process, hence we do them all at once. *Note: if one wanted to compile URGR_004 directly, the first spatial intersection would not be necessary.*

We start by doing the spatial intersection between ET01 and LAUs. 
*Note: we add an extensive cleaning steps to ensure that the resultsing first intersection only has polygons or multipolygons gemoetries as it is a requirement from `exact_extract`.*

```{r}
#| eval: false
#| code-summary: "Spatial intersection ET01 and LAU"

## Intersect ET01 with LAUs - block processing to avoid memory issues

# Block intersections
fylke_nb <- dbGetQuery(conn_ug, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

walk(seq_along(fylke_nb), function(i){

  # Input data  
  et <- as_duckspatial_df(paste0("cln_gk_et01_", fylke_nb[i]), conn = conn_ug) 
  
  lau <- as_duckspatial_df("cln_lau", conn = conn_ug) %>%
          filter(Fylke_nummer == !!fylke_nb[i])

  
  # Intersection
  ddbs_intersection(lau,
                    et,
                    conn = conn_ug,
                    conn_x = conn_ug,
                    conn_y = conn_ug) %>% 
    
    ddbs_make_valid() %>%
    
    # Need to remove linestrings and geometrycollection for extact_extract
    mutate(geometry = ST_CollectionExtract(geometry, 3L)) %>% # clean up geometrycollections
    filter(!ST_IsEmpty(geometry)) %>% # remove empty geometry resulting from the step above
    filter(ddbs_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>% # keep only polygons and multipolygons
    
    ddbs_area(new_column = "area_m2", name = paste0("lau_et01_", fylke_nb[i]), conn = conn_ug, overwrite = TRUE)
  
})

# Verify table is in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

## Check valididty
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "lau_et01_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}

## Export the temporary layers, just in case of crash
path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers/ET01_LAU_3035.gpkg")

for (lyr in fylke_nb){
  
as_duckspatial_df(paste0("lau_et01_", lyr), conn = conn_ug) %>%
                ddbs_collect() %>%
    
    st_write(path_gpkg,
             layer = lyr,
             append = TRUE)
}

## Read intermediate layer (if needed)
# path_gpkg <- paste0(str_remove(tempdir(), "Rtmp\\w*"), "URGR_tmp_layers/ET01_LAU_3035.gpkg")
# layers <- st_layers(path_gpkg)$name
# 
# walk(layers, function(lyr) {
# 
#   ddbs_data <- ddbs_open_dataset(path_gpkg, layer = lyr, conn = conn_ug)
# 
#   ddbs_write_table(data = ddbs_data,
#                    conn = conn_ug,
#                    name = paste0("lau_et01_", lyr),
#                    overwrite = TRUE)
# 
#   message("Wrote layer: ", lyr)
# })


## Clean duckdb database
tables <- dbListTables(conn_ug)

tables <- keep(tables, startsWith(tables, "temp_"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}


## Clean the R environment
rm(list = setdiff(ls(), 
                  c("file_imd",
                    "imd_crs",
                    "pull",
                    "conn_ug")))

```


Now we do the spatial intersection with the ET01 in LAUs and IMD data. *Note: No need to wrap the raster because the virtual raster can be passed through workers. But we need to store the result from the previous intersection in the R environment because workers can't read from the duckdb databse due to the locking of the main session.*
```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Spatial intersection ET01-LAU and IMD data"
#| 

## Merge then back together and store them in R environment for parallel processing at the next step
tables <- dbListTables(conn_ug)
tables <- keep(tables, startsWith(tables, "lau_et01_"))

lau_et01_all <- map(tables, function(tbl) {

  as_duckspatial_df(tbl, conn = conn_ug) %>%
    ddbs_collect()}) %>%
  bind_rows() #%>%
  #select(!c(fid)) # if from intermediate layer


## Fylke number for parallelisation
fylke_nb <- dbGetQuery(conn_ug, "SELECT DISTINCT Fylke_nummer FROM cln_lau") %>%
              unname() %>%
              unlist()

yrs_imd <- str_extract(file_imd, ".{4}(?=.tif)")

## Retrieve IMD per year and fylke
daemons(5)

lau_et01_imd <- map(seq_along(file_imd), function(yr){
  
  # Year
  year <- yrs_imd[yr]
  
  message("Processing: ", year)
  
    # Parallelise
  imd_yr <- map(seq_along(fylke_nb), in_parallel(function(fyk){
    
    # Libraries needed
    library(duckspatial)
    library(exactextractr)
    library(dplyr)
    library(terra)
    library(duckdb)
    library(DBI)

    # et01 for filke fyk
    lau_et01 <- filter(lau_et01_all,
                       Fylke_nummer == fylke_nb[fyk])
    
    # IMD input
    imd <- rast(file_imd[yr])
    
    # Run exact_extract
    exact_extract(imd, lau_et01, fun = c("mean"), append_cols = TRUE) %>%
      mutate(`Accounting period` = year)
    
  },
  file_imd = file_imd,
  year = year,
  fylke_nb = fylke_nb,
  lau_et01_all = lau_et01_all,
  yr = yr))
  
})

daemons(0)


# Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "imd_crs",
                    "lau_et01_all",
                    "lau_et01_imd")))

```

Now, we compile the four remaining URAQ indicators: URGR_001 to URGR_004. During the extraction of the raster cell values, there will be the creation of "NA" values. This means that some polygon will have a imperviousness share of "NA". This is due to some municipal polygons, partially overlap with NA raster cells.We remove these NAs when performing the mean.

```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Compile URGR_001"

## URGR national level - area-weighted mean
urgr_001_nat <- map(seq_along(lau_et01_imd), function(x){
                      lau_et01_imd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 1) %>%
                        mutate(non_imp = 100 - mean) %>%
                        summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

urgr_001_nat <- urgr_001_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_001",
                         `Accounting area` = "Norway",
                         .before = `Share of urban green/blue spaces (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## URGR municipal level
urgr_001_mun <- map(seq_along(lau_et01_imd), function(x){
                        lau_et01_imd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 1) %>%
                          mutate(non_imp = 100 - mean) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                          relocate(LAU_NAME, .after = GISCO_ID) %>%
                          rename("Accounting area (municipality name)"= LAU_NAME,
                                 "Accounting area (municipality number)"= GISCO_ID) }) 
  
urgr_001_mun <- urgr_001_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_001",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "imd_crs",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun")))
```

```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Compile URGR_002"

## URGR national level - area-weighted mean
urgr_002_nat <- map(seq_along(lau_et01_imd), function(x){
                      lau_et01_imd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 2) %>%
                        mutate(non_imp = 100 - mean) %>%
                        summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                  `Accounting period` = unique(`Accounting period`))}) 

urgr_002_nat <- urgr_002_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_002",
                         `Accounting area` = "Norway",
                         .before = `Share of urban green/blue spaces (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## URGR municipal level
urgr_002_mun <- map(seq_along(lau_et01_imd), function(x){
                        lau_et01_imd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 2) %>%
                          mutate(non_imp = 100 - mean) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                          relocate(LAU_NAME, .after = GISCO_ID) %>%
                          rename("Accounting area (municipality name)"= LAU_NAME,
                                 "Accounting area (municipality number)"= GISCO_ID) }) 
  
urgr_002_mun <- urgr_002_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_002",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun")))
```

```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Compile URGR_003"

## URGR national level - area-weighted mean
urgr_003_nat <- map(seq_along(lau_et01_imd), function(x){
                      lau_et01_imd[[x]] %>%
                        bind_rows() %>%
                        filter(DEGURBA == 3) %>%
                        mutate(non_imp = 100 - mean) %>%
                        summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`))}) 

urgr_003_nat <- urgr_003_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_003",
                         `Accounting area` = "Norway",
                         .before = `Share of urban green/blue spaces (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
  
  

## URGR municipal level
urgr_003_mun <- map(seq_along(lau_et01_imd), function(x){
                        lau_et01_imd[[x]] %>%
                          bind_rows() %>%
                          filter(DEGURBA == 3) %>%
                          mutate(non_imp = 100 - mean) %>%
                          group_by(LAU_NAME, GISCO_ID) %>%
                          summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`)) %>%
                          relocate(LAU_NAME, .after = GISCO_ID) %>%
                          rename("Accounting area (municipality name)"= LAU_NAME,
                                 "Accounting area (municipality number)"= GISCO_ID) }) 
  
urgr_003_mun <- urgr_003_mun %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_003",
                         .before = `Accounting area (municipality number)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area (municipality name)`)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun")))
```

```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Compile URGR_004"

## URGR national level - area-weighted mean
urgr_004_nat <- map(seq_along(lau_et01_imd), function(x){
                      lau_et01_imd[[x]] %>%
                      bind_rows() %>%
                      mutate(non_imp = 100 - mean) %>%
                      summarise("Share of urban green/blue spaces (%)" = round2(stats:::weighted.mean(non_imp, area_m2, na.rm =TRUE), digits = 2),
                                    `Accounting period` = unique(`Accounting period`))}) 

urgr_004_nat <- urgr_004_nat %>%
                  bind_rows() %>%
                  mutate(`Indicator` = "Urban green areas",
                         ID = "URGR_004",
                         `Accounting area` = "Norway",
                         .before = `Share of urban green/blue spaces (%)`) %>%
                  relocate(`Accounting period`, .after = `Accounting area`)
       
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun",
                    "urgr_004_nat",
                    "urgr_004_mun")))
```

### 9.2 URGR_005
**Target area** = the LAU whose `DEGURBA` is `1`.
**Target feature** = the area of any ecosystem with green cover.
**Compilation process** = spatial intersection between **target area** and **target feature**. Compared with URGR_001-004, only one year is available as the Grunkart has only one iteration.

We start by identifying the city LAUs (`DEGURBA = 1`) and their adjacent towns and suburb (`DEGURBA = 2`), Then we intersect them with the "green/blue cover ecosystems" that are not in ET01. We will do a separate spatial intersection for "small-grained" ecosystem of ET01 that should be considered as "green/blue cover ecosystems" (*i.e.,* Cemeteries and Urban green spaces).

```{r}
#| eval: false
#| code-summary: "Identify cities and adjacent towns & suburbs"

## Select LAU with DEGURBA = 1 and the adjacent DEGURBA = 2

# LAU 1
lau_1 <- as_duckspatial_df("cln_lau", conn = conn_ug) %>%
          filter(DEGURBA == 1) %>%
          ddbs_collect()

# LAU 2
lau_2 <- as_duckspatial_df("cln_lau", conn = conn_ug) %>%
          filter(DEGURBA == 2) %>%
          ddbs_collect()

# LAU 1  and adjacent LAU 2
lau_2_adj_id <- st_touches(lau_1, lau_2) %>%
                unlist() %>%
                unique()

lau_2_adj <- lau_2[lau_2_adj_id, ]

lau_urgr5 <- rbind(lau_1, lau_2_adj) %>%
                ddbs_write_table("lau_urgr5", conn = conn_ug)

## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun",
                    "urgr_004_nat")))

```

```{r}
#| eval: false
#| code-summary: "Spatial intersection LAU and non-ET01 ecosystem types"

## Intersect ET01 with LAUs - block processing to avoid memory issues

# Block intersections
fylke_nb <- dbGetQuery(conn_ug, "SELECT DISTINCT Fylke_nummer FROM lau_urgr5") %>%
              unname() %>%
              unlist()

walk(seq_along(fylke_nb), function(i){

  # Input data  
  et <- as_duckspatial_df(paste0("cln_gk_not_et01_", fylke_nb[i]), conn = conn_ug) 
  
  lau <- as_duckspatial_df("lau_urgr5", conn = conn_ug) %>%
          filter(Fylke_nummer == !!fylke_nb[i])

  
  # Intersection
  ddbs_intersection(lau,
                    et,
                    conn = conn_ug,
                    conn_x = conn_ug,
                    conn_y = conn_ug) %>% 
    
    ddbs_make_valid() %>%
    
    # Need to remove linestrings and geometrycollection for extact_extract
    mutate(geometry = ST_CollectionExtract(geometry, 3L)) %>% # clean up geometrycollections
    filter(!ST_IsEmpty(geometry)) %>% # remove empty geometry resulting from the step above
    filter(ddbs_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>% # keep only polygons and multipolygons
    
    ddbs_area(new_column = "area_m2", name = paste0("lau_urgr5_not_et01_", fylke_nb[i]), conn = conn_ug, overwrite = TRUE)
  
})

# Verify table is in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

## Check valididty
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "lau_urgr5_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}


## Clean duckdb database
tables <- dbListTables(conn_ug)

tables <- keep(tables, startsWith(tables, "temp"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}


## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun",
                    "urgr_004_nat")))

```

```{r}
#| eval: false
#| code-summary: "Spatial intersection LAU and green/blue ET01 ecosystem types"

## Intersect ET01 with LAUs - block processing to avoid memory issues

# Block intersections
fylke_nb <- dbGetQuery(conn_ug, "SELECT DISTINCT Fylke_nummer FROM lau_urgr5") %>%
              unname() %>%
              unlist()

walk(seq_along(fylke_nb), function(i){

  # Input data  
  et <- as_duckspatial_df(paste0("cln_gk_urb_green_features_", fylke_nb[i]), conn = conn_ug) 
  
  lau <- as_duckspatial_df("lau_urgr5", conn = conn_ug) %>%
          filter(Fylke_nummer == !!fylke_nb[i])

  
  # Intersection
  ddbs_intersection(lau,
                    et,
                    conn = conn_ug,
                    conn_x = conn_ug,
                    conn_y = conn_ug) %>% 
    
    ddbs_make_valid() %>%
    
    # Need to remove linestrings and geometrycollection for extact_extract
    mutate(geometry = ST_CollectionExtract(geometry, 3L)) %>% # clean up geometrycollections
    filter(!ST_IsEmpty(geometry)) %>% # remove empty geometry resulting from the step above
    filter(ddbs_geometry_type(geometry) %in% c("POLYGON", "MULTIPOLYGON")) %>% # keep only polygons and multipolygons
    
    ddbs_area(new_column = "area_m2", name = paste0("lau_urgr5_et01_", fylke_nb[i]), conn = conn_ug, overwrite = TRUE)
  
})

# Verify table is in my duckdb conn_pm database
ddbs_list_tables(conn_ug)

## Check valididty
tables <- dbListTables(conn_ug) %>%
            keep(~ startsWith(.x, "lau_urgr5_et01_"))

for(tbl in tables){
  
  message("Processing: ", tbl)
  
  # Checks
  valid_check <- ddbs_is_valid(tbl, conn = conn_ug) %>%
                    filter(!is_valid) %>%
                    ddbs_collect() %>%
                    dim()

  empty_check <- ddbs_is_empty(tbl, conn = conn_ug) %>%
                    filter(is_empty == TRUE) %>%
                    ddbs_collect() %>%
                    dim()

  simple_check <- ddbs_is_simple(tbl, conn = conn_ug) %>%
                    filter(!is_simple) %>%
                    ddbs_collect() %>%
                    dim()
  
  # Print - should all have 0 rows
  print(valid_check)
  print(empty_check)
  print(simple_check)

  
}


## Clean duckdb database
tables <- dbListTables(conn_ug)

tables <- keep(tables, startsWith(tables, "temp"))

for (tbl in tables) {

  if(substr(tbl, 1, 9) == "temp_view"){

    dbExecute(conn_ug, sprintf("DROP VIEW IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }else{

    dbExecute(conn_ug, sprintf("DROP TABLE IF EXISTS %s", dbQuoteIdentifier(conn_ug, tbl)))

  }

}


## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "conn_ug",
                    "file_imd",
                    "lau_et01_all",
                    "lau_et01_imd",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun",
                    "urgr_004_nat")))

```

Now we can compile URGR_005.

```{r}
#| eval: true
#| message: false
#| results: false
#| code-summary: "Compile URGR_005"

## List of tables
tables <- dbListTables(conn_ug)
tables <- keep(tables, startsWith(tables, "lau_urgr5"))[-1]

## Pull data together
urgr_005_data <- map(tables,
                     function(x){
                       as_duckspatial_df(x, conn = conn_ug) %>%
                       ddbs_collect()}) %>%
                  st_drop_geometry() %>%
                  bind_rows()


## Total area LAUs
lau_area <- as_duckspatial_df("lau_urgr5", conn = conn_ug) %>%
              ddbs_area("area_m2") %>%
              ddbs_collect() %>%
              st_drop_geometry() %>%
              select(GISCO_ID, area_m2)


## URGR national level - area-weighted mean
urgr_005_nat <-  data.frame(`Indicator` = "Urban green areas",
                            `ID` = "URGR_005",
                            `Accounting area` = "Norway",
                            `Accounting period` = "2024", # We assume GK is from 2024
                            `Share of urban green/blue spaces (%)` = round2(sum(urgr_005_data$area_m2)/sum(lau_area$area_m2)*100, digits = 2),
                            check.names = FALSE)
       

## Close duckdb database
ddbs_stop_conn(conn_ug)
                  
## Clean the R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "urgr_001_nat",
                    "urgr_001_mun",
                    "urgr_002_nat",
                    "urgr_002_mun",
                    "urgr_003_nat",
                    "urgr_003_mun",
                    "urgr_004_nat",
                    "urgr_005_nat")))
```


## 10. Results
<!--# Repeat the final results here. Typically this is a map or table of indicator values.
This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either. -->

### National-level results


```{r tbl-national}
#| eval: true
#| warning: false
#| code-summary: "Code to create a comparative URGR table"
#| tbl-cap: 'Ecosystem condition variables NO_URGR from 1 to 5.'

# Merge all tables
res_national <- bind_rows(urgr_001_nat,
                          urgr_002_nat,
                          urgr_003_nat,
                          urgr_004_nat,
                          urgr_005_nat)


# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "urgr_001_mun",
                    "urgr_002_mun",
                    "urgr_003_mun",
                    "res_national")))

# Show table
datatable(res_national, filter = "top") %>%
  formatRound(columns = "Share of urban green/blue spaces (%)", digits = 2)

```

### Municipal results
URGR_004 was not presented at municipal level as it corresponds to either URGR_001-2-3 depending on the degree of urbanisation the municipalit has been classified with by Eurostat.

```{r tbl-regional}
#| eval: true
#| warning: false
#| code-summary: "Code to create a comparative URGR table"
#| tbl-cap: 'Ecosystem condition variables URGR from 1 to 3, divided into municipalities.'

# Merge all tables
res_municipal <- bind_rows(urgr_001_mun,
                           urgr_002_mun,
                           urgr_003_mun)


# Clean R environment
rm(list = setdiff(ls(), 
                  c("pull",
                    "res_municipal",
                    "res_national")))

# Show table
datatable(res_municipal, filter = "top") %>%
  formatRound(columns = "Share of urban green/blue spaces (%)", digits = 2)


```



## 11. Export file
Code to export the result tables above.

```{r export}
#| eval: false
#| code-summary: "Code to export the result tables above"

#CSV
write.table(res_national, "URGR_NO_national_2018-2024.csv")
write.table(res_municipal, "URGR_NO_regional_2018-2024.csv")

#Excel
write_xlsx(res_national, "URGR_NO_national_2018-2024.xlsx")
write_xlsx(res_municipal, "URGR_NO_regional_2018-2024.xlsx")

```

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
