---
title: "Connected natural forest habitat"
subtitle: |
  NO_FC_NAT (Forest)  
# delete the line with the second indicatorID (line 5) 
# if it's not relevant for you.
# In the parentheses you can add a word to distinguish multiple indicators 
# from each other, for example the ecosystem type that they cover
indicatorID: 
  - NO_FC_NAT
# If you are presenting multiple indicators, and have multiple 
# indicator IDs, then add them as a list here, and make sure you add
# the same number of list entries, 
# and in the same order that corresponds to the indicator IDs,  
# to the meta data fields where the indicators differ. 
# For example, if the indicators cover different ecosystem types, 
# add the ecosystems as a list under 'Ecosystems', but if they are in the
# same realm, same country etc., then you only need to write the realm and
# country once. 
indicatorName: 
  - Connected natural forest habitat # add more by adding another line starting '-'
verbatimName: 
  - Sammenhengende naturskoghabitat # the indicator name that us used in the native language
folderName: XXX_YYY_ZZZ # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
ECT: C1 - Landscape and seascape characteristics # e.g. A2 - Chemical State characteristics
Realm: # Please copy and paste Realms, Biomes and Ecostsem names from here: https://github.com/NINAnor/ecRxiv/wiki#ecosystem-typology 
  - Terrestrial (T) # e.g. Freshwater (F)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome # e.g. F1 Rivers and streams biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T2.2 Deciduous temperate forests # e.g. F1.3 Freeze-thaw streams
verbatimEcosystem: skog # add name of the ecosystem or naturetype according to any relevant local or national typology. E.g. fjell, skog, ... 
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First draft # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/XX_YYYY_ZZZ
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Aggregate - Transform - Scale # e.g. Scale - Transform - Aggregate (length-weighted arithmetic mean) - Truncate
Uncertainties: # What kind of quantitative unceratinties are reported alongside the point estimates. Can be either inferrential (i.e. bootstrapped confidence intervals or standard errors) or descriptive (variation, standard deviation or percentiles from the distribution of the raw indicator values).
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: false
author:
  - name: Bram Van Moorter              # Enter name
    email: bram.van.moorter@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: First Last              #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Van Moorter, B. and Last, F.  
date: June 1, 2026 # Enter date
data_availability: gold
code_reproducibility: bronze
open_science_badge: bronze
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)

library(terra)
library(ggplot2)
library(viridis)

```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 01 Jun. 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->
Connected natural forest habitat is an ecosystem condition indicator for forest. It represents the amount of natural-forest habitat that remains connected through the surrounding landscape, given the current mapped forest extent and forest-type composition. The ecological motivation is that many forest-associated species groups, including certain fungi, lichens, insects, bryophytes and other organisms linked to natural and less intensively managed forest, depend not only on local habitat quality but also on the spatial continuity of suitable forest habitat. The indicator is not a single-species dispersal model. It is a broad structural-functional indicator of potential ecological connectivity for communities associated with natural forest.

2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->
The indicator combines public spatial datasets describing forest extent, forest type and forest naturalness. Forest type is represented by three partly distinct networks: spruce, pine and broadleaf forest. For each network, habitat quality is calculated from forest-type contribution and naturalness, while permeability describes how easily organisms can move through both matching and non-matching forest. Connectivity is calculated using ConScape in Julia on an HPC cluster. The resulting connected-habitat rasters for spruce, pine and broadleaf forest are summed to produce a forest-level connected-habitat layer. Equivalent connected habitat is calculated by summing connected habitat within each reporting unit and taking the square root of the sum. Current equivalent connected habitat is divided by reference equivalent connected habitat to obtain the normalised indicator.

3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
An indicator value of 1 signifies that the currently mapped forest landscape has the same equivalent connected natural-forest habitat as the reference condition, where all currently existing forest is assigned maximum naturalness. Values decreasing toward zero reflect loss of connected natural-forest habitat caused by reduced naturalness and/or spatial isolation of suitable natural-forest habitat. The indicator does not measure loss of forest extent, because forest extent and forest-type composition are fixed in the reference condition.

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->
In the current national aggregation, the indicator value for connected natural forest habitat is 0.406. This means that current equivalent connected natural-forest habitat is approximately 41% of the modelled reference value under the current assumptions. The indicator is still incomplete, as uncertainty is not represented and the indicator has not yet been ecologically validated.


## 2. About the underlying data

<!--# Give a general introduction the datasets you have used, and provide more specific details in the sub-chapters.  -->

The indicator is based on publicly available spatial datasets describing forest extent, dominant tree species, and forest naturalness. These datasets are combined into harmonised raster inputs used to calculate habitat quality, permeability, and connected habitat. The full workflow from original vector and raster source data to harmonised analysis rasters is not rerun in this document. Some source datasets, such as forest-cover require computationally intensive rasterisation and alignment before they can be used in the connectivity model. This document therefore treats the harmonised rasters as input layers and documents how they are used in the indicator calculation.

#### Forest extent

The forest extent is derived from the National base-map for area accounting (i.e. [Nasjonalt grunnkart for arealanalyse](https://kartkatalog.geonorge.no/metadata/grunnkart-for-bruk-i-arealregnskap-testversjon-1/28c28e3a-d88f-4a34-8c60-5efe6d56a44d)). This base-map has been developed in collaboration between NIBIO, Statistics Norway, the Norwegian Mapping Authority and the Norwegian Environment Agency, and is intended as a common data basis for various types of area analyses. The base-map is based on existing data from the public map basis (DOK), which shows land resource and land use data, and does not involve new mapping. For forest extent, we selected the polygons from the base-map with `okosystemtype_1 == "Forest and woodlands"`. 

#### Dominant tree species

The dominant tree species is derived from the forest resources map (Skogressurskart $16 \times 16$ meter raster - [SR16R](https://kartkatalog.geonorge.no/metadata/sr16-skogressurskart-16x16-meter-raster/5de45872-f534-4e97-840e-3cfd8db04398)). SR16 is a comprehensive dataset that provides an overview of the distribution and characteristics of the Norway's forest resources. SR16R is a raster map. The dataset is produced through automatic processes as a combination of existing maps (AR5), terrain models, 3D remote sensing data (photogrammetry and laser) and national forest areas. The layer 'SRTRESLAG' from SR16R contains the dominant tree species, as: 1=spruce ('gran'), 2=pine ('furu'), and 3=broadleaf ('lauv').


#### Forest naturalness

Forest naturalness is derived from the 'Proximity of natural forest' ([Naturskog](https://kartkatalog.geonorge.no/metadata/naturskog-v1/a0062ac4-8ee0-408f-9373-4c8b8c3088d8)). The map 'Natural forest proximity' provides knowledge about possible occurrences of forest area with a lower or higher degree of natural forest proximity on a 1-7 scale. Natural forest proximity is expressed through a combination of the three NiN 3.0 variables: deadwood proportion (KM-DA), deadwood variation (KM-DV) and succession on wooded land (AD-ST). Natural forest proximity is divided into seven classes, where increasing values indicate increasing degrees of natural forest proximity.


### 2.1 Spatial and temporal resolution and extent

<!--# Describe the temporal and spatial resolution and extent of the data used -->

The analysis was conducted on a $50 \times 50$ meter grid aligned with Statistics Norway (Statistisk sentralbyra - SSB), all original data sources were aggregated to this resolution. We used the most recent data available in 2026.     

For **Forest extent**, the base-map is a vector map.

For **Dominant tree species**, the SR16R is a $16 \times 16$ meter raster map.

The **Forest naturalness** is a $16 \times 16$ meter raster map based on the SR16R. 


### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

For **Forest extent**, the vector base-map is rasterized as the proportion of a $50 \times 50$ meter covered by forest.

For **Dominant tree species**, the $16 \times 16$ meter raster map was resampled as the proportion of each $50 \times 50$ pixel covered by $16 \times 16$ pixels of each dominant tree species (i.e. spruce, pine, broadleaf).

For **Forest naturalness**, the $16 \times 16$ meter raster map was resampled as the average proximity to natural forest within each $50 \times 50$ pixel.


### 2.3 Original units

<!--# What are the original units for the most relevant  variables in the data-->

For **forest extent**, we selected the polygons from the base-map with `okosystemtype_1 == "Forest and woodlands"`. 

For **dominant tree species**, the layer 'SRTRESLAG' from SR16R contains the dominant tree species, as: 1=spruce ('gran'), 2=pine ('furu'), and 3=broadleaf ('lauv').

The **proximity to natural forest** is on a 1 to 7 scale, with 1 is least natural and 7 is most like natural forest. The categories 6 and 7 are extremely rare (< 0.1 % of the landscape), therefore we truncated the scale and merged categories 5,6, and 7. Thus, we truncated the scale to a 1-5 scale for forest naturalness. 


### 2.4 Instructions for citing, using and accessing data

<!--# Is the data openly available? If not, how can one access it? What are the key references to the datasets?   -->

These data are publicly accessible:

The base-map for **forest extent** is available here: https://kartkatalog.geonorge.no/metadata/grunnkart-for-bruk-i-arealregnskap-testversjon-1/28c28e3a-d88f-4a34-8c60-5efe6d56a44d

The SR16R for **dominant tree species** is available here: https://kartkatalog.geonorge.no/metadata/sr16-skogressurskart-16x16-meter-raster/5de45872-f534-4e97-840e-3cfd8db04398

And, the **forest naturalness** is available here: 
https://kartkatalog.geonorge.no/metadata/naturskog-v1/a0062ac4-8ee0-408f-9373-4c8b8c3088d8

### 2.5 Additional comments about the dataset

<!--# Optional. Text here -->

While the source data are publicly available (see above), we take the harmonised $50 times 50$ meter rasters aligned with the Statistics Norway grid as our input. While relatively trivial, some of these harmonization steps are fairly time consuming for the whole of Norway. 

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->

C1 - Landscape and seascape characteristics. This indicator fits here as it describes the spatial, structural, and compositional properties of an environment; it highlights how habitats are interconnected.

<!--#
### 3.1.1 Other condition typologies

 Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->

Intact natural forest

### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

Forest naturalness or components thereof.

### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->

The forest naturalness is the main variable in the indicator, which is affected/determined by forestry practices.

### 3.5 Spatial aggregation

<!-- Describe in word the same aggregation pathway presented in the top yaml (the metadata). If you have scaled against the reference value, it is especially important to know if you did any avergaing or summations etc. beforehand. An example text for a rather simple case could be: "The variable is scaled at the initial polygon level and the point estimate is a area-weighted arithmetic mean of all polygins. There is no transformation, and truncation is not needed as the indicator becomes naturally bound between 0 and 1".  -->

The 'connected habitat' raster is aggregated at the different administrative levels (e.g. municipal, county, etc.) -- **aggregation**. These aggregated values are square-root transformed to obtain the 'equivalent connected habitat' (ECH; Saura et al., Van Moorter et.) for the different administrative units -- **transformation**. The ECH is then normalized on a 0-1 scale using the ECH from the reference condition -- **scaling**.

## 4. Reference condition and levels

### 4.1 Reference condition

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

The reference condition is forest in its natural condition.

### 4.2 Reference levels

<!--# 

If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ refers to the reference level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.

 -->

We are working with a truncated naturalness scale 1-5 (from the original 1-7 scale), hence the reference level of a forest in its natural condition corresponds to a 'proximity to natural forest' value of 5.


#### 4.2.1 Spatial resolution and validity

<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->

Resolution is $16 \times 16$, as to original forest naturalness data described above. While high naturalness makes for a valid reference condition, maintaining constant forest species composition ignores the natural occurring forest succession. In coniferous boreal forests, the coniferous species are the apex species, while early colonizing species after natural disturbances are typically broadleaf trees. This shifting distribution of early-succession broadleaf forests is ignored for the current reference condition.  

## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

There are two main sources of uncertainty: first, uncertainty in the input data, and second, uncertainty stemming from modelling choices. The forest extent and tree species are mapped quite accurately -- the accuracy of the best model ranged between 74% for spruce and 87% for deciduous forest [@breidenbach2021national]. The proximity to natural forest had a considerably higher uncertainty with a 54% accuracy [@hauglin2025naturskogkart]. In addition to the uncertainties associated to the input data, modeling choices also affect the indicator. Modeling choices involve the choice of input variables to characterize habitat quality and permeability (incl. tree species specific networks), but also the way connectivity is characterized, for instance, using a combination of the number and cost of paths in the Randomized Shortest Paths framework, and the distance decay function [@vanmoorter2021defining]. 

### 5.2 Reported uncertainties
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->

The uncertainties from the input data and all modeling choices will affect the indicator; however, a sensitivity analysis to quantify and propagate these uncertainties has not been developed, yet. 

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

```{r}
#| echo: false
#| warning: false
#| message: false

library(here)
library(terra)
library(sf)

focal_municipality <- NA # set NA for whole of Norway
focal_municipality <- c("0301") # e.g. Oslo: "0301"

if (grepl("/data/", getwd())){
  pth <- "/data/R/GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/"
}else{
  pth <-  "R:/GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/"
}

if (is.na(focal_municipality[1])){
    msk <- st_read(paste0(pth, "Processed/Norge_land.shp"))
}else{
    msk <- st_read(paste0(pth, "/Converted/Norway_Municipalities/Norge_Kommuner_2023.shp"))
    msk <- msk[msk$kommunenum %in% focal_municipality,]
}

focus <- function(x){
  mask(crop(x, msk), msk)
}
```
The model is built from three main input layers: forest extent, dominant tree species layers, and forest naturalness. Together, these layers define where the model considers forest habitat to be present, what kind of forest it is, and how close that forest is to a natural condition.

The forest extent defines the spatial extent of the habitat. Habitat quality is only assigned within areas classified as forest or woodland in the base-map. Areas outside this forest mask are not treated as habitat, although they still receive a baseline matrix permeability in the movement model.

The dominant tree species layers are used to distinguish between broad forest-associated communities. We used the tree species layer (`treslag`) from SR16 raster (16x16 meter pixels) to classify forest into spruce, pine, and broadleaf components. These are not mutually exclusive biological categories in the model. Instead, each forest cell receives a weight for how strongly it contributes to each of the three dominant tree species networks.

The naturalness layer represents forest condition. We use the `naturskognaerhet` layer, with categories 5, 6, and 7 merged into a single highest naturalness category. The resulting 1-to-5 naturalness values are scaled from 0.2 to 1, where 0.2 represents the least natural forest category included in the model, and 1 represents the most natural forest condition.

These three layers are then combined to create habitat quality and habitat permeability for each forest type. Forest type determines which forest-associated community a cell is most relevant for, while naturalness modifies the strength of that contribution.


### 7.1 Forest extent

<!--# Import and prepare the first dataset for analysis. Change the header from Dataset A to the name of the actual dataset. -->

First, habitat quality is only defined within the main ecosystem forest from the base-map (`okosystemtype_1 == "Forest and woodlands"`). Note: rasterization to the Statistics Norway grid was done previously.


```{r}
#| echo: false
#| warning: false
#| message: false

forest_ext <- rast(here("data", "forest_pct_NO.tif"))
forest_ext <- focus(ifel(forest_ext>0, 1, 0))
plot(forest_ext, main="Forest extent")
```


### 7.2 Dominant tree species

<!--# Repeat the process for each dataset -->

Within the forest extent, we set the quality based on the dominant tree species (`treslag`) from the SR16 raster. 
The SR16 raster has 16x16 meter pixels with the dominant tree species (i.e. spruce, pine, and broadleaf), we previously computed the proportion of each 50x50 meter pixel covered by each tree species:  

```{r}
#| echo: false
#| warning: false
#| message: false

tree_species <- c("spruce", "pine", "broadleaf")
DTS <- focus(rast(here("data", paste0(tree_species,"_pct_NO.tif"))))/100
names(DTS) <- tree_species
plot(DTS)
```

### 7.3 Forest naturalness

<!--# Repeat the process for each dataset -->

Finally, the value of the forest to support species was represented through its naturalness, or proximity to natural forest (i.e. naturskognaerhet; see: https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/arealplanlegging/kart-over-naturskog/). The original naturalness scale ranges from 1 to 7. We merged categories 5, 6, and 7 into a single highest category, because categories 6 and 7 are exceedingly rare. The resulting naturalness layer therefore ranges from 1 to 5. Dividing by 5 gives the model-scale naturalness variable, N, ranging from 0.2 to 1.

```{r}
#| echo: false
#| warning: false
#| message: false

nat <- focus(rast(here("data", "naturalness_pct_NO.tif")))/100
plot(nat, main="Naturalness (naturskogsnaerhet)", legend=T)
```

## 8. Spatial units

<!--# Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpreted at. Potential spatial delineation data (maksing layers etc.) should be introduced under chapert 7. We recommend using the SEEA EA terminology of Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). -->

We used a spatial resolution of $50 \times 50$ meter aligned with the grid from Statistics Norway for our analysis.

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

The analysis consists of six components: (1) computation of the habitat quality, (2) computation of the habitat permeability, (3) computation of the reference conditions, (4) network computation on high-performance cluster, (5) integration across dominant tree species, and (6) computation of indicators for different administrative units. We describe these individual steps in detail below.

### 9.1 Habitat quality

Habitat quality describes how suitable each cell is as habitat for each dominant tree species network. In this model, habitat quality depends on two things: forest type and naturalness.

Forest type determines which ecological network the cell mainly contributes to. For example, spruce-dominated forest contributes strongly to the spruce-forest network, while pine-dominated forest contributes strongly to the pine-forest network. Mixed forest contributes partly to several networks. The dominant tree species proportions used for the three dominant tree species networks were shown before. These values translate the SR16 tree-type information into estimated proportions of spruce, pine, and broadleaf forest within each 50-by-50 m pixel. Thus, a pixel covered completely by spruce gets a value of 1 for spruce, similarly for the other forest types. Hence, summed over the forest types, we get the proportion of a forest pixel that is covered by 16-by-16 m dominant tree species pixels, which is maximal 1 and can be smaller than 1. 

Naturalness then modifies this dominant tree species contribution. A forest cell with high naturalness receives a higher quality value than a similar forest cell with low naturalness. 

Thus, within the forest mask, for each forest type, the final habitat quality is calculated as:

$$Q_k = T_k * N$$

where $Q_k$ is the habitat quality for dominant tree species $k$, $T_k$ is the proportion of the pixel covered by this dominant tree species (DTS), and $N$ is the naturalness value. This means that a cell must have both a relevant forest type and sufficient naturalness to receive high habitat quality.

```{r}
#| echo: false
#| warning: false
#| message: false

quality <- forest_ext * DTS * nat
names(quality) <- tree_species
quality <- focus(ifel(is.na(quality), 0, quality))

for (i in c(1:3)){
  writeRaster(quality[[i]], here("data", paste0(tree_species[i], "_forest_quality.tif")), overwrite=T)
}

plot(quality)
```
The resulting quality maps should be interpreted as type-specific habitat layers. They indicate where the landscape contains forest that is both compositionally relevant and sufficiently natural for each forest-associated community. High-quality cells are expected to function more strongly as habitat nodes in the connectivity model. Low-quality cells may still contribute to movement, but they contribute less as habitat.

### 9.2 Habitat permeability

Habitat permeability describes how easily organisms can move through each cell. It is related to habitat quality, but it represents a different ecological role. Habitat quality describes how suitable a cell is as habitat, while permeability describes how much the cell facilitates movement through the landscape.

In the model, permeability is less strongly tied to exact forest type than habitat quality. The reasoning is that a forest cell that is not the preferred habitat type may still be easier to move through than open or non-forest matrix. For example, pine forest may not be high-quality habitat for spruce-forest-associated communities, but it may still provide more suitable movement conditions than non-forest land.

We therefore start with a matrix_perm permeability for the non-forest matrix: $B = 0.25$.

This value means that areas outside the forest habitat extent are not treated as completely impermeable. They still allow some movement, but less than forested areas:

```{r}
#| echo: false
#| warning: false
#| message: false

matrix_perm = 0.25
```
Within the forest extent, permeability is based on the same broad dominant tree species information used for habitat quality, but the dominant tree species proportions are compressed using a cube-root transformation. Thus, just as for habitat quality, as more of the matching forest type is available the permeability increases. However, the compression implies that even a little of the matching forest type contributes disproportionally to permeability. 

This reflects the idea that the amount of matching forest types matters less for movement than for habitat use. A cell with little of the 'right' forest type may be poor habitat, but it can still function as part of the movement landscape. Furthermore, even in the absence of the matching forest type, we assign a minimum value of 0.01 to all pixels with forest.

We compressed the dominant tree species contribution to permeability using a cube-root transformation (given a minimum value of a proportion of 0.01 for the focal type within forests, even when the focal forest type was not recorded):
```{r}
#| echo: false
#| warning: false
#| message: false

x <- c(0.01, seq(0.05,1,by=0.05))
plot(x, x^(1/3), xlab="prop. focal species", ylab="focal permeability", ylim=c(0,1)); abline(0,1)
```
Thus, the compressed permeability value $C_k$ is:
$$C_k = max(0.01, T_k)^{1/3}$$
where $T_k$ is the dominant tree species proportion and $C_k$ is the compressed permeability value, with 0.01 being the minimum value within the forest (even in the absence of the focal type: $T_k=0$).

```{r}
#| echo: false
#| warning: false
#| message: false

# compute the amount of forest across types:
forest <- app(DTS, fun = sum, na.rm = TRUE)

# set minimum value (1 percent):
DTS_perm <- ifel(forest>0 & DTS==0, 1/100, DTS)

# cube-root tranform
DTS_perm <- DTS_perm^(1/3)

plot(DTS_perm)
```

Within the forest, each forest type receives at least a small permeability contribution for all forest-associated networks. This means that even if a forest cell is not classified as containing forest type $k$, it is still treated as somewhat more permeable than non-forest matrix simply because it is forest. The cube-root transformation then compresses the differences among the amount of matching forest type, so that low availability of matching forest type can still contribute to movement, while high-availability of the matching forest type remain most permeable.

Final permeability is calculated by adding a forest-related permeability contribution on top of the baseline matrix permeability:

$$P_k = B + 0.75 * C_k * N$$

where $P_k$ is the final permeability for dominant tree species network $k$, $B$ is the baseline matrix permeability, $C_k$ is the compressed dominant tree species permeability value, and $N$ is naturalness scaled from 0.2 to 1.

The matrix_perm permeability is set to 0.25. This means that the non-forest matrix is not treated as completely impermeable, but as having a low background permeability.

The forest contribution is multiplied by naturalness, so more natural forest receives higher permeability than less natural forest. However, naturalness ranges from 0.2 to 1, so even forest with low naturalness remains more permeable than the non-forest matrix. If naturalness were allowed to approach 0, low-naturalness forest would effectively fall back to the baseline matrix permeability, which would remove the distinction between degraded forest and non-forest. 

The factor 0.75 is used because the baseline matrix permeability is 0.25. Together, these keep permeability less than one (permeability is capped at 0.99 for numerical reasons), when $C_k$ and $N$ are at their maximum values.


```{r}
#| echo: true

perm <- ((DTS_perm * nat) * 0.75) + matrix_perm
perm <- ifel(perm>0.99, 0.99, perm)
perm <- focus(ifel(is.na(perm), matrix_perm, perm))

for (i in c(1:3)){
  writeRaster(perm[[i]], here("data", paste0(tree_species[i], "_forest_permeability.tif")), overwrite=T)
}

plot(perm)
```

#### Recap of the permeability calculation

The permeability calculation has four steps.

First, each forest cell receives an estimated dominant tree species proportion, $T_k$, for each dominant tree species network $k$. This is the same basic dominant tree species information used in the habitat quality calculation.

Second, this dominant tree species proportion is transformed into a compressed permeability value:

$$C_k = max(0.01, T_k)^{1/3}$$

The minimum value of 0.01 means that all forest cells receive at least a small forest-related permeability contribution for each dominant tree species network, even when that forest type is not present in the cell. The cube-root transformation compresses differences among the proportion of a forest type, so pixels with relatively little of the matching forest type are still treated as relatively permeable, while a large amount of the matching forest type remain most permeable. Note that the minimum value for forest is 0.01, thus the permeability of a forest pixels is $0.75 * 0.01^{1/3} * N \leq 0.162$ higher than the baseline ($0.25$). In other words, the contribution from a non-matching forest pixel is about 65% above the matrix (0.412 vs. 0.25 in the matrix), if it is highly natural.

Third, the compressed dominant tree species permeability is multiplied by naturalness, N. The original naturalness layer is reduced to five ordered categories and rescaled by dividing by 5, giving values from 0.2 to 1. 

Fourth, the forest-related permeability contribution is added on top of the baseline matrix permeability:

$$P_k = B + 0.75 * C_k * N$$

where B = 0.25. The factor 0.75 ensures that when $C_k$ = 1 and N = 1, the final permeability reaches 1 (in practice capped to 0.99 for numerical reasons). Non-forest matrix has permeability 0.25, while forest cells receive additional permeability depending on forest type and naturalness.

In short, the permeability layer still gives the highest permeability to forest that both matches the focal dominant tree species network and has high naturalness, but is less sensitive to the amount of the matching dominant tree species.

### 9.3 Reference condition

For ecosystem accounting, the current condition of an ecosystem is usually interpreted relative to a reference condition. The reference condition represents the state against which current ecosystem condition can be compared. In this analysis, we define the reference condition as a landscape where all existing forest has maximum naturalness.

This does not mean that the whole landscape is assumed to become forest. The forest extent and dominant tree species composition are kept the same as in the current-condition model. What changes is the naturalness value. In the reference condition, naturalness is set to 1 for all forest cells. This does not imply that all forest is old forest, also a young forest can be highly natural.

The reference condition therefore answers a specific question: what would the connectivity model return if the same forest landscape had natural forest condition everywhere within the forest mask?

This reference is useful because it separates two things that are otherwise combined in the current-condition outputs. The first is the spatial structure and composition of the forest landscape: where forest occurs, and whether it is spruce, pine, or broadleaf forest. The second is the condition of that forest, represented by naturalness. By setting naturalness to 1, the reference condition keeps the forest landscape fixed but removes degradation in forest condition.

To compute the reference condition, we recreate the habitat quality and permeability layers without reducing them by current naturalness. For habitat quality, this means:

$$Q_{k,ref} = T_k$$
because N = 1 in the reference condition.

```{r}
#| echo: true

quality_ref <- forest_ext * DTS
names(quality_ref) <- tree_species
quality_ref <- focus(ifel(is.na(quality_ref), 0, quality_ref))

for (i in c(1:3)){
  writeRaster(quality_ref[[i]], here("data", paste0(tree_species[i], "_forest_ref_quality.tif")), overwrite=T)
}

plot(quality_ref)
```
For permeability, the reference condition is:
$$P_{k,ref} = B + 0.75 * C_k$$
again because N = 1.

```{r}
#| echo: true

perm_ref <- DTS_perm * 0.75 + matrix_perm
perm_ref <- ifel(perm_ref>0.99, 0.99, perm_ref)
perm_ref <- focus(ifel(is.na(perm_ref), matrix_perm, perm_ref))

for (i in c(1:3)){
  writeRaster(perm_ref[[i]], here("data", paste0(tree_species[i], "_forest_ref_permeability.tif")), overwrite=T)
}

plot(perm_ref)
```
The reference-condition connectivity output is not intended as a prediction of a realistic future landscape. It is a model-based benchmark. It represents the connectivity that would be expected if the existing forest landscape were in the best naturalness condition represented by the input data.


### 9.4 Network computation

We used ConScape to compute the connectivity outputs from the habitat quality and permeability layers. ConScape is a Julia library for landscape connectivity modelling that can compute connected-habitat and movement-flow-related metrics for large raster landscapes.

The model was run separately for each dominant tree species network: spruce, pine, and broadleaf forest. For each network, the corresponding habitat quality layer describes where suitable habitat is located, while the permeability layer describes how easily movement can occur through the landscape.

#### Distance calibration


#### Distance decay

ConScape uses ecological distance to describe how strongly cells are connected through the permeability surface. For the connected-habitat computation, this distance is transformed into a proximity value between 0 and 1. In this analysis we used a simple exponential distance-decay function:

$$k(d) = exp(-\alpha \times d)$$

where $k(d)$ is the proximity, or remaining connection strength, at ecological distance $d$, and $\alpha$ controls how quickly connection strength declines with distance.

A smaller $\alpha$ means that habitat influence extends farther through the landscape. A larger $\alpha$ means that connectivity is more local. In this document, the distance-decay function should be interpreted as a generic connectivity assumption for forest-associated communities, not as a dispersal kernel for a specific species.

```{r}
#| echo: false

alpha = 1 / 1000
df <- data.frame(d=c(0:5000))
df$k <- exp(-alpha*df$d)

ggplot(data=df, aes(x=d, y=k)) +
  geom_line()+theme_classic()
```
The decay is applied to ecological distance rather than simple Euclidean distance. This means that the effective connection between two cells depends both on how far apart they are and on the permeability of the landscape between them. High-permeability areas maintain stronger connections over distance, while low-permeability areas reduce connection strength more quickly.

#### Run on high-performance cluster


### 9.5 Integration across dominant tree species

ConScape produces several connectivity outputs. In this document we focus on three types of output: connected habitat, corridors, and reference state for ecosystem accounting.

Connected habitat describes where habitat contributes to the broader connected network. A cell with high connected habitat is not only suitable habitat locally, but is also well connected to other suitable habitat through the surrounding landscape.

Corridors describe areas that are important for movement between habitat areas. These areas may or may not be high-quality habitat themselves. Their importance comes from their role in linking other parts of the network.

We first load the connected-habitat maps for the three dominant tree species networks:

```{r}
#| echo: false

func <- paste0(tree_species, "_functionality.tif")
func <- focus(c(rast(here("data/outputs", func[1])),rast(here("data/outputs", func[2])),rast(here("data/outputs", func[3]))))
plot(sqrt(func), col = viridis(100, option = "viridis"))
```

These maps show connected habitat separately for spruce, pine, and broadleaf forest. Differences among the maps reflect the distribution of each forest type, the naturalness of those forests, and the permeability of the surrounding landscape.

To obtain a single forest-level connected-habitat layer, we sum the outputs across the three dominant tree species networks:

```{r}
#| echo: true

func <- app(focus(func), sum)
plot(sqrt(func), col = viridis(100, option = "viridis"), main="Connected habitat")
```

This combined layer should be interpreted as an aggregate indicator of connected natural forest habitat. It is not a separate model run for a generic forest type. Instead, it summarizes the type-specific networks into one forest-level map.

We use the same approach for corridors. First, we load the corridor outputs for the three dominant tree species networks:

```{r}
#| echo: false

corr <- paste0(tree_species, "_corridors.tif")
corr <- focus(c(rast(here("data/outputs", corr[1])),rast(here("data/outputs", corr[2])),rast(here("data/outputs", corr[3]))))
plot(corr, col = viridis(100, option = "magma"))
```

These maps show where each dominant tree species network depends on movement through the landscape. A high corridor value does not necessarily mean that the cell is high-quality habitat. It means that the cell contributes to connecting habitat elsewhere.

We then sum the corridor outputs across forest types to obtain a forest-level corridor layer:
```{r}
#| echo: true

corr <- app(corr, sum)
plot(corr, col = viridis(100, option = "magma"), main="Corridors")
```

The forest-level corridor map highlights areas that contribute to movement across one or more dominant tree species networks. These areas are useful to distinguish from connected habitat, because they may represent connective landscape rather than habitat nodes.


### 9.6 Connectivity indicators

The reference condition allows the current connectivity output to be expressed as a condition indicator. We calculate this by comparing connected habitat under the current condition with connected habitat under the reference condition.

The connected habitat under the reference condition aggregated across forest types is:
```{r}
#| echo: true

ref <- paste0(tree_species, "_ref_functionality.tif")
ref <- focus(c(rast(here("data/outputs", ref[1])),rast(here("data/outputs", ref[2])),rast(here("data/outputs", ref[3]))))
ref <- app(ref, sum)
plot(sqrt(ref), col = viridis(100, option = "viridis"), main="Reference connected habitat")
```
This map shows the square root of the cell-level connected-habitat raster. It is a diagnostic visualization, not a map of ECH. ECH is calculated only after summing connected habitat across the accounting area.

For the summary indicator, we use equivalent connected habitat (ECH) rather than raw connected habitat. ECH is a summary statistic for an area, not a cell-level raster quantity. It is calculated from the total connected habitat within the area:

$$ECH = \sqrt{\sum H}$$

where H is the connected habitat metric, also known as habitat functionality.

Because ECH is calculated after aggregation, it cannot be mapped directly at the pixel level. However, we often plot the square root of the connected-habitat raster as a visual transformation. This root-transformed map is not ECH, because $\sqrt{\sum H}$ is not the same as $\sum \sqrt{H}$. It is only a diagnostic visualization that puts the cell-level connected-habitat values on a scale that is closer to the area units of ECH and is easier to visually interpret.

The connectivity condition indicator is calculated as the ratio between current and reference ECH:

$$I_{ECH} = \frac{ECH_{current}}{ECH_{reference}}$$

or equivalently:

$$I_{ECH} = \frac{\sqrt{\sum H_{current}}}{\sqrt{\sum H_{reference}}}$$

Values close to 1 indicate that current equivalent connected habitat is close to the reference condition. Lower values indicate that current forest condition reduces connected habitat relative to the reference.

```{r}
#| echo: true

EAI <- sqrt(global(func, "sum", na.rm=T)[1,1])/sqrt(global(ref, "sum", na.rm=T)[1,1])
(EAI <- round(EAI, 2))
```

The equivalent connected habitat transformation is mainly used for the summary indicator, because it gives a value in area units. 

For spatial diagnosis, we map the cell-wise ratio between root-transformed current and reference connected habitat:

$$I_{cell} = \frac{\sqrt{H_{current}}}{\sqrt{H_{reference}}}$$
This map is not the ecosystem-accounting indicator itself. The accounting indicator is calculated after aggregating connected habitat across the accounting area. However, the root-transformed ratio is visually closer to the ECH logic than the raw connected-habitat ratio, because ECH is also based on a square-root transformation. The map should therefore be interpreted as a diagnostic layer showing where current connected habitat is low relative to the reference condition, not as a pixel-level ECH indicator.

```{r}
#| echo: true

rindicator <- (sqrt(func) + 0.001) / (sqrt(ref) + 0.001)

plot(ifel(rindicator>1, 1, rindicator),
  col = rev(viridis(100, option = "turbo")),
  main = "Cell-wise root-transformed current/reference ratio")
```

A small constant is added in the code to avoid undefined values where the reference layer is zero or very close to zero.

#### Aggregation across administrative units

```{r}
#| echo: true

if (grepl("/data/", getwd())){
  data_folder <- "/data/R/GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/Converted/"
}else{
  data_folder <- "R:/GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/Converted/"
}

admin_levels <- paste0("Norway_", c("Municipalities", "County", "Country"))
```

```{r}
#| echo: true

municipality_shp <- st_read(paste0(data_folder, admin_levels[1],"/Norge_Kommuner_2023.shp"))

tmp <- terra::extract(func, municipality_shp, fun="sum", na.rm=T)
municipality_shp$naturalness_variable_km2 <- sqrt(tmp[,2])
#plot(municipality_shp["naturalness_variable_km2"])

tmp <- terra::extract(ref, municipality_shp, fun="sum", na.rm=T)
municipality_shp$naturalness_ref_km2 <- sqrt(tmp[,2])
#plot(municipality_shp["naturalness_ref_km2"])

municipality_shp$naturalness_indicator <- municipality_shp$naturalness_variable_km2/municipality_shp$naturalness_ref_km2
plot(municipality_shp["naturalness_indicator"])
#st_write(municipality_shp, "outputs/municipal_account.shp", append=F)
```

```{r}
#| echo: true

county_shp <- st_read(paste0(data_folder, admin_levels[2],"/Norge_Fylker_2023.shp"))

tmp <- terra::extract(func, county_shp, fun="sum", na.rm=T)
county_shp$naturalness_variable_km2 <- sqrt(tmp[,2])
#plot(county_shp["naturalness_variable_km2"])

tmp <- terra::extract(ref, county_shp, fun="sum", na.rm=T)
county_shp$naturalness_ref_km2 <- sqrt(tmp[,2])
#plot(county_shp["naturalness_ref_km2"])

county_shp$naturalness_indicator <- county_shp$naturalness_variable_km2/county_shp$naturalness_ref_km2
plot(county_shp["naturalness_indicator"])
#st_write(county_shp, "outputs/county_account.shp", append=F)
```

```{r}
#| echo: true

naturalness_variable_km2 <- sqrt(global(func, "sum", na.rm=T))[1,1]
naturalness_variable_km2 # 246955.5

naturalness_ref_km2 <- sqrt(global(ref, "sum", na.rm=T))[1,1]
naturalness_ref_km2 # 608554.6

naturalness_indicator <- naturalness_variable_km2/naturalness_ref_km2
naturalness_indicator # 0.4058066
```


## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->





## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [Initials]
* Methodology: [Initials]
* Data curation: [Initials]
* Formal analysis: [Initials]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [Initials]
* Visualization: [Initials]
* Resources: [Initials]
* Writing original draft: [Initials]
* Writing review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
