---
title: "Alien coniferous tree species"
subtitle: |
  NO_AATS_002
indicatorID: 
  - NO_AATS_002
indicatorName: 
  - Alien coniferous tree species # add more by adding another line starting '-'
folderName: NO_AATS_002
country: Norway
continent: Europe
ECT: B2 Structural state characteristics
Realm: 
  - Terrestrial (T) # e.g. Freshwater (F)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome # e.g. F1 Rivers and streams biome
Ecosystem: 
  - T2.1 Boreal/montane forests # e.g. F1.3 Freeze-thaw streams
yearAdded: 2025 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.002" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Updated version to new template # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_AATS_002
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Aggregate (NFI plots to county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
  docx:
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad              # Enter name
    email: erik.framstad@nina.no  # Enter email
    affiliations:
        name: Norwegian Institute for Nature Research # Enter affiliations
  - name: Ken Olav Storaunet       
    affiliations:             
        name: Norwegian Institute of Bioeconomy Research # Enter affiliations
  - name: Oliver Snoksrud      
    affiliations:             
        name: Norwegian Institute of Bioeconomy Research # Enter affiliations
  - name: Aud Helen Halbritter             
    affiliations:             
        name: University of Bergen # Enter affiliations
AuthorList: Framstad, E., Storaunet, K.O., Snoksrud, O. and Halbritter, A. H.
date: May 11, 2026 # Enter date
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
bibliography: references.bib
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
library(ggplot2)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```

<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 26 Oct. 2025 - Original PR
- 11 May. 2026 - Updated version to new template
- 8 Jun. 2026 - Change indicator from absence to share of alien coniferous tree species; national refs X0 = 50%, X100 = 0%; ecTools workflow
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:

Many non-native or alien tree species are considered to have a high or very high risk for native forest ecosystems.
Intact ecosystems should not contain alien coniferous tree species of high ecological risk.
This indicator is specified as the percentage share of alien coniferous tree species in the forest stand.
Alien coniferous species represent the most abundant alien tree species considered to have a high ecological risk by @artsdatabanken2023-fremmedartslista.
Data for this indicator come from the National Forest Inventory (NFI), where all trees larger than 5 cm at breast height (dbh) have been recorded within the 250 m² NFI plots since the mid-1980s (@viken2021-feltinstruks).
The planting of alien tree species in forestry is the most important anthropogenic driver for this indicator.

2. **Methods**:

All trees larger than 5 cm at breast height (dbh) are recorded and identified to species for each 250 m² NFI inventory plot.
There are about 12,000 NFI plots in forest ecosystems, distributed on a regular grid, sampled over a 5-year period with a representative sampling scheme every year.
For each reporting period, the indicator is the area-weighted mean percentage of alien coniferous species across NFI plots in the selected geographic unit.
The share is derived from NFI variables for introduced Norway spruce and Scots pine.
Results are calculated for rolling 5-year periods (2010–2014, 2015–2019, 2020–2024) and scaled to 0–1 against national reference levels.

3. **Interpretation**:

An indicator value of 1 corresponds to 0% alien coniferous species (best condition, X100).
An indicator value of 0 corresponds to 50% or more alien coniferous species (worst condition, X0).
Values between 0 and 1 reflect intermediate shares on a linear scale between the national reference levels.
The relationship between indicator values and ecosystem condition is assumed to be linear at this stage.

4. **Results and development needs**:

At national level, scaled indicator values are high and close to the upper reference in all three reporting periods (@sec-results, @tbl-aats-national-results).
The national median is 0.97 in all three periods (approximate 95% interval 0.97–0.97).
These values correspond to a low area-weighted mean share of alien coniferous species nationally (about 1.5–1.6%).
The indicator is ready for use but should not be used at finer spatial resolution than counties.
Development needs include refinement of the relationship between indicator values and ecosystem condition (linear vs. non-linear).

### 1.2 Native language summary - Norwegian

**Bakgrunn og begrunnelse**:
Fremmede treslag kan utgjøre en høy eller svært høy risiko for naturlige skogøkosystemer.
Intakte økosystemer bør ikke inneholde fremmede bartrær med høy økologisk risiko.
Indikatoren er angitt som andel (%) fremmede bartrær i bestanden, basert på innført gran og furu i NFI.

**Metoder**:
For hver rapporteringsperiode beregnes arealvektede gjennomsnitt for fylker, regioner og nasjonalt nivå med `au_areal`.
Verdiene skaleres lineært mot nasjonale referansenivå.

**Tolkning**:
Indikatorverdi 1 tilsvarer 0 % fremmede bartrær (X100).
Indikatorverdi 0 tilsvarer 50 % eller mer fremmede bartrær (X0).

**Resultater og utviklingsbehov**:
På nasjonalt nivå er de skalerte indikatorverdiene høye og nær øvre referansenivå i alle tre rapporteringsperioder.
Medianverdien er 0,97 i alle tre perioder (omtrentlig 95 %-intervall ca. 0,97–0,97).
Dette tilsvarer en lav arealvektet andel fremmede bartrær nasjonalt (omtrent 1,5–1,6 %).
Over tid er den nasjonale utviklingen i praksis stabil.
Indikatoren er klar til bruk, men bør ikke brukes på finere oppløsning enn fylker.

## 2. About the underlying data

### 2.1 Spatial and temporal resolution and extent

Data are available from 2000, when the underlying variables (`gran_pr`, `intro_gran_pr`, `intro_furu_pr`, `komm_m_nat_gran`) were first recorded.
Annual estimates are possible due to representative sampling every year.
The analyses use rolling 5-year reporting periods from 2010 onward (2010-2014, 2015-2019, and 2020-2024).
The data are representative for all forest land within the selected region.
Spatial resolution should be limited to counties.

### 2.2 Sampling design

All trees larger than 5 cm at breast height (dbh) are recorded and identified to species for each 250 m² NFI inventory plot (@viken2021-feltinstruks).
There are about 12,000 NFI plots in forest ecosystems, distributed on a regular grid, with varying coverage for forest below/above the coniferous forest line and in Finnmark county.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.

### 2.3 Original units

The indicator is specified as the percentage share of alien coniferous species, derived from the percentage volume (cutting class 3–5) or crown cover (cutting class 2) of introduced Norway spruce and Scots pine (NFI variables `intro_gran_pr`, `intro_furu_pr`).
At plot level this is a percentage (0–100); aggregated values are area-weighted means of that percentage.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The share of alien coniferous tree species is considered ...
Hence, this indicator represents a B2 Structural state characteristics.

### 3.2 Ecosystem condition characteristic

The share of alien coniferous tree species is considered an aspect of the composition of biodiversity.
The indicator represents a characteristic of Biodiversity.

### 3.3 Collinearities with other indicators

There may be some positive correlation with indicators for alien plant species in the field layer, although those indicators use different data and spatial support.

### 3.4 Impact factors

Natural disturbances and long-term shifts in climate conditions may increase the opportunity for alien species to spread and get established in otherwise intact forest ecosystems.
The planting of alien tree species in forestry is the most important anthropogenic driver for this indicator, although such planting has been reduced lately.
Harvested alien tree species may to some extent be replaced by native species.
Other planting of exotic trees in parks, gardens, as windbreaks etc will establish a source of such species that may spread into forest ecosystems.
Disturbances associated with forestry and other land use will also facilitate the natural establishment of alien species.
Human-induced climate change will also facilitate the spread and establishment of alien species.

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each reporting period, non-scaled values for county, region, and Norway are obtained by pooling all included forest plots in the unit: the area-weighted mean of the plot-level alien coniferous share (%).
Where NFI county labels are split into parts, those labels are merged to a single county name before aggregation.
Scaling and truncation are applied after non-scaled county aggregates have been calculated: each county is scaled against the national lower reference level X0 (50% alien conifers) and upper reference level X100 (0% alien conifers) using a linear scaling function, and values outside [0, 1] are truncated at that step.
Regional and national scaled point estimates are area-weighted means of the already scaled county values, using each county's represented area as weight.
The indicator is likely to have a high frequency of NFI plots without any recorded alien coniferous species, so results should not be interpreted at finer resolution than counties.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems, where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances.

### 4.2 Reference levels

Intact forest ecosystems should not have alien coniferous tree species.
The upper reference value (X~100~) is therefore 0% alien coniferous species (best condition).
The lower reference value (X~0~) is set to 50% alien coniferous species (worst condition).
The same X0 and X100 are applied when scaling each county before regional and national aggregation.
For lack of better knowledge, a linear relationship is assumed between X0 and X100.
Hence, indicator values are scaled between 0 and 1 based on a linear scaling function; scaled indicator values outside [0, 1] are truncated at scaling.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined as one national value for alien coniferous share (%).
Although indicator values from NFI plots may technically be aggregated at finer spatial units than counties, scaling at such finer levels may result in biases.


## 5. Uncertainties
### 5.1 General uncertainties

Uncertainties relate to the variability in the frequency of alien species among NFI plots and recording errors for tree species within NFI plots.
There is also uncertainty related to the setting of the reference values and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties are not quantified.

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within each county and recalculating the area-weighted mean alien coniferous share.
For scaled county values, bootstrap draws are transformed using the same national reference levels X0 and X100 as the point estimates (truncation at scaling).
For scaled regional and national values, uncertainty is propagated from the scaled county bootstrap draws.
The resulting distributions are summarised with median and quantile intervals.


## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Alien coniferous tree species

All trees larger than 5 cm dbh are recorded and identified for each 250 m² NFI plot according to methods described in (@viken2021-feltinstruks).
For this indicator, data for the NFI variables `gran_pr`, `intro_gran_pr`, `intro_furu_pr`, and `komm_m_nat_gran` are used for the reporting periods defined in section 9.1.
Plots with missing species-percentage values are included in aggregated estimates with 0% share and full plot area weight.
Aggregated area-weighted data must be used for appropriate spatial units such as counties, regions or the whole country.
As the NFI plots vary in density for forest below/above the coniferous forest line and in Finnmark, each type of plot represents a different part of all forest land.
In addition, some NFI plots are divided in two parts in the data.
These variations in plot representativity must be accounted for in aggregated estimates.
Data from NFI contain an area-weighting variable to be used in aggregating NFI data for spatial units larger than individual NFI plots.

```{r load-data}
#| code-summary: "Load data"

# This export wraps each physical line in an extra pair of double quotes, so a
# single read_csv() sees one column spanning the whole file. Read once as one
# character column per line (readr unwraps the outer quotes), then parse the
# inner CSV text.
lsk_inner_csv <- read_csv(
  here("data/LSK_data_okologisk_tilstand_2026.csv"),
  col_names = FALSE,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8"),
  col_types = cols(.default = col_character())
) |>
  pull(1) |>
  paste(collapse = "\n")

lsk_data_raw <- read_csv(
  I(lsk_inner_csv),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, probably counties (i.e., de facto Ecosystem Assets).

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-aats-derived-vars}

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

We filter out Finnmark plots marked for removal (`merket_finnm == 0`), keep data from 2010 onward, and standardise county names by merging split NFI labels to county level.
Missing (`NA`) values for `gran_pr`, `intro_gran_pr`, and `intro_furu_pr` are set to 0 before aggregation, so these plots remain in the area basis (`au_areal`) with 0% alien coniferous share (missing values occur mainly in harvest class 1 stands).
The plot-level indicator depends on `komm_m_nat_gran`: when `n` (no natural spruce), the share is `pmin(gran_pr + intro_gran_pr + intro_furu_pr, 100)`; when `j` (natural spruce present), it is `pmin(intro_gran_pr + intro_furu_pr, 100)`.
For each included plot, the area-weighted contribution to later aggregation is `indicator_continuous * au_areal`.

```{r aats-create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(merket_finnm == 0) |>
  filter(sesong >= 2010) |>
  mutate(
    gran_pr = dplyr::coalesce(gran_pr, 0),
    intro_gran_pr = dplyr::coalesce(intro_gran_pr, 0),
    intro_furu_pr = dplyr::coalesce(intro_furu_pr, 0),
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    fylke_name = normalize_county_parts(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    alien_conif_share = dplyr::case_when(
      komm_m_nat_gran == "n" ~ gran_pr + intro_gran_pr + intro_furu_pr,
      TRUE ~ intro_gran_pr + intro_furu_pr
    ),
    indicator_continuous = pmin(alien_conif_share, 100),
    indicator_continuous_weighted = indicator_continuous * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r aats-select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    merket_finnm,
    au_areal,
    komm_m_nat_gran,
    gran_pr,
    intro_gran_pr,
    intro_furu_pr,
    alien_conif_share,
    indicator_continuous,
    indicator_continuous_weighted,
    region_fylke,
    region_code,
    region_name,
    fylke_name,
    fylke_name_clean
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values {#sec-aats-non-scaled}

For each reporting period, we calculate non-scaled indicator values separately for counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.

```{r aats-calculate-all-periods}
#| code-summary: "Calculate non-scaled values"

aats_results <- purrr::imap(
  reporting_periods,
  ~ calculate_period_indicators(lsk_data, .y, .x)
) |>
  purrr::compact()

aats_results_county_all_periods   <- purrr::map_dfr(aats_results, "county")
aats_results_region_all_periods   <- purrr::map_dfr(aats_results, "region")
aats_results_national_all_periods <- purrr::map_dfr(aats_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted non-scaled mean alien coniferous share (%), giving a distribution of 1000 draws per county.

```{r aats-county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(here::here("data/LSK_data_okologisk_tilstand_2026.csv")), file.mtime(here::here("R/NO_AATS_002.qmd")))

county_draws_long <- purrr::imap_dfr(aats_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-aats-county-nonscaled) summarises non-scaled bootstrap draws for all reporting periods: the median area-weighted alien coniferous share per county.

```{r tbl-aats-county-nonscaled}
#| tbl-cap: "Non-scaled alien coniferous tree species share (%) by county for all reporting periods (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aats_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Non-scaled value (%)` = non_scaled_median,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-aats-reference-values}

National reference levels are X0 = 50% alien coniferous species (worst condition) and X100 = 0% alien coniferous species (best condition).

```{r aats-reference-values}
#| code-summary: "National reference levels"

x0 <- 50
x100 <- 0

```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 50%, X100 = 0%; truncation at scaling).
@fig-aats-norm shows the mapping.

```{r aats-county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aats_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-aats-norm}
#| fig-cap: "Linear normalisation of alien coniferous tree species share (%) to indicator values (0–1), using national reference levels X0 = 50% and X100 = 0%. Points are 500 sampled bootstrap draws, pooled across all reporting periods."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 5
#| fig-width: 6

norm_plot_data <- county_draws_scaled |>
  dplyr::slice_sample(n = min(500, nrow(county_draws_scaled)))

aats_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::labs(
    x = "Alien coniferous species share (%)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
aats_norm_plot$layers[[1]]$aes_params$size <- 3

suppressMessages(
  ggsave(
    filename = here::here("img/aats_norm_plot.png"),
    plot = aats_norm_plot,
    scale = 1.5
  )
)

aats_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r aats-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

aats_mc_cache_deps <- c(
  here::here("data/LSK_data_okologisk_tilstand_2026.csv"),
  here::here("R/functions.R"),
  here::here("R/NO_AATS_002.qmd")
)
aats_mc_rds_regional <- here::here("data/aats_mc_regional.rds")

if (
  file.exists(aats_mc_rds_regional) &&
  file.mtime(aats_mc_rds_regional) >= max(file.mtime(aats_mc_cache_deps), na.rm = TRUE)
) {
  aats_mc_regional <- readRDS(aats_mc_rds_regional)
} else {
  aats_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(aats_mc_regional, aats_mc_rds_regional)
}
```

```{r aats-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

aats_mc_regional_for_national <- aats_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

aats_mc_rds_national <- here::here("data/aats_mc_national.rds")

if (
  file.exists(aats_mc_rds_national) &&
  file.mtime(aats_mc_rds_national) >= max(
    c(file.mtime(aats_mc_cache_deps), file.mtime(aats_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  aats_mc_national <- readRDS(aats_mc_rds_national)
} else {
  aats_mc_national <- ecTools::ec_upscale(
    data        = aats_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(aats_mc_national, aats_mc_rds_national)
}
```

```{r aats-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

aats_mc_combined <- dplyr::bind_rows(
  aats_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  aats_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- aats_results_county_all_periods |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- aats_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- aats_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- aats_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    aats_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- aats_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    aats_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  aats_results_county_all_periods |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x0,
      original_value_pct = round(indicator_value, 2),
      reference_value_pct = x0,
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      aats_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      aats_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county level, truncation is evaluated by comparing the non-scaled point estimate to X0 (50%).
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.

```{r aats-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_pct, original_value_pct) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded X0 (50%) and required truncation.\n")
}

```

The time series below (@fig-aats-timeSeries) summarises regional and national values.

```{r fig-aats-timeSeries}
#| fig-cap: "Indicator values for alien coniferous tree species by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

aats_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
aats_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
aats_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(aats_region_levels)),
  aats_region_levels
)

aats_plot_data <- aats_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!aats_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = aats_region_levels))

ggplot2::ggplot(
  aats_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = aats_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to X0 (50% alien coniferous share) and 1 corresponds to X100 (0% alien coniferous share).

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-aats-county-results to browse county rows across all reporting periods.

```{r tbl-aats-county-results}
#| tbl-cap: 'County indicator values for alien coniferous tree species by reporting period. Values are scaled to 0–1 using national X0 and X100. Uncertainty is derived from 1000 bootstrap draws. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Alien coniferous tree species",
    ID = "NO_AATS_002",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-aats-region-results}
#| tbl-cap: 'Regional indicator values for alien coniferous tree species by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Alien coniferous tree species",
    ID = "NO_AATS_002",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-aats-national-results}
#| tbl-cap: 'National indicator values for alien coniferous tree species by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Alien coniferous tree species",
    ID = "NO_AATS_002",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`aats_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  aats_mc_combined,
  here::here("data/results_NO_AATS_002.csv"),
  row.names = FALSE
)
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [EF, KOS]
* Methodology: [EF, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::