---
title: "Share of forest area without clear-cutting"
subtitle: |
  NO_AIFH_001
indicatorID: 
  - NO_AIFH_001
indicatorName: 
  - Share of forest area without clear-cutting
verbatimName: 
  - Arealandel ikke flatehogd skog
folderName: NO_AIFH_001
country: Norway
continent: Europe
ECT: C1 Landscape and Seascape Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2024 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: ecTools workflow; national refs X0 = 100% clear-cut area share, X100 = 0% clear-cut area share
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_AIFH_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
  docx:
    embed-resources: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.
date: June 9, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 09 Jun. 2026 - NO_AIFH_001, ecTools workflow
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:

<!-- To be added. -->

2. **Methods**:

Data come from the National Forest Inventory (NFI) LSK extract.
At plot level, clear-cutting is recorded in the NFI variable `FLATEHOGD`.
This variable is only available from 2020 onward, so the indicator is calculated for one reporting period: 2020–2024.
Non-scaled values are estimated by bootstrapping counties, and the resulting draws are scaled and aggregated to regional and national levels.

3. **Interpretation**:

An indicator value of 1 means that 0% of the forest area is clear-cut (best condition, X100).
An indicator value of 0 means that 100% of the forest area is clear-cut (worst condition, X0).

4. **Results and development needs**:

<!-- To be added. -->

### 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**:

<!-- To be added. -->

2. **Metoder**:

Dataene kommer fra Landsskogtakseringens LSK-uttrekk.
På flate-nivå registreres flatehogst i variabelen `FLATEHOGD`.
Variabelen finnes bare fra og med 2020, så indikatoren beregnes for én rapporteringsperiode: 2020–2024.
Ikke-skalerte verdier estimeres ved bootstrap på fylkesnivå, og trekningene skaleres og aggregeres til region- og riksnivå.

3. **Tolkning**:

En indikatorverdi på 1 tilsvarer 0 % arealandel med flatehogst (beste tilstand, X100).
En indikatorverdi på 0 tilsvarer 100 % arealandel med flatehogst (verste tilstand, X0).

4. **Resultater og utviklingsbehov**:

<!-- To be added. -->

## 2. About the underlying data

<!-- To be added. -->

### 2.1 Spatial and temporal extent, resolution, and representativity

The underlying variable is only available from 2020.
The indicator is therefore reported for a single five-year period, 2020–2024.
Earlier periods may be added if `FLATEHOGD` becomes available for those years.

### 2.2 Sampling design

<!-- To be added. -->

### 2.3 Original units

At plot level, the NFI variable `FLATEHOGD` uses code 1 for clear-cut plots, code 2 for no clear-cutting, and code 3 for undetermined status.
Plots with codes 1, 2, or 3 are retained; only missing values are excluded.
Aggregated values are the area-weighted mean proportion of forest area that is clear-cut.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

C1 Landscape and Seascape Characteristics

### 3.2 Ecosystem condition characteristic

<!-- To be added. -->

### 3.3 Collinearities with other indicators

<!-- To be added. -->

### 3.4 Impact factors

<!-- To be added. -->

### 3.5 Spatial aggregation

The spatial aggregation pathway is: aggregate (NFI plots to county) – [scale – truncate] – aggregate (geographic region) – aggregate (national level).
Non-scaled values are area-weighted means of the plot-level `FLATEHOGD` variable.
Scaling uses national reference levels X0 = 100% clear-cut area share and X100 = 0% clear-cut area share (proportions 1 and 0 on the original scale).

## 4. Reference condition and levels

### 4.1 Reference condition

<!-- To be added. -->

### 4.2 Reference levels

X0 is set to 100% clear-cut area share (worst condition) and X100 to 0% clear-cut area share (best condition).
Values are scaled linearly between X0 and X100, with truncation where values fall outside the reference range (@sec-aifh-reference-values).

#### 4.2.1 Spatial resolution and validity

National reference levels are applied when scaling each county before regional and national aggregation.

## 5. Uncertainties

### 5.1 General uncertainties

<!-- To be added. -->

### 5.2 Reported uncertainties

County non-scaled values are bootstrapped with 1000 draws per county per period.
County draws are scaled using the national X0 and X100.
Uncertainty is propagated when aggregating to regional and national levels.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Share of forest area with clear-cut

The NFI records clear-cutting status in the variable `FLATEHOGD` (column name after `clean_names()`: `flatehogd`).
Code 1 identifies clear-cut plots, code 2 identifies plots without clear-cutting, and code 3 indicates undetermined status.
Plots with codes 1, 2, or 3 are retained; only missing values are excluded.
Codes 2 and 3 are both treated as no clear-cutting in the indicator calculation.
`flatehogd` is available from 2020 onward, and analyses use the reporting period 2020–2024.
Both productive and unproductive forest plots (`arealtype` 1 and 12) are included.


```{r load-data}
#| code-summary: "Load data"

lsk_data_file <- here("data/LSK_data_okologisk_tilstand_2026-06-03.csv")

lsk_data_raw <- read_csv(
  lsk_data_file,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties, regions and nationally.
Where counties are split into parts, those parts are merged to a single county name before aggregation.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-aifh-derived-vars}

This section prepares the NFI plot data for indicator calculation.
We define one five-year reporting period (2020–2024).

```{r reporting-periods}
#| code-summary: "Define reporting periods"

min_indicator_year <- 2020L

reporting_periods <- list(
  # "2010-2014" = 2010:2014,  # enable when data is available
  # "2015-2019" = 2015:2019,  # enable when data is available
  "2020-2024" = 2020:2024
)

```

We filter the raw dataset to inventory years from 2020 onward (`sesong >= min_indicator_year`) and include productive and unproductive forest plots (`arealtype` 1 and 12).
Plots with non-missing `flatehogd` (codes 1, 2, or 3) are retained, and split NFI county labels are standardised by merging county parts to a single county name.
Code 1 is recoded to 1 (clear-cut); codes 2 and 3 are recoded to 0 (no clear-cutting).
For each included plot, the area-weighted contribution to later aggregation is `indicator_continuous * au_areal`.

```{r aifh-create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(sesong >= min_indicator_year) |>
  filter(arealtype %in% c(1, 12)) |> 
  filter(!is.na(flatehogd)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    fylke_name = normalize_county_parts(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = dplyr::if_else(flatehogd == 1, 1, 0),
    indicator_continuous_weighted = indicator_continuous * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r aifh-select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    fylke_name,
    fylke_name_clean,
    flatehogd,
    indicator_continuous,
    indicator_continuous_weighted
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values {#sec-aifh-non-scaled}

For the reporting period (2020–2024), we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted mean of all included plots: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
Plots with code 3 contribute their area to the denominator and are treated as no clear-cutting (0) in the numerator.

```{r calculate-aifh-all-periods}
#| code-summary: "Calculate non-scaled values"

 aifh_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 aifh_results_county_all_periods   <- purrr::map_dfr(aifh_results, "county")
 aifh_results_region_all_periods   <- purrr::map_dfr(aifh_results, "region")
 aifh_results_national_all_periods <- purrr::map_dfr(aifh_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each bootstrap replicate computes the area-weighted share of clear-cut forest area on the non-scaled scale, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(lsk_data_file), file.mtime(here::here("R/NO_AIFH_001.qmd")))

county_draws_long <- purrr::imap_dfr(aifh_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-aifh-county-nonscaled) summarises non-scaled bootstrap draws for the reporting period 2020–2024.
It shows the median area-weighted share of clear-cut forest area per county.

```{r tbl-aifh-county-nonscaled}
#| tbl-cap: "Non-scaled share of clear-cut forest area by county for the reporting period 2020–2024 (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aifh_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    non_scaled_median_pct = non_scaled_median * 100
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Non-scaled value (proportion)` = non_scaled_median,
    `Non-scaled value (%)` = non_scaled_median_pct,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (proportion)`,
    decimals = 4
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-aifh-reference-values}

National reference levels are X0 = 100% clear-cut area share and X100 = 0% clear-cut area share (proportions 1 and 0 on the original scale).

```{r aifh-reference-values}
#| code-summary: "National reference levels"

x0 <- 1
x100 <- 0


```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 1, X100 = 0; truncation at scaling).
@fig-aifh-norm shows the mapping.

```{r county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    aifh_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-aifh-norm}
#| fig-cap: "Linear normalisation of clear-cut area share (proportion) to indicator values (0–1), using national reference levels X0 = 100% and X100 = 0%. Points are 500 sampled bootstrap draws for the reporting period 2020–2024."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 5
#| fig-width: 6

norm_plot_data <- county_draws_scaled |>
  dplyr::slice_sample(n = min(500, nrow(county_draws_scaled)))

aifh_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::labs(
    x = "Clear-cut area share (proportion)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
aifh_norm_plot$layers[[1]]$aes_params$size <- 3

suppressMessages(
  ggsave(
    filename = here::here("img/aifh_norm_plot.png"),
    plot = aifh_norm_plot,
    scale = 1.5
  )
)

aifh_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r aifh-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

aifh_mc_cache_deps <- c(
  lsk_data_file,
  here::here("R/functions.R"),
  here::here("R/NO_AIFH_001.qmd")
)
aifh_mc_rds_regional <- here::here("data/aifh_mc_regional.rds")

if (
  file.exists(aifh_mc_rds_regional) &&
  file.mtime(aifh_mc_rds_regional) >= max(file.mtime(aifh_mc_cache_deps), na.rm = TRUE)
) {
  aifh_mc_regional <- readRDS(aifh_mc_rds_regional)
} else {
  aifh_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(aifh_mc_regional, aifh_mc_rds_regional)
}
```

```{r aifh-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

aifh_mc_regional_for_national <- aifh_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

aifh_mc_rds_national <- here::here("data/aifh_mc_national.rds")

if (
  file.exists(aifh_mc_rds_national) &&
  file.mtime(aifh_mc_rds_national) >= max(
    c(file.mtime(aifh_mc_cache_deps), file.mtime(aifh_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  aifh_mc_national <- readRDS(aifh_mc_rds_national)
} else {
  aifh_mc_national <- ecTools::ec_upscale(
    data        = aifh_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(aifh_mc_national, aifh_mc_rds_national)
}
```

```{r aifh-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

aifh_mc_combined <- dplyr::bind_rows(
  aifh_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  aifh_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- aifh_results_county_all_periods |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- aifh_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- aifh_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- aifh_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    aifh_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- aifh_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    aifh_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  aifh_results_county_all_periods |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value < x100,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = 0,
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      aifh_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      aifh_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county level, truncation is evaluated by comparing the non-scaled point estimate to X100 (0% clear-cut area share).
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r aifh-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_pct, original_value_pct) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values fell below X100 (0% clear-cut area share) and required truncation.\n")
}

```

The figure below (@fig-aifh-timeSeries) summarises regional and national indicator values for the reporting period 2020–2024.
When additional periods are enabled in `reporting_periods`, the same plot will show values across periods.

```{r fig-aifh-timeSeries}
#| fig-cap: "Indicator values for share of forest area without clear-cutting by region for the reporting period 2020–2024. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

aifh_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
aifh_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
aifh_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(aifh_region_levels)),
  aifh_region_levels
)

aifh_plot_data <- aifh_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!aifh_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = aifh_region_levels))

ggplot2::ggplot(
  aifh_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = aifh_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county, regional, and national scales for the reporting period 2020–2024.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to X0 (100% clear-cut area share) and 1 corresponds to X100 (0% clear-cut area share).

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-aifh-county-results to browse county rows for the reporting period 2020–2024.

```{r tbl-aifh-county-results}
#| tbl-cap: 'County indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Values are scaled to 0–1 using national X0 and X100. Uncertainty is derived from 1000 bootstrap draws. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-aifh-region-results}
#| tbl-cap: 'Regional indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-aifh-national-results}
#| tbl-cap: 'National indicator values for share of forest area without clear-cutting for the reporting period 2020–2024. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Share of forest area without clear-cutting",
    ID = "NO_AIFH_001",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`aifh_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  aifh_mc_combined,
  here::here("data/results_NO_AIFH_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
