---
title: "Bilberry cover by area"
subtitle: |
  NO_BLAA_001  
indicatorID: 
  - NO_BLAA_001
indicatorName: 
  - NO_BLAA
folderName: NO_BLAA
country: Norway
continent: Europe
ECT: B3 Biotic functional characteristic
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BLAA_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county/sub-county) - Scale - Aggregate (geographic region) - Aggregate (national level) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen 
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
AuthorList: Storaunet, K. O., Snoksrud, O., Halbritter, A. H., and Framstad, E. 
date: May 7, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 07 May. 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
Bilberry (*Vaccinium myrtillus*) is a widely distributed key component of boreal forest vegetation.
It provides structure to the field layer and important food resources for a wide range of invertebrates, birds and mammals.
It is particularly associated with mature, moderately moist forests with intermediate crown cover, which is typical for much of intact forests under natural dynamics.
Hence, the cover (%) of bilberry plants is positively associated with ecosystem condition. 

2. **Methods**:
Data for this indicator come from the National Forest Inventory (NFI), where bilberry cover has been recorded (with current methods) for each NFI plot since 2008.
Data are combined for each full inventory cycle of 5 years, starting with 2010, with the last available period covering data from 2020-2024.
The recorded mean cover of bilberries is compared to reference values for regions, based on expert assessments supported by data on bilberry cover from NFI plots with near-intact conditions, regional distribution of dominating tree species, and general assessments of natural forest dynamics.
The indicator is equivalent to the same indicator used in the Norwegian Nature Index.
The indicator is ready for use.
The indicator should not be used at finer spatial resolution than applied to reference values (regions).

3. **Interpretation**:
An indicator value of 1 signifies that the bilberry cover is at least as extensive as expected in the reference condition.
Values decreasing towards zero reflect a deteriorating condition towards complete lack of bilberry plants.

4. **Results and development needs**:
For the current reporting period (2020-2024) the indicator value is 0.88, with a somewhat lower values for region West (0.76).
The indicator value increases over time nationally and for all regions, except for a low decline (from a high value) in region South since the previous period.
There are no specific development needs, except for a possible review of reference values and the scaling function.

## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: 

2. **Metoder**: 

3. **Tolkning**:

4. **Resultater og utviklingsbehov**: 



## 2. About the underlying data

This indicator uses data from the Norwegian National Forest Inventory (NFI, Landsskogtakseringen).
The bilberry cover by area is recorded for each NFI inventory plot.
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the bilberry cover for NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal resolution and extent

Data are available from 2010, when the bilberry cover by area was first recorded based on current methods.
Annual estimates are possible due to representative sampling every year.
For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. The data are representative for all forest land of Norway.
Spatial resolution should be limited to the resolution applied to reference values, that is counties or sub-counties for scaling and regions for aggregated reporting.

### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

### 2.3 Original units

The indicator is specified as the percentage (%) of the area covered by bilberry on the NFI plot.
In the analyses, this value is converted to a proportion before area weighting, aggregation, and scaling.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

Bilberry plants are important food resources for a range of invertebrate and vertebrate species. Hence, the indicator is classified as a B3 Biotic functional characteristic.

### 3.2 Ecosystem condition characteristic

Bilberry is a key component in boreal forest ecosystems and is characterised as Functionally important species and biophysical structures. 

### 3.3 Collinearities with other indicators

Bilberry cover is related to tree cover density (crown cover), with the highest values at intermediate crown cover. Correlation with other indicators related to old forest is likely. 

### 3.4 Impact factors

Natural disturbances from climate variability (drought, storms) and pathogens (fungi, insects) will affect bilberry cover.
Human impacts with major negative effects on bilberry cover are various forestry activities, especially open harvest types such as clear-cutting, short rotation, dense planting, and whole-tree harvesting.
Nitrogen fertilisation or pollution will reduce the health and viability of bilberry plants. 
Human-induced climate change with warmer/drier summers or thawing/freezing episodes in winter/early spring is also likely to have negative effects on bilberry cover.

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county/sub-county) – Scale – Aggregate (geographic region) – Aggregate (national level) - Truncate.
For each NFI plot we first convert bilberry cover from percent to proportion and create `indicator_continuous_weighted = indicator_cover * au_areal`.
Non-scaled indicator values for each reporting period are calculated separately for each geographic unit as area-weighted arithmetic means of all NFI plots within that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
This is done for counties or sub-counties, regions, and at national level.
Scaling is applied after these non-scaled county or sub-county aggregates have been calculated: each county or sub-county is scaled against its own lower reference level X0 (0% cover) and upper reference level X100 using a linear scaling function without truncation.
Regional and national scaled values are area-weighted means of these scaled county or sub-county values, using each county or sub-county's represented area as weight; mean values that are larger than the reference value are then truncated to 0–1 both at the regional and national level.

## 4. Reference condition and levels

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances.  

### 4.1 Reference condition

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels

The lower reference level X0 is set to 0% bilberry cover and the upper reference level X100 is set separately for each county or sub-county.
X0 denotes the worst possible condition and X100 denotes the best possible condition on the original cover scale.
Between X0 and X100, indicator values are scaled linearly to a 0-1 indicator scale.
Scaled values may exceed 1 at county or sub-county level when bilberry cover exceeds the county-specific X100; truncation to 1 is applied only after area-weighted aggregation to regional and national level.
The county and sub-county X100 values used in the analyses are listed in @sec-blaa-reference-values.
X0 is 0 for all units.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at for regional groupings of counties or sub-counties, not as one national value.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties
### 5.1 General uncertainties

Reference levels are held fixed in this workflow (their own uncertainties are not propagated). Regional and national values are area-weighted summaries of counties, so they inherit both model uncertainty at county level and any mismatch between county boundaries and ecological variation within regions. 

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in bilberry cover among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within the relevant geographic unit and recalculating the area-weighted mean.
This is done separately for counties or sub-counties, regions, and the national level.
For scaled county or sub-county values, bootstrap draws are transformed with the same county-specific linear scaling function used for the point estimate, without truncation.
For scaled regional and national values, uncertainty is propagated from the scaled county bootstrap draws; each Monte Carlo replicate is truncated to the 0–1 condition scale after aggregation.
The resulting distributions are summarised with median and quantile intervals.

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->
::: {#refs}
:::

## 7. Import and prepare datasets

The bilberry cover by area is recorded for each NFI plot in the LSK export.
For this indicator we use data from each full 5-year inventory cycle, starting in 2010.

```{r load-data}
#| code-summary: "Load data"

# Load main data file
# This export wraps each physical line in an extra pair of double quotes, so a
# single read_csv() sees one column spanning the whole file. Read once as one
# character column per line (readr unwraps the outer quotes), then parse the
# inner CSV text.
lsk_inner_csv <- read_csv(
  here("data/LSK_data_okologisk_tilstand_2026.csv"),
  col_names = FALSE,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8"),
  col_types = cols(.default = col_character())
) |>
  pull(1) |>
  paste(collapse = "\n")

lsk_data_raw <- read_csv(
  I(lsk_inner_csv),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, e.g., regions.

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

Next we filter the raw dataset to include both productive and unproductive forest plots (`arealtype` 1 and 12), remove records missing in the `blaabaer_gjsn` variable, and standardise the county names.
NFI bilberry cover is recorded on a 0-100 percent scale and is converted to a proportion before aggregation and scaling.
Area-weighted values are created `indicator_continuous_weighted = indicator_cover * au_areal` (section 9.1).

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    # productive and unproductive forest plots
    arealtype %in% c(1, 12)
  ) |>
  # remove records with missing data on bilberry cover
  filter(!is.na(blaabaer_gjsn)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_blaa_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),

    indicator_cover = blaabaer_gjsn / 100,
    indicator_continuous_weighted = indicator_cover * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    blaabaer_gjsn,
    indicator_continuous_weighted
  )

```


### 9.2 Calculate non-scaled area-weighted indicator values

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
County or sub-county estimates use the plot-level `fylke_name` field; regional estimates use `region_code`; the national estimate uses all plots in the period.

```{r calculate-blaa-all-periods}
#| code-summary: "Calculate non-scaled values"

 blaa_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 blaa_results_county_all_periods   <- purrr::map_dfr(blaa_results, "county")
 blaa_results_region_all_periods   <- purrr::map_dfr(blaa_results, "region")
 blaa_results_national_all_periods <- purrr::map_dfr(blaa_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted mean non-scaled indicator value, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr file.mtime(here::here("R/functions.R"))

county_draws_long <- purrr::imap_dfr(blaa_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, county_part, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-blaa-county-nonscaled) summarises non-scaled bootstrap draws for all reporting periods: the median area-weighted bilberry cover per county or sub-county. Use the search box and page controls to browse rows.

```{r tbl-blaa-county-nonscaled}
#| tbl-cap: "Non-scaled, area-weighted bilberry cover by county or sub-county for all reporting periods (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, county_part, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    blaa_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    non_scaled_median_pct = non_scaled_median * 100
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    `Sub-county` = county_part,
    Region = region_name,
    `Non-scaled value (proportion)` = non_scaled_median,
    `Non-scaled value (%)` = non_scaled_median_pct,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (proportion)`,
    decimals = 4
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = `Total area (ha)`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-blaa-reference-values}

County or sub-county reference values are imported manually.

```{r blaa-reference-values}
#| code-summary: "Reference values by county or sub-county"

# X100 reference values (% bilberry cover by area)
blaa_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuV", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_pct = c(
    12.5, 12.5, 12.5, 10, 12.5, 10, 12.5, 10, 12.5,
    12.5, 10, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5,
    10, 5, 5, 2.5
  )
) |>
  dplyr::mutate(x100 = x100_pct / 100)

x0 <- 0
county_reference <- blaa_reference_values |>
  dplyr::select(county_part, x100)

```

Each county's 1000 non-scaled draws from section 9.3 are linearly scaled with `ec_scale_linear()` (X0 = 0, county-specific X100) without truncation.
@fig-blaa-norm shows the mapping for each sub-county.

```{r county-draws-scaled}
#| code-summary: "Scale county draws (no truncation); derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_scale_linear(
      non_scaled_value,
      x0 = x0,
      x100 = x100
    ),
    nation = "National"
  )

# Summarise scaled draws per county → county-level results table (10.1)
county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, county_part, total_area) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    blaa_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-blaa-norm}
#| fig-cap: "Linear scaling of bilberry cover proportion to the indicator scale (without truncation), one panel per sub-county reference unit. Points are 100 sampled bootstrap draws per unit, pooled across all reporting periods; values may exceed 1 when cover exceeds X100."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 8
#| fig-width: 8

norm_plot_data <- county_draws_scaled |>
  dplyr::group_by(county_part) |>
  dplyr::slice_sample(n = 100) |>
  dplyr::ungroup()

blaa_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::facet_wrap(~ norm_plot_data$county_part, scales = "free_x") +
  ggplot2::labs(
    x = "Bilberry cover (proportion)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
blaa_norm_plot$layers[[1]]$aes_params$size <- 3

ggsave(
  filename = here::here("img/blaa_norm_plot.png"),
  plot     = blaa_norm_plot,
  scale    = 1.5
)

blaa_norm_plot
```

The scaled draws are passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.
Truncation to the 0–1 condition scale is applied to regional and national Monte Carlo draws after aggregation.

```{r blaa-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

blaa_mc_cache_deps <- c(
  here::here("data/LSK_data_okologisk_tilstand_2026.csv"),
  here::here("R/functions.R"),
  here::here("R/NO_BLAA_001.qmd")
)
blaa_mc_rds_regional <- here::here("data/blaa_mc_regional.rds")

if (
  file.exists(blaa_mc_rds_regional) &&
  file.mtime(blaa_mc_rds_regional) >= max(file.mtime(blaa_mc_cache_deps), na.rm = TRUE)
) {
  blaa_mc_regional_raw <- readRDS(blaa_mc_rds_regional)
} else {
  blaa_mc_regional_raw <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(blaa_mc_regional_raw, blaa_mc_rds_regional)
}

blaa_mc_regional <- blaa_mc_regional_raw |>
  dplyr::mutate(
    sampled_mean_raw = sampled_mean,
    sampled_mean = ec_truncate_indicator(sampled_mean)
  )
```

```{r blaa-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

blaa_mc_regional_for_national <- blaa_mc_regional_raw |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

blaa_mc_rds_national <- here::here("data/blaa_mc_national.rds")

if (
  file.exists(blaa_mc_rds_national) &&
  file.mtime(blaa_mc_rds_national) >= max(
    c(file.mtime(blaa_mc_cache_deps), file.mtime(blaa_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  blaa_mc_national_raw <- readRDS(blaa_mc_rds_national)
} else {
  blaa_mc_national_raw <- ecTools::ec_upscale(
    data        = blaa_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(blaa_mc_national_raw, blaa_mc_rds_national)
}

blaa_mc_national <- blaa_mc_national_raw |>
  dplyr::mutate(
    sampled_mean_raw = sampled_mean,
    sampled_mean = ec_truncate_indicator(sampled_mean)
  )
```

```{r blaa-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

blaa_mc_combined <- dplyr::bind_rows(
  blaa_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  blaa_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- blaa_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::filter(!is.na(x100)) |>
  dplyr::mutate(
    scaled_value = ec_scale_linear(indicator_value, x0 = x0, x100 = x100)
  )

region_scaled_agg_all_periods <- blaa_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    scaled_value_raw = median(sampled_mean_raw),
    scaled_value = median(sampled_mean),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- blaa_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    scaled_value_raw = median(sampled_mean_raw),
    scaled_value = median(sampled_mean),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- blaa_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    blaa_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- blaa_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    blaa_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      blaa_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value_raw > 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(period, level, region_code, region_name, fylke_name,
                  reference_value_pct, exceeds_threshold, original_value_pct),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      blaa_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value_raw > 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(period, level, region_code, region_name, fylke_name,
                  reference_value_pct, exceeds_threshold, original_value_pct)
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

Truncation is applied only after regional and national aggregation.
The tables below report cases where the area-weighted aggregated scaled value exceeded 1 before truncation (`scaled_value_raw > 1`).
The following tables summarise how often this occurred in each period and list the affected units.

```{r report-truncation}
#| code-summary: "Report truncation"

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(
      period,
      level,
      location,
      reference_value_pct,
      original_value_pct
    ) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original Value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original Value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their reference and required truncation.\n")
}

```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
County values are linearly scaled without truncation; regional and national values are truncated to 0–1 after area-weighted aggregation (section 9.4).
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).

The time series below (@fig-blaa-timeSeries) summarises regional and national values.

```{r fig-blaa-timeSeries}
#| fig-cap: "Indicator values for bilberry cover by area by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

blaa_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
blaa_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
blaa_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(blaa_region_levels)),
  blaa_region_levels
)

blaa_plot_data <- blaa_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!blaa_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = blaa_region_levels))

ggplot2::ggplot(blaa_plot_data,
                ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = blaa_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(nrow = 2,
                                   override.aes = list(size = 0.7, linewidth = 0.45)),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-county-results-table}
#| tbl-cap: 'County or sub-county indicator values for bilberry cover by area by reporting period. Values are linearly scaled with county-specific X0 and X100 without truncation. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Bilberry cover by area",
    ID = "NO_BLAA_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    `Sub-county` = .data$county_part,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-region-results-table}
#| tbl-cap: 'Regional indicator values for bilberry cover by area by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region per period), with truncation to 0–1 applied after aggregation. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the truncated MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Bilberry cover by area",
    ID = "NO_BLAA_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-national-results-table}
#| tbl-cap: 'National indicator values for bilberry cover by area by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws), with truncation to 0–1 applied after aggregation. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the truncated MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Bilberry cover by area",
    ID = "NO_BLAA_001",
    `Accounting area` = .data$region,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`blaa_mc_combined`: `year`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).
This mirrors the export format used in the MOOS indicator.

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  blaa_mc_combined,
  here::here("data/results_NO_BLAA_001.csv"),
  row.names = FALSE
)
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
