---
title: "Old forest"
subtitle: |
  NO_BGSK_001
indicatorID: 
  - NO_BGSK_001
indicatorName: 
  - Old forest
verbatimName:  
  - Gammel skog
folderName: NO_BGSK_001
country: Norway
continent: Europe
ECT: C1 - Landscape and Seascape Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T2.2 Deciduous temperate forests
verbatimEcosystem: 
  - T2.1 Boreal/subalpin skog
  - T2.2 Temperert edelauvskog
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "000.001"
VersionComment: Initial version
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BGSK_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
Precision: Approximate 95% bootstrap interval (2.5th–97.5th percentile of 1000 bootstrap draws) as inferential uncertainty around the point estimate (median). At county or sub-county level, bootstrapping is applied directly to NFI plot data; at regional and national level, uncertainty is propagated from scaled county bootstrap draws.
format:  
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Ulrika Jansson Asplund
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Mari Jönsson
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Erik Framstad
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Ken Olaf Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
AuthorList: Halbritter, A. H., Jansson Asplund, U., Jönsson, M., Framstad, E., Snoksrud, O., and Storaunet, K. O.
date: February 12, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
bibliography: references.bib
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
---

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull <- dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
if (file.exists(here::here("R/functions.R"))) {
  source(here::here("R/functions.R"))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> Recommended citation: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()
```
:::

::: {.callout-tip collapse="true"}
## Log
- 12 Feb. 2026 - Initial qmd from ecRxiv template; ECT C1; county X100 table from NI; NI column name pending confirmation
:::

<hr />

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->

2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->

3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
An indicator value of 1 signifies ...
Values decreasing toward zero reflect...

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->


### 1.2 Native language summary - Norwegian

**Bakgrunn og begrunnelse**:

**Metoder**:

**Tolkning**:

**Resultater og utviklingsbehov**:


## 2. About the underlying data

The old forest share of area is recorded from plots in the Norwegian National Forest Inventory (NFI, Landsskogtakseringen).
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the old forest share of area for NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal resolution and extent

Data are available from 2010, when the the old forest share of area was first recorded based on current methods.
Annual estimates are possible due to representative sampling every year.
For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010.
The data are representative for all forest land of Norway.
Spatial resolution should be limited to the resolution applied to reference values, that is counties or sub-counties for scaling and regions for aggregated reporting.

### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

### 2.3 Original units

Each forest inventory plot carries a binary presence variable `bgs` (0/1): whether biologically old forest occurs on that plot, and `au_areal` is the forest area on each plot (ha).
At county (fylke), region, or national level, the non-scaled share of old forest is the ratio of total old forest area on plots, divided by total forest area on all forest plots in the unit.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

### 2.5 Additional comments about the dataset

<!--# Optional. Text here -->


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

C1 - Landscape and Seascape Characteristics

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->

### 3.1.1 Other condition typologies

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->

### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->


### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county/sub-county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each reporting period, non-scaled values for county (fylke), region, and Norway are obtained by pooling all forest plots in the unit: the total forest area on plots where old forest is present, divided by the total forest area on all forest plots in that unit.
This is done for counties or sub-counties, regions, and at national level.
Scaling and truncation are applied only after these non-scaled county or sub-county aggregates have been calculated: each county or sub-county is scaled against its own lower reference level X0 (0% share) and upper reference level X100 using a linear scaling function, and values above 1 are truncated to 1 at that step.
Regional and national scaled point estimates are area-weighted means of the already scaled county or sub-county values, using each county or sub-county's represented area as weight.
Non-scaled regional and national point estimates are calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances.

### 4.2 Reference levels

The lower reference level X0 is set to 0% old forest share of area and the upper reference level X100 is set separately for each county or sub-county.
X0 denotes the worst possible condition and X100 denotes the best possible condition on the original share of area scale.
Between X0 and X100, indicator values are scaled linearly to a 0-1 indicator scale.
Scaled values above 1 are truncated to 1 when old forest share of area exceeds the county-specific X100.
The county and sub-county X100 values used in the analyses are listed in @sec-bgsk-reference-values.
X0 is 0 for all units.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or sub-county resolution.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties

### 5.1 General uncertainties

Reference levels are held fixed in this workflow (their own uncertainties are not propagated). Regional and national values are area-weighted summaries of counties, so they inherit both model uncertainty at county level and any mismatch between county boundaries and ecological variation within regions.

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in old forest share of area among forest plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling forest plots with replacement within the relevant geographic unit and recomputing the forest-area share.
This is done separately for counties or sub-counties, regions, and the national level.

For scaled county or sub-county values, bootstrap draws are transformed with the same county-specific reference levels as the point estimates (truncation at scaling).
For scaled regional and national values, uncertainty is propagated from the scaled county bootstrap draws.
The resulting distributions are summarised with median and quantile intervals.


## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

The old forest area share is built from each forest inventory plot: binary presence `bgs` (biologically old forest, 0/1).
For this indicator data from each full 5-year inventory cycle, starting in 2010, are used.

```{r load-data}
#| code-summary: "Load data"

# This export wraps each physical line in an extra pair of double quotes, so a
# single read_csv() sees one column spanning the whole file. Read once as one
# character column per line (readr unwraps the outer quotes), then parse the
# inner CSV text.
lsk_inner_csv <- read_csv(
  here("data/LSK_data_okologisk_tilstand_2026.csv"),
  col_names = FALSE,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8"),
  col_types = cols(.default = col_character())
) |>
  pull(1) |>
  paste(collapse = "\n")

lsk_data_raw <- read_csv(
  I(lsk_inner_csv),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, e.g., regions.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-bgsk-derived-vars}

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.


```{r reporting-periods}
#| code-summary: "Define reporting periods"

reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

The analysis includes both productive and unproductive forest plots (`arealtype` 1 and 12).
The LSK export contains the binary variable `bgs` (0/1) for biologically old forest on each forest plot.
`bgs` is the plot-level indicator on forest area (`au_areal`, ha): 1 = old forest present on the plot, 0 = not.
We store the per-plot numerator as `indicator_continuous_weighted = bgs * au_areal`; area-weighted aggregation to fylke, region, and national units is in @sec-bgsk-non-scaled.

```{r bgsk-create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  dplyr::filter(arealtype %in% c(1, 12)) |>
  dplyr::mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    bgs = as.numeric(bgs),
    indicator_continuous_weighted = bgs * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r bgsk-select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    bgs,
    indicator_continuous_weighted
  )

```


### 9.2 Calculate non-scaled area-weighted indicator values {#sec-bgsk-non-scaled}

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the share of inventoried forest area on plots with old forest present is

$$\frac{\sum_i (\mathrm{bgs}_i \cdot a_i)}{\sum_i a_i}$$

where $a_i$ is plot forest area (`au_areal`, ha).


```{r bgsk-calculate-all-periods}
#| code-summary: "Calculate non-scaled values"

bgsk_results <- purrr::imap(
  reporting_periods,
  ~ calculate_period_indicators(lsk_data, .y, .x)
) |>
  purrr::compact()

bgsk_results_county_all_periods   <- purrr::map_dfr(bgsk_results, "county")
bgsk_results_region_all_periods   <- purrr::map_dfr(bgsk_results, "region")
bgsk_results_national_all_periods <- purrr::map_dfr(bgsk_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted non-scaled share of old forest, giving a distribution of 1000 draws per county.

```{r bgsk-county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(here::here("data/LSK_data_okologisk_tilstand_2026.csv")), file.mtime(here::here("R/NO_BGSK_001.qmd")))

county_draws_long <- purrr::imap_dfr(bgsk_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, county_part, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-bgsk-county-nonscaled) summarises non-scaled bootstrap draws for all reporting periods: the median area-weighted old forest share per county or sub-county. Use the search box and page controls to browse rows.

```{r tbl-bgsk-county-nonscaled}
#| tbl-cap: "Non-scaled old forest share of forest area by county or sub-county for all reporting periods (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, county_part, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    bgsk_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    non_scaled_median_pct = non_scaled_median * 100
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    `Sub-county` = county_part,
    Region = region_name,
    `Non-scaled value (proportion)` = non_scaled_median,
    `Non-scaled value (%)` = non_scaled_median_pct,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (proportion)`,
    decimals = 4
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-bgsk-reference-values}

County or sub-county reference levels are listed below.

```{r bgsk-reference-values}
#| code-summary: "Reference values by county or sub-county"

bgsk_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuV", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_pct = c(
    50, 50, 50, 45, 50, 45, 50, 45, 50, 50, 45, 45, 45, 40, 40,
    40, 40, 45, 45, 45, 30, 30, 30
  )
) |>
  dplyr::mutate(x100 = x100_pct / 100)

x0 <- 0
county_reference <- bgsk_reference_values |>
  dplyr::select(county_part, x100)

knitr::kable(bgsk_reference_values |> dplyr::select(county_part, `X100 (%)` = x100_pct))

```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 0, county-specific X100; truncation at scaling).
@fig-bgsk-norm shows the mapping for each sub-county.

```{r bgsk-county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, county_part, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    bgsk_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-bgsk-norm}
#| fig-cap: "Linear normalisation of old forest share (proportion of forest area) to indicator values (0–1), one panel per sub-county reference unit. Points are 100 sampled bootstrap draws per unit, pooled across all reporting periods."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 8
#| fig-width: 8

norm_plot_data <- county_draws_scaled |>
  dplyr::group_by(county_part) |>
  dplyr::slice_sample(n = 100) |>
  dplyr::ungroup()

bgsk_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::facet_wrap(~ norm_plot_data$county_part, scales = "free_x") +
  ggplot2::labs(
    x = "Old forest share (proportion of forest area)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
bgsk_norm_plot$layers[[1]]$aes_params$size <- 3

suppressMessages(
  ggsave(
    filename = here::here("img/bgsk_norm_plot.png"),
    plot = bgsk_norm_plot,
    scale = 1.5
  )
)

bgsk_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r bgsk-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region (compute once, save RDS)"

bgsk_mc_cache_deps <- c(
  here::here("data/LSK_data_okologisk_tilstand_2026.csv"),
  here::here("R/functions.R"),
  here::here("R/NO_BGSK_001.qmd")
)
bgsk_mc_rds_regional <- here::here("data/bgsk_mc_regional.rds")

if (
  file.exists(bgsk_mc_rds_regional) &&
  file.mtime(bgsk_mc_rds_regional) >= max(file.mtime(bgsk_mc_cache_deps), na.rm = TRUE)
) {
  bgsk_mc_regional <- readRDS(bgsk_mc_rds_regional)
} else {
  bgsk_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(bgsk_mc_regional, bgsk_mc_rds_regional)
}
```

```{r bgsk-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

bgsk_mc_regional_for_national <- bgsk_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

bgsk_mc_rds_national <- here::here("data/bgsk_mc_national.rds")

if (
  file.exists(bgsk_mc_rds_national) &&
  file.mtime(bgsk_mc_rds_national) >= max(
    c(file.mtime(bgsk_mc_cache_deps), file.mtime(bgsk_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  bgsk_mc_national <- readRDS(bgsk_mc_rds_national)
} else {
  bgsk_mc_national <- ecTools::ec_upscale(
    data        = bgsk_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(bgsk_mc_national, bgsk_mc_rds_national)
}
```

```{r bgsk-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

bgsk_mc_combined <- dplyr::bind_rows(
  bgsk_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  bgsk_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- bgsk_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::filter(!is.na(x100)) |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- bgsk_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- bgsk_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- bgsk_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    bgsk_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- bgsk_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    bgsk_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  county_scaled_all_periods |>
    dplyr::left_join(
      bgsk_reference_values |> dplyr::select(county_part, reference_value_pct = x100_pct),
      by = "county_part"
    ) |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      bgsk_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      bgsk_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county or sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r bgsk-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_pct, original_value_pct) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their reference and required truncation.\n")
}

```

The time series below (@fig-bgsk-timeSeries) summarises regional and national values.

```{r fig-bgsk-timeSeries}
#| fig-cap: "Indicator values for biologically old forest share by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

bgsk_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
bgsk_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
bgsk_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(bgsk_region_levels)),
  bgsk_region_levels
)

bgsk_plot_data <- bgsk_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!bgsk_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = bgsk_region_levels))

ggplot2::ggplot(
  bgsk_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = bgsk_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```

## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-bgsk-county-results to browse county rows across all reporting periods.

```{r tbl-bgsk-county-results}
#| tbl-cap: 'County or sub-county indicator values for biologically old forest share by reporting period. Values are scaled to 0–1 using county-specific X0 and X100 via ec_normalise(). Uncertainty is derived from 1000 bootstrap draws scaled with the same function (section 9.3–9.4). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Biologically old forest share",
    ID = "NO_BGSK_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    `Sub-county` = .data$county_part,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-bgsk-region-results}
#| tbl-cap: 'Regional indicator values for biologically old forest share by reporting period. Scaled values and uncertainty are derived from ecTools::ec_upscale() Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Biologically old forest share",
    ID = "NO_BGSK_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-bgsk-national-results}
#| tbl-cap: 'National indicator values for biologically old forest share by reporting period. Scaled values and uncertainty are derived from ecTools::ec_upscale() Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "C1",
    Indicator = "Biologically old forest share",
    ID = "NO_BGSK_001",
    `Accounting area` = .data$region,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`bgsk_mc_combined`: `period`, `region`, `sampled_mean`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  bgsk_mc_combined,
  here::here("data/results_NO_BGSK_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

CRediT statement

* Conceptualization: [EF, KOS]
* Methodology: [EF, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]

MeRIT attribution

This part supplements the CRediT statement with more granularity for method section.

Study design
The study was designed by [AB, CD].

Data collection / fieldwork
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

Laboratory procedures
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

Data processing
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

Statistical analysis
All analyses were conducted by [CD] using [software], with verification by [AB].

Visualization
Figures were produced by [KL], with input from [CD].

Reproducibility
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info
```{r sessionInfo}
#| echo: false
sessionInfo()
```
:::
