---
title: "Biologically old forest as share of forest area"
subtitle: |
  [NO_BOFA_001]
indicatorID: 
  - NO_BOFA_001
indicatorName: 
  - Biologically old forest (NI, 0/1)
folderName: NO_BOFA_001
country: Norway
continent: Europe
ECT: C1 - Landscape and Seascape Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "000.001"
VersionComment: Initial version
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BOFA_001
Normalised: Yes
SpatialAggregationPathway: Scale - Transform - Aggregate (area-weighted mean of scaled county values) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.
date: February 12, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
bibliography: references.bib
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
---

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull <- dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
if (file.exists(here::here("R/functions.R"))) {
  source(here::here("R/functions.R"))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> Recommended citation: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class",
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added",
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>
  kable()
```
:::

::: {.callout-tip collapse="true"}
## Log
- 12 Feb. 2026 - Initial qmd from ecRxiv template; ECT C1; county X100 table from NI; NI column name pending confirmation
:::

<hr />

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->

2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->

3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
An indicator value of 1 signifies ...
Values decreasing toward zero reflect...

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->

## 2. About the underlying data

The old forest share of area is recorded for each NFI inventory plot.
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the old forest share of area for NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal resolution and extent

Data are available from 2010, when the the old forest share of area was first recorded based on current methods.
Annual estimates are possible due to representative sampling every year.
For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010.
The data are representative for all productive forest land of Norway.
Spatial resolution should be limited to the resolution applied to reference values, that is counties or sub-counties for scaling and regions for aggregated reporting.

### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

### 2.3 Original units

Each forest inventory plot carries a binary presence variable `OLD_FOREST_BINARY_01` (0/1): whether biologically old forest occurs on that plot in the Nature Index sense, and `au_areal` is the inventoried forest area (ha) on the plot.
At county (fylke), region, or national level, the non-scaled share of old forest is not the simple mean of the binary across plots; it is the ratio of forest areas: total forest area on plots where `OLD_FOREST_BINARY_01 = 1`, divided by total forest area on all forest plots in the unit after restricting to forest plots and the relevant reporting period.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2025-10-20, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

### 2.5 Additional comments about the dataset

<!--# Optional. Text here -->


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->

### 3.1.1 Other condition typologies

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->

### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->


### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->

### 3.5 Spatial aggregation

The spatial aggregation pathway is Scale - Aggregate - Truncate.
At the plot level, the inventory records whether biologically old forest is present or absent on each forest plot, and each plot has an associated forest area in hectares. When areas are combined upward, only the forest area on plots where old forest is present contributes to the old-forest side of the ratio; every forest plot in the unit still contributes its forest area to the overall forest total.
For each reporting period, non-scaled values for county (fylke), region, and Norway are obtained by pooling all forest plots in the unit: the total forest area on plots where old forest is present, divided by the total forest area on all forest plots in that unit. The result is the share of inventoried forest area that falls on plots classified as having old forest.
This is done for counties or sub-counties, regions, and at national level.
Scaling is applied only after these non-scaled county or sub-county aggregates have been calculated.
Each county or sub-county is scaled against its own lower reference level X0 (0% cover) and upper reference level X100 using a linear scaling function; scaled values above 1 are truncated to 1.
Regional and national scaled point estimates are not obtained by scaling a single aggregated non-scaled value with one reference level.
They are area-weighted means of the already scaled county or sub-county values, using each county or sub-county's represented area as weight.
Non-scaled regional and national point estimates are still calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels

The lower reference level X0 is set to 0% old forest share of area and the upper reference level X100 is set separately for each county or sub-county.
X0 denotes the worst possible condition and X100 denotes the best possible condition on the original share of area scale.
Between X0 and X100, indicator values are scaled linearly to a 0-1 indicator scale.
Scaled values above 1 are truncated to 1 when old forest share of area exceeds the county-specific X100.
The county and sub-county X100 values used in the analyses are listed in @sec-bofa-reference-values.
X0 is 0 for all units.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or sub-county resolution.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties

### 5.1 General uncertainties

<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in old forest share of area among forest plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling forest plots with replacement within the relevant geographic unit and recomputing the forest-area share.
This is done separately for counties or sub-counties, regions, and the national level.

For scaled county or sub-county values, the bootstrap mean, median, standard error, confidence interval, and quartiles from the non-scaled bootstrap are transformed with the same county-specific linear scaling function used for the point estimate.
This is an approximation when many bootstrap replicates exceed X100, because truncation is not applied inside each bootstrap replicate before summarising the scaled distribution.
For scaled regional and national values, uncertainty is estimated by bootstrapping plots, calculating county or sub-county non-scaled values in each replicate, scaling each county or sub-county with its own reference levels, and then taking the area-weighted mean of the scaled county or sub-county values within each region or for the whole country.
The resulting bootstrap distribution is summarised with mean, median, standard error, confidence interval, and quartiles.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

R package versions for this indicator are managed with `renv`.
The project lockfile `renv.lock` records the package versions used to render the document.
After cloning the repository, run `renv::restore()` in the project root before rendering.

The old forest area share is built from each forest inventory plot: binary presence `OLD_FOREST_BINARY_01` and forest area `au_areal` (ha), then aggregated to counties (fylke), regions, and Norway.
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used. Aggregated estimates use sums of forest area within each reporting unit so that each plot contributes in proportion to its inventoried forest area.
As the NFI grid varies in density for forest below/above the coniferous forest line and in Finnmark, each plot type represents a different part of all forest land.
These variations in representativity must be accounted for in aggregated estimates. 
Normally, data from NFI will contain an area variable such as `au_areal` to be used when aggregating to spatial units larger than individual plots. 

```{r load-data}
#| eval: true
#| code-summary: "Load data"

# Load main data file
lsk_data_raw <- read_csv(here("data/LSK_data_naturregnskap.csv"), 
                     show_col_types = FALSE,
                     locale = locale(encoding = "UTF-8")) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, e.g., regions.

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.
We retain both productive and unproductive forest plots (`arealtype` 1 and 12), and then create the derived variables used in subsequent aggregation steps.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| eval: false
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

The LSK extract must contain a binary column (0/1) for old forest on each forest plot; in the code below that source column is referenced by name through `OLD_FOREST_BINARY_01_COL` (replace the string once the export is confirmed).
`OLD_FOREST_BINARY_01` is the plot-level binary on forest area (`au_areal`, ha): 1 = old forest present on the plot, 0 = not. For aggregation to fylke (and then region / Norway), each unit’s non-scaled share is forest area on plots with old forest divided by total forest area in the unit: \(\sum (\texttt{OLD\_FOREST\_BINARY\_01} \times \texttt{au\_areal}) / \sum \texttt{au\_areal}\). That is implemented as `indicator_continuous_weighted = OLD_FOREST_BINARY_01 * au_areal` so the existing `calculate_period_indicators()` / `calculate_indicator()` logic (sums of weighted area over sums of `au_areal`) reproduces this ratio. Bootstraps resample plots within the unit and recompute the same ratio.
If the NI binary means 'part of the plot' rather than 'entire inventoried forest polygon', attributing the full `au_areal` to the numerator when `OLD_FOREST_BINARY_01 = 1` is an approximation; then a finer-area column would be needed instead of `au_areal` for the numerator.

```{r bofa-create-derived-variables}
#| eval: false
#| cache: false
#| code-summary: "Create derived variables"

# Replace with the actual LSK column name (see data dictionary / names(lsk_data_raw)).
OLD_FOREST_BINARY_01_COL <- "REPLACE_WITH_CONFIRMED_LSK_COLUMN_NAME"

lsk_data <- lsk_data_raw |>
  dplyr::filter(arealtype %in% c(1, 12)) |>
  dplyr::filter(!is.na(.data[[OLD_FOREST_BINARY_01_COL]])) |>
  dplyr::mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),

    # Plot-level binary; county/region/national share = sum(OLD_FOREST_BINARY_01 * au_areal) / sum(au_areal)
    OLD_FOREST_BINARY_01 = as.numeric(.data[[OLD_FOREST_BINARY_01_COL]]),
    indicator_continuous_weighted = OLD_FOREST_BINARY_01 * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r bofa-select-downstream-fields}
#| eval: false
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    OLD_FOREST_BINARY_01,
    indicator_continuous_weighted
  )

```

The indicator is scaled against county-specific reference levels in @sec-bofa-reference-values.


### 9.2 Calculate non-scaled area-weighted indicator values

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the share of inventoried forest area on plots with old forest present: sum of `indicator_continuous_weighted` (= `OLD_FOREST_BINARY_01 * au_areal`) divided by sum of `au_areal` over all forest plots in that unit (implemented via the shared `calculate_indicator()` helper).
County or sub-county estimates use the plot-level `fylke_name` field; regional estimates use `region_code`; the national estimate uses all plots in the period.

```{r bofa-calculate-all-periods}
#| eval: false
#| code-summary: "Calculate non-scaled values"

bofa_results <- purrr::imap(
  reporting_periods,
  ~ calculate_period_indicators(lsk_data, .y, .x)
) |>
  purrr::compact()

bofa_results_county_all_periods   <- purrr::map_dfr(bofa_results, "county")
bofa_results_region_all_periods   <- purrr::map_dfr(bofa_results, "region")
bofa_results_national_all_periods <- purrr::map_dfr(bofa_results, "national")

```

### 9.3 Bootstrap uncertainty estimation for non-scaled indicator values

Uncertainty in the non-scaled indicator estimates is quantified using bootstrapping.
For each 5-year period and geographic unit, NFI plots in that unit are resampled with replacement and the area-weighted non-scaled indicator value is recalculated in each replicate.
This is done separately for counties or sub-counties, regions, and the national level.
The resulting bootstrap distributions are summarised with mean, median, standard error, 95% confidence interval, and quartiles.
These non-scaled uncertainty measures are used directly for reporting and, at county or sub-county level, as input to the scaled uncertainty calculations in section 9.4.

```{r bofa-bootstrap-all-periods}
#| eval: false
#| code-summary: "Bootstrap uncertainty"

bofa_bootstrap <- bofa_results |>
  purrr::imap(
    ~ list(
      period = .y,
      data = .x$data,
      county = .x$county,
      region = .x$region,
      national = .x$national,
      national_boot = bootstrap_indicator(.x$data, n_bootstrap = 1000),
      county_boot = purrr::map(
        split(.x$data, .x$data$fylke_name),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
      ),
      region_boot = purrr::map(
        split(.x$data, .x$data$region_code),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
      )
    )
  )

bofa_results_county_non_scaled_uncertainty <- purrr::map_dfr(
  bofa_bootstrap,
  ~ {
    county_summary <- purrr::imap_dfr(
      .x$county_boot,
      function(boot, county_name) {
        dplyr::tibble(
          fylke_name = county_name,
          mean = boot$mean,
          indicator_value = boot$median,
          se = boot$se,
          ci_lower = boot$ci_lower,
          ci_upper = boot$ci_upper,
          q1 = boot$q1,
          median = boot$median,
          q3 = boot$q3
        )
      }
    )

    .x$county |>
      dplyr::select(fylke_name, county_part, period, total_area, n_plots) |>
      dplyr::left_join(county_summary, by = "fylke_name")
  }
)

bofa_results_region_non_scaled_uncertainty <- purrr::map_dfr(
  bofa_bootstrap,
  ~ {
    region_summary <- purrr::imap_dfr(
      .x$region_boot,
      function(boot, region_code) {
        dplyr::tibble(
          region_code = region_code,
          mean = boot$mean,
          indicator_value = boot$median,
          se = boot$se,
          ci_lower = boot$ci_lower,
          ci_upper = boot$ci_upper,
          q1 = boot$q1,
          median = boot$median,
          q3 = boot$q3
        )
      }
    )

    .x$region |>
      dplyr::select(region_code, period, total_area, n_plots) |>
      dplyr::left_join(region_summary, by = "region_code")
  }
)

bofa_results_national_non_scaled_uncertainty <- purrr::map_dfr(
  bofa_bootstrap,
  ~ .x$national |>
    dplyr::mutate(
      mean = .x$national_boot$mean,
      indicator_value = .x$national_boot$median,
      se = .x$national_boot$se,
      ci_lower = .x$national_boot$ci_lower,
      ci_upper = .x$national_boot$ci_upper,
      q1 = .x$national_boot$q1,
      median = .x$national_boot$median,
      q3 = .x$national_boot$q3
    )
)

```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-bofa-reference-values}

County or sub-county reference levels are listed below.
For each county or sub-county, the non-scaled point estimate is scaled with `scale_indicator()` using X0 = 0 and the matching X100.
The function applies linear scaling between X0 and X100 and truncates values above 1 when old forest share of area exceeds X100.
Regional and national scaled point estimates are area-weighted means of the scaled county or sub-county values, weighted by each unit's represented area.
Non-scaled regional and national point estimates are also retained for reporting and truncation summaries.
For scaled county or sub-county uncertainty, the non-scaled bootstrap summaries from section 9.3 are transformed with the same county-specific scaling function.
For scaled regional and national uncertainty, each bootstrap replicate resamples plots, calculates county or sub-county non-scaled values, scales each county or sub-county with its own reference levels, and then aggregates the scaled county or sub-county values within each region or for the whole country before the bootstrap distribution is summarised.

```{r bofa-reference-values}
#| eval: false
#| code-summary: "Reference values by county or sub-county"

bofa_county_x100_pct <- tibble::tribble(
  ~county_part, ~x100_pct,
  "Øs", 50, "OA", 50, "HeS", 50, "HeN", 45, "OpS", 50, "OpN", 45,
  "BuS", 50, "BuN", 45, "Ve", 50, "TeS", 50, "TeN", 45, "AA", 45, "VA", 45,
  "Ro", 40, "Ho", 40, "SF", 40, "MR", 40, "ST", 45, "NT", 45,
  "NoS", 45, "NoN", 30, "Tr", 30, "Fi", 30
) |>
  dplyr::mutate(x100 = x100_pct / 100)

x0 <- 0
county_reference <- bofa_county_x100_pct |>
  dplyr::select(county_part, x100)

knitr::kable(bofa_county_x100_pct, caption = "X100 (%) by sub-county for NO_BOFA_001")

```
```{r bofa-scale-and-aggregate}
#| eval: false
#| code-summary: "Scale and aggregate"

county_scaled_all_periods <- bofa_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = scale_indicator(indicator_value, x0, x100)
  )

county_scaled_with_uncertainty <- bofa_results_county_non_scaled_uncertainty |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_mean = scale_indicator(mean, x0, x100),
    scaled_median = scale_indicator(indicator_value, x0, x100),
    scaled_ci_lower = scale_indicator(ci_lower, x0, x100),
    scaled_ci_upper = scale_indicator(ci_upper, x0, x100),
    scaled_q1 = scale_indicator(q1, x0, x100),
    scaled_q3 = scale_indicator(q3, x0, x100)
  )

region_non_scaled_agg_all_periods <- bofa_results_region_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

region_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_non_scaled_agg_all_periods <- bofa_results_national_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

national_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National")

truncation_summary <- dplyr::bind_rows(
  county_scaled_all_periods |>
    dplyr::left_join(
      bofa_county_x100_pct |> dplyr::select(county_part, reference_value_pct = x100_pct),
      by = "county_part"
    ) |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      region_non_scaled_agg_all_periods |>
        dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      national_non_scaled_agg_all_periods |>
        dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county or sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the scaled aggregate equals 1 after area-weighted aggregation of scaled county or sub-county values.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r bofa-report-truncation}
#| eval: false
#| code-summary: "Report truncation"

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(
      period,
      level,
      location,
      reference_value_pct,
      original_value_pct
    ) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original Value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original Value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their reference and required truncation.\n")
}

```

The bootstrap workflow above is used to estimate uncertainty for scaled regional and national values.
County scaled uncertainty is obtained by applying the county-specific scaling function to the non-scaled bootstrap summaries from section 9.3.

```{r bofa-bootstrap-scaled-uncertainty}
#| eval: false
#| code-summary: "Bootstrap uncertainty for scaled values"

region_scaled_uncertainty <- purrr::imap_dfr(
  bofa_results,
  function(period_result, period_name) {
    period_data <- period_result$data

    if (nrow(period_data) == 0) {
      return(NULL)
    }

    bootstrap_region_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    ) |>
      dplyr::rename(
        boot_mean = mean,
        boot_median = median
      ) |>
      dplyr::mutate(period = period_name)
  }
)

national_scaled_uncertainty <- purrr::imap_dfr(
  bofa_results,
  function(period_result, period_name) {
    period_data <- period_result$data

    if (nrow(period_data) == 0) {
      return(NULL)
    }

    boot_result <- bootstrap_national_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    )

    dplyr::tibble(
      period = period_name,
      region = "National",
      boot_mean = boot_result$mean,
      boot_median = boot_result$median,
      se = boot_result$se,
      ci_lower = boot_result$ci_lower,
      ci_upper = boot_result$ci_upper,
      q1 = boot_result$q1,
      q3 = boot_result$q3
    )
  }
)

region_scaled_agg_with_uncertainty <- region_scaled_agg_all_periods |>
  dplyr::left_join(region_scaled_uncertainty, by = c("period", "region_code")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

national_scaled_agg_with_uncertainty <- national_scaled_agg_all_periods |>
  dplyr::left_join(national_scaled_uncertainty, by = c("period", "region")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

```

## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
All scaled values are on a 0-1 scale, where 0 corresponds to X0 (0% old forest share of area) and 1 corresponds to the county-specific X100 for counties or sub-counties, or to an area-weighted aggregate of scaled county or sub-county values for regions and the whole country.
County or sub-county tables report bootstrap means and medians, standard errors, confidence intervals, and quartiles derived from the non-scaled bootstrap and transformed with the sub-county-specific scaling function.
Regional and national tables report bootstrap summaries obtained from the scale-then-aggregate bootstrap workflow in section 9.4.

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-bofa-county-results}
#| eval: false
#| tbl-cap: 'County or sub-county indicator values for biologically old forest share of area by reporting period. Values are scaled to 0-1 using county-specific X0 and X100. Uncertainty is obtained by applying the county-specific scaling function to the non-scaled bootstrap summaries from section 9.3.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code)
  ) |>
  dplyr::select(period, fylke_name, region_name, scaled_mean, scaled_median, se, scaled_ci_lower, scaled_ci_upper,
    scaled_q1, scaled_q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    scaled_ci_lower = round(scaled_ci_lower, 3),
    scaled_ci_upper = round(scaled_ci_upper, 3),
    scaled_q1 = round(scaled_q1, 3),
    scaled_q3 = round(scaled_q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = scaled_ci_lower,
    `CI Upper` = scaled_ci_upper,
    `Q1` = scaled_q1,
    `Q3` = scaled_q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  DT::datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, lengthMenu = c(10, 25, 50, 100))
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-bofa-region-results}
#| eval: false
#| tbl-cap: 'Regional indicator values for biologically old forest share of area by reporting period. Scaled point estimates are area-weighted means of scaled county or sub-county values. Uncertainty is estimated by bootstrapping plots, scaling counties or sub-counties in each replicate, and aggregating the scaled values within each region.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::select(period, region_name, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    Region = region_name,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-bofa-national-results}
#| eval: false
#| tbl-cap: 'National indicator values for biologically old forest share of area by reporting period. Scaled point estimates are area-weighted means of scaled county or sub-county values. Uncertainty is estimated by bootstrapping plots, scaling counties or sub-counties in each replicate, and aggregating the scaled values nationally.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::select(period, region, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    `Spatial Level` = region,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## 11. Export file

```{r export}
#| eval: false
```

## 12. Author contributions

CRediT statement

* Conceptualization: [EF, KOS]
* Methodology: [EF, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]

MeRIT attribution

This part supplements the CRediT statement with more granularity for method section.

Study design
The study was designed by [AB, CD].

Data collection / fieldwork
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

Laboratory procedures
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

Data processing
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

Statistical analysis
All analyses were conducted by [CD] using [software], with verification by [AB].

Visualization
Figures were produced by [KL], with input from [CD].

Reproducibility
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info
```{r sessionInfo}
#| echo: false
sessionInfo()
```
:::
