---
title: "Brown bear abundance"
subtitle: |
  [NO_BBAB_001]
indicatorID: 
  - NO_BBAB_001
indicatorName: 
  - Brown bear abundance [Brunbjørnbestand]
verbatimName: 
  - Brunbjørnbestand
folderName: NO_BBAB_001
country: Norway
continent: Europe
ECT: B3. Functional state characteristics
Realm: 
  - Terrestrial (T) # e.g. Freshwater (F)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First version on ecRxiv, but this indicator is already published and described in the Norwegian Nature Index
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BBAB_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (population model, large-carnivore management region) - [Scale - Truncate] - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud H. Halbritter
    email: aud.halbritter@uib.no
    affiliations:
      - id: myID
        name: University of Bergen
AuthorList: Halbritter, A. H.
date: May 15, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull
library(NIcalc)

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
if (file.exists(here::here("R/spatial_regions.R"))) {
  source(here::here("R/spatial_regions.R"))
}
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- May 15, 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
The brown bear (_Ursus arctos_) is a large omnivorous mammal found in forested and mountainous areas of Norway, mainly in the east, Trøndelag, and north.
As a top predator and scavenger it plays a functional role in boreal and montane forest ecosystems.
Population trends reflect habitat quality, connectivity, and anthropogenic pressures such as habitat loss, traffic mortality, conflict management, and licensed hunting.
Norway has long-running, DNA-based monitoring of brown bears, which provides a relatively solid basis for using the species as an indicator of ecosystem condition in relevant ecosystems.
The indicator is part of the Norwegian Nature Index and is documented on [Naturindeksen](https://naturindeks.no/Indicators/brunbjoern).

2. **Methods**:
Estimated brown bear abundance (or an equivalent population index) is derived from the national DNA-monitoring programme and associated statistical models, not from harvest statistics.
Non-lethal genetic sampling (scats, hair, and other material) is used to identify individuals; population size is estimated and reported for large-carnivore management regions (_forvaltningsregioner for rovvilt_, regions 5–8 in Norway).
Values are normalised to a 0–1 indicator scale using reference levels (_X~100~_) defined for those management regions, with _X~0~_ representing the worst condition.
For Naturindeks reporting, scaled values are aggregated to five geographic regions and to the national level using area weights.

3. **Interpretation**:
An indicator value of 1 means that the brown bear population in the reporting unit is at least as large as expected under the reference condition.
Values decreasing toward zero indicate that the population index is further below the reference, reflecting negative change relative to the reference state.

4. **Results and development needs**:
The indicator is operational in Naturindeksen and can be used for reporting ecosystem condition where brown bear is a selected species indicator.

## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: Brunbjørn (_Ursus arctos_) er et stort, altetende pattedyr som finnes i skog- og fjellområder i Norge, særlig i øst, Trøndelag og nord. Som rovdyr og åtseleter har arten en viktig funksjon i boreale og fjellnære skogøkosystemer. Bestandsutviklingen speiler habitatkvalitet, sammenheng i landskapet og menneskelig påvirkning, blant annet tap og fragmentering av habitat, trafikkdødelighet, skadeforvaltning og lisensjakt. Norge har langvarig, DNA-basert overvåking av brunbjørn [@broseth2024_bb], som gir et relativt godt grunnlag for å bruke arten som indikator på økologisk tilstand. Indikatoren inngår i Naturindeksen og er beskrevet på [Naturindeksen](https://naturindeks.no/Indicators/brunbjoern).

2. **Metoder**: Estimert brunbjørnbestand (eller et tilsvarende bestandsindeks) bygger på det nasjonale DNA-overvåkingsprogrammet og tilhørende statistiske modeller, ikke på jaktstatistikk. Genetisk materiale (ekskrement, hår m.m.) brukes til å identifisere individer; bestandsstørrelse rapporteres for forvaltningsregioner for rovvilt. Verdiene normaliseres til en indikatorskala 0–1 med referanseverdier (_X~100~_) knyttet til disse regionene, og _X~0~_ som verste tilstand. For Naturindeks aggregeres skalerte verdier videre til fem geografiske regioner og nasjonalt med arealvekter.

3. **Tolkning**: En indikatorverdi på 1 betyr at brunbjørnbestanden i rapporteringsenheten er minst like stor som forventet under referansetilstanden. Verdier som nærmer seg null, viser at bestandsindekset ligger lenger under referansen og at tilstanden har utviklet seg negativt i forhold til referansen.

4. **Resultater og utviklingsbehov**: Indikatoren er operativ i Naturindeksen.


## 2. About the underlying data

The primary data source for this indicator is the national programme for DNA-based monitoring of brown bears in Norway, coordinated under the large carnivore monitoring framework and reported annually (e.g. [@broseth2024_bb]).
For Naturindeks reporting, prepared indicator objects (population estimates or indices with uncertainty, plus reference values) are obtained from the Naturindeks database or equivalent exports.

### 2.1 Spatial and temporal resolution and extent

Indicator values are available for the assessment years in _Naturindeksen_: 1990, 2000, 2010, 2014, 2019, and 2024.
To produce indicator estimates for these years, input from the DNA-monitoring programme is combined over a short window (typically two to three years), as specified in the Naturindeks indicator protocol.

The brown bear is tied to large-carnivore management regions (_forvaltningsregioner for rovvilt_).
Regional and national monitoring and management use regions 5–8 (Hedmark; Møre og Romsdal and Trøndelag; Nordland; Troms and Finnmark; see [Rovdata](https://rovdata.no/Brunbj%C3%B8rn/Bestandsstatus.aspx)).
In Naturindeksen, indicator and reference values are stored at this management-region scale (as IC units in the database).

### 2.2 Sampling design

Monitoring relies on opportunistic and targeted collection of DNA samples from the field (hunters, managers, researchers, and the public), followed by genetic analysis to assign samples to individuals and sexes.
Coverage is extensive but not a classical random plot design: detection depends on sample submission, field effort, and bear presence. The method is designed for capture–recapture–style inference on identified individuals rather than complete census. 
Biases can arise from uneven sampling intensity between areas and years, misidentified samples, and incomplete detection of individuals.
Full methodological detail is given on [Naturindeksen](https://naturindeks.no/Indicators/brunbjoern) and in the annual monitoring reports [@broseth2024_bb].

### 2.3 Original units

At the monitoring stage, the core quantities are numbers of genetically identified individuals (and derived abundance or density estimates) from DNA profiles, reported nationally and by forvaltningsregion for rovvilt.
For Naturindeks, these are translated into indicator variables (expected values with uncertainty distributions) stored in the NI database before normalisation to 0–1.

### 2.4 Instructions for citing, using and accessing data

Raw genetic monitoring results and annual reports should be cited from the NINA monitoring publications (e.g. [@broseth2024_bb]).
Indicator point estimates and uncertainty quantiles for Naturindeks are available through the [Naturindeks download pages](https://naturindeks.no/DownloadData) and the indicator data repository [@nater_ni2025_2025].
Methodological documentation for this species is linked from [Naturindeksen – brunbjørn](https://naturindeks.no/Indicators/brunbjoern).

<!-- ### 2.5 Additional comments about the dataset -->

<!--# Optional. Text here -->

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

B3 – Functional state characteristics. Brown bear is assigned to functional composition because population status reflects the presence and role of a large carnivore in the ecosystem (trophic regulation, carcass subsidy, and related ecological processes), not only the occurrence of a single species name in a species list.

<!-- 
### 3.1.1 Other condition typologies -->

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

The ecosystem characteristic represented is a functionally intact large-carnivore component of boreal and montane forest ecosystems: an ecosystem where brown bear occurs at abundances consistent with the reference condition and can fulfil its ecological roles without being depressed far below sustainable levels.

### 3.3 Collinearities with other indicators

The indicator may covary with other large carnivore or mammal indicators in the same regions (e.g. wolf, wolverine, lynx) and with indicators linked to similar forest and mountain habitats.
Including several tightly correlated predator indicators in the same composite index without accounting for dependence could overweight the predator guild signal.

### 3.4 Impact factors

Important anthropogenic pressures include habitat alteration and fragmentation, infrastructure and traffic mortality, human–wildlife conflict and removals, licensed hunting and management actions, and climate-related changes affecting food resources and denning. 
Natural variation in reproduction and survival also affects the population index.

### 3.5 Spatial aggregation

Following the metadata pathway Aggregate (population model, large-carnivore management region) – [Scale – Truncate] – Aggregate (national level): population estimates or indices from the monitoring/model workflow are available per forvaltningsregion for rovvilt (IC unit in Naturindeksen). Each unit is normalised against _X~0~_ and _X~100~_ for that region (linear scaling with truncation to [0, 1]).

Regional (east, south, west, central, north) and national indicator values are then obtained as area-weighted combinations of the scaled management-region values, using weights from the Naturindeks spatial framework.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition for brown bear in Naturindeksen is a boreal/montane forest ecosystem with minimal human impact and a climate similar to the reference period used in the Norwegian Nature Index framework [@nybo_fagsystem_2017].

### 4.2 Reference levels

_X~100~_ is set per large-carnivore management region (IC unit in Naturindeksen), using suitable area and density within each _rovviltregion_ and expert assessment: it represents the expected brown bear population (or equivalent index) under reference conditions in that unit.

_X~0~_ is defined as no brown bears in the unit (functional absence).


#### 4.2.1 Spatial resolution and validity

<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->

Reference levels and indicator values apply at the scale of forvaltningsregioner for rovvilt (IC units).
Aggregated results for the five Naturindeks geographic regions and for Norway are summaries of those units with area weights.

## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

### 5.2 Reported uncertainties
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->



## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

### 7.1 Naturindeks

Run the download chunk once when you have API credentials. Credentials are read from `R/_pw.R` (not committed to git). The export is saved under `data/` (also gitignored) so the document can be rendered without calling the API on every build.

```{r bbab-import-api}
#| eval: false
#| code-summary: "Download brown bear data from Naturindeks API"
source(here::here("R", "_pw.R"))

bbabImport <- NIcalc::importDatasetApi(
  username = myUser,
  password = myPwd,
  eco = NULL,
  indic = c("Brunbjørn"),
  year = c("1990", "2000", "2010", "2014", "2019", "2024")
)

saveRDS(bbabImport, here::here("data/bbabImport.rds"))
```

```{r bbab-load-import}
#| include: false
#| cache: false
#| code-summary: "Load saved Naturindeks import"

bbabImport <- readRDS(here::here("data/bbabImport.rds"))
```


## 8. Spatial units

The primary reporting units are forvaltningsregioner for rovvilt (large-carnivore management regions 5–8), which function as ecosystem-relevant IC units for brown bear.
Aggregated results for the five Naturindeks geographic regions (NI units: east, south, west, central, north) and for mainland Norway are built from those management-region values with area weights.
Indicator values should be read at management-region scale for detail, and at geographic-region or national scale for overview.

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

Not relevant, as this is the work of a single author. I would, however, like to acknowledge the work of those that created the indicator on _Naturindeks_ [@nater_naturindeks_2025].

::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
