---
title: "Bilberry cover by area"
subtitle: |
  NO_BBCA_001  
indicatorID: 
  - NO_BBCA_001
indicatorName: 
  - NO_BBCA
folderName: NO_BBCA
country: Norway
continent: Europe
ECT: B1 Biotic compositional characteristic
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_BBCA_001
Normalised: Yes
SpatialAggregationPathway: Scale - Aggregate - Truncate # e.g. Scale - Transform - Aggregate (length-weighted arithmetic mean) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen 
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.  
date: May 7, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 07 May. 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->
2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->
3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
An indicator value of 1 signifies ...
Values decreasing toward zero reflect...

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->



## 2. About the underlying data

The bilberry cover by area is recorded for each NFI inventory plot. 
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the billberry cover for NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal resolution and extent

Data are available from 2010, when the bilberry cover by area was first recorded based on current methods. Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. The data are representative for all productiver forest land of Norway. Spatial resolution should be limited to the resolution applied to reference values, that is counties or part county for scaling and regions for aggregated reporting.

### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

### 2.3 Original units

The indicator is specified as the percentage (%) of the area covered by bilberry on the NFI plot.
In the analyses, this value is converted to a proportion before area weighting, aggregation, and scaling.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2025-10-20, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

### 2.5 Additional comments about the dataset

<!--# Optional. Text here -->

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->

### 3.1.1 Other condition typologies

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->

### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->


### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->

### 3.5 Spatial aggregation

The spatial aggregation pathway is Scale - Aggregate - Truncate.
For each NFI plot we first convert bilberry cover from percent to proportion and create `indicator_continuous_weighted = indicator_cover * au_areal`.
Non-scaled indicator values for each reporting period are calculated separately for each geographic unit as area-weighted arithmetic means of all NFI plots within that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
This is done for counties or part county, regions, and at national level.
Scaling is applied only after these non-scaled county or county-part aggregates have been calculated.
Each county or part county is scaled against its own lower reference level X0 (0% cover) and upper reference level X100 using a linear scaling function; scaled values above 1 are truncated to 1.
Regional and national scaled point estimates are not obtained by scaling a single aggregated non-scaled value with one reference level.
They are area-weighted means of the already scaled county or county-part values, using each county or county part's represented area as weight.
Non-scaled regional and national point estimates are still calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->

### 4.2 Reference levels

The lower reference level X0 is set to 0% bilberry cover and the upper reference level X100 is set separately for each county or part county.
X0 denotes the worst possible condition and X100 denotes the best possible condition on the original cover scale.
Between X0 and X100, indicator values are scaled linearly to a 0-1 indicator scale.
Scaled values above 1 are truncated to 1 when bilberry cover exceeds the county-specific X100.
The county and county-part X100 values used in the analyses are listed in @sec-bbca-reference-values.
X0 is 0 for all units.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or county-part resolution, not as one national value.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or county-part unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or county-part values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or county-part reference levels allow.

## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in bilberry cover among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within the relevant geographic unit and recalculating the area-weighted mean.
This is done separately for counties or part county, regions, and the national level.
For scaled county or part county values, the bootstrap mean, median, standard error, confidence interval, and quartiles from the non-scaled bootstrap are transformed with the same county-specific linear scaling function used for the point estimate.
This is an approximation when many bootstrap replicates exceed X100, because truncation is not applied inside each bootstrap replicate before summarising the scaled distribution.
For scaled regional and national values, uncertainty is estimated by bootstrapping plots, calculating county or county-part non-scaled values in each replicate, scaling each county or county part with its own reference levels, and then taking the area-weighted mean of the scaled county or county-part values within each region or for the whole country.
The resulting bootstrap distribution is summarised with mean, median, standard error, confidence interval, and quartiles.

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->
::: {#refs}
:::

## 7. Import and prepare datasets

R package versions for this indicator are managed with `renv`.
The project lockfile `renv.lock` records the package versions used to render the document.
After cloning the repository, run `renv::restore()` in the project root before rendering.

The bilberry cover by area is quantified for each NFI plot.
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used. Aggregated area-weighted data must be used for appropriate spatial units such as counties, regions or the whole country.
As the NFI plots vary in density for forest below/above the coniferous forest line and in Finnmark, each type of plot represents a different part of all forest land.
In addition, some NFI plots are divided in two parts in the data.
These variations in plot representativity must be accounted for in aggregated estimates. 
Normally, data from NFI will contain an area-weighting variable to be used in aggregating NFI data for spatial units larger than individual NFI plots. 

```{r load-data}
#| code-summary: "Load data"

# Load main data file
lsk_data_raw <- read_csv(here("data/LSK_data_naturregnskap.csv"), 
                     show_col_types = FALSE,
                     locale = locale(encoding = "UTF-8")) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, e.g., regions.

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.
We retain both productive and unproductive forest plots (`arealtype` 1 and 12), and then create the derived variables used in subsequent aggregation steps.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

Next we filter the raw dataset to include only the two relevant forest area types, remove records missing in the `blaabaer_gjsn` variable, standardise the geography fields.
NFI bilberry cover is recorded on a 0-100 percent scale and is converted to a proportion before aggregation and scaling.
Area-weighted values are created using `au_areal`.

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    # productive and unproductive forest plots
    arealtype %in% c(1, 12)
  ) |>
  # remove records with missing data on bilberry cover
  filter(!is.na(blaabaer_gjsn)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),

    indicator_cover = blaabaer_gjsn / 100,
    indicator_continuous_weighted = indicator_cover * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    blaabaer_gjsn,
    indicator_continuous_weighted
  )

```

The indicator is scaled against county-specific reference levels in section 9.4.


### 9.2 Calculate non-scaled area-weighted indicator values

For each reporting period, we calculate non-scaled indicator values separately for counties or county parts, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
County or county-part estimates use the plot-level `fylke_name` field; regional estimates use `region_code`; the national estimate uses all plots in the period.

```{r calculate-bbca-all-periods}
#| code-summary: "Calculate non-scaled values"

 bbca_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 bbca_results_county_all_periods   <- purrr::map_dfr(bbca_results, "county")
 bbca_results_region_all_periods   <- purrr::map_dfr(bbca_results, "region")
 bbca_results_national_all_periods <- purrr::map_dfr(bbca_results, "national")

```

### 9.3 Bootstrap uncertainty estimation for non-scaled indicator values

Uncertainty in the non-scaled indicator estimates is quantified using bootstrapping.
For each 5-year period and geographic unit, NFI plots in that unit are resampled with replacement and the area-weighted non-scaled indicator value is recalculated in each replicate.
This is done separately for counties or part county, regions, and the national level.
The resulting bootstrap distributions are summarised with mean, median, standard error, 95% confidence interval, and quartiles.
These non-scaled uncertainty measures are used directly for reporting and, at county or county-part level, as input to the scaled uncertainty calculations in section 9.4.

```{r bootstrap-bbca-all-periods}
#| code-summary: "Bootstrap uncertainty"

# Prepare bootstrap objects per reporting period
 bbca_bootstrap <- bbca_results |>
   purrr::imap(
     ~ list(
       period = .y,
       data = .x$data,
       county = .x$county,
       region = .x$region,
       national = .x$national,
      national_boot = bootstrap_indicator(.x$data, n_bootstrap = 1000),
       county_boot = purrr::map(
         split(.x$data, .x$data$fylke_name),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
       ),
       region_boot = purrr::map(
         split(.x$data, .x$data$region_code),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
       )
     )
   )

# County-level summaries with uncertainty
 bbca_results_county_non_scaled_uncertainty <- purrr::map_dfr(
   bbca_bootstrap,
   ~ {
     county_summary <- purrr::imap_dfr(
       .x$county_boot,
       function(boot, county_name) {
         dplyr::tibble(
           fylke_name = county_name,
          mean = boot$mean,
           indicator_value = boot$median,
           se = boot$se,
           ci_lower = boot$ci_lower,
           ci_upper = boot$ci_upper,
           q1 = boot$q1,
           median = boot$median,
           q3 = boot$q3
         )
       }
     )

     .x$county |>
       dplyr::select(fylke_name, county_part, period, total_area, n_plots) |>
       dplyr::left_join(county_summary, by = "fylke_name")
   }
 )

# Region-level summaries with uncertainty
 bbca_results_region_non_scaled_uncertainty <- purrr::map_dfr(
   bbca_bootstrap,
   ~ {
     region_summary <- purrr::imap_dfr(
       .x$region_boot,
       function(boot, region_code) {
         dplyr::tibble(
           region_code = region_code,
          mean = boot$mean,
           indicator_value = boot$median,
           se = boot$se,
           ci_lower = boot$ci_lower,
           ci_upper = boot$ci_upper,
           q1 = boot$q1,
           median = boot$median,
           q3 = boot$q3
         )
       }
     )

     .x$region |>
       dplyr::select(region_code, period, total_area, n_plots) |>
       dplyr::left_join(region_summary, by = "region_code")
   }
 )

# National-level summaries with uncertainty
 bbca_results_national_non_scaled_uncertainty <- purrr::map_dfr(
   bbca_bootstrap,
   ~ .x$national |>
     dplyr::mutate(
      mean = .x$national_boot$mean,
       indicator_value = .x$national_boot$median,
       se = .x$national_boot$se,
       ci_lower = .x$national_boot$ci_lower,
       ci_upper = .x$national_boot$ci_upper,
       q1 = .x$national_boot$q1,
       median = .x$national_boot$median,
       q3 = .x$national_boot$q3
     )
 )

```


### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-bbca-reference-values}

County or part county reference levels are listed below.
For each county or part county, the non-scaled point estimate is scaled with `scale_indicator()` using X0 = 0 and the matching X100.
The function applies linear scaling between X0 and X100 and truncates values above 1 when bilberry cover exceeds X100.
Regional and national scaled point estimates are area-weighted means of the scaled county or part county values, weighted by each unit's represented area.
Non-scaled regional and national point estimates are also retained for reporting and truncation summaries.
For scaled county or part county uncertainty, the non-scaled bootstrap summaries from section 9.3 are transformed with the same county-specific scaling function.
For scaled regional and national uncertainty, each bootstrap replicate resamples plots, calculates county or part county non-scaled values, scales each county or part county with its own reference levels, and then aggregates the scaled county or part county values within each region or for the whole country before the bootstrap distribution is summarised.

```{r bbca-reference-values}
#| code-summary: "Reference values by county or county part"

# X100 reference values (% bilberry cover by area)
bbca_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuN", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_pct = c(
    12.5, 12.5, 12.5, 10, 12.5, 10, 12.5, 10, 12.5,
    12.5, 10, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5, 12.5,
    10, 5, 5, 2.5
  )
) |>
  dplyr::mutate(x100 = x100_pct / 100)

x0 <- 0
county_reference <- bbca_reference_values |>
  dplyr::select(county_part, x100)

```

```{r scale-and-aggregate}
#| code-summary: "Scale and aggregate"

# Scale county values (point estimates) for all periods
county_scaled_all_periods <- bbca_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = scale_indicator(indicator_value, x0, x100)
  )

# Uncertainty for county scaled: apply linear scaling to non-scaled CIs and quartiles
county_scaled_with_uncertainty <- bbca_results_county_non_scaled_uncertainty |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_mean = scale_indicator(mean, x0, x100),
    scaled_median = scale_indicator(indicator_value, x0, x100),
    scaled_ci_lower = scale_indicator(ci_lower, x0, x100),
    scaled_ci_upper = scale_indicator(ci_upper, x0, x100),
    scaled_q1 = scale_indicator(q1, x0, x100),
    scaled_q3 = scale_indicator(q3, x0, x100)
  )

# Aggregate non-scaled values to regional level (point estimates) for all periods
region_non_scaled_agg_all_periods <- bbca_results_region_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

# Aggregate scaled county values to regional level (point estimates) for all periods
region_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code))

# Aggregate non-scaled values to national level (point estimates) for all periods
national_non_scaled_agg_all_periods <- bbca_results_national_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

# Aggregate scaled county values to national level (point estimates) for all periods
national_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National")

# Check for truncation: count county values exceeding their county-specific reference
truncation_summary <- dplyr::bind_rows(
  county_scaled_all_periods |>
    dplyr::left_join(
      bbca_reference_values |> dplyr::select(county_part, reference_value_pct = x100_pct),
      by = "county_part"
    ) |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      region_non_scaled_agg_all_periods |>
        dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      national_non_scaled_agg_all_periods |>
        dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_pct,
      exceeds_threshold,
      original_value_pct
    )
)

# Summary of truncation by period and level
truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

# Values that were truncated (for reporting)
truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county or part county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the scaled aggregate equals 1 after area-weighted aggregation of scaled county or part county values.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r report-truncation}
#| code-summary: "Report truncation"

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(
      period,
      level,
      location,
      reference_value_pct,
      original_value_pct
    ) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original Value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original Value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their reference and required truncation.\n")
}

```

The bootstrap workflow above is used to estimate uncertainty for scaled regional and national values.
County scaled uncertainty is obtained by applying the county-specific scaling function to the non-scaled bootstrap summaries from section 9.3.

```{r bootstrap-scaled-uncertainty}
#| code-summary: "Bootstrap uncertainty for scaled values"

# Bootstrap uncertainty for scaled-then-aggregated regions per period
region_scaled_uncertainty <- purrr::imap_dfr(
  bbca_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    
    if (nrow(period_data) == 0) return(NULL)
    
    bootstrap_region_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    ) |>
      dplyr::rename(
        boot_mean = mean,
        boot_median = median
      ) |>
      dplyr::mutate(period = period_name)
  }
)

# Bootstrap uncertainty for scaled-then-aggregated national per period
national_scaled_uncertainty <- purrr::imap_dfr(
  bbca_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    
    if (nrow(period_data) == 0) return(NULL)
    
    boot_result <- bootstrap_national_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    )
    
    dplyr::tibble(
      period = period_name,
      region = "National",
      boot_mean = boot_result$mean,
      boot_median = boot_result$median,
      se = boot_result$se,
      ci_lower = boot_result$ci_lower,
      ci_upper = boot_result$ci_upper,
      q1 = boot_result$q1,
      q3 = boot_result$q3
    )
  }
)

# Attach uncertainty to scaled aggregated values and retain both bootstrap mean and median
region_scaled_agg_with_uncertainty <- region_scaled_agg_all_periods |>
  dplyr::left_join(region_scaled_uncertainty, by = c("period", "region_code")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

national_scaled_agg_with_uncertainty <- national_scaled_agg_all_periods |>
  dplyr::left_join(national_scaled_uncertainty, by = c("period", "region")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county or part county, regional, and national scales for each reporting period.
All scaled values are on a 0-1 scale, where 0 corresponds to X0 (0% bilberry cover) and 1 corresponds to the county-specific X100 for counties or part counties, or to an area-weighted aggregate of scaled county or part county values for regions and the whole country.
County or part county tables report bootstrap means and medians, standard errors, confidence intervals, and quartiles derived from the non-scaled bootstrap and transformed with the partcounty-specific scaling function.
Regional and national tables report bootstrap summaries obtained from the scale-then-aggregate bootstrap workflow in section 9.4.

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-county-results-table}
#| tbl-cap: 'County or county-part indicator values for bilberry cover by area by reporting period. Values are scaled to 0-1 using county-specific X0 and X100. Uncertainty is obtained by applying the county-specific scaling function to the non-scaled bootstrap summaries from section 9.3. The table includes bootstrap mean and median estimates, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code)
  ) |>
  dplyr::select(period, fylke_name, region_name, scaled_mean, scaled_median, se, scaled_ci_lower, scaled_ci_upper, 
         scaled_q1, scaled_q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    scaled_ci_lower = round(scaled_ci_lower, 3),
    scaled_ci_upper = round(scaled_ci_upper, 3),
    scaled_q1 = round(scaled_q1, 3),
    scaled_q3 = round(scaled_q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = scaled_ci_lower,
    `CI Upper` = scaled_ci_upper,
    `Q1` = scaled_q1,
    `Q3` = scaled_q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  DT::datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, lengthMenu = c(10, 25, 50, 100))
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-region-results-table}
#| tbl-cap: 'Regional indicator values for bilberry cover by area by reporting period. Scaled point estimates are area-weighted means of scaled county or county-part values. Uncertainty is estimated by bootstrapping plots, scaling counties or county parts in each replicate, and aggregating the scaled values within each region. The table includes the bootstrap mean and median, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::select(period, region_name, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    Region = region_name,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-national-results-table}
#| tbl-cap: 'National indicator values for bilberry cover by area by reporting period. Scaled point estimates are area-weighted means of scaled county or county-part values. Uncertainty is estimated by bootstrapping plots, scaling counties or county parts in each replicate, and aggregating the scaled values nationally. The table includes the bootstrap mean and median, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::select(period, region, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    `Spatial Level` = region,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
