---
title: "Large trees"
subtitle: |
  NO_DLTA_001
indicatorID: 
  - NO_DLTA_001
indicatorName: 
  - Large trees
verbatimName: 
  - Arealandel med forekomst av store trær
folderName: NO_DLTA_001
country: Norway
continent: Europe
ECT: B2 Structural state characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2024 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: ecTools workflow with X0 = 0% and county/sub-county-specific X100
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_DLTA_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county/sub-county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
  docx:
    embed-resources: true
execute: 
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Ulrika Jansson Asplund
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Mari Jönsson
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Ken Olaf Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
AuthorList: Halbritter, A. H., Jansson Asplund, U., Framstad, E., Jönsson, M., Snoksrud, O., and Storaunet, K. O.
date: June 9, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 08 Nov. 2025 - Original PR (NO_LTFA_001)
- 09 Jun. 2026 - NO_DLTA_001: ecTools workflow, national refs X0 = 0%, X100 are county specific
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
This indicator represents the share of forest area with trees of all species with a diameter at breast height of at least 40 cm.
In forests shaped by natural disturbances and succession, large trees are expected to occur to a much higher extent than today.
Historical harvesting of large trees and short harvest rotations in current forestry tend to prevent trees from growing very large.


2. **Methods**:
Data for this indicator come from the National Forest Inventory (NFI), where the diameter at breast height (dbh) of all trees larger than 5 cm dbh within the 250 m² NFI plot has been recorded by current methods since the mid-1990s [@viken2021-feltinstruks].
For each reporting period, the indicator is the area-weighted mean of this binary variable across NFI plots in the selected geographic unit.

3. **Interpretation**:
An indicator value of 1 corresponds to the county or sub-county X100 share of forest area with large trees or more (best condition).
An indicator value of 0 corresponds to 0% of the forest area with large trees (worst condition, X0).
Values between 0 and 1 reflect intermediate shares on a linear scale between X0 and the matching county or sub-county X100.

4. **Results and development needs**:
At national level, scaled indicator values are well below the upper reference in all three reporting periods (@sec-results, @tbl-dlta-national-results).
The national median increases from 0.16 (2010–2014) to 0.19 (2020–2024); approximate 95% intervals are narrow (about 0.15–0.20).
This corresponds to a non-scaled area-weighted share of forest with large trees of about 10–12% nationally, with a slight increase over time.
The indicator is ready for use but should not be used at finer spatial resolution than counties.


### 1.2 Native language summary - Norwegian

**Bakgrunn og begrunnelse**:
Indikatoren angir andelen av skogarealet med trær av alle arter med diameter i brysthøyde (dbh) minst 40 cm.
I skog preget av naturlige forstyrrelser og suksesjon forventes store trær i langt større utstrekning enn i dag.
Historisk hogst av store trær og korte omløpstider i dagens skogbruk hindrer trær i å vokse svært store.

**Metoder**:
Data kommer fra Landsskogtakseringen (NFI), der dbh er registrert for alle trær > 5 cm dbh på hver 250 m² takstflate [@viken2021-feltinstruks].
For hver rapporteringsperiode beregnes arealvektede gjennomsnitt for fylker, regioner og nasjonalt nivå.

**Tolkning**:
Indikatorverdi 1 tilsvarer fylkes-/delfylkes-X100 eller mer av skogarealet med store trær.
Indikatorverdi 0 tilsvarer 0 % (X0).
Verdier mellom 0 og 1 reflekterer mellomliggende andeler på en lineær skala mot gjeldende X100.

**Resultater og utviklingsbehov**:
På nasjonalt nivå ligger de skalerte verdiene under øvre referansenivå i alle tre rapporteringsperioder.
Medianverdien øker fra 0,16 (2010–2014) til 0,19 (2020–2024); omtrentlig 95 %-intervall er smalt (ca. 0,15–0,20).
Dette tilsvarer en arealvektet andel skog med store trær på om lag 10–12 % nasjonalt, med en svak økning over tid.
Indikatoren er klar til bruk, men bør ikke brukes på finere oppløsning enn fylker.

## 2. About the underlying data

The diameter at breast height (dbh) of all trees larger than 5 cm dbh is recorded for each 250 m² NFI inventory plot [@viken2021-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the proportion of NFI plots with at least one tree with dbh ≥ 40 cm for included NFI plots in any selected region.
Both productive and unproductive forest (`arealtype` 1 and 12) are included.
Plots with missing `stort_tre` are excluded before aggregation.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020–2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal extent, resolution, and representativity

Data are available from the mid-1990s, when current recording methods were established.
Annual estimates are possible due to representative sampling every year.
For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010.
The data are representative for productive and unproductive forest land within the selected region.
Spatial resolution should be limited to counties.

### 2.2 Sampling design

All trees larger than 5 cm at breast height (dbh) are recorded and identified to species for each 250 m² NFI inventory plot [@viken2021-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying coverage for forest below/above the coniferous forest line and in Finnmark county.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.

### 2.3 Original units

The indicator is specified as the proportion of NFI plots with at least one tree with dbh ≥ 40 cm.
At plot level this is a binary variable (0/1); aggregated values are area-weighted means of that proportion (equivalent to the share of forest area with large trees).

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The indicator represents the proportion of forest area with large trees.
Large trees are an important structural property related to forests under long-term, rather stable conditions.
This indicator represents a B2 Structural state characteristic.

### 3.2 Ecosystem condition characteristic

Large trees are indicative of forests under large-scale, long-term and rather stable conditions and represent important habitat for many species.
The indicator is characterised as Functionally important species and biophysical structures.

### 3.3 Collinearities with other indicators

There may be positive correlations with indicators for old forest, e.g., biologically old forest, succession of old deciduous trees, and large trees volume.

### 3.4 Impact factors

Frequent natural disturbances, such as rock slides, fires, floods, or windthrows, will reduce the frequency of large trees (although large pines will often survive fires).
Historical forestry practices of harvesting the largest trees or current clear-cutting with frequent recurrent harvesting will prevent the development of large trees.
Human-induced climate change may result in more rapid tree growth with faster attainment of the dbh criterion for large trees.
However, such climate change may also induce more frequent disturbances, drought events, or increased pressure of pathogens, all generally reducing the life span of forest trees.

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each reporting period, non-scaled values for county, region, and Norway are obtained by pooling all included forest plots in the unit: the area-weighted mean of the plot-level large-tree presence.
Where NFI county labels are split into parts, those parts are retained as separate accounting units for scaling.
Scaling and truncation are applied at county or sub-county level using X0 (0%) and county or sub-county-specific X100 values with a linear scaling function; values outside [0, 1] are truncated at scaling.
Regional and national scaled values are derived from scaled county bootstrap draws.
The indicator is likely to have a high frequency of NFI plots without large trees, so results should not be interpreted at finer resolution than counties.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

Intact forest ecosystems are expected to contain a substantial share of forest area with large trees.
The lower reference value (X0) is therefore 0% of forest area with large trees (worst condition).
The upper reference value (X100) is defined separately for each county or sub-county accounting unit (@sec-dlta-reference-values).
For lack of better knowledge, a linear relationship is assumed between X0 and X100 within each unit.
Hence, indicator values are scaled between 0 and 1 based on a linear scaling function; scaled indicator values outside [0, 1] are truncated at scaling.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or sub-county resolution for the share of forest area with large trees.
Although indicator values from NFI plots may technically be aggregated at finer spatial units than counties, scaling at such finer levels may result in biases.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability in the frequency of large trees among NFI plots and measurement errors for trees within NFI plots.
There is also uncertainty related to the setting of the reference values and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties are not quantified.

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within each county and recalculating the area-weighted mean share of forest area with large trees.
For scaled county or sub-county values, bootstrap draws are transformed using the same unit-specific reference levels X0 and X100 as the point estimates (truncation at scaling).
For scaled regional and national values, uncertainty is propagated from the scaled county bootstrap draws.
The resulting distributions are summarised with median and quantile intervals.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Proportion of forest with large trees

The diameter at breast height (dbh) is measured for all trees > 5 cm dbh for each 250 m² NFI plot according to methods described in [@viken2021-feltinstruks].
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used.
The variable `stort_tre` is binary: 1 if the plot has at least one tree with dbh ≥ 40 cm, 0 otherwise.
Both productive and unproductive forest plots (`arealtype` 1 and 12) are included.


```{r load-data}
#| cache: false
#| code-summary: "Load data"

lsk_data_file <- here("data/LSK_data_okologisk_tilstand_2026-06-03.csv")
dlta_cache_deps <- c(
  lsk_data_file,
  here::here("R/functions.R"),
  here::here("R/NO_DLTA_001.qmd")
)

lsk_data_raw <- read_csv(
  lsk_data_file,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties or sub-counties, regions and nationally.
Where counties are split into parts in the NFI data, those parts are retained as separate accounting units before scaling.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-dlta-derived-vars}

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

We filter the raw dataset to include both productive and unproductive forest plots (`arealtype` 1 and 12) and standardise NFI county labels while retaining split county or sub-county units.
The plot-level indicator is the binary NFI variable `stort_tre` (1 = at least one tree with dbh ≥ 40 cm, 0 = no large trees).
For each included plot, the area-weighted contribution to later aggregation is `indicator_continuous * au_areal`.

```{r dlta-create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(arealtype %in% c(1, 12)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = stort_tre,
    indicator_continuous_weighted = indicator_continuous * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r dlta-select-downstream-fields}
#| cache: false
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    stort_tre,
    indicator_continuous,
    indicator_continuous_weighted
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values {#sec-dlta-non-scaled}

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.

```{r calculate-dlta-all-periods}
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))
#| code-summary: "Calculate non-scaled values"

 dlta_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 dlta_results_county_all_periods   <- purrr::map_dfr(dlta_results, "county")
 dlta_results_region_all_periods   <- purrr::map_dfr(dlta_results, "region")
 dlta_results_national_all_periods <- purrr::map_dfr(dlta_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted non-scaled large trees share, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))

county_draws_long <- purrr::imap_dfr(dlta_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, county_part, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-dlta-county-nonscaled) summarises non-scaled bootstrap draws for all reporting periods: the median area-weighted large trees share per county.

```{r tbl-dlta-county-nonscaled}
#| tbl-cap: "Non-scaled share of forest area with large trees by county for all reporting periods (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, county_part, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dlta_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    non_scaled_median_pct = non_scaled_median * 100
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    `Sub-county` = county_part,
    Region = region_name,
    `Non-scaled value (proportion)` = non_scaled_median,
    `Non-scaled value (%)` = non_scaled_median_pct,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (proportion)`,
    decimals = 4
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (%)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-dlta-reference-values}

County or sub-county reference levels are listed below.
X0 is 0% for all units; X100 is the share of forest area with large trees at best condition for each county or sub-county accounting unit.

```{r dlta-reference-values}
#| code-summary: "Reference values by county or sub-county"

dlta_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuV", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_pct = c(
    70, 75, 70, 55, 65, 60, 70, 60, 75,
    70, 60, 65, 65, 60, 60, 60, 65,
    60, 60, 60, 55, 55, 50
  )
) |>
  dplyr::mutate(x100 = x100_pct / 100)

x0 <- 0
county_reference <- dlta_reference_values |>
  dplyr::select(county_part, x100, x100_pct)

```

Each county or sub-county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 0, unit-specific X100; truncation at scaling).
@fig-dlta-norm shows the mapping with one panel per county or sub-county accounting unit.

```{r county-draws-scaled}
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, county_part, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dlta_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-dlta-norm}
#| fig-cap: "Linear normalisation of the share of forest area with large trees (proportion) to indicator values (0–1), using X0 = 0% and county/sub-county-specific X100. One panel per county or sub-county accounting unit; points are sampled bootstrap draws pooled across reporting periods."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 6
#| fig-width: 8

norm_plot_data <- county_draws_scaled |>
  dplyr::group_by(county_part) |>
  dplyr::group_modify(~{
    n_sample <- min(30L, nrow(.x))
    dplyr::slice_sample(.x, n = n_sample, replace = FALSE)
  }) |>
  dplyr::ungroup()

dlta_norm_plot <- ec_norm_plot(
  norm_plot_data$non_scaled_value,
  norm_plot_data$scaled_value
) +
  ggplot2::facet_wrap(~norm_plot_data$county_part, scales = "free_x") +
  ggplot2::labs(
    x = "Share of forest area with large trees (proportion)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme_bw(base_size = 11)
dlta_norm_plot$layers[[1]]$aes_params$size <- 2.5

suppressMessages(
  ggsave(
    filename = here::here("img/dlta_norm_plot.png"),
    plot = dlta_norm_plot,
    scale = 1.5
  )
)

dlta_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r dlta-ec-upscale-regional}
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))
#| code-summary: "ec_upscale: county to region"

dlta_mc_rds_regional <- here::here("data/dlta_mc_regional.rds")

dlta_mc_regional <- ecTools::ec_upscale(
  data        = county_draws_scaled,
  variable    = scaled_value,
  weight      = total_area,
  start_units = fylke_name,
  end_units   = region_code,
  year        = period,
  end_units_name = "region_code",
  n = 1000
)
saveRDS(dlta_mc_regional, dlta_mc_rds_regional)
```

```{r dlta-ec-upscale-national}
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))
#| code-summary: "ec_upscale: region to national"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

dlta_mc_regional_for_national <- dlta_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

dlta_mc_rds_national <- here::here("data/dlta_mc_national.rds")

dlta_mc_national <- ecTools::ec_upscale(
  data        = dlta_mc_regional_for_national,
  variable    = sampled_mean,
  weight      = region_area,
  start_units = region_code,
  end_units   = nation,
  year        = period,
  end_units_name = "nation",
  n = 1000
)
saveRDS(dlta_mc_national, dlta_mc_rds_national)
```

```{r dlta-mc-combined}
#| cache: true
#| cache.extra: !expr list(deps_mtime = max(file.mtime(dlta_cache_deps), na.rm = TRUE))
#| code-summary: "Combine regional and national MC draws; build summary tables"

dlta_mc_combined <- dplyr::bind_rows(
  dlta_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  dlta_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- dlta_results_county_all_periods |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- dlta_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- dlta_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- dlta_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    dlta_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- dlta_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    dlta_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  dlta_results_county_all_periods |>
    dplyr::inner_join(county_reference, by = "county_part") |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = x100_pct,
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      dlta_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      dlta_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_pct = round(indicator_value * 100, 2),
      reference_value_pct = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_pct, exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

At county or sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r dlta-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_pct, original_value_pct) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (%)` = reference_value_pct,
      `Original value (%)` = original_value_pct
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (%)`, `Original value (%)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded the county/sub-county X100 reference and required truncation.\n")
}

```

The time series below (@fig-dlta-timeSeries) summarises regional and national values.

```{r fig-dlta-timeSeries}
#| fig-cap: "Indicator values for density of large trees by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

dlta_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
dlta_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
dlta_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(dlta_region_levels)),
  dlta_region_levels
)

dlta_plot_data <- dlta_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!dlta_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = dlta_region_levels))

ggplot2::ggplot(
  dlta_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = dlta_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to X0 (0% of forest area with large trees) and 1 corresponds to the county or sub-county X100.

### 10.1 County-level results (scaled with uncertainty)

Use the search box and page controls in @tbl-dlta-county-results to browse county rows across all reporting periods.

```{r tbl-dlta-county-results}
#| tbl-cap: 'County or sub-county indicator values for density of large trees by reporting period. Values are scaled to 0–1 using X0 and county/sub-county-specific X100 via ec_normalise(). Uncertainty is derived from 1000 bootstrap draws scaled with the same function. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Density of large trees",
    ID = "NO_DLTA_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    `Sub-county` = .data$county_part,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-dlta-region-results}
#| tbl-cap: 'Regional indicator values for density of large trees by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Density of large trees",
    ID = "NO_DLTA_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-dlta-national-results}
#| tbl-cap: 'National indicator values for density of large trees by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Density of large trees",
    ID = "NO_DLTA_001",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`dlta_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  dlta_mc_combined,
  here::here("data/results_NO_DLTA_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
