---
title: "Dead wood total volume per hectare"
subtitle: |
  NO_DWTV_001
indicatorID: 
  - NO_DWTV_001
indicatorName: 
  - NO_DWTV 
folderName: NO_DWTV_001
country: Norway
continent: Europe
ECT: B1 Biotic compositional characteristic
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_DWTV_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Scale - Aggregate (area-weighted arithmetic mean) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.
date: May 12, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 12. May 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:


2. **Methods**:
Data for this indicator come from the NFI, where standing and lying dead wood volume per hectare (`dw_prha`) is recorded for each plot.
For each reporting period, area-weighted means of `dw_prha` are calculated for counties or sub-counties, regions and nationally using `au_areal`.
Plots with missing dead wood volume are excluded before aggregation.
Each county or sub-county is scaled against its own reference level before regional and national values are calculated as area-weighted means of scaled county or sub-county values.

3. **Interpretation**:


4. **Results and development needs**:


## 2. About the underlying data

The volume of standing and lying dead wood > 10 cm in diameter is recorded for each NFI inventory plot [@viken2021-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an area-weighted mean of dead wood volume per hectare on included plots.
Plots with missing dead wood volume are excluded before aggregation.
Results are reported for counties or sub-counties, regions and nationally.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI. 

### 2.1 Spatial and temporal extent, resolution, and representativity

Data are available from 2010, when the volume of lying dead wood was first recorded based on current methods. A different method for quantifying the volume of dead wood was applied during the 1994-1998 forest inventory. Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. The data are representative for all productive forest land of Norway. Aggregated reporting uses counties, regions and the whole country.

### 2.2 Sampling design


### 2.3 Original units

The indicator is specified as dead wood total volume per hectare (m³ ha⁻¹), comprising standing and lying dead wood.
Aggregated values are area-weighted means of plot-level volume per hectare.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2025-10-20, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.

### 2.5 Additional comments about the dataset

None

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)



### 3.1.1 Other condition typologies

None

### 3.2 Ecosystem condition characteristic



### 3.3 Collinearities with other indicators



### 3.4 Impact factors



### 3.5 Spatial aggregation

For each reporting period, plot-level dead wood volume per hectare is aggregated with area weights to county or sub-county, regional and national units.
County or sub-county codes are assigned from NFI county labels using the same county and sub-county units as in related forest indicators.
Only plots with non-missing `dw_prha` enter the aggregation.
For each geographic unit, the non-scaled indicator value is the area-weighted mean of plot-level dead wood volume per hectare, calculated as the sum of `dw_prha * au_areal` divided by the sum of `au_areal` over the included plots in that unit.
Each county or sub-county is scaled against its own reference levels X0 = 0 m³ ha⁻¹ and county-specific X100.
Regional and national scaled values are area-weighted means of the scaled county or sub-county values.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

The reference value is based on ...
Reference values are expert assessments without estimated uncertainty.
The lower reference level X0 is set to 0 m³ ha⁻¹ and the upper reference level X100 is set separately for each county or sub-county.
The county and sub-county X100 values used in the analyses are listed in section 9.4.
For lack of better knowledge, the relationship between indicator values and ecosystem condition is assumed to be linear for values between X0 and X100.
Hence, indicator values are scaled between 0 and 1 based on a linear scaling function between X0 and X100; scaled indicator values above 1 are truncated to 1.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or sub-county resolution, not as one national value.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability of dead wood volume among NFI plots and to the measurements of this volume for each NFI plot.
This combined uncertainty is quantified by bootstrapping.
There is also uncertainty related to the setting of the upper reference value and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties are not quantified.

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in dead wood volume among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within the relevant geographic unit and recalculating the area-weighted mean.
This is done separately for counties or sub-counties, regions, and the national level.
For scaled county or sub-county values, the bootstrap mean, median, standard error, confidence interval, and quartiles from the non-scaled bootstrap are transformed with the same county-specific linear scaling function used for the point estimate.
This is an approximation when many bootstrap replicates exceed X100, because truncation is not applied inside each bootstrap replicate before summarising the scaled distribution.
For scaled regional and national values, uncertainty is estimated by bootstrapping plots, calculating county or sub-county non-scaled values in each replicate, scaling each county or sub-county with its own reference levels, and then taking the area-weighted mean of the scaled county or sub-county values within each region or for the whole country.
The resulting bootstrap distribution is summarised with mean, median, standard error, confidence interval, and quartiles.
The reported point estimate is the bootstrap median.
Standard errors, confidence intervals, and quartiles are calculated from the same bootstrap replicate distribution.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

R package versions for this indicator are managed with `renv`.
The project lockfile `renv.lock` records the package versions used to render the document.
After cloning the repository, run `renv::restore()` in the project root before rendering.

### 7.1 Dead wood volume per hectare

The volume of standing and lying dead wood per hectare is quantified for each NFI plot according to methods described in [@viken2021-feltinstruks].
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used.
The plot-level indicator is dead wood total volume per hectare (`dw_prha`).
Aggregated values are area-weighted means of `dw_prha` for counties or sub-counties, regions and the whole country using `au_areal`.
Plots with missing dead wood volume are excluded before aggregation.
As the NFI plots vary in density for forest below/above the coniferous forest line and in Finnmark, each type of plot represents a different part of all forest land.
In addition, some NFI plots are divided in two parts in the data.
These variations in plot representativity must be accounted for in aggregated estimates.
Normally, data from NFI will contain an area-weighting variable to be used in aggregating NFI data for spatial units larger than individual NFI plots.

```{r load-data}
#| code-summary: "Load data"

# Load main data file
lsk_data_raw <- read_csv(here("data/LSK_data_naturregnskap.csv"),
                     show_col_types = FALSE,
                     locale = locale(encoding = "UTF-8")) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties or sub-counties, regions and nationally.
County or sub-county codes are assigned from NFI county labels before aggregation.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.
We retain both productive and unproductive forest plots (`arealtype` 1 and 12), exclude plots with missing dead wood volume, assign county or sub-county codes, and create the derived variables used in subsequent aggregation steps.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

Next we filter the raw dataset to include only the two relevant forest area types, remove records missing `dw_prha`, and standardise the geography fields.
Area-weighted values are created using `au_areal`.

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    arealtype %in% c(1, 12)
  ) |>
  filter(!is.na(dw_prha)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = dw_prha,
    indicator_continuous_weighted = dw_prha * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    dw_prha,
    indicator_continuous,
    indicator_continuous_weighted
  )

```

The indicator is scaled against county or sub-county reference levels in section 9.4.

### 9.2 Calculate non-scaled area-weighted indicator values

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
County or sub-county estimates use the plot-level `fylke_name` field; regional estimates use `region_code`; the national estimate uses all plots in the period.

```{r calculate-dwtv-all-periods}
#| code-summary: "Calculate non-scaled values"

 dwtv_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 dwtv_results_county_all_periods   <- purrr::map_dfr(dwtv_results, "county")
 dwtv_results_region_all_periods   <- purrr::map_dfr(dwtv_results, "region")
 dwtv_results_national_all_periods <- purrr::map_dfr(dwtv_results, "national")

```

### 9.3 Bootstrap uncertainty estimation for non-scaled indicator values

Uncertainty in the non-scaled indicator estimates is quantified using bootstrapping.
For each 5-year period and geographic unit, NFI plots in that unit are resampled with replacement and the area-weighted non-scaled indicator value is recalculated in each replicate.
This is done separately for counties or sub-counties, regions, and the national level.
The resulting bootstrap distributions are summarised with mean, median, standard error, 95% confidence interval, and quartiles.
These non-scaled uncertainty measures are used directly for reporting and, at county or sub-county level, as input to the scaled uncertainty calculations in section 9.4.

```{r bootstrap-dwtv-all-periods}
#| code-summary: "Bootstrap uncertainty"

# Prepare bootstrap objects per reporting period
 dwtv_bootstrap <- dwtv_results |>
   purrr::imap(
     ~ list(
       period = .y,
       data = .x$data,
       county = .x$county,
       region = .x$region,
       national = .x$national,
      national_boot = bootstrap_indicator(.x$data, n_bootstrap = 1000),
       county_boot = purrr::map(
         split(.x$data, .x$data$fylke_name),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
       ),
       region_boot = purrr::map(
         split(.x$data, .x$data$region_code),
        ~ bootstrap_indicator(.x, n_bootstrap = 1000)
       )
     )
   )

# County-level summaries with uncertainty
 dwtv_results_county_non_scaled_uncertainty <- purrr::map_dfr(
   dwtv_bootstrap,
   ~ {
     county_summary <- purrr::imap_dfr(
       .x$county_boot,
       function(boot, county_name) {
         dplyr::tibble(
           fylke_name = county_name,
          mean = boot$mean,
           indicator_value = boot$median,
           se = boot$se,
           ci_lower = boot$ci_lower,
           ci_upper = boot$ci_upper,
           q1 = boot$q1,
           median = boot$median,
           q3 = boot$q3
         )
       }
     )

     .x$county |>
       dplyr::select(fylke_name, county_part, period, total_area, n_plots) |>
       dplyr::left_join(county_summary, by = "fylke_name")
   }
 )

# Region-level summaries with uncertainty
 dwtv_results_region_non_scaled_uncertainty <- purrr::map_dfr(
   dwtv_bootstrap,
   ~ {
     region_summary <- purrr::imap_dfr(
       .x$region_boot,
       function(boot, region_code) {
         dplyr::tibble(
           region_code = region_code,
          mean = boot$mean,
           indicator_value = boot$median,
           se = boot$se,
           ci_lower = boot$ci_lower,
           ci_upper = boot$ci_upper,
           q1 = boot$q1,
           median = boot$median,
           q3 = boot$q3
         )
       }
     )

     .x$region |>
       dplyr::select(region_code, period, total_area, n_plots) |>
       dplyr::left_join(region_summary, by = "region_code")
   }
 )

# National-level summaries with uncertainty
 dwtv_results_national_non_scaled_uncertainty <- purrr::map_dfr(
   dwtv_bootstrap,
   ~ .x$national |>
     dplyr::mutate(
      mean = .x$national_boot$mean,
       indicator_value = .x$national_boot$median,
       se = .x$national_boot$se,
       ci_lower = .x$national_boot$ci_lower,
       ci_upper = .x$national_boot$ci_upper,
       q1 = .x$national_boot$q1,
       median = .x$national_boot$median,
       q3 = .x$national_boot$q3
     )
 )

```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-dwtv-reference-values}

County or sub-county reference levels are listed below.
For each county or sub-county, the non-scaled point estimate is scaled with `scale_indicator()` using X0 = 0 and the matching X100.
The function applies linear scaling between X0 and X100 and truncates values above 1 when volume exceeds X100.
Regional and national scaled point estimates are area-weighted means of the scaled county or sub-county values, weighted by each unit's represented area.
Non-scaled regional and national point estimates are also retained for reporting and truncation summaries.
For scaled county or sub-county uncertainty, the non-scaled bootstrap summaries from section 9.3 are transformed with the same county-specific scaling function.
For scaled regional and national uncertainty, each bootstrap replicate resamples plots, calculates county or sub-county non-scaled values, scales each county or sub-county with its own reference levels, and then aggregates the scaled county or sub-county values within each region or for the whole country before the bootstrap distribution is summarised.

```{r dwtv-reference-values}
#| code-summary: "Reference values by county or sub-county"

# X100 reference values (m3/ha)
dwtv_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuN", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_m3_ha = c(
    50, 50, 50, 45, 50, 45, 50, 45, 50,
    50, 45, 45, 45, 40, 40, 40, 40, 45, 45,
    45, 45, 30, 30
  )
) |>
  dplyr::mutate(x100 = x100_m3_ha)

x0 <- 0
county_reference <- dwtv_reference_values |>
  dplyr::select(county_part, x100)

```

```{r scale-and-aggregate}
#| code-summary: "Scale and aggregate"

# Scale county values (point estimates) for all periods
county_scaled_all_periods <- dwtv_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = scale_indicator(indicator_value, x0, x100)
  )

# Uncertainty for county scaled: apply linear scaling to non-scaled CIs and quartiles
county_scaled_with_uncertainty <- dwtv_results_county_non_scaled_uncertainty |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::left_join(
    county_scaled_all_periods |>
      dplyr::select(period, fylke_name, scaled_value),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    scaled_mean = scale_indicator(mean, x0, x100),
    scaled_median = scale_indicator(indicator_value, x0, x100),
    scaled_ci_lower = scale_indicator(ci_lower, x0, x100),
    scaled_ci_upper = scale_indicator(ci_upper, x0, x100),
    scaled_q1 = scale_indicator(q1, x0, x100),
    scaled_q3 = scale_indicator(q3, x0, x100)
  )

# Aggregate non-scaled values to regional level (point estimates) for all periods
region_non_scaled_agg_all_periods <- dwtv_results_region_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

# Aggregate scaled county values to regional level (point estimates) for all periods
region_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code))

# Aggregate non-scaled values to national level (point estimates) for all periods
national_non_scaled_agg_all_periods <- dwtv_results_national_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

# Aggregate scaled county values to national level (point estimates) for all periods
national_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National")

# Check for truncation: count county values exceeding their county-specific reference
truncation_summary <- dplyr::bind_rows(
  county_scaled_all_periods |>
    dplyr::left_join(
      dwtv_reference_values |> dplyr::select(county_part, reference_value_m3_ha = x100_m3_ha),
      by = "county_part"
    ) |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_m3_ha = round(indicator_value, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_m3_ha,
      exceeds_threshold,
      original_value_m3_ha
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      region_non_scaled_agg_all_periods |>
        dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_m3_ha,
      exceeds_threshold,
      original_value_m3_ha
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      national_non_scaled_agg_all_periods |>
        dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period,
      level,
      region_code,
      region_name,
      fylke_name,
      reference_value_m3_ha,
      exceeds_threshold,
      original_value_m3_ha
    )
)

# Summary of truncation by period and level
truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

# Values that were truncated (for reporting)
truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_m3_ha))

```

At county or sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the scaled aggregate equals 1 after area-weighted aggregation of scaled county or sub-county values.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r report-truncation}
#| code-summary: "Report truncation"

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  cat("\n### Values that were truncated:\n")
  truncated_display <- truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_m3_ha, original_value_m3_ha) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference Value (m3/ha)` = reference_value_m3_ha,
      `Original Value (m3/ha)` = original_value_m3_ha
    )
  print(truncated_display)
} else {
  cat("\nNo values exceeded the reference level and required truncation.\n")
}

```

```{r bootstrap-scaled-uncertainty}
#| code-summary: "Bootstrap uncertainty for scaled values"

# Bootstrap uncertainty for scaled-then-aggregated regions per period
region_scaled_uncertainty <- purrr::imap_dfr(
  dwtv_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    
    if (nrow(period_data) == 0) return(NULL)
    
    bootstrap_region_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    ) |>
      dplyr::rename(
        boot_mean = mean,
        boot_median = median
      ) |>
      dplyr::mutate(period = period_name)
  }
)

# Bootstrap uncertainty for scaled-then-aggregated national per period
national_scaled_uncertainty <- purrr::imap_dfr(
  dwtv_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    
    if (nrow(period_data) == 0) return(NULL)
    
    boot_result <- bootstrap_national_scaled_agg(
      period_data,
      x0 = x0,
      county_reference = county_reference,
      n_bootstrap = 1000
    )
    
    dplyr::tibble(
      period = period_name,
      region = "National",
      boot_mean = boot_result$mean,
      boot_median = boot_result$median,
      se = boot_result$se,
      ci_lower = boot_result$ci_lower,
      ci_upper = boot_result$ci_upper,
      q1 = boot_result$q1,
      q3 = boot_result$q3
    )
  }
)

# Attach uncertainty to scaled aggregated values and retain both bootstrap mean and median
region_scaled_agg_with_uncertainty <- region_scaled_agg_all_periods |>
  dplyr::left_join(region_scaled_uncertainty, by = c("period", "region_code")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

national_scaled_agg_with_uncertainty <- national_scaled_agg_all_periods |>
  dplyr::left_join(national_scaled_uncertainty, by = c("period", "region")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

```


## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
All scaled values are on a 0-1 scale, where 0 corresponds to X0 (0 m³ ha⁻¹) and 1 corresponds to the county-specific X100 for counties or sub-county units, or to an area-weighted aggregate of scaled county or sub-county values for regions and the whole country.
County or sub-county tables report bootstrap means and medians, standard errors, confidence intervals, and quartiles derived from the non-scaled bootstrap and transformed with the county-specific scaling function.
The reported point estimate is the bootstrap median.
Standard errors, confidence intervals, and quartiles are calculated from the same bootstrap replicate distribution and are not centred on the bootstrap mean.
Regional and national tables report bootstrap means and medians together with standard errors, confidence intervals, and quartiles from the scale-then-aggregate bootstrap workflow in section 9.4.

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-county-results-table}
#| tbl-cap: 'County or sub-county indicator values for dead wood total volume per hectare by reporting period. Values are scaled to 0-1 using county-specific X0 and X100. Uncertainty is obtained by applying the county-specific scaling function to the non-scaled bootstrap summaries from section 9.3. The reported point estimate is the bootstrap median. The table includes the scaled point estimate, bootstrap mean and median estimates, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code)
  ) |>
  dplyr::select(period, fylke_name, region_name, scaled_value, scaled_mean, scaled_median, se, scaled_ci_lower, scaled_ci_upper, 
         scaled_q1, scaled_q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    scaled_ci_lower = round(scaled_ci_lower, 3),
    scaled_ci_upper = round(scaled_ci_upper, 3),
    scaled_q1 = round(scaled_q1, 3),
    scaled_q3 = round(scaled_q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = scaled_ci_lower,
    `CI Upper` = scaled_ci_upper,
    `Q1` = scaled_q1,
    `Q3` = scaled_q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  DT::datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, lengthMenu = c(10, 25, 50, 100))
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-region-results-table}
#| tbl-cap: 'Regional indicator values for dead wood total volume per hectare by reporting period. Scaled point estimates are area-weighted means of scaled county or sub-county values. Uncertainty is estimated by bootstrapping plots, scaling counties or sub-county units in each replicate, and aggregating the scaled values within each region. The reported point estimate is the bootstrap median. The table includes the scaled point estimate, bootstrap mean and median, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::select(period, region_name, scaled_value, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    Region = region_name,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-national-results-table}
#| tbl-cap: 'National indicator values for dead wood total volume per hectare by reporting period. Scaled point estimates are area-weighted means of scaled county or sub-county values. Uncertainty is estimated by bootstrapping plots, scaling counties or sub-county units in each replicate, and aggregating the scaled values nationally. The reported point estimate is the bootstrap median. The table includes the scaled point estimate, bootstrap mean and median, standard error (SE), 95% confidence intervals (CI Lower, CI Upper), and quartiles (Q1, Q3).'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::select(period, region, scaled_value, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    `Spatial Level` = region,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


## 11. Export file

```{r export}
#| eval: false
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
