---
title: "Dead wood total volume per hectare"
subtitle: |
  NO_DWTO_001
indicatorID: 
  - NO_DWTO_001
indicatorName: 
  - NO_DWTO 
folderName: NO_DWTO_001
country: Norway
continent: Europe
ECT: B1 Biotic compositional characteristic
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_DWTO_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to sub-county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
  docx:
    embed-resources: true
    code-fold: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.
date: May 12, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 12. May 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:


2. **Methods**:
Data for this indicator come from the NFI, where standing and lying dead wood volume per hectare is recorded for each plot.
For each reporting period, area-weighted means of dead wood are calculated for sub-counties, regions and nationally.
Each sub-county is scaled against its own reference level before regional and national values are calculated as area-weighted means of scaled sub-county values.

3. **Interpretation**:
An indicator value of 1 corresponds to dead wood volume per hectare at the sub-county reference level X100 (m³ ha⁻¹).
Values decrease toward zero as dead wood volume falls below that reference.

4. **Results and development needs**:
The indicator is ready for use at sub-county resolution and above.
It should not be interpreted at finer spatial resolution than the sub-county reference levels allow.

### 1.2 Native language summary - Norwegian

**Bakgrunn og begrunnelse**:


**Metoder**:
Data kommer fra Landsskogtakseringen (NFI).
Indikatoren er død vedvolum per hektar på hver takstflate.
Arealvektede gjennomsnitt beregnes for delfylker, regioner og nasjonalt nivå.

**Tolkning**:
Indikatorverdier skaleres mot delfylkesvis referansenivå X0 = 0 m³ ha⁻¹ og delfylkespesifikk X100 (m³ ha⁻¹).

**Resultater og utviklingsbehov**:
Indikatoren er klar til bruk på delfylkesnivå og oppover.


## 2. About the underlying data

The volume of standing and lying dead wood > 10 cm in diameter is recorded for each NFI inventory plot [@viken2021-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an area-weighted mean of dead wood volume per hectare on included plots.
Plots with missing dead wood volume are excluded before aggregation.
Results are reported for counties or sub-counties, regions and nationally.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI. 

### 2.1 Spatial and temporal extent, resolution, and representativity

Data are available from 2010, when the volume of lying dead wood was first recorded based on current methods. A different method for quantifying the volume of dead wood was applied during the 1994-1998 forest inventory. Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. The data are representative for all productive forest land of Norway. Aggregated reporting uses counties, regions and the whole country.

### 2.2 Sampling design


### 2.3 Original units

The indicator is specified as dead wood total volume per hectare (m³ ha⁻¹), comprising standing and lying dead wood.
Aggregated values are area-weighted means of plot-level volume per hectare.

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.


## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

B1 Biotic compositional characteristic



### 3.2 Ecosystem condition characteristic



### 3.3 Collinearities with other indicators



### 3.4 Impact factors



### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to sub-county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each reporting period, non-scaled values for sub-county, region, and Norway are obtained by pooling all included forest plots in the unit: the area-weighted mean of plot-level `dw_prha`.
Sub-county codes are assigned from NFI county labels before aggregation.
Scaling and truncation are applied after non-scaled sub-county aggregates have been calculated: each sub-county is scaled against its own lower reference level X0 (0 m³ ha⁻¹) and upper reference level X100 (m³ ha⁻¹) using a linear scaling function, and values above 1 are truncated to 1 at that step.
Regional and national scaled point estimates are area-weighted means of the already scaled sub-county values, using each sub-county's represented area as weight.
Non-scaled regional and national point estimates are calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

Reference values are expert assessments without estimated uncertainty.
The lower reference level X0 is set to 0 m³ ha⁻¹ and the upper reference level X100 is set separately for each sub-county.
The sub-county X100 values used in the analyses are listed in @sec-dwto-reference-values.
For lack of better knowledge, the relationship between indicator values and ecosystem condition is assumed to be linear for values between X0 and X100.
Hence, indicator values are scaled between 0 and 1 based on a linear scaling function between X0 and X100; scaled indicator values above 1 are truncated to 1.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined at county or sub-county resolution, not as one national value.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values rather than one common X100 for the aggregated non-scaled value.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability of dead wood volume among NFI plots and to the measurements of this volume for each NFI plot.
This combined uncertainty is quantified by bootstrapping.
There is also uncertainty related to the setting of the upper reference value and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties are not quantified.

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in dead wood volume among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within each county or sub-county and recalculating the area-weighted mean of `dw_prha`.
For scaled county or sub-county values, bootstrap draws are transformed with `ecTools::ec_normalise()` using the same county or sub-county-specific reference levels as the point estimates (truncation at scaling).
For scaled regional and national values, uncertainty is propagated from the scaled sub-county bootstrap draws with `ecTools::ec_upscale()`.
The resulting distributions are summarised with median and quantile intervals.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Dead wood volume per hectare

The volume of standing and lying dead wood per hectare is quantified for each NFI plot according to methods described in [@viken2021-feltinstruks].
For this indicator data from each full 5-yr inventory cycle, starting in 2010, are used.

```{r load-data}
#| code-summary: "Load data"

# Load main data file
# This export wraps each physical line in an extra pair of double quotes, so a
# single read_csv() sees one column spanning the whole file. Read once as one
# character column per line (readr unwraps the outer quotes), then parse the
# inner CSV text.
lsk_inner_csv <- read_csv(
  here("data/LSK_data_okologisk_tilstand_2026.csv"),
  col_names = FALSE,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8"),
  col_types = cols(.default = col_character())
) |>
  pull(1) |>
  paste(collapse = "\n")

lsk_data_raw <- read_csv(
  I(lsk_inner_csv),
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```


## 8. Spatial units

Aggregated indicator values are reported for counties or sub-counties, regions and nationally.
County or sub-county codes are assigned from NFI county labels before aggregation.
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates.

## 9. Analyses

### 9.1 Prepare data and create derived variables {#sec-dwto-derived-vars}

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

The plot-level indicator is dead wood total volume per hectare (`dw_prha`).
We filter the raw dataset to include both productive and unproductive forest plots (`arealtype` 1 and 12), remove records missing `dw_prha`, and standardise the geography fields.
Area-weighted values are created using `au_areal`.

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    arealtype %in% c(1, 12)
  ) |>
  filter(!is.na(dw_prha)) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = dw_prha,
    indicator_continuous_weighted = dw_prha * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream fields"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    dw_prha,
    indicator_continuous,
    indicator_continuous_weighted
  )

```


### 9.2 Calculate non-scaled area-weighted indicator values {#sec-dwto-non-scaled}

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.

```{r calculate-dwto-all-periods}
#| code-summary: "Calculate non-scaled values"

 dwto_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 dwto_results_county_all_periods   <- purrr::map_dfr(dwto_results, "county")
 dwto_results_region_all_periods   <- purrr::map_dfr(dwto_results, "region")
 dwto_results_national_all_periods <- purrr::map_dfr(dwto_results, "national")

```

### 9.3 Bootstrap draws per sub-county

For each sub-county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted non-scaled mean of `dw_prha`, giving a distribution of 1000 draws per sub-county.

```{r dwto-county-draws-long}
#| code-summary: "Bootstrap draws per sub-county (1000 per sub-county per period)"
#| cache: true
#| cache.extra: !expr list(file.mtime(here::here("R/functions.R")), file.mtime(here::here("data/LSK_data_okologisk_tilstand_2026.csv")), file.mtime(here::here("R/NO_DWTO_001.qmd")))

county_draws_long <- purrr::imap_dfr(dwto_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, county_part, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-dwto-subcounty-nonscaled) summarises non-scaled bootstrap draws for all reporting periods: the median area-weighted dead wood volume per hectare per sub-county.

```{r tbl-dwto-subcounty-nonscaled}
#| tbl-cap: "Non-scaled dead wood volume per hectare (m³ ha⁻¹) by sub-county for all reporting periods (median of 1000 bootstrap draws)."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, county_part, region_code, total_area) |>
  dplyr::summarise(
    non_scaled_median = median(non_scaled_value),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dwto_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    `Sub-county` = county_part,
    Region = region_name,
    `Non-scaled value (m³ ha⁻¹)` = non_scaled_median,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Non-scaled value (m³ ha⁻¹)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-dwto-reference-values}

Sub-county reference levels are listed below.

```{r dwto-reference-values}
#| code-summary: "Reference values by sub-county"

dwto_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuV", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_m3_ha = c(
    60, 60, 60, 50, 60, 50, 60, 50, 60,
    60, 50, 50, 50, 50, 50, 50, 50, 42.5, 42.5,
    42.5, 28, 28, 23
  )
) |>
  dplyr::mutate(x100 = x100_m3_ha)

x0 <- 0
county_reference <- dwto_reference_values |>
  dplyr::select(county_part, x100)

knitr::kable(
  dwto_reference_values |>
    dplyr::select(`Sub-county` = county_part, `X100 (m³ ha⁻¹)` = x100_m3_ha)
)

```

Each sub-county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 0, sub-county-specific X100; truncation at scaling).
@fig-dwto-norm shows the mapping for each sub-county.

```{r dwto-county-draws-scaled}
#| code-summary: "Scale sub-county draws; derive uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = x0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, county_part, total_area, region_code) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    dwto_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-dwto-norm}
#| fig-cap: "Linear normalisation of dead wood volume per hectare (m³ ha⁻¹) to indicator values (0–1), one panel per sub-county reference unit. Points are 100 sampled bootstrap draws per unit, pooled across all reporting periods."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 8
#| fig-width: 8

norm_plot_data <- county_draws_scaled |>
  dplyr::group_by(county_part) |>
  dplyr::slice_sample(n = 100) |>
  dplyr::ungroup()

dwto_norm_plot <- ggplot2::ggplot(
  norm_plot_data,
  ggplot2::aes(
    x = non_scaled_value,
    y = scaled_value,
    fill = scaled_value
  )
) +
  ggplot2::geom_point(
    stroke = NA, size = 3, alpha = 0.7, shape = 21
  ) +
  ggplot2::facet_wrap(~ county_part, scales = "free_x") +
  ggplot2::scale_fill_gradient(low = "red", high = "green") +
  ggplot2::scale_y_continuous(breaks = seq(0, 1, by = 0.2)) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Dead wood volume (m³ ha⁻¹)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))

suppressMessages(
  ggsave(
    filename = here::here("img/dwto_norm_plot.png"),
    plot = dwto_norm_plot,
    scale = 1.5
  )
)

dwto_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per sub-county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r dwto-ec-upscale-regional}
#| code-summary: "ec_upscale: sub-county to region (compute once, save RDS)"

dwto_mc_cache_deps <- c(
  here::here("data/LSK_data_okologisk_tilstand_2026.csv"),
  here::here("R/functions.R"),
  here::here("R/NO_DWTO_001.qmd")
)
dwto_mc_rds_regional <- here::here("data/dwto_mc_regional.rds")

if (
  file.exists(dwto_mc_rds_regional) &&
  file.mtime(dwto_mc_rds_regional) >= max(file.mtime(dwto_mc_cache_deps), na.rm = TRUE)
) {
  dwto_mc_regional <- readRDS(dwto_mc_rds_regional)
} else {
  dwto_mc_regional <- ecTools::ec_upscale(
    data        = county_draws_scaled,
    variable    = scaled_value,
    weight      = total_area,
    start_units = fylke_name,
    end_units   = region_code,
    year        = period,
    end_units_name = "region_code",
    n = 1000
  )
  saveRDS(dwto_mc_regional, dwto_mc_rds_regional)
}
```

```{r dwto-ec-upscale-national}
#| code-summary: "ec_upscale: region to national (compute once, save RDS)"

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

dwto_mc_regional_for_national <- dwto_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

dwto_mc_rds_national <- here::here("data/dwto_mc_national.rds")

if (
  file.exists(dwto_mc_rds_national) &&
  file.mtime(dwto_mc_rds_national) >= max(
    c(file.mtime(dwto_mc_cache_deps), file.mtime(dwto_mc_rds_regional)),
    na.rm = TRUE
  )
) {
  dwto_mc_national <- readRDS(dwto_mc_rds_national)
} else {
  dwto_mc_national <- ecTools::ec_upscale(
    data        = dwto_mc_regional_for_national,
    variable    = sampled_mean,
    weight      = region_area,
    start_units = region_code,
    end_units   = nation,
    year        = period,
    end_units_name = "nation",
    n = 1000
  )
  saveRDS(dwto_mc_national, dwto_mc_rds_national)
}
```

```{r dwto-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

dwto_mc_combined <- dplyr::bind_rows(
  dwto_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  dwto_mc_national |> dplyr::select(period, region = nation, sampled_mean)
)

county_scaled_all_periods <- dwto_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::filter(!is.na(x100)) |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- dwto_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- dwto_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- dwto_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    dwto_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- dwto_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    dwto_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  dwto_results_county_all_periods |>
    dplyr::left_join(
      dwto_reference_values |> dplyr::select(county_part, reference_value_m3_ha = x100_m3_ha),
      by = "county_part"
    ) |>
    dplyr::left_join(county_reference, by = "county_part") |>
    dplyr::mutate(
      level = "Sub-county",
      exceeds_threshold = indicator_value > x100,
      original_value_m3_ha = round(indicator_value, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name, county_part,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      dwto_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      fylke_name = NA_character_,
      county_part = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name, county_part,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      dwto_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_,
      county_part = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name, county_part,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_m3_ha))

```

At sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r dwto-report-truncation}
#| code-summary: "Report truncation"
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "Sub-county" ~ paste(fylke_name, county_part, sep = " / "),
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(period, level, location, reference_value_m3_ha, original_value_m3_ha) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (m³ ha⁻¹)` = reference_value_m3_ha,
      `Original value (m³ ha⁻¹)` = original_value_m3_ha
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (m³ ha⁻¹)`, `Original value (m³ ha⁻¹)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their sub-county reference and required truncation.\n")
}

```

The time series below (@fig-dwto-timeSeries) summarises regional and national values.

```{r fig-dwto-timeSeries}
#| fig-cap: "Indicator values for dead wood total volume per hectare by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

dwto_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
dwto_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
dwto_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(dwto_region_levels)),
  dwto_region_levels
)

dwto_plot_data <- dwto_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!dwto_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = dwto_region_levels))

ggplot2::ggplot(
  dwto_plot_data,
  ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)
) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = dwto_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(
      nrow = 2,
      override.aes = list(size = 0.7, linewidth = 0.45)
    ),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```


## 10. Results {#sec-results}

The following tables present scaled indicator results at sub-county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).
All scaled values are on a 0–1 scale, where 0 corresponds to X0 (0 m³ ha⁻¹) and 1 corresponds to the sub-county-specific X100 for sub-county units.

### 10.1 Sub-county results (scaled with uncertainty)

Use the search box and page controls in @tbl-dwto-subcounty-results to browse sub-county rows across all reporting periods.

```{r tbl-dwto-subcounty-results}
#| tbl-cap: 'Sub-county indicator values for dead wood total volume per hectare by reporting period. Values are scaled to 0–1 using sub-county-specific X0 and X100. Uncertainty is derived from 1000 bootstrap draws. The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Dead wood total volume",
    ID = "NO_DWTO_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    `Sub-county` = .data$county_part,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-dwto-region-results}
#| tbl-cap: 'Regional indicator values for dead wood total volume per hectare by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Dead wood total volume",
    ID = "NO_DWTO_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-dwto-national-results}
#| tbl-cap: 'National indicator values for dead wood total volume per hectare by reporting period. Scaled values and uncertainty are derived from Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B1",
    Indicator = "Dead wood total volume",
    ID = "NO_DWTO_001",
    `Accounting area` = "National",
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  )
```


## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`dwto_mc_combined`: `period`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  dwto_mc_combined,
  here::here("data/results_NO_DWTO_001.csv"),
  row.names = FALSE
)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [ES, KOS]
* Methodology: [ES, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [AHH]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].

::: {.callout-tip collapse="true"}
## Session Info

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
