---
title: "Nitrogen impact on vegetation"
subtitle: |
  [NO_FUNI_004] (forests) 
# delete the line with the second indicatorID (line 5) 
# if it's not relevant for you.
# In the parentheses you can add a word to distinguish multiple indicators 
# from each other, for example the ecosystem type that they cover
indicatorID: 
  - [NO_FUNI_004]
# If you are presenting multiple indicators, and have multiple 
# indicator IDs, then add them as a list here, and make sure you add
# the same number of list entries, 
# and in the same order that corresponds to the indicator IDs,  
# to the meta data fields where the indicators differ. 
# For example, if the indicators cover different ecosystem types, 
# add the ecosystems as a list under 'Ecosystems', but if they are in the
# same realm, same country etc., then you only need to write the realm and
# country once. 
indicatorName: 
  - Nitrogen impact on vegetation
verbatimName: 
  - Nitrogenpåvirkning på vegetasjon # the indicator name that us used in the native language
folderName: XXX_YYY_ZZZ # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
ECT: A2 - Chemical State characteristics
Realm: # Please copy and paste Realms, Biomes and Ecostsem names from here: https://github.com/NINAnor/ecRxiv/wiki#ecosystem-typology 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T2.2 Deciduous temperate forests
  - T2.3 Oceanic temp rainforests
verbatimEcosystem:
  - Skog
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: Comment here # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/XX_YYYY_ZZZ
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Scale/Truncate - Aggregate (zero-one inflated beta-regression) # e.g. Scale - Transform - Aggregate (length-weighted arithmetic mean) - Truncate
Uncertainties: 95% confidence intervals for the indicator values derived from the beta-regression standard errors. # What kind of quantitative unceratinties are reported alongside the point estimates. Can be either inferrential (i.e. bootstrapped confidence intervals or standard errors) or descriptive (variation, standard deviation or percentiles from the distribution of the raw indicator values).
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Joachim Töpper              # Enter name
    email: joachim.topper@nina.no  # Enter email
    affiliations: NINA
      - id: myID
        name: Norwegian institute for nature research # Enter affiliations
AuthorList: Töpper, J.  
date: June 6, 2026
data_availability: Openly available and accessible at Miljøidirektoratets kartkatalog
code_reproducibility: Fully documented on github and executable in R
open_science_badge: Data Integrated, Documented and executable
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(sf)
library(ecTools)
library(glmmTMB)
library(MASS)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 01 Jan. 1901 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**: This indicator represents plant community responses to nitrogen. It represents the lowest score of deviations from an optimal state either to higher or to lower nitrogen levels. 

2. **Methods**: The data comes from field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019). We combine this with functional trait values from Tyler et al. (2021) and plant strategies towards stressors from Grime (1974). Indicator values are based on calculating community-weighted means (Diekmann 2003, Töpper et al. 2018) or cumulative abundance (see Framstad et al. 2022) of functional trait values for plant community data. Deviations in such an indicator beyond a certain range of indicator values indicate a directional shift in the plant community. As reference communities we use ‘generalized species lists’ underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020, see Evju et al. 2023 for naturally open lowland ecosystems). For each major ecosystem type a separate set of indicators, related to the main drivers of change today, is computed. The lowest scoring functional plant community indicator is then reported as the ‘Functional plant community index’.

3. **Interpretation**:
An indicator value of 1 indicates no discernible difference in the functional plant composition due to nitrogen impacts. 
Values decreasing toward zero reflect a shift in species composition and community trait values away from the reference state, either as a response to increased or decreased nitrogen.

4. **Results and development needs**:

XXXX

To improve this indicator in the future, the perhaps most pressing issue it to regionalise the generalised species list, as currently they have a bias towards southhern and eastern Norway. The indicator values also need validation against independent datasets.




## 2. About the underlying data
We use three types of data for building functional plant community indicators for ecological condition:

- as observation data  we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (Halvorsen et al. 2020)
- Swedish functional plant indicator data from Tyler et al. (2021) and indicator values for plant strategies from Grime (1974)

For details see point 7 below.

### 2.1 Spatial and temporal resolution and extent
The ANO monitoring data available now represent one cycle of sampling from 2019-2024 comprising ca. 1000 sites across Norway.
ANO has no defined end date. 2025 is the first year of repeated sampling for about 20% of the approximately 1000 ANO sites for all of mainland Norway (i.e. the Ecosystem Accounting Area). Annual sampling is spatially representative, in principle allowing to compute area-representative estimates for single years but also multiple-yr-reporting periods. The indicator at hand is calculated based on the ANO data from 2020-2024.

### 2.2 Sampling design
At every ANO site, the the plant community is recorded (presence and percent cover of every vascular plant species) in the central 1 sqm plot of each of systematically arranged 18 monitoring points.

### 2.3 Original units
Abundance data for vascular plants in ANO is given as % cover within a 1 m2 sample quadrat. This is transformed to relative abundance data per quadrat. Each ANO quadrat is classified to its respective ecosystem type according to ‘Nature in Norway’ (NiN; <https://artsdatabanken.no/NiN>). Environmental indicator values from Tyler et al. (2021) are given on an ordinal scale from 1 (lowest affinity to the respective variable) to n (highest affinity to the respective variable), the number of levels (n) varying between environmental indicators.


### 2.4 Instructions for citing, using and accessing data
The ANO data are openly available at https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054

### 2.5 Additional comments about the dataset
The ANO program is run by the Norwegian Environment Agency. On their web sites you may also find some additional summary data. https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)
The functional plant community indicator for nitrogen reflects the availability of nutrients and soil chemical properties and thus represents the ECT class 'A2 Chemical characteristics'.

### 3.2 Ecosystem condition characteristic
The functional plant community indicator for nitrogen reflects the availability of nutrients and soil chemical properties and thus represents the characteristic 'Abiotic factors'.

### 3.3 Collinearities with other indicators
Different functional plant community indicators for different environmental variables (light, moisture, ...) might show a certain degree of co-linearity with one another as they are derived from the very same plant community data.
Functional plant community indicators can potentially exhibit co-linearities with other vegetation-oriented indicators, such as NDVI, though this depends greatly on which functional plant community indicator ends up as the leading one (i.e. the one with the lowest scaled score).


### 3.4 Impact factors
The plant community composition in forest ecosystems is subject to changes with climate warming and precipitation changes as well as pollution and disturbance from human infrastructure.

### 3.5 Spatial aggregation
The variable is scaled at the initial ANO-point level and the indicator estimate is an arithmetic mean of all ANO-points. There is no transformation, but indicator values are truncated at 1.

## 4. Reference condition and levels

### 4.1 Reference condition
In Norway, the reference condition for natural ecosystems has been one of largely intact ecosystems without large-scale human modifications (Nybø and Evju 2017). For semi-natural ecosystems, human influence is allowed, and indeed a prerequisite, but then in the sense of appropriate, stable and traditional land-use practices. Thus, in terms of alien species, the reference condition is defined as an ecosystem without any alien plant species.

### 4.2 Reference levels
Reference and limit values (scaling values) for functional plant community indicators are derived as the 0.025, 0.5, and 0.975 quantiles (lower limit values, reference value, upper limit value, respectively) from distributions of community weighted means of the respective plant indicators originating from bootstrapped generalized species lists.
Bootstraps were set up with resampling of the original generalized species lists 1000 times. In every sample a random 1/3 of the original generalized species list was sampled (as ANO points on average contain roughly a third of the potential species richness in the generalized species lists -> local variation in species composition thus is large and needs to be reflected in the bootstraps), with the exception of dominant species (abundance value of 5 or higher in the original scale of the generalized species lists), which were retained in every sample.
All reference distributions, and thus all scaling values are unique to NiN mapping units at a scale of 1:5000.

#### 4.2.1 Spatial resolution and validity
The same reference levels apply to all types of ecosystems and all regions.

## 5. Uncertainties
### 5.1 General uncertainties
Uncertainties relate partly to sampling and observer variability (and possible observer bias) in the recording of the plant cover data. Presumably, there is also uncertainty related to the scoring of indicator values for species, and this will probably vary among species (cf. Tyler et al. 2021 and Grime 1974). These potential sources uncertainty can only be assumed and not quantified.
Spatial variability as a source of uncertainty can be quantified. For every mean or median indicator value (after scaling) for every region (or any other delimited area of interest) we can calculate its corresponding standard error or confidence interval quantiles as a measure of spatial uncertainty for a geographical area.
There is some uncertainty related to the ecosystem categorization at the field sites (see Ullerud et al. 2018 and Naas et al. 2023), but since ANO monitoring does not include spatial delineation of ecosystem polygons this problem is considered smaller than in spatial ecosystem mapping which these references mainly relate to.
Other uncertainties connected to the indicator arise from community weighted means calculated from plant communities of different species richness in the observation data; since no uncertainty measure for these community-weighted means is calculated and propagated, the actual uncertainty of functional community signals for the single community observations is unknown.


### 5.2 Reported uncertainties
For this indicator, we report 95% confidence intervals as inferential uncertainty based on a statistical model (zero-one inflated beta-regression).

## 6. References
Artsdatabanken (2023, 11. august). Fremmede arter i Norge - med økologisk risiko 2023. http://www.artsdatabanken.no/lister/fremmedartslista/2023.

Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

Evju, M., Olsen, S.L., Prestø, T., Vange, V., Bratli, H. & Töpper, J. 2023. Ecological condition of naturally open areas below the forest line. Background, proposed indicators, and knowledge needs. NINA Report 2341. Norwegian Institute for Nature Research.

Framstad, E., Kolstad, A. L., Nybø, S., Töpper, J. & Vandvik, V. 2022. The condition of forest and mountain ecosystems in Norway. Assessment by the IBECA method. NINA Report 2100. Norwegian Institute for Nature Research.

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E. (2020). Towards a systematics of ecodiversity: The EcoSyst framework. Global Ecology and Biogeography, 29(11), 1887-1906. doi:10.1111/geb.13164

Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik. V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2020. Setting reference levels and limits for good ecological condition in terrestrial ecosystems. Insights from a case study based on the IBECA approach. Ecological Indicators 116: 106492.

Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2021. An index-based assessment of ecological condition and its links to international frameworks. Ecological Indicators 124: 107252.

Naas, A.E., Halvorsen, R., Horvath, P., Wollan, A.K., Bratli, H., Brynildsrud, K. et al. (2023) What explains inconsistencies in field-based ecosystem mapping? Applied Vegetation Science, 26, e12715. Available from: https://doi.org/10.1111/avsc.12715

Nybø, S. & Evju, M. (red.) 2017. Fagsystem for fastsetting av god økologisk tilstand. Forslag fra et ekspertråd, https://regjeringen.no/no/dokument/rapportar-og-planar/id438817/.

Tingstad, L., Evju, M., Sickel, H., & Töpper, J. 2019. Utvikling av nasjonal arealrepresentativ naturovervåking (ANO). Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag. NINA Rapport 1642.

Töpper, J., Velle, L.G. & Vandvik, V. 2018. Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition). NINA Report 1529b. Norwegian Institute for Nature Research.

Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923

Ullerud HA, Bryn A, Halvorsen R, Hemsing LØ. Consistency in land-cover mapping: Influence of field workers, spatial scale and classification system. Appl Veg Sci. 2018; 21: 278–288. https://doi.org/10.1111/avsc.12368


## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

### 7.1 Dataset A
Import ANO testing data:
First, download from  Miljødirektoratet...
```{r, eval = F, echo = F}
# download 
url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", mode="wb") 
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb")
```

...then upload from local storage
The geodatabase contains several layers; 'ANO_SurveyPoint' contains site and point data including alien plant species cover.
```{r}
#Sys.setlocale("LC_ALL", "no_NB.utf8") #works with æøå or use "Norwegian"
ANO.sp<- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb",
                   layer="ANO_Art", quiet = T)
ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb",
                   layer="ANO_SurveyPoint", quiet = T)
#head(ANO.sp)
#head(ANO.geo)

```

### 7.2. Dataset B
Import Reference data from NiN:
```{r}
# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
# str(Eco_State)
```

The generalized species lists underlying the ecosystem categorization in NiN (Halvorsen et al. 2020) represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the ‘expected combination of frequency and cover’ of occurrence

1 = < 1/32

2 = 1/32 - 1/8

3 = 1/8 - 3/8

4 = 3/8 - 4/5

5 = 3/8 - 4/5 + dominance

6 = > 4/5

For the purpose of this project, these steps are simplified to maximum expected combination of frequency and cover, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.



### 7.3. Dataset C
Import plant indicator values from Tyler et al. (2021):
```{r}
ind.Tyler <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/ind.Tyler.RDS")
```

The Swedish plant indicator values dataset published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we use indicator data for light, heat requirement, moisture, pH, nitrogen, phosphorus, grazing/mowing, and soil disturbance as these are thought to be subject to potential change due to ongoing pressures in the respective ecosystems (see details above under 'Summary').


## 8. Spatial units
The spatial units for functional plant community indicators are governed by the ANO data, which consist of 1000 randomly chosen sites in Norway. Each site is a 500 x 500 m grid cell with 18 monitoring points of 250sqm and a central 1 x 1 m vegetation plot. The basic unit for which observational community data exist, and for which every functional plant community indicator is computed, is this central 1 x 1 m vegetation plot. Aggregation of these basic spatial units to higher ones - like sites, municipalities, counties, regions, or the national level - should be done with appropriate consideration of the spatial structure of the data, i.e. imbalances and clustering (highly unequal number of points of an ecosystem type between sites).


## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [Initials]
* Methodology: [Initials]
* Data curation: [Initials]
* Formal analysis: [Initials]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [Initials]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
