---
title: "Vegetation indicator for nitrogen"
subtitle: |
  [NO_FUNI_001] (open lowland) 
indicatorID: 
  - [NO_FUNI_001]
indicatorName: 
  - Vegetation indicator for nitrogen
verbatimName: 
  - Vegetasjonsindikator for nitrogen # the indicator name that us used in the native language
folderName: NO_FUNI_001 # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
ECT: A2 - Chemical State characteristics
Realm: # Please copy and paste Realms, Biomes and Ecostsem names from here: https://github.com/NINAnor/ecRxiv/wiki#ecosystem-typology 
  - Terrestrial (T)
Biome: 
  - T3 Shrublands and shrubby woodlands biome
  - T4 Savannas and grasslands biome
  - T7 Intensive land-use biome
Ecosystem: 
  - T4.5 Temperate grasslands
  - T3.4 Rocky pavements
  - T3.2 Seas dry temp shrublands
  - T7.5 Semi-natural old fields
verbatimEcosystem:
  - åpen lavland
yearAdded: 2026
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_FUNI_001
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Scale - Aggregate (zero-one inflated beta-regression) 
Precision: 95% confidence intervals for the true value of the spatially aggregated indicator values derived from the beta-regression standard errors. # What kind of quantitative unceratinties are reported alongside the point estimates. Can be either inferrential (i.e. bootstrapped confidence intervals or standard errors) or descriptive (variation, standard deviation or percentiles from the distribution of the raw indicator values).
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: false
author:
  - name: Francesca Jaroszynska              # Enter name
    email: francesca.jaroszynska@nina.no  # Enter email
    affiliations:
      - id: myID
        name: Norwegian institute for nature research # Enter affiliations
  - name: Joachim Töpper
    affiliations:
      - ref: myID     
AuthorList: Jaroszynska, F. and Töpper, J.
date: September 14, 2026
data_availability: silver
code_reproducibility: silver
open_science_badge: silver
callout-icon: false
lightbox: true
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE 
bibliography: references.bib
codeReviewers:
  - name: Aud Halbritter
    email: aud.halbritter@uib.no
    affiliations:
      - id: reviewerInst
        name: University of Bergen
    reviewType: code review
    github: 
---
  
<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->
  
<!--# Load all you dependencies here -->
  
```{r setup}
#| include: false
#| cache: false
# load packages
library(tidyverse)
library(pwr)
library(lme4)
library(simr)
library(readxl)
library(formattable)
library(sf)
library(tmap)
library(WorldFlora)
library(zen4R)
library(tidylog)
library(rgbif)
library(here)
library(ecTools)
library(glmmTMB)
library(knitr)
library(kableExtra)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(tidylog::filter)
conflicts_prefer(tidylog::select)
conflicts_prefer(tidylog::mutate)
conflicts_prefer(tidylog::rename)
conflicts_prefer(tidylog::semi_join)
conflicts_prefer(tidylog::left_join)
conflicts_prefer(tidylog::full_join)
conflicts_prefer(tidylog::distinct)
conflicts_prefer(tidylog::summarise)
conflicts_prefer(tidylog::group_by)
conflicts_prefer(tidylog::pivot_longer)
conflicts_prefer(tidylog::pivot_wider)
conflicts_prefer(tidylog::count)
pull<-dplyr::pull
library(arrow)
library(gt)
knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->
  
```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
  
  
> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`


::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```

:::
  
::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
Presentation of Nitrogen indicator for ANO, GRUK and ASO datasets

:::
  
<hr />
  
<!--# Document you work below.  -->
  
## 1. Summary
  
1. **Background and rationale**: 
Nitrogen is a limiting macronutrient in many plant communities and terrestrial ecosystems. Fertilisation by nitrogen, for example from manual application or airborne deposition, can increase plant productivity, change the community composition, and most likely also decrease the species diversity. This indicator represents plant community responses to increased nitrogen. It represents the  deviations from an optimal state and towards higher nitrogen levels. 

2. **Methods**:
The data originate from field sampled plant community data collected through three monitoring programmes: 1. the Norwegian nature monitoring program ANO (Tingstad et al. 2019), 2. GRUK, an open calcareous grasslands monitoring programme (Evju & Vang 2026), and finally 3. the ASO monitoring programme (Bär et al. 2025). We combine these with plant functional trait values from Tyler et al. (2021). Indicator values are based on calculating community-weighted means (Diekmann 2003, Töpper et al. 2018) or cumulative abundance (see Framstad et al. 2022) of functional trait values for plant community data. Deviations in such an indicator beyond a certain range of indicator values indicate a directional shift in the plant community. As reference communities we use ‘generalized species lists’ underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020, see Evju et al. 2023 for naturally open lowland ecosystems). For each major ecosystem type a separate set of indicators, related to the main drivers of change today, is computed. For open lowland ecosystems these are the vegetation indicators for nitrogen, light, moisture, pH, phosphorus, mowing/grazing, and soil disturbance. This file documents the 1-sided nitrogen indicator for open lowland.

3. **Interpretation**:
An indicator value of 1 indicates no discernible difference in the functional plant composition due to impacts of increased nitrogen.
Values decreasing toward zero reflect a shift in species composition and community trait values away from the reference state as a response to increased nitrogen.

4. **Results and development needs**:
The indicator for Norway is 0.74 with 95%CI of 0.59-0.84, which means that the condition for this indicator is neither significantly lower nor higher than 0.6, the threshold for good ecological condition. The generally large spatial heterogeneity indicates that local results vary greatly across the country and within regions a range of sites showing changes towards communities with higher nitrogen affinity/tolerance. As for the regions, Northern Norway (0.66) and Central Norway (0.7) show clearly lower indicator values than the other regions (>0.77).

To improve this indicator in the future, the perhaps most pressing issue it to regionalise the generalised species list, as currently they have a bias towards southern and eastern Norway. The indicator values also need validation against independent datasets.

Access to further years of data for the ASO monitoring scheme would strengthen its temporal and spatial representation, and thus the indicator's robustness.



## 2. About the underlying data {#sec-about}
We use three types of data for building functional plant community indicators for ecological condition:
  
  - as observation data  we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019), the ASO monitoring scheme (Bär et al. 2025), and the GRUK monitoring scheme (Evju and Vang 2026).
  - as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (Halvorsen et al. 2020)
- Swedish functional plant indicator data from Tyler et al. (2021)

For details see point 7 below.

### 2.1 Spatial and temporal resolution and extent
The ANO monitoring data available now represent one cycle of sampling from 2019-2024 comprising ca. 1000 sites across Norway.
ANO has no defined end date. 2025 is the first year of repeated sampling for about 20% of the approximately 1000 ANO sites for all of mainland Norway (i.e. the Ecosystem Accounting Area). Annual sampling is spatially representative, in principle allowing to compute area-representative estimates for single years but also multiple-yr-reporting periods. The indicator at hand is calculated based on the ANO data from 2020-2024.

The GRUK monitoring data follows a similar five-year sampling cycle from 2020-2024. Monitoring sites were randomly selected across the Oslo fjord basin, resulting in ~80 sites surveyed annually. The sites comprise open, shallow-soil calcareous grassland in three sub-areas - the inner, middle and outer Oslo fjord region.

The ASO monitoring dataset is spatially representative of Norway, with each year's sampling pool representative at a national and regional scale. ASO follows a 5-year monitoring cycle, though only data from 2022 are represented here as species abundances were only accessible to us for that year.

### 2.2 Sampling design
At every ANO site, the plant community is recorded (presence and percent cover of every vascular plant species) in the central 1 sqm plot of each of systematically arranged 18 monitoring points.

Within each GRUK site, one or several polygons matching the NiN mapping units T2-C-7 and T2-C-8 were delineated, and 5, 10, or 15 0.5 x 0.5 m permanent plots established, randomly selected from a 10 x 10 m grid within the polygons. All vascular plant species were recorded, including their percentage cover (visual estimate). 

ASO is structured using a nested design consisting of ASO areas (10x10 km), ASO plots (500x500 m), and semi-natural meadows. Within each semi-natural meadow up to three transects were established, the length of each chosen to reflect the size of the meadow. 

### 2.3 Original units
Abundance data for vascular plants in ANO is given as % cover within a 1 m2 sample quadrat, and within a 0.5 x 0.5 m sample quadrat for GRUK. This is transformed to relative abundance data per quadrat. Each ANO and GRUK quadrat is classified to its respective ecosystem type according to ‘Nature in Norway’ (NiN; <https://artsdatabanken.no/NiN>). Species diversity in ASO was recorded using cover and frequency classes (cf. modified measurement scale A7 for NiN2). Environmental indicator values from Tyler et al. (2021) are given on an ordinal scale from 1 (lowest affinity to the respective variable) to n (highest affinity to the respective variable), the number of levels (n) varying between environmental indicators.




### 2.4 Instructions for citing, using and accessing data
The ANO data are openly available at https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054. See @sec-about for detail on the other datasets.  

### 2.5 Additional comments about the dataset
The ANO, ASO and GRUK programs are run by the Norwegian Environment Agency. On their web sites you may also find some additional summary data. https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/
https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/semi-naturlig-eng/
https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/overvaking-av-apen-grunnlendt-kalkmark/
  
## 3. Indicator properties
  
### 3.1 Ecosystem Condition Typology Class (ECT)
The functional plant community indicator for nitrogen reflects the availability of nutrients and soil chemical properties and thus represents the ECT class 'A2 Chemical characteristics'.

### 3.2 Ecosystem condition characteristic
The functional plant community indicator for nitrogen reflects the availability of nutrients and soil chemical properties and thus represents the characteristic 'Abiotic factors'.

### 3.3 Collinearities with other indicators
Different functional plant community indicators for different environmental variables (light, moisture, ...) might show a certain degree of co-linearity with one another as they are derived from the very same plant community data.
Functional plant community indicators can potentially exhibit co-linearities with other vegetation-oriented indicators, such as NDVI.


### 3.4 Impact factors
The plant community composition in open lowland ecosystems is subject to changes with climate warming and precipitation changes as well as pollution (including nitrogen fertilisation) and disturbance from human infrastructure.

### 3.5 Spatial aggregation
The variable is scaled, transformed (piece-wise linear) and truncated at the initial ANO-point level and the indicator estimate is a beta-regression-modelled mean of all ANO-, ASO, and GRUK-units.

## 4. Reference condition and levels

### 4.1 Reference condition
In Norway, the reference condition for natural ecosystems has been one of largely intact ecosystems without large-scale human modifications (Nybø and Evju 2017). For semi-natural ecosystems, human influence is allowed, and indeed a prerequisite, but then in the sense of appropriate, stable and traditional land-use practices. In terms of nitrogen, the reference condition is defined by a reference community.

### 4.2 Reference levels
Reference and limit values (scaling values) for this one-sided functional plant community indicator are derived as the 0.5 and 0.95 quantiles (reference value and upper limit value, respectively) from distributions of community weighted means of the respective plant indicators originating from bootstrapped generalized species lists.
Bootstraps were set up with resampling of the original generalized species lists 1000 times. In every sample a random 1/3 of the original generalized species list was sampled (as ANO points on average contain roughly a third of the potential species richness in the generalized species lists -> local variation in species composition thus is large and needs to be reflected in the bootstraps), with the exception of dominant species (abundance value of 5 or higher in the original scale of the generalized species lists), which were retained in every sample.
All reference distributions, and thus all scaling values are unique to NiN mapping units at a scale of 1:5000.

#### 4.2.1 Spatial resolution and validity
The same reference levels apply to all types of ecosystems and all regions.

## 5. Uncertainties
### 5.1 General uncertainties
Uncertainties relate partly to sampling and observer variability (and possible observer bias) in the recording of the plant cover data. Presumably, there is also uncertainty related to the scoring of indicator values for species, and this will probably vary among species (cf. Tyler et al. 2021). These potential sources uncertainty can only be assumed and not quantified.
Spatial variability as a source of uncertainty can be quantified. For every mean or median indicator value (after scaling) for every region (or any other delimited area of interest) we can calculate its corresponding standard error or confidence interval quantiles as a measure of spatial uncertainty for a geographical area.
There is some uncertainty related to the ecosystem categorization at the field sites (see Ullerud et al. 2018 and Naas et al. 2023), but since ANO monitoring does not include spatial delineation of ecosystem polygons this problem is considered smaller than in spatial ecosystem mapping which these references mainly relate to. Similarly, this analysis is not crucially dependent on the exact delineation of the polygons, and the plots and transects in GRUK and ASO placed and assessed within the polygons hold little uncertainty of not being the target types of the two monitoring programmes.
Other uncertainties connected to the indicator arise from community weighted means calculated from plant communities of different species richness in the observation data; since no uncertainty measure for these community-weighted means is calculated and propagated, the actual uncertainty of functional community signals for the single community observations is unknown.

There is some uncertainty about the appropriateness of using (Halvorsen et al. 2020) as a reference dataset and a reference plant community for different nature types, and especially for different geographical and environenmetal regions of Norway. The dataset has a known bias towards south and eastern Norway, and may not reflect the natural plant communities equally well in the western and northern parts.

### 5.2 Reported uncertainties
For this indicator, we report 95% confidence intervals as inferential uncertainty based on a statistical model (zero-one inflated beta-regression).

## 6. References
Artsdatabanken (2023, 11. august). Fremmede arter i Norge - med økologisk risiko 2023. http://www.artsdatabanken.no/lister/fremmedartslista/2023.

Bär, A., Solbu, E., Vassvik, L., Velle, L.G. & L. Johansen 2025. Arealrepresentativ Overvåking av Semi-naturlig eng (ASO). Resultater fra 1. omløp og anbefalinger for veien. (NIBIO RAPPORT, 11/149). Norsk institutt for bioøkonomi.  

Bär, Johansen, Vesterbukt, Svalheim, Daugstad, Hinderaker, Thorvaldsen, Albertsen, Carlsen, Grenne, Maurset, Øien, Heimstad, Lyngstad, Hasvik, Ombler, Wehn, Moen, Blütecher, Leh, Davidsen, Paulsen, Vassvik, Sundsdal, Velle, Solstad, Bele, Solbu, Winther, Henriksen, Nesheim-Hauge, Høydal (2025). Overvåking av semi-naturlig eng (ASO). Version 1.8. Norwegian Institute of Bioeconomy Research. Sampling event dataset https://doi.org/10.15468/gq6wa5 accessed via GBIF.org on 2026-09-11.

Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

Evju, M., Olsen, S.L., Prestø, T., Vange, V., Bratli, H. & Töpper, J. 2023. Ecological condition of naturally open areas below the forest line. Background, proposed indicators, and knowledge needs. NINA Report 2341. Norwegian Institute for Nature Research.

Evju M, & Vang R 2026. Overvåking av åpen grunnlendt kalkmark i Oslofjordområdet. Version 1.20. Norwegian Institute for Nature Research. Occurrence dataset. https://ipt.nina.no/resource?r=gruk_2020&v=1.20

Framstad, E., Kolstad, A. L., Nybø, S., Töpper, J. & Vandvik, V. 2022. The condition of forest and mountain ecosystems in Norway. Assessment by the IBECA method. NINA Report 2100. Norwegian Institute for Nature Research.

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E. (2020). Towards a systematics of ecodiversity: The EcoSyst framework. Global Ecology and Biogeography, 29(11), 1887-1906. doi:10.1111/geb.13164

Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik. V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2020. Setting reference levels and limits for good ecological condition in terrestrial ecosystems. Insights from a case study based on the IBECA approach. Ecological Indicators 116: 106492.

Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2021. An index-based assessment of ecological condition and its links to international frameworks. Ecological Indicators 124: 107252.

Naas, A.E., Halvorsen, R., Horvath, P., Wollan, A.K., Bratli, H., Brynildsrud, K. et al. (2023) What explains inconsistencies in field-based ecosystem mapping? Applied Vegetation Science, 26, e12715. Available from: https://doi.org/10.1111/avsc.12715

Nybø, S. & Evju, M. (red.) 2017. Fagsystem for fastsetting av god økologisk tilstand. Forslag fra et ekspertråd, https://regjeringen.no/no/dokument/rapportar-og-planar/id438817/.

Tingstad, L., Evju, M., Sickel, H., & Töpper, J. 2019. Utvikling av nasjonal arealrepresentativ naturovervåking (ANO). Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag. NINA Rapport 1642.

Töpper, J., Velle, L.G. & Vandvik, V. 2018. Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition). NINA Report 1529b. Norwegian Institute for Nature Research.

Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923

Ullerud HA, Bryn A, Halvorsen R, Hemsing LØ. Consistency in land-cover mapping: Influence of field workers, spatial scale and classification system. Appl Veg Sci. 2018; 21: 278–288. https://doi.org/10.1111/avsc.12368


## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

### 7.1. Dataset A
Import plant indicator values from Tyler et al. (2021):
```{r}
#ind_tyler <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/ind.Tyler.RDS")
ind_tyler <- readRDS(paste0(here(), "/data/indicators/ind.Tyler.RDS"))
```

The Swedish plant indicator values dataset published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we use indicator data for mowing and grazing as this is thought to be subject to potential change due to ongoing pressures in the respective ecosystems (see details above under 3.4 'Impact factors').


### 7.2 Dataset B
   backbone
```{r, eval = F, echo = F}
# download WFO plant list from zenodo
#download_zenodo("https://doi.org/10.5281/zenodo.20782718", files = "_DwC_backbone_R.zip", path = "/Users/francesca.jaroszynsk/OneDrive - NINA/nina_projects/ANO/lowlands/ecRxiv_veg_indicators/data")

wfo_backbone <- read_rds("P:/41201785_okologisk_tilstand_2022_2023/data/WFO_backbone/wfo_backbone.RDS")
```

### 7.3. Dataset C
Import Reference data from NiN:
```{r}

```

The generalized species lists underlying the ecosystem categorization in NiN (Halvorsen et al. 2020) represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the ‘expected combination of frequency and cover’ of occurrence

1 = < 1/32
2 = 1/32 - 1/8
3 = 1/8 - 3/8
4 = 3/8 - 4/5
5 = 3/8 - 4/5 + dominance
6 = > 4/5

For the purpose of this project, these steps are simplified to maximum expected combination of frequency and cover, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
  
### 7.4 Dataset D
Import ANO testing data:
First, download from  Miljødirektoratet (downloaded spring 2026).
```{r, eval = F, echo = F}
# download 
#url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
#download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", mode="wb") 
#unzip("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb")
```

...then upload from local storage
The geodatabase contains several layers; 'ANO_SurveyPoint' contains site and point data including alien plant species cover.
```{r loadData}
#Sys.setlocale("LC_ALL", "no_NB.utf8") #works with æøå or use "Norwegian"
source(paste0(here(), "/R/cleaning_ano.R"))

#ano_species <- read_rds(file = paste0(here(), "/data/cache/ano_species.RDS"))
#ano_geo <- read_rds(file = paste0(here(), "/data/cache/ano_geo.RDS"))

#head(ano_species)
#head(ano_geo)

```



### 7.5 Dataset E
Import GRUK data
```{r, eval = FALSE, echo = F}


```

### 7.6 Dataset F
  Import ASO data

```{r, eval = F, echo = F}

```


## 8. Spatial units
The spatial units for functional plant community indicators are governed by the ANO data, which consist of 1000 randomly chosen sites in Norway. Each site is a 500 x 500 m grid cell with 18 monitoring points of 250sqm and a central 1 x 1 m vegetation plot. The basic unit for which observational community data exist, and for which every functional plant community indicator is computed, is this central 1 x 1 m vegetation plot. Aggregation of these basic spatial units to higher ones - like sites, municipalities, counties, regions, or the national level - should be done with appropriate consideration of the spatial structure of the data, i.e. imbalances and clustering (highly unequal number of points of an ecosystem type between sites).

## 9. Analyses

<!--# 
  
  Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->
  
#### Data handling
  - Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)


```{r spatial_data, warnings = FALSE}
```


```{r dataHandling_Tyler, warning=F, message=F, results='hide', eval = FALSE} 


# 2.0 Plant indicator data


```

```{r dataHandling_GRUK, warning = FALSE, eval = FALSE}

# 2.1 GRUK data


```

```{r dataHandling_ASO, warning = FALSE, eval=FALSE}
# 2.2 ASO

```

```{r dataHandling_ANO, warning = FALSE, eval = FALSE}
#### 2.3 ANO monitoring data

```

```{r dataHandling_NiN, warning = FALSE, eval = FALSE}


rm(wfo_backbone)

```

This leaves us with the monitoring data including plant indicators (ano_all, gruk_all, aso_all) and the reference data including plant indicators (nin_sp_ind):
  
  <!-- Print head of data set with horizontal scrolling -->
```{r ano-sp-tab, echo=F}
head(ANO.sp.ind) |>
  kable("html",caption = "Species occurrence data set with attached plant trait data.") |> 
  kable_styling("striped") |> scroll_box(width = "100%")
```

```{r wetland_mountain_forest_seminatural-ref-data-tab, echo=FALSE}
head(nin_sp_ind) |> 
  kable("html", caption = "Reference data set for semi-natural ecosystems") |> 
  kable_styling("striped") |> scroll_box(width = "100%")

```

<br />
  
  For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
In order to get distributions of these metrics rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

<br />
  
##### Bootstrap functions for frequency abundance and cumulative density
  - function to calculate community weighted means (indBoot.freq() ) or cumulative density (indBoot.HeatOverhang() ) of selected indicator values (ind)
- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
- with a given number of iterations (iter),
- with species given a certain minimum abundance occurring in all bootstraps (obl), and
- with a given re-sampling ratio of the original species list (rat)
- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)

```{r bootstrapping}


```

```{r}
#colnames(NiN.forest.cov)
# 1st column is the species
# 2nd-27th column is the abundances of sp in different ecosystem types
# 28th-29th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 0.8 or more (dominant species) must be included in each sample
# each sample re-samples 1/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
```

Running the bootstraps:
```{r boot2, eval = F}




```

```{r, include = F}
# Data from cache
lowland_ref_cov <- readRDS(paste0(here::here(), "/data/cache/lowland_ref_cov.RDS"))
```

```{r}
head(lowland_ref_cov[[1]]) |>
  kable("html", caption = "Table showing the first 6 rows of the bootstrapped data set.") |> kable_styling("striped") |> scroll_box(width = "100%")
```

This results in an R-list for forest ecosystem, with a data frame for every selected plant indicator and each data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.

Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:
  
  - Median = reference values
- 0.025 and 0.975 quantiles = lower and upper limit values
- min and max of the respective indicators scale = min/max values



```{r scalingValues, attr.output='style="max-height: 300px;"', results='hide'}

# every NiN-type is represented by one 'generalisert artsliste'
# some NiN-types are represented by two such species lists
# in some cases two NiN-types are represented by the same species list

```

```{r}
lowland_ref_cov_val |>
  kable("html", caption = "Reference values (Rv and threshold values (Gv) for each indicator and nature type combination") |> kable_styling("striped") |> scroll_box(width = "100%", height = "300px")
```


```{r merging datasets}

```

Once test data and the scaling values from the reference data are in place, we can calculate CWMs of the selected indicator for the community data for each dataset and scale them against the scaling values from the reference distribution. Note that we scale each plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type. A plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

For scaling the data against the reference we use the normalisation function from package ecTools.

```{r}
#| warning: false

#### 1. calculating scaled and non-truncated values for the indicators based on the datasets ####
ref_grunn <- unique(sub("\\-.*", "", as.character(lowland_ref_cov_val$grunn)))
combined_species <- combined_species |> filter(!is.na(Nitrogen))
lowland_ref_cov_val <- lowland_ref_cov_val |> filter(Ind %in% c("Nitrogen1", "Nitrogen2"))


for (i in seq_len(nrow(combined_lowlands))) {
  tryCatch({
    
    ds       <- combined_lowlands$dataset[i]
    hoved_i  <- as.character(combined_lowlands$hovedtype_rute[i])
    global_i <- as.character(combined_lowlands$globalid_chr[i])
    nin_i    <- as.character(combined_lowlands$kartleggingsenhet[i])
    
    # 1) hovedtype must exist in ref
    cond_hoved <- !is.na(hoved_i) && hoved_i %in% ref_grunn
    if (!isTRUE(cond_hoved)) next
    
    # 2) species data for this row and dataset
    dat <- combined_species |>
      dplyr::filter(
        dataset == ds,
        parentglobalid == global_i
      ) |>
      dplyr::select(art_dekning, Nitrogen)
    
    if (nrow(dat) == 0) next
    
    combined_lowlands$richness[i] <- nrow(dat)
    dat <- dat[!is.na(dat$Nitrogen), ]
    if (nrow(dat) == 0) next
    
    # 3) CWM Nitrogen
    val <- sum(dat$art_dekning * dat$Nitrogen, na.rm = TRUE) /
      sum(dat$art_dekning, na.rm = TRUE)
    
    ## Nitrogen1
    idx1 <- lowland_ref_cov_val$Ind == "Nitrogen1" &
      lowland_ref_cov_val$grunn == nin_i
    idx1[is.na(idx1)] <- FALSE
    
    if (sum(idx1) == 1) {
      ref1    <- as.numeric(lowland_ref_cov_val$Rv[idx1][[1]])
      lim1    <- as.numeric(lowland_ref_cov_val$Gv[idx1][[1]])
      maxmin1 <- as.numeric(lowland_ref_cov_val$maxmin[idx1][[1]])
      
      if (!any(is.na(c(ref1, lim1, maxmin1)))) {
        combined_lowlands$Nitrogen1[i] <- ec_normalise(
          variable = val,
          x0   = maxmin1,
          x60  = lim1,
          x100 = ref1,
          fun  = "linear"
        )
      }
    }
    
    ## Nitrogen2
    idx2 <- lowland_ref_cov_val$Ind == "Nitrogen2" &
      lowland_ref_cov_val$grunn == nin_i
    idx2[is.na(idx2)] <- FALSE
    
    if (sum(idx2) == 1) {
      ref2    <- as.numeric(lowland_ref_cov_val$Rv[idx2][[1]])
      lim2    <- as.numeric(lowland_ref_cov_val$Gv[idx2][[1]])
      maxmin2 <- as.numeric(lowland_ref_cov_val$maxmin[idx2][[1]])
      
      if (!any(is.na(c(ref2, lim2, maxmin2)))) {
        combined_lowlands$Nitrogen2[i] <- ec_normalise(
          variable = val,
          x0   = maxmin2,
          x60  = lim2,
          x100 = ref2,
          fun  = "linear"
        )
      }
    }
    
  }, error = function(e) {
    cat("ERROR at row", i, ":", conditionMessage(e), "\n")
  })
}


combined_lowlands <- combined_lowlands |>
  dplyr::mutate(
    Nitrogen  = Nitrogen2
  )



```

```{r}
head(combined_lowlands) |>
  kable("html", caption = "Plots for ecosystems with scaled indicator values in the far-right columns.") |> 
  kable_styling("striped") |> scroll_box(width = "100%")
```


We can visualise the normalisation with a generic figure (@fig-norm).

```{r fig-norm}
#| fig-cap: 'A figure illustrating the normalisation process.The placement of the x60 values are arbitrary in this figure and does not give a representative picture of the relationship between the variable and indicator values.'

var <- seq(0, 100, by = 0.5)
var2 <- seq(101, 200, by = 0.5)

ind <- ec_normalise(var, x0 = 0, x100 = 100, x60 = 50)
ind2 <- ec_normalise(var2, x100 = 101, x0 = 200, x60 = 150)

ec_norm_plot(
  variable = c(var, var2),
  indicator = c(ind, ind2)) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200),
    labels = c("lowest possible\nvalue", "2.5%", "50%", "97.5%", "highest possible\nvalue"),
    expand = expansion(add = 30)
  ) +
  labs(x = "Variable values as percentiles of \nbootstrapped reference values")
```

```{r}
#| eval: false
#| include: false

ec_norm_plot(
  variable = c(var, var2),
  indicator = c(ind, ind2)
) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200),
    labels = c("lowest possible\nvalue", "2.5%", "50%", "97.5%", "highest possible\nvalue"),
    expand = expansion(add = 30)
  ) +
  labs(x = "Variabelverdier som persentiler ift. \nbootstrappede referanse verdier",
       y = "Indikatorverdi")

ggsave(here::here("img/norm_plot.png"))
```

#### Calculating national and regional indices

We can calculate national and region-wise indicator values using a mixed-effects, zero-one inflated beta-regression analysis. Since the indicator values are on a scale from 0-1 the data per definition dont follow a Gaussian distribution, hence the zero-one inflated beta-regression. Since the underlying monitoring data are spatially aggregated (several points per site), not all points are independent, hence the mixed-effects model.

The uncertainty around the national and region indicator values is derived from the same statistical model.

The national and regional estimates are computed separately.

```{r tmbs}
#| warning: false
## Fit ordered beta regression null-model with a random effect
# Norway

Nitrogen_no <- glmmTMB(
  Nitrogen ~ 1 + (1|dataset/nest_1),
  data = combined_lowlands,
  family = ordbeta
)


summary(Nitrogen_no)

# regions
Nitrogen_re <- glmmTMB(
  Nitrogen ~ 0 + region + (1|dataset/nest_1),
  data = combined_lowlands,
  family = ordbeta
)


summary(Nitrogen_re)
```


## 10. Results

The model estimates and standard errors are on a logit scale, but we can calculate indicator values and confidence intervals (CI) by transforming back from logit to values between 0-1
```{r}

NO_FUNI_001_table <- tibble(
  Region = c(
    "Norway",
    "Central Norway",
    "Eastern Norway",
    "Northern Norway",
    "Southern Norway",
    "Western Norway"
  ),
  indicator_value = c(
    plogis(summary(Nitrogen_no)$coef$cond[, 1]),
    plogis(summary(Nitrogen_re)$coef$cond[, 1])
  ),
  CI_upper = c(
    plogis(
      summary(Nitrogen_no)$coef$cond[, 1] + 1.96 * summary(Nitrogen_no)$coef$cond[, 2]
    ),
    plogis(
      summary(Nitrogen_re)$coef$cond[, 1] + 1.96 * summary(Nitrogen_re)$coef$cond[, 2]
    )
  ),
  CI_lower = c(
    plogis(
      summary(Nitrogen_no)$coef$cond[, 1] - 1.96 * summary(Nitrogen_no)$coef$cond[, 2]
    ),
    plogis(
      summary(Nitrogen_re)$coef$cond[, 1] - 1.96 * summary(Nitrogen_re)$coef$cond[, 2]
    )
  )
)
NO_FUNI_001_table

#write.table(
#    NO_FUNI_001_table,
#    file = paste0(here(), "/data/NO_FUNI_005_table.txt"),
#    sep = "\t",
#    row.names = FALSE,
#    quote = FALSE)

```

To express the results as a distribution, we need to simulate mean estimates based on the models intercept and associated standard error.
```{r}

## Norway
# Fixed-effect intercept and SE
b0 <- fixef(Moisture_no)$cond["(Intercept)"]
se_b0 <- sqrt(vcov(Moisture_no)$cond["(Intercept)", "(Intercept)"])
# Dispersion parameter (For glmmTMB this is usually on the response/positive scale via sigma() )
phi <- sigma(Moisture_no)


# draw estimates and backtransform from logit-scale
eta_draws <- rnorm(1000, mean = b0, sd = se_b0)
mu_draws  <- plogis(eta_draws)


## regions
b_re <- fixef(Moisture_re)$cond
vc_re <- vcov(Moisture_re)$cond
phi <- sigma(Moisture_re)

eta_draws_reg <- MASS::mvrnorm(
  n = 1000,
  mu = b_re,
  Sigma = vc_re
)

mu_draws_reg <- plogis(eta_draws_reg)


# combining Moisture_no_sim and Moisture_re_sim
mu <- mu_draws_reg |>
  as_tibble() |>
  add_column(Norway = mu_draws) |>
  pivot_longer(everything(), names_to = "part") |>
  mutate(part = case_when(
    part == "regionCentral_Norway" ~ "C",
    part == "regionEastern_Norway" ~ "E",
    part == "regionNorthern_Norway" ~ "N",
    part == "regionWestern_Norway" ~ "W",
    part == "regionSouthern_Norway" ~ "S",
    .default = part
  )) #|>
  #add_column(year = "2024")
```


```{r tbl-results}
#| tbl-cap: 'Indicator values and 95% confidence intervals for NO_FUMO_005, calculated for Norwegian open lowland ecosystems.'
mu |>
  group_by(part) |>
  summarise(
    indicator = mean(value),
    low = round(quantile(value, probs = 0.025), 2),
    high = round(quantile(value, probs = .975), 2)
  ) |>
  unite(col = "95% CI", c(low, high), sep = " - ") |>
  gt::gt()
```

## 11. Export file

<!--# 
  
  Optional: Display the code (dont execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
arrow::write_parquet(mu,"data/results_NO_FUMO_004.parquet")
```

Demonstrating how to read back the data:

```{r readingBack}
arrow::read_parquet(here::here("data/results_NO_FUMO_004.parquet"))
```

## 12. Author contributions



**CRediT statement**

* Conceptualization: [JT]
* Methodology: [JT,FJ]
* Data curation: [FJ]
* Formal analysis: [FJ,JT]
* Investigation: [FJ,JT]




**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [JT].

**Data processing**
Raw data were curated and cleaned by [FJ].

**Statistical analysis**
All analyses were conducted by [FJ and JT] , with verification and minor edits by [AH].

**Visualization**
Figures were produced by [FJ].

**Reproducibility**
All code was reviewed and validated by [AH].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This workflow uses renv. Run `renv::restore()` to get started.

:::