---
title: "Glacier area"
subtitle: "[NO_GLAC_001]"
indicatorID: 
  - NO_GLAC_001
indicatorName: 
  - Glacier area
verbatimName: Breareal
folderName: NO_GLAC_001
country: Norway
continent: Europe
ECT: C1 - Landscape and Seascape Characteristics
Realm: 
  - Terrestrial
Biome: 
  - T6 Polar-alpine (cryogenic) biome
Ecosystem: 
  - T6.1 Permanent snow
verbatimEcosystem: Fjell
yearAdded: 2025 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "1.0.0." # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: Last version was 000.003. This update has removed redundtant sensitivity analyses and simplified the workflow.
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_GLAC_001
Normalised: Yes
SpatialAggregationPathway: 'sp. agg (geographic region) - [scale - trunc. - trans.] - sp. agg. (Norway, weighted by initial glacier area)'
Uncertainties: Inferrential
AuthorList: Kolstad, A.
data_availability: silver
code_reproducibility: silver
open_science_badge: silver
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Anders L. Kolstad
    email: anders.kolstad@nina.no
    affiliations:
      - id: myID
        name: Norwegian Institute for Nature Research 
date: May 15, 2025
callout-icon: false
lightbox: true
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
hide: FALSE
---

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(sf)
knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
-   March 26, 2025 - Original PR
-   April 25, 2025 - New reference levels
-   May 15, 2025 - revision following code review
-   November 25, 2025 - restructure (moving metadata into top yaml) and add open science badges (ALK)
-   June 5, 2026 - New version
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->
Glaciers represents important structural elements in alpine ecosystems,
providing melt water and local climate control. It is also a unique
nature type and habitat in itself, with few, but very specialised, species living on
or in association with the ice. Glaciers are permanently ice covered
areas and they are sensitive to climate change, especially climate
warming, which melts the ice. 
2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->
In Norway we have a rather detailed and
precise map of the glacier cover or extent from the period 1947-1985. Here I use
this data as a representation of the climatic reference condition (pre
climate warming), and compare this with an updated glacier cover map
from 2018-2019 to create a normalised ecosystem condition indicator. 
Glacier area and glacier area change is calculated for each of the five geographical regions in Norway, and for Norway as a whole. The aggregation from the regional to the national estimate is based on a weighted average using the initial glacier area as weights. 
Indicator uncertainty originates from two sources: a 3% measurement error, 
based on the reported uncertainties
in the mapping method, and the spatial variation across strata.


3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
The indicator value represents the negative ecological effect of melting glacier ice. 
Most glaciers are found in western and northern Norway, and due to the way the regions are weighted by their initial glacer extent, the national estimate is determined mainly by the changes in glacer area in these regions. 
The indicator does NOT directly represent the area of remaining glacier ice.
An indicator value of 1 signifies that the glacier extent is equal to, or greater then it was in the reference condition, defined by the climatic period between 1947 and 1985. 
Values decreasing toward zero reflect a reduction an the positive impacts of glacier and melt water, driven by the reduction in glacer area and volume.

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->
The indicator value for Norway as a whole in 2024 (using data from 2018-2019) is 0.62 [95% CI = 0.58-0.66], which indicates that glacier area and volume have been reduced considerably, to the point where this change starts to negtatively effects the integrity of mountain ecosystems. 
Future updates will require new maps of glacier cover in Norway, which is highly possible to produce from satellite imagery.

## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: 
Isbreer utgjør viktige strukturelle elementer i alpine økosystemer ved å tilføre smeltevann og bidra til lokal klimaregulering. De representerer også en unik naturtype og et eget habitat, med få, men høyt spesialiserte arter som lever på eller i tilknytning til isen. Isbreer er permanent isdekte områder og er svært følsomme for klimaendringer, særlig klimaoppvarming, som fører til smelting av isen.


2. **Metoder**: 
I Norge har vi relativt detaljerte og presise kart over bredekning fra perioden 1947–1985. Disse dataene brukes her som en representasjon av den klimatiske referansetilstanden (før den markerte klimaoppvarmingen), og sammenlignes med oppdaterte kart over bredekning fra 2018–2019 for å beregne en normalisert indikator for økosystemtilstand.
Breal og endringer i breareal beregnes for hver av de fem geografiske regionene i Norge, samt for landet som helhet. Aggregeringen fra regionale til nasjonale estimater er basert på et vektet gjennomsnitt, der det opprinnelige brearealet brukes som vekter.
Usikkerheten i indikatoren har to hovedkilder: en målefeil på 3 %, basert på rapporterte usikkerheter i kartleggingsmetodene, og den romlige variasjonen mellom regionene.

3. **Tolkning**: 
Indikatorverdien representerer de negative økologiske effektene av smeltende breis. De fleste isbreene finnes i Vest-Norge og Nord-Norge, og fordi regionene vektes etter sitt opprinnelige breareal, er det de observerte endringene i disse regionene som i størst grad bestemmer den nasjonale indikatorverdien.
En indikatorverdi på 1 betyr at breutbredelsen er lik eller større enn den var under referansetilstanden, definert som den klimatiske perioden mellom 1947 og 1985. Verdier som nærmer seg null reflekterer en reduksjon i de positive økologiske effektene av breer og smeltevann, drevet av nedgang i breareal og brevolum.
Indikatoren representerer ikke direkte hvor stort areal som fortsatt er dekket av breis.

4. **Resultater og utviklingsbehov**: 
Indikatorverdien for hele Norge i 2024 (basert på data fra 2018–2019) er 0,62 [95 % KI = 0,58–0,66]. Dette indikerer at breareal og brevolum er betydelig redusert, til et nivå der endringene begynner å påvirke integriteten til fjelløkosystemene negativt.
Fremtidige oppdateringer av indikatoren vil kreve nye kart over bredekningen i Norge. Slike kart kan med trolig produseres ved hjelp av satellittdata.


## 2. About the underlying data

<!--# Describe the data you have used in more detail, it's origin, biases, availabilit ect.-->

I mainly make use of two datasets on glacier extent in Norway. One
covering the years 1947 to 1985 [@winsvold2014], and one covering
2018-2019 [@andreassen2022]. The first dataset is made to represent the
climatic reference condition (pre-climate change). It as a digitalised
version of the N50 map, made based on the interpretation of aerial
photos. The newer map is a classification based on Sentinel imagery. The
uncertainties in both maps is said to be 3% (interpreted here as sd =
mean × 0.03), but these are several generalisations embedded in this
single number.

### 2.1 Spatial and temporal resolution and extent

<!--# Describe the temporal and spatial resolution and extent of the data used -->

The data covers all of mainland Norway (defined here as the ecosystem
accounting area, EAA), and two time steps (see above).

### 2.2 Sampling design

The sampling design is a complete sampling (wall-to-wall).

### 2.3 Original units

<!--# What are the original units for the most relevant  variables in the data-->

The original data is a map with presence or absence of glaciers. 
Early in the workflow, I summarise these as summed area (m^2^) per strata.

### 2.4 Instructions for citing, using and accessing data
The glacier cover maps are distributed under CC-BY 4.0. We have made a
simplified (less columns) version of the dataset available on github in
order to make the workflow more reproducible.

### 2.5 Additional comments about the dataset

<!--# Text here -->

I know of no concrete plans to update the glacier cover map, but with
of remote sensing, this is indeed possible to to, and quite
easy as well probably.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

Area of glaciers is categorised as describing a *A1 - Physical state characteristic*. on the importance of melt water, and hence glacer volume.
The indicator has previous been under C1 - Landscape characteristics, with a stronger emphasis on glaciers as landscape elements and a habitat.


### 3.2 Ecosystem condition characteristic

The indicator is meant to represent a characteristic of ecosystems in good condition where the climate is as it was under the normal climate period 1961-1990, and where the extent, volume, and hence ecological importance of glaciers is in equilibrium with this climate.

### 3.3 Collinearities with other indicators

The indicator is in many ways unique, but will likely be collinear with
other indicators that are similarly affected by climate warming, suce as snow depth.

### 3.4 Impact factors

Glaciers are mainly affected by climate change, especially climate
warming [@zemp2025].

### 3.5 Spatial aggregation
The map data is prepared (clipped) to the spatial scale of five geographic regions in Norway. Then the variable is normalised before being aggregated to a national estimate.
$$
sp. agg (geographic region) - [scale - trunc. - trans.] - sp. agg. (Norway)
$$

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition of mountain ecosystems are defined, in general
terms, as one with minimal human impact. For the climate, the definition
is more precisely defined as the time period 1961-1990, which is the
last 30 year normal period before climate change became really
prominent.

### 4.2 Reference levels

The reference level X~100~ is defined for each strata uniquely based on the glacier
extent in 1947-1985. X~0~ equates to an absence of glaciers.
X~60~ is defined as 30% loss of glaciers (70% remaining). 
This value was chosen after a discussion between several aline ecologist, and is documented [here](https://github.com/NINAnor/ECA_2026/issues/43).
The indicator is designed to reflect the effects of a reduced supply of melt water.
The supply of melt water from glaciers is more dependent on the glacer volume than the glacers extent or cover.
Glacer area scale non-linearely with glacier volume such that $V = A^1.36$ [@bahr1997physical].
This means that a 30% reduction in glacier areas equates to approx. a 38% reduction in volume. 
A 60% reduction in area would imply an approx. 71% less volume.

```{r}
#| results: false
v1 <- 100^1.36
v2 <- 70^1.36
v3 <- 40^1.36
(v1-v2) / v1 # 0.3843508
(v1-v3) / v1 # 0.7123916

```

The X~60~value therefore ensures that the indicator puts more weight on the ecological effect of a reduced supply of melt water from glacers then if there was no X~60~value defined.  




```{r X60}
scalar <- 0.7
```



#### 4.2.1 Spatial resolution and validity

The reference levels are defined for each region, and can therefore only be
used to normalise variables at that scale. However, reference levels can
easily be recalculated for any spatial scale.

## 5. Uncertainties
### 5.1 General uncertainties

The reference condition, using the climate in 1961-1990, introduces some uncertainty, or at least some arbitrariness, when we consider that most glaciers in Norway were melted away in the 
warm period between 7500 and 6000 year before present.

### 5.2 Reported uncertainties
The reported uncertainties reflect the uncertainties around the true parameter means, i.e. they are inferrential uncertainties for the true value of the indicator for the given area and time combination.  
These uncertainties originates from the spatial variation and the 3% measurement error in the dataset [@andreassen2022]. 


## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

### 7.1 Dataset A - Glacier cover 1947-1985

This data is based on a combination of landsat imagery and digitized
topographic maps [@winsvold2014]. We use this map as our reference
condition and assume this glacier cover represents the climatic period
1961-1990.

The dataset is manually downloaded.

URL: <https://nve.brage.unit.no/nve-xmlui/handle/11250/2831053>

```{r readCache1}
#| code-summary: "Read data"
dat_old <- readRDS(here::here("data/glacial_extent_old.rds"))
```


### 7.2 Dataset B - Glacier cover 2018-2019

This is the map of glacier cover measured in the period 2018-2019
[@andreassen2022].

URL: <https://nve.brage.unit.no/nve-xmlui/handle/11250/2836926>

```{r readCache2}
#| code-summary: "Read data"
dat_new <- readRDS(here::here("data/glacial_extent_new.rds"))
```

### 7.3 Dataset C - Geographical regions

I import a delineation of five geographical regions in Norway that I
will use as an intermediate aggregation level and for defining strata.

```{r greg2}
#| code-fold: true
#| code-summary: "Importing geographical regions"

greg <- readRDS(here::here("data/regions.rds")) |>
  mutate(region = case_when(
    id == 3 ~ "Oestlandet",
    .default = region
  ))

```


## 8. Spatial units

The five geographic regions in Norway are defined as ecosystem assets.
Mainland Norway is the ecosystem accounting area.


## 9. Analyses

We need the initial and current glacier area per region.

```{r tbl-areas}
#| tbl-cap: 'Initial (1947-1985) and current (2018-2019) glacer area for fiver geographic regions in Norway.'

dat_old2 <-  dat_old |>
  st_union()|>
  st_sf(geometry = _) |>  
  st_intersection(greg) |>
  mutate(Initial_area_km2 = units::drop_units(st_area(geometry)*10^-6))

dat_new2 <-  dat_new |>
  st_union()|>
  st_sf(geometry = _) |>  
  st_intersection(greg) |>
  mutate(Current_area_km2 = units::drop_units(st_area(geometry)*10^-6))

dat_comb <- dat_old2 |>
  st_drop_geometry() |>
  left_join(dat_new2 |> st_drop_geometry()) |>
  select(region, Initial_area_km2, Current_area_km2)

dat_comb |>
  mutate(Initial_area_km2 = round(Initial_area_km2),
         Current_area_km2 = round(Current_area_km2) ) |>
  gt::gt()


```


### 9.1 Calculating indicator values (distributions) for the regions

The next steps are a follows:

1.  Add 3% methodological uncertainty to the data at the regional level by
    sampling *n* numbers from a normal distribution with sd =
    mean × 0.03.
2.  Normalise the variable (now a prob. dist.) at the strata level using a
    piecewise linear transformation.
3.  Aggregate to a national estimate using monte carolo propagation.


The papers describing the glacier extent data report an estimated 3%
uncertainty with their method [@andreassen2022]. I assume this to mean
the standard deviation is 3% of the mean. I start by replacing the point
estimates in each region with a probability
distribution centered on the point estimate, and with sd = mean × 0.03.
In ecosystem accounting it is not common to allow uncertainty in the
reference levels, but here it is quite easy to implement, and I
therefore see no reason not to.


```{r}
#| warning: false
#| code-fold: true
#| code-summary: "Add measurement error"

dat_comb2 <- dat_comb |>
  # adding 3% uncertainty
  rowwise() |>
  mutate(
    init = list(rnorm(1000, Initial_area_km2, Initial_area_km2*0.03)), 
    curr = list(rnorm(1000, Current_area_km2, Current_area_km2*0.03))
    )

```

Then I normalise, using a piecewise linear transformation, and get
similar probability distribution for indicator values. @tbl-normalise
shows the dataset with indicator values (just the median value) at the
reginal level. 
The normalisation uses a piecewise linear transformation, with truncation (@fig-normplot).

```{r tbl-normalise}
#| tbl-cap: 'Table of ecosystem assets, with variable values, X100 values (glacier area 1947-1985), indicator values and quartiles.'
#| code-fold: true
#| code-summary: "Normalise variable"

dat_comb3 <- dat_comb2 |>
  # unlist and create new rows for each random pair of current and old glacier area estimate
  unnest_longer(c(
    curr,
    init), 
    keep_empty = T) |>
  rename(variable = curr, 
         x100 = init) |>
  mutate(
    # define x60 value as 30% glacier loss (70% of X100)
    x60 = x100*scalar,
    # normalise using piecewise linear transformation
    indicator =  ecTools::ec_normalise(
      variable = variable,
      x100 = x100,
      x60 = x60 
    ))
```

```{r fig-normplot}
#| fig-cap: 'Figure showing the general scaling function used for this indicator.'

var <- seq(0,100,.2)
ind <- ecTools::ec_normalise(var, x100=100, x60=70)
ec_norm_plot(var, ind)
```

```{r export}
#| include: false
#| eval: false
ec_norm_plot(var, ind, lan = "norsk") +
  labs(x="Breareal som % av opprinnelig breareal", y = "Indikatorverdi")
ggsave(here::here("img/skaleringsfunksjone.png"))
```


### Aggregate
Then I propagate the uncertainties and a get an estimate for Norway as a whole.

```{r}
dat_comb4 <- dat_comb3 |>
  add_column(part = "Norway") |>
  add_column(year = 2024)
```
```{r agg}
#| eval: false
mc_propagated <-  ecTools::ec_upscale(
  data = dat_comb4,
  variable = indicator,
  year = year,
  weight = Initial_area_km2,
  start_units = region,
  end_units = part,
  end_units_name = "part",
  n = 1000
)

out <- dat_comb4 |>
  select(part = region, year, sampled_mean = indicator) |>
  bind_rows(mc_propagated) |>
  mutate(part = case_when(
    part == "Midt-Norge" ~ "C",
    part == "Nord-Norge" ~ "N",
    part == "Oestlandet" ~ "E",
    part == "Sørlandet" ~ "S",
    part == "Vestlandet" ~ "W",
    part == "Norway" ~ "Norway",
    .unmatched = "error"
  ))

saveRDS(out, here::here("data/out.rds"))
```

```{r}
#| include: false
out <- readRDS(here::here("data/out.rds"))
```

## Results
```{r summaryStats}
#| warning: false
#| message: false
summary_out <- out |>
  group_by(part, year) |>
  summarise(
    median = round(median(sampled_mean), 2),
    p_025 = round(quantile(sampled_mean, probs = 0.025), 2),
    p_25 = round(quantile(sampled_mean, probs = 0.25), 2),
    p_75 = round(quantile(sampled_mean, probs = 0.75), 2),
    p_975 = round(quantile(sampled_mean, probs = 0.975), 2),
  ) |>
  add_column(
    ECT = "A1",
    Indicator = "Glacier area",
    Ecosystem = "Mountains",
    ID = "NO_GLAC_001"
  )
```



```{r tbl-results}
#| warning: false
#| message: false
#| tbl-cap: 'Indicator values and uncertainty estimates for NO_GLAC_001.'

summary_out |>
  unite(col = "CI_95", c(p_025, p_975), sep = " - ") |>
  unite(col = "IQR", c(p_25, p_75), sep = " - ") |>
  select(
    ECT,
    Indicator,
    ID,
    "Accounting area" = part,
    Ecosystem,
    "Closing year" = year,
    "Indicator value" = median,
    "95% CI" = CI_95,
    "50% CI" = IQR
  ) |>
  knitr::kable()
```


## 11. Export file

The indicator values are exported and can be found under `data/results_NO_GLAC_001_v1.0.0.csv`.

```{r export}
#| eval: false
#| code-summary: 'Export data'
write.csv(out, here::here("data/results_NO_GLAC_001_v1.0.0.csv"), row.names = F)
```

## 12. Author contributions

This is the work of a single author, although I would like to acknowledge interllectual contributions from the ecosystem group that was part of the assessment of the Norwegian mountain ecosystems for 2024.



::: {.callout-tip collapse="true"}
## Session Info

This workflow uses `renv`. Run `renv::restore()` to recreate the working environment. 
 
:::