---
title: "Calculation of ecological indicators index"
subtitle: |
  Ecological index for forest and mountain
indicatorID:
  - NO_INDEX_001
indicatorName:
  - Index
folderName: NO_INDEX_001
country: Norway
continent: Europe
ECT: NA
Realm:
  - Terrestrial (T)
Biome:
  - NA
Ecosystem:
  - NA
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "1.0.0"
VersionComment: Phase 1 equal-weight composite index from published indicator MC distributions
url: https://github.com/NINAnor/ecRxiv
Normalised: NA
SpatialAggregationPathway: Resample - Aggregate (equal-weight mean) - Summarise
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Halbritter, A. H.
date: July 1, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
  - text: Add a review
    icon: github
    href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)

if (!requireNamespace("ecTools", quietly = TRUE)) {
  remotes::install_github("NINAnor/ecTools", upgrade = "never")
}
if (!requireNamespace("arrow", quietly = TRUE)) {
  install.packages("arrow")
}
if (!requireNamespace("MetBrewer", quietly = TRUE)) {
  install.packages("MetBrewer")
}
library(ecTools)

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/index_functions.R"))
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

## 1. Aim

NO_INDEX_001 combines published ecRxiv indicator uncertainty distributions into composite indices for **forest** and **mountain** ecosystems, separately by region, for Norway, and for ECT classes.
Values near 1 indicate better ecological condition relative to reference; values toward 0 indicate deterioration.

## 2. Datasets {#sec-datasets}

First, we check which indicators are included in each index and where their results can be found on ecRxiv.

**Indicator lists** Forest indicators are listed in `data/indicators_2026.csv`; mountain indicators follow [ECA_2026 #104](https://github.com/NINAnor/ECA_2026/issues/104) in `data/indicators_mountain.csv`.
Each row specifies the indicator name, ecRxiv ID, and ECT class assignment.

```{r refresh-sources}
#| cache: false
# Refresh manifest from local git (upstream main + submission-* branches)
invisible(tryCatch(
  scan_git_indicator_sources(git_remote = "upstream"),
  error = function(e) invisible(NULL)
))
```

**Source manifest** `scan_git_indicator_sources()` searches the local `upstream` git remote (main and `submission-*` branches) for published result files (CSV, RDS, parquet) and writes `data/indicator_sources.csv`.
Optional manual or fork entries can be added in `data/indicator_sources_extra.csv` if results are not yet on `upstream`.
This manifest is used internally when resolving result paths; it is not shown in the document. 
If the scan finds few entries, run `git fetch upstream` from the project root to refresh local branch refs.

**Registry** `build_indicator_registry()` joins the indicator lists with the manifest, probes each candidate URL or local path, and records whether results are available (`match_status`: `matched` or `no_results`).
The tables below show the outcome for forest and mountain.
Only `matched` indicators proceed to the analysis in @sec-analyses.

```{r registry}
#| cache: false
n_sim <- 1000
n_years <- 2

registry_forest <- build_indicator_registry(
  here("data/indicators_2026.csv"),
  ecosystem = "forest",
  write_path = here("data/indicators_registry_forest.csv")
)

registry_mountain <- build_indicator_registry(
  here("data/indicators_mountain.csv"),
  ecosystem = "mountain",
  write_path = here("data/indicators_registry_mountain.csv")
)
```

```{r registry-status}
#| cache: false
# Compact included / not-included overview (instead of long coverage lists)
registry_status_table <- function(registry, caption) {
  registry |>
    dplyr::transmute(
      ECT = .data$ect,
      Indicator = .data$indicatorName,
      ID = .data$indicatorID,
      Status = dplyr::case_when(
        .data$match_status == "matched" ~ "Included",
        .data$match_status == "no_results" ~ "No results yet",
        TRUE ~ "Missing ID"
      )
    ) |>
    dplyr::arrange(.data$ECT, .data$Status, .data$Indicator) |>
    knitr::kable(caption = caption)
}

registry_status_table(registry_forest, "Forest indicators — inclusion status")
```

```{r registry-status-mountain}
#| cache: false
registry_status_table(registry_mountain, "Mountain indicators — inclusion status")
```

## 3. Analyses {#sec-analyses}

Forest and mountain indices are calculated separately using the same workflow (`calculate_index()` in `R/index_functions.R`).
The indicator registries built in @sec-datasets are the input to this section.

### 3.1 Loading and harmonising indicator draws

For each `matched` indicator in the registry, `load_registry_draws()` fetches the resolved results file and harmonises it to a common structure (`year`, `part`, `value`).
Supported source formats include:

- standard CSV (`year`, `part`, `sampled_mean`);
- period-based CSV or RDS (`period`, `region`/`region_code`, `sampled_mean`);
- regional RDS with companion national RDS (`*_mc_regional.rds` + `*_mc_national.rds`);
- parquet (`year`, `part`, `value` or `sim`).

Region labels are mapped to NI regions (`C`, `E`, `S`, `W`, `N`, `Norway`) via `data/region_lookup.csv`.
Each row in the loaded data is one Monte Carlo draw for a given indicator, reporting year, and region.

### 3.2 Reporting periods

Indicators do not all use the same time axis.
Some publish single calendar years (`year`); others publish multi-year **reporting periods** such as `2010-2014` or `2020-2024` (`period`).
These period labels also differ between indicators (length and start/end years are not standardised).

For aggregation we map each reporting period to a single calendar year using the **period end year** (`period_to_year()`, e.g. `2020-2024` → `2024`).
Where the source already provides `year`, that value is kept as is. For each indicator we then retain the **last two** of these mapped years (`select_reporting_years()`), which for most current forest indicators corresponds to **2019** and **2024**.
Indicators may therefore appear in different years in the index, depending on their own reporting schedule.

### 3.3 Resampling and aggregation

Aggregation uses `ecTools::ec_upscale()` with equal weights **across indicators within an ECT class**, and **across ECT classes within the index** (`n_sim = 1000` by default).
Steps 1–2 are run separately for each year and each of six parts: the five NI regions (`C`, `E`, `S`, `W`, `N`) and Norway.

1. **Indicator → ECT** — within a part, one draw is resampled per indicator within each ECT class, then combined with equal weights to give an ECT-level draw (`aggregate_to_ect()`).
2. **ECT → index** — within the same part, one draw is resampled per ECT class, then combined with equal weights to give an overall index draw for that part (`aggregate_to_index()`).

**Norway is not recomputed by equally averaging the five regions.**
Most ecRxiv indicators already publish their own national ("Norway") draws alongside the five regions, and that published value may already reflect region weighting that is specific to the indicator.
For example, the glacier area indicator ([NO_GLAC_001](https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_GLAC_001)) down-weights southern Norway in its own national value, because glaciers make up a much smaller share of that region (see [ECA_2026 #76](https://github.com/NINAnor/ECA_2026/issues/76#issuecomment-4864265018)).
We therefore always use an indicator's own published Norway draws when available, and only fall back to an equal-weight upscale across its five regions (`fill_missing_national_draws()`) for indicators that do not publish a national value.
Norway is then treated as just another part in steps 1–2 above, so indicator-specific national weighting propagates through the ECT and index levels rather than being overwritten by our own equal-weight recomputation.

This propagates uncertainty from individual indicators through ECT classes to the composite index without assuming independence beyond the structure encoded in the published indicator MC distributions.

Weighting **between** indicators, ECT classes, or regions (e.g. by area share) is not yet implemented in this Phase 1 version; all weights described above are equal (`weight = 1`) except for the indicator-specific national weighting already embedded in a small number of published national values (currently `NO_GLAC_001`).

### 3.4 Summaries

The full MC distributions (all regions) are stored for export.
For reporting we summarise each ECT class and the overall index by the **median** and an approximate **95% interval** (2.5% and 97.5% quantiles).
Regional indices are calculated in steps 1–2 above; national values are obtained in step 3.

```{r run-forest}
#| cache: false
forest_index <- calculate_index(
  registry_forest |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)
```

```{r run-mountain}
#| cache: false
mountain_index <- calculate_index(
  registry_mountain |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)
```

```{r validation}
#| cache: false
# Single-indicator test: with one matched indicator, the index for any part
# (including Norway, using that indicator's own published national draws)
# equals that indicator's own draws for the same part.
if (sum(registry_mountain$match_status == "matched") >= 1) {
  one_reg <- registry_mountain |> filter(match_status == "matched") |> slice(1)
  one_out <- calculate_index(one_reg, n_sim = 200, n_years = 1)
  part_chk <- one_out$indicator_draws |>
    distinct(part) |>
    arrange(part) |>
    slice(1) |>
    pull(part)
  val_one <- one_out$summaries |>
    filter(level == "Index", part == part_chk) |>
    slice(1)
  src <- one_out$indicator_draws |>
    filter(part == part_chk) |>
    summarise(median = median(value), .groups = "drop")
  stopifnot(abs(val_one$median - src$median) < 0.05)
}
```

## 4. Results

Summary tables report ECT classes and the overall index for **Norway** only (both retained reporting years).
Regional indices (`C`, `E`, `S`, `W`, `N`) are calculated and shown in the figures below; full draws including regions are also written to the export files in @sec-export.

```{r forest-summary}
#| cache: false
forest_index$summaries |>
  filter(part == "Norway") |>
  arrange(year, level, id) |>
  select(year, level, id, median, q025, q975) |>
  kable(digits = 3, caption = "Forest index — Norway")
```

```{r mountain-summary}
#| cache: false
mountain_index$summaries |>
  filter(part == "Norway") |>
  arrange(year, level, id) |>
  select(year, level, id, median, q025, q975) |>
  kable(digits = 3, caption = "Mountain index — Norway")
```

```{r forest-plot}
#| cache: false
#| fig-height: 4
#| fig-width: 9
plot_index_forest(
  forest_index$summaries,
  title = "NO_INDEX_001 — forest"
)
```

```{r mountain-plot}
#| cache: false
#| fig-height: 3.5
#| fig-width: 9
plot_index_forest(
  mountain_index$summaries,
  title = "NO_INDEX_001 — mountain"
)
```

## 5. Export file {#sec-export}

```{r export}
#| cache: false
export_index_results(
  forest_index$distributions,
  here("data/results_NO_INDEX_001_forest_v1.0.0.csv")
)

export_index_results(
  mountain_index$distributions,
  here("data/results_NO_INDEX_001_mountain_v1.0.0.csv")
)
```

::: {.callout-tip collapse="true"}
## Session Info

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
