---
title: "Calculation of ecological condition indices"
subtitle: |
  NO_IDEX_001 - for forest and mountain ecosystems
indicatorID:
  - NO_IDEX_001
indicatorName:
  - Index
folderName: NO_IDEX_001
country: Norway
continent: Europe
ECT: NA
Realm:
  - Terrestrial (T)
Biome:
  - NA
Ecosystem:
  - NA
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "1.0.0"
VersionComment: Phase 1 equal-weight composite index from published indicator MC distributions
url: https://github.com/NINAnor/ecRxiv
Normalised: NA
SpatialAggregationPathway: Resample - Aggregate (equal-weight mean) - Summarise
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Halbritter, A. H.
date: July 1, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
code-links:
  - text: Add a review
    icon: github
    href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
codeReviewers:
  - name: Anders Kolstad
    email: anders.kolstad@nina.no
    affiliations:
      - id: reviewerInst
        name: NINA
    reviewType: code review
    github: anders-kolstad
---

```{r setup}
#| include: false
#| cache: false
#if (file.exists("renv/activate.R")) {
#  source("renv/activate.R")
#} else if (file.exists("../renv/activate.R")) {
#  source("../renv/activate.R")
#}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)

if (!requireNamespace("ecTools", quietly = TRUE)) {
  remotes::install_github("NINAnor/ecTools", upgrade = "never")
}
if (!requireNamespace("arrow", quietly = TRUE)) {
  install.packages("arrow")
}
if (!requireNamespace("MetBrewer", quietly = TRUE)) {
  install.packages("MetBrewer")
}
library(ecTools)

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/index_functions.R"))
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

::: {.callout-tip collapse="true"}

## Log

<!--# Update this log with short messages for each update -->
- 1 July 2026 - First draft of calculation
- 14 August 2026 - Renamed to NO_IDEX_001, update indicator list, imputation of missing reporting years, direct national aggregation, and update figures
- 31 August 2026 - Finalize results and figures for forest and mountain index
:::

<hr />

## 1. Aim

NO_IDEX_001 combines published ecRxiv indicator uncertainty distributions into composite indices. Version 1.0.0 contains indices for **forest** and **mountain** ecosystems, for year 2024, separately by region, for Norway, and for ECT (Ecosystem Condition Typology) classes.
Values near 1 indicate better ecological condition relative to reference; values toward 0 indicate deterioration.

## 2. Datasets {#sec-datasets}

First, we check which indicators are included in each index and where their results can be found on ecRxiv.
The overview tables at the end of this section list every matched indicator with its national (Norway) value for 2024 (median and approximate 95% interval).
Indicators without published uncertainty show the point estimate only.

**Indicator lists.** Forest indicators are listed in `data/indicators_2026.csv`; mountain indicators follow [ECA_2026 #104](https://github.com/NINAnor/ECA_2026/issues/104) in `data/indicators_mountain.csv`.
Each row specifies the indicator name, ecRxiv ID, and ECT class assignment.

**Source manifest.** `scan_git_indicator_sources()` searches the local `upstream` git remote (main and `submission-*` branches) for published result files (CSV, RDS, parquet) and writes `data/indicator_sources.csv`.
Optional manual or fork entries can be added in `data/indicator_sources_extra.csv` if results are not yet on `upstream`.
This manifest is used internally when resolving result paths; it is not shown in the document. 
If the scan finds few entries, run `git fetch upstream` from the project root to refresh local branch refs.

```{r refresh-sources}
#| cache: false
# Refresh manifest from local git (upstream main + submission-* branches)
invisible(tryCatch(
  scan_git_indicator_sources(git_remote = "upstream"),
  error = function(e) invisible(NULL)
))
```

**Registry.** `build_indicator_registry()` joins the indicator lists with the manifest and probes each candidate URL or local path.
Only `matched` indicators proceed to the analysis in @sec-analyses.
The tables below give an overview of matched indicators and their national (Norway) values for 2024.
Full index results are reported in @sec-results-forest and @sec-results-mountain.

```{r registry}
#| cache: false
n_sim <- 1000
n_years <- 2
report_years <- c(2014, 2019, 2024)
max_gap <- 5 # max gap in years to fill missing reporting years

registry_forest <- build_indicator_registry(
  here("data/indicators_2026.csv"),
  ecosystem = "forest",
  write_path = here("data/indicators_registry_forest.csv")
)

registry_mountain <- build_indicator_registry(
  here("data/indicators_mountain.csv"),
  ecosystem = "mountain",
  write_path = here("data/indicators_registry_mountain.csv")
)
```

```{r readForestCache}
#| include: false
#| cache: false
forest_index <- readRDS(here::here("data/forest_index_170826.rds"))
```

```{r readMountainCache}
#| include: false
#| cache: false
mountain_index <- readRDS(here::here("data/mountain_index_170826.rds"))
```

```{r target-years}
#| include: false
target_years <- tail(sort(report_years), n_years)
```

```{r registry-indicator-overview-forest}
#| cache: false
#| dependson: ["readForestCache"]
registry_indicator_table(
  registry_forest,
  forest_index,
  year = 2024,
  caption = "Forest indicators included in NO_IDEX_001 — Norway 2024"
)
```

```{r registry-indicator-overview-mountain}
#| cache: false
#| dependson: ["readMountainCache"]
registry_indicator_table(
  registry_mountain,
  mountain_index,
  year = 2024,
  caption = "Mountain indicators included in NO_IDEX_001 — Norway 2024"
)
```

## 3. Analyses {#sec-analyses}

Forest and mountain indices are calculated separately using the same workflow (`calculate_index()` in `R/index_functions.R`).
The indicator registries built in @sec-datasets are the input to this section.

### 3.1 Loading and harmonising indicator draws

For each `matched` indicator in the registry, `load_registry_draws()` fetches the resolved results file and harmonises it to a common structure (`year`, `part`, `value`).
Supported source formats include:

- standard CSV (`year`, `area` or `part`/`region`, `value` or `sampled_mean`/`sampled_value`);
- period-based CSV or RDS (`period`, `area`/`region`/`region_code`, `value` or `sampled_mean`/`sampled_value`);
- regional RDS with companion national RDS (`*_mc_regional.rds` + `*_mc_national.rds`);
- parquet (`year`, `part`, `value` or `sim`).

Region labels are mapped to NI regions (`C`, `E`, `S`, `W`, `N`, `Norway`) via `data/region_lookup.csv`.
Each row in the loaded data is one Monte Carlo draw for a given indicator, reporting year, and region.

### 3.2 Reporting periods

Indicators do not all use the same time axis.
Some publish single calendar years (`year`); others publish multi-year **reporting periods** such as `2010-2014` or `2020-2024` (`period`).
These period labels also differ between indicators (length and start/end years are not standardised).

For aggregation we map each reporting period to a single calendar year using the **period end year** (`period_to_year()`, e.g. `2020-2024` → `2024`).
Where the source already provides `year`, that value is kept as is.

We use fixed national reporting years: `report_years <- c(2014, 2019, 2024)`, and retain the last `n_years` of these (`2019` and `2024` with the current settings).

Some indicators only publish one of the two retained report years — for example a first submission with only `2024` data and nothing for `2019`.
For indicators that are missing one of the retained report years, `impute_missing_indicator_years()` carries its nearest other retained year's draws forward or backward to fill the gap, so the same indicator set contributes to every reported year.
This is bounded by `max_gap <- 5`: if no other retained year is within `max_gap` of the missing one, the indicator stays excluded for that year rather than being filled with a stale value. With `report_years <- c(2014, 2019, 2024)` this simply means an indicator's nearest retained year must be within one reporting cycle (5 years) to be carried forward or backward.

### 3.3 Resampling and aggregation

Aggregation uses `ecTools::ec_upscale()` with equal weights **across indicators within an ECT class**, and **across ECT classes within the index** (`n_sim = 1000` by default).
Steps 1–2 are run separately for each year and each of six areas: the five NI regions (`C`, `E`, `S`, `W`, `N`) and Norway.

Note that some indicators will have missing values for some regions. This will be describes in the workflow documentation for each indicator.

1. **Indicator → ECT** — within an area, one draw is resampled per indicator within each ECT class, then combined with equal weights to give an ECT-level draw (`aggregate_to_ect()`).
2. **ECT → index** — within the same area, one draw is resampled per ECT class, then combined with equal weights to give an overall index draw for that area (`aggregate_to_index()`).

**The national index goes through the same two steps above, same as the regions.**
All indicators publish their own national ("Norway") draws alongside the five regions, so each indicator's value going into step 1 for the Norway area is simply its own published national draw, not a recomputation by us from the five regions.
This matters because a published national value can reflect region weighting that is specific to that indicator - for example, the glacier area indicator ([NO_GLAC_001](https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_GLAC_001)) down-weights southern Norway in its own national value, because glaciers make up a much smaller share of that region (see [ECA_2026 #76](https://github.com/NINAnor/ECA_2026/issues/76#issuecomment-4864265018)).
As a fallback, indicators without a published national value get a synthetic one via equal-weight upscaling across the five regions (`fill_missing_national_draws()`).

This propagates uncertainty from individual indicators through ECT classes to the composite index without assuming independence beyond the structure encoded in the published indicator MC distributions.

No weighting is done between indicators or ECT classes: all indicators and ECT classes have equal weight (`weight = 1`).
Individual indicators define their own regional weights to obtain their national estimate.
Usually this weight is defined by the ecosystem extant for each region.
There is one notable exception: `NO_GLAC_001`'s own published national value used the glacier area in the reference period as the regional weight, with the result of down-weighting southern Norway in the national estimate.
We use these national draws as-is rather than overriding it with an equal-weight regional average.

### 3.3.1 Direct national aggregation (no ECT step)

We also calculate a direct national index that skips the ECT step: every matched indicator is resampled directly to the index with equal weight (`aggregate_indicators_to_index_direct()`), rather than first being equal-weighted within its ECT class.
This means an ECT class with many indicators has proportionally more influence on this version of the index than it does in the ECT-based workflow, where each ECT class counts equally regardless of how many indicators it contains.

This direct aggregation is computed independently, and reported alongside the ECT-based index (tagged `"Index (direct)"` in the results) so the two can be compared. Note that the indicator ID NO_IDEX_001 retains to the hierarchically aggregated version. 

### 3.4 Summaries

The full MC distributions (all areas) are stored for export.
For reporting we summarise each ECT class and the overall index by the **median** and an approximate **95% interval** (2.5% and 97.5% quantiles).
Regional and national (Norway) indices are both calculated by the same steps 1–2 above, since Norway is treated as just another area (see @sec-analyses).

```{r run-forest}
#| cache: true
#| code-summary: 'Calculate forest index'
#| cache.extra: !expr digest::digest(registry_forest)
#| eval: false
forest_index <- calculate_index(
  registry_forest |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years,
  report_years = report_years,
  max_gap = max_gap,
  include_direct_index = TRUE
)

saveRDS(forest_index, here::here("data/forest_index_170826.rds"))
```


```{r run-mountain}
#| cache: true
#| code-summary: 'Calculate mountain index'
#| cache.extra: !expr digest::digest(registry_mountain)
#| eval: false
mountain_index <- calculate_index(
  registry_mountain |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years,
  report_years = report_years,
  max_gap = max_gap,
  include_direct_index = TRUE
)

saveRDS(mountain_index, here::here("data/mountain_index_170826.rds"))
```

```{r tbl-checkForest}
#| tbl-cap: 'Table showing the number of indicators used in the index, and how many of these were either observed or imputed.' 
forest_index$coverage |>
  select(year, part, "# indicators" = n_indicators, "# indicators imputed" = n_indicators_imputed, "# indicators observed" = n_indicators_observed) |>
  gt::gt()
```

```{r validation}
#| cache: false
# Single-indicator test: with one matched indicator, the index for any area
# (including Norway, using that indicator's own published national draws)
# equals that indicator's own draws for the same area.
if (sum(registry_mountain$match_status == "matched") >= 1) {
  one_reg <- registry_mountain |> filter(match_status == "matched") |> slice(1)
  one_out <- calculate_index(one_reg, n_sim = 200, n_years = 1, report_years = report_years, max_gap = max_gap)
  part_chk <- one_out$indicator_draws |>
    distinct(part) |>
    arrange(part) |>
    slice(1) |>
    pull(part)
  val_one <- one_out$summaries |>
    filter(level == "Index", part == part_chk) |>
    slice(1)
  src <- one_out$indicator_draws |>
    filter(part == part_chk) |>
    summarise(median = median(value), .groups = "drop")
  stopifnot(abs(val_one$median - src$median) < 0.05)
}
```

## 4. Results

Forest and mountain indices are reported separately below.
Each subsection gives a full summary table (all indicators, ECT classes, and index totals for every area and reporting year) followed by figures.

### 4.1 Forest {#sec-results-forest}

```{r forest-summary-full}
#| cache: false
#| dependson: ["readForestCache", "target-years"]
index_figure_summary_table(
  forest_index,
  registry_forest,
  years = target_years,
  caption = "Forest index — all areas and reporting years"
)
```

```{r forest-plot-detailed}
#| cache: false
#| fig-height: 11
#| fig-width: 9
for (yr in target_years) {
  print(plot_index_detailed(
    forest_index, yr,
    title = "NO_IDEX_001 — forest",
    registry = registry_forest
  ))
}
```

```{r forest-plot-detailed-national}
#| cache: false
#| fig-height: 7
#| fig-width: 7
for (yr in target_years) {
  print(plot_index_detailed(
    forest_index, yr,
    title = "NO_IDEX_001 — forest",
    national_only = TRUE,
    registry = registry_forest
  ))
}
```

### 4.2 Mountain {#sec-results-mountain}

```{r mountain-summary-full}
#| cache: false
#| dependson: ["readMountainCache", "target-years"]
index_figure_summary_table(
  mountain_index,
  registry_mountain,
  years = target_years,
  caption = "Mountain index — all areas and reporting years"
)
```

```{r mountain-plot-detailed}
#| cache: false
#| fig-height: 8
#| fig-width: 9
for (yr in target_years) {
  print(plot_index_detailed(
    mountain_index, yr,
    title = "NO_IDEX_001 — mountain",
    registry = registry_mountain
  ))
}
```

```{r mountain-plot-detailed-national}
#| cache: false
#| fig-height: 5.5
#| fig-width: 7
for (yr in target_years) {
  print(plot_index_detailed(
    mountain_index, yr,
    title = "NO_IDEX_001 — mountain",
    national_only = TRUE,
    registry = registry_mountain
  ))
}
```

English and Norwegian versions of the same figures are also written to `img/` (not shown here) for reuse.

```{r save-figures}
#| include: false
#| cache: false
#| code-summary: "Save EN and NB figures to img/"
save_index_detailed_figures(
  forest_index,
  ecosystem = "forest",
  years = target_years,
  registry = registry_forest,
  height_regional = 11,
  height_national = 7
)
save_index_detailed_figures(
  mountain_index,
  ecosystem = "mountain",
  years = target_years,
  registry = registry_mountain,
  height_regional = 8,
  height_national = 5.5
)
```

## 5. Export file {#sec-export}

The exported files contain the full MC distributions for every level (`"ECT"`, `"Index"`, and the alternative `"Index (direct)"`, see @sec-analyses), area, and year — identifiable via the `level` and `id` columns.

```{r export}
#| cache: false
export_index_results(
  forest_index$distributions,
  here("data/results_NO_IDEX_001_forest_v1.0.0.csv")
)

export_index_results(
  mountain_index$distributions,
  here("data/results_NO_IDEX_001_mountain_v1.0.0.csv")
)
```

::: {.callout-tip collapse="true"}
## Session Info

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
