---
title: "Calculation of ecological indicators index"
subtitle: |
  NO_INDEX_001 (forest and mountain)
indicatorID:
  - NO_INDEX_001
indicatorName:
  - Composite index
folderName: NO_INDEX_001
country: Norway
continent: Europe
ECT: Composite
Realm:
  - Terrestrial (T)
Biome:
  - T2 Temperate-boreal forests and woodlands biome
  - T7 Polar/alpine biome
Ecosystem:
  - T2.1 Boreal/montane forests
  - T7.1 Alpine grasslands and shrublands
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "1.0.0"
VersionComment: Phase 1 equal-weight composite index from published indicator MC distributions
url: https://github.com/NINAnor/ecRxiv
Normalised: Yes
SpatialAggregationPathway: Resample - Aggregate (equal-weight mean) - Summarise
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Halbritter, A. H.
date: July 1, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
  - text: Add a review
    icon: github
    href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)

if (!requireNamespace("ecTools", quietly = TRUE)) {
  remotes::install_github("NINAnor/ecTools", upgrade = "never")
}
if (!requireNamespace("arrow", quietly = TRUE)) {
  install.packages("arrow")
}
library(ecTools)

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/index_functions.R"))
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

## 1. Aim

NO_INDEX_001 combines published ecRxiv indicator uncertainty distributions into composite indices for **forest** and **mountain** ecosystems, separately by region and for Norway and for ECT classes. Values near 1 indicate better ecological condition relative to reference; values toward 0 indicate deterioration.

## 2. Datasets {#sec-datasets}

First, we check which indicators are included in each index and where their results can be found on ecRxiv.

**Indicator lists.** Forest indicators are listed in `data/indicators_2026.csv`; mountain indicators follow [ECA_2026 #104](https://github.com/NINAnor/ECA_2026/issues/104) in `data/indicators_mountain.csv`. Each row specifies the indicator name, ecRxiv ID, and ECT class assignment.

**Source manifest.** `scan_git_indicator_sources()` searches the local `upstream` git remote (main and `submission-*` branches) for published result files (CSV, RDS, parquet) and writes `data/indicator_sources.csv`. Manual or fork entries can be added in `data/indicator_sources_extra.csv`.

**Registry.** `build_indicator_registry()` joins the indicator lists with the manifest, probes each candidate URL or local path, and records whether results are available (`match_status`: `matched` or `no_results`). The tables below show the outcome for forest and mountain. Only `matched` indicators proceed to the analysis in @sec-analyses.

```{r refresh-sources}
#| cache: false
# Refresh manifest from local git (upstream main + submission-* branches)
tryCatch(
  scan_git_indicator_sources(git_remote = "upstream"),
  error = function(e) invisible(NULL)
)
```

```{r registry}
#| cache: false
n_sim <- 1000
n_years <- 3

registry_forest <- build_indicator_registry(
  here("data/indicators_2026.csv"),
  ecosystem = "forest",
  write_path = here("data/indicators_registry_forest.csv")
)

registry_mountain <- build_indicator_registry(
  here("data/indicators_mountain.csv"),
  ecosystem = "mountain",
  write_path = here("data/indicators_registry_mountain.csv")
)

registry_forest |>
  select(ecosystem, ect, indicatorName, indicatorID, match_status, results_ref, results_source) |>
  kable(caption = "Forest indicator registry")
```

```{r registry-mountain}
registry_mountain |>
  select(ecosystem, ect, indicatorName, indicatorID, match_status, results_ref, results_source) |>
  kable(caption = "Mountain indicator registry (ECA_2026 #104)")
```

## 3. Analyses {#sec-analyses}

Forest and mountain indices are calculated separately using the same workflow (`calculate_index()` in `R/index_functions.R`). The steps below describe phase 1 (equal weights). The indicator registries built in @sec-datasets are the input to this section.

### 3.1 Loading and harmonising indicator draws

For each `matched` indicator in the registry, `load_registry_draws()` fetches the resolved results file and harmonises it to a common structure (`year`, `part`, `value`). Supported source formats include:

- standard CSV (`year`, `part`, `sampled_mean`);
- period-based CSV or RDS (`period`, `region`/`region_code`, `sampled_mean`);
- regional RDS with companion national RDS (`*_mc_regional.rds` + `*_mc_national.rds`);
- parquet (`year`, `part`, `value` or `sim`).

Region labels are mapped to NI regions (`C`, `E`, `S`, `W`, `N`, `Norway`) via `data/region_lookup.csv`. Each row in the loaded data is one Monte Carlo draw for a given indicator, reporting year, and region.

### 3.2 Reporting years

For each indicator we retain the **last three reporting years** available (`select_reporting_years()`). Indicators therefore contribute to different calendar years depending on their own reporting schedule.

### 3.3 Resampling and aggregation

Aggregation uses `ecTools::ec_upscale()` with equal weights throughout (phase 1). For each year, region, and simulation (`n_sim = 1000` by default):

1. **Indicator → ECT** — one draw is resampled per indicator within each ECT class, then combined with equal weights to give an ECT-level draw (`aggregate_to_ect()`).
2. **ECT → index** — one draw is resampled per ECT class, then combined with equal weights to give an overall index draw (`aggregate_to_index()`).
3. **Region → Norway** — regional ECT and index draws are upscaled to national values with equal weights across the five NI regions (`aggregate_to_national()`).

This propagates uncertainty from individual indicators through ECT classes to the composite index without assuming independence beyond the structure encoded in the published indicator MC distributions.

### 3.4 Summaries and coverage

The full MC distributions are stored for export. For reporting we summarise each ECT class and the overall index by the **median** and an approximate **95% interval** (2.5% and 97.5% quantiles). A coverage table records how many indicators contributed to each year–region combination.

```{r run-forest}
forest_index <- calculate_index(
  registry_forest |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)

forest_index$coverage |>
  kable(caption = "Forest index — indicator coverage by year and region")
```

```{r run-mountain}
mountain_index <- calculate_index(
  registry_mountain |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)

mountain_index$coverage |>
  kable(caption = "Mountain index — indicator coverage by year and region")
```

```{r validation}
# Single-indicator test: index equals that indicator when only one is matched
if (sum(registry_mountain$match_status == "matched") >= 1) {
  one_id <- registry_mountain |> filter(match_status == "matched") |> slice(1)
  one_reg <- one_id
  one_out <- calculate_index(one_reg, n_sim = 200, n_years = 1)
  val_one <- one_out$summaries |> filter(level == "Index", part == "Norway") |> slice(1)
  src <- one_out$indicator_draws |>
    filter(part == "Norway") |>
    summarise(median = median(value), .groups = "drop")
  stopifnot(abs(val_one$median - src$median) < 0.05)
}
```

## 4. Results

```{r forest-summary}
forest_index$summaries |>
  filter(part == "Norway", year == max(year)) |>
  arrange(level, id) |>
  kable(digits = 3, caption = "Forest index — Norway (latest year)")
```

```{r mountain-summary}
mountain_index$summaries |>
  filter(part == "Norway", year == max(year)) |>
  arrange(level, id) |>
  kable(digits = 3, caption = "Mountain index — Norway (latest year)")
```

```{r forest-plot}
#| fig-height: 8
plot_index_forest(
  forest_index$summaries |> filter(part == "Norway"),
  title = "NO_INDEX_001 — forest (Norway)"
)
```

```{r mountain-plot}
#| fig-height: 8
plot_index_forest(
  mountain_index$summaries |> filter(part == "Norway"),
  title = "NO_INDEX_001 — mountain (Norway)"
)
```

## 5. Export file

```{r export}
#| cache: false
export_index_results(
  forest_index$distributions,
  here("data/results_NO_INDEX_001_forest_v1.0.0.csv")
)

export_index_results(
  mountain_index$distributions,
  here("data/results_NO_INDEX_001_mountain_v1.0.0.csv")
)
```

::: {.callout-tip collapse="true"}
## Session Info

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
