---
title: "Calculation of ecological indicators index"
subtitle: |
  NO_INDEX_001 (forest and mountain)
indicatorID:
  - NO_INDEX_001
indicatorName:
  - Composite index
folderName: NO_INDEX_001
country: Norway
continent: Europe
ECT: Composite
Realm:
  - Terrestrial (T)
Biome:
  - T2 Temperate-boreal forests and woodlands biome
  - T7 Polar/alpine biome
Ecosystem:
  - T2.1 Boreal/montane forests
  - T7.1 Alpine grasslands and shrublands
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete
Version: "1.0.0"
VersionComment: Phase 1 equal-weight composite index from published indicator MC distributions
url: https://github.com/NINAnor/ecRxiv
Normalised: Yes
SpatialAggregationPathway: Resample - Aggregate (equal-weight mean) - Summarise
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
AuthorList: Halbritter, A. H.
date: July 1, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
  - text: Add a review
    icon: github
    href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
---

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(conflicted)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)

if (!requireNamespace("ecTools", quietly = TRUE)) {
  remotes::install_github("NINAnor/ecTools", upgrade = "never")
}
library(ecTools)

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/index_functions.R"))
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```

:::

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`

## 1. Summary

1. **Background and rationale**: NO_INDEX_001 combines published ecRxiv indicator uncertainty distributions into composite indices for **forest** and **mountain** ecosystems, separately by region and for Norway and for ECT classes.

2. **Methods**: For each reporting period and region we (i) load MC draws from indicator `results_*_v*.csv` files on ecRxiv main (with fallback to submission branches and the source manifest), (ii) keep the last three reporting years per indicator, (iii) resample one draw per indicator per simulation using `ecTools::ec_upscale()`, (iv) aggregate to ECT classes (equal weights), and (v) compute an overall index as the unweighted mean of available ECT values.

3. **Interpretation**: Values near 1 indicate better ecological condition relative to reference; values toward 0 indicate deterioration.

4. **Results and development needs**: ...

## 2. Datasets

```{r refresh-sources}
#| cache: false
# Refresh manifest from local git (upstream main + submission-* branches)
tryCatch(
  scan_git_indicator_sources(git_remote = "upstream"),
  error = function(e) invisible(NULL)
)
```

```{r registry}
#| cache: false
n_sim <- 1000
n_years <- 3

registry_forest <- build_indicator_registry(
  here("data/indicators_2026.csv"),
  ecosystem = "forest",
  write_path = here("data/indicators_registry_forest.csv")
)

registry_mountain <- build_indicator_registry(
  here("data/indicators_mountain.csv"),
  ecosystem = "mountain",
  write_path = here("data/indicators_registry_mountain.csv")
)

registry_forest |>
  select(ecosystem, ect, indicatorName, indicatorID, match_status, results_ref, results_source) |>
  kable(caption = "Forest indicator registry")
```

```{r registry-mountain}
registry_mountain |>
  select(ecosystem, ect, indicatorName, indicatorID, match_status, results_ref, results_source) |>
  kable(caption = "Mountain indicator registry (draft)")
```

## 3. Analyses

```{r run-forest}
forest_index <- calculate_index(
  registry_forest |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)

forest_index$coverage |>
  kable(caption = "Forest index — indicator coverage by year and region")
```

```{r run-mountain}
mountain_index <- calculate_index(
  registry_mountain |> filter(match_status == "matched"),
  n_sim = n_sim,
  n_years = n_years
)

mountain_index$coverage |>
  kable(caption = "Mountain index — indicator coverage by year and region")
```

```{r validation}
# Single-indicator test: index equals that indicator when only one is matched
if (sum(registry_mountain$match_status == "matched") >= 1) {
  one_id <- registry_mountain |> filter(match_status == "matched") |> slice(1)
  one_reg <- one_id
  one_out <- calculate_index(one_reg, n_sim = 200, n_years = 1)
  val_one <- one_out$summaries |> filter(level == "Index", part == "Norway") |> slice(1)
  src <- one_out$indicator_draws |>
    filter(part == "Norway") |>
    summarise(median = median(value), .groups = "drop")
  stopifnot(abs(val_one$median - src$median) < 0.05)
}
```

## 4. Results

```{r forest-summary}
forest_index$summaries |>
  filter(part == "Norway", year == max(year)) |>
  arrange(level, id) |>
  kable(digits = 3, caption = "Forest index — Norway (latest year)")
```

```{r mountain-summary}
mountain_index$summaries |>
  filter(part == "Norway", year == max(year)) |>
  arrange(level, id) |>
  kable(digits = 3, caption = "Mountain index — Norway (latest year)")
```

```{r forest-plot}
#| fig-height: 8
plot_index_forest(
  forest_index$summaries |> filter(part == "Norway"),
  title = "NO_INDEX_001 — forest (Norway)"
)
```

```{r mountain-plot}
#| fig-height: 8
plot_index_forest(
  mountain_index$summaries |> filter(part == "Norway"),
  title = "NO_INDEX_001 — mountain (Norway)"
)
```

## 5. Export file

```{r export}
#| cache: false
export_index_results(
  forest_index$distributions,
  here("data/results_NO_INDEX_001_forest_v1.0.0.csv")
)

export_index_results(
  mountain_index$distributions,
  here("data/results_NO_INDEX_001_mountain_v1.0.0.csv")
)
```

::: {.callout-tip collapse="true"}
## Session Info

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
