---
title: "Lynx abundance"
subtitle: |
  [NO_LYAB_001]
indicatorID: 
  - NO_LYAB_001
indicatorName: 
  - Lynx abundance [Gaupebestand]
verbatimName: 
  - Gaupebestand
folderName: NO_LYAB_001
country: Norway
continent: Europe
ECT: B3 Functional state characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First version on ecRxiv; indicator described in the Norwegian Nature Index
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_LYAB_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (population model, large-carnivore management region) - [Scale - Truncate] - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud H. Halbritter
    email: aud.halbritter@uib.no
    affiliations:
      - id: myID
        name: University of Bergen
AuthorList: Halbritter, A. H.
date: May 15, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull
library(NIcalc)

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
if (file.exists(here::here("R/spatial_regions.R"))) {
  source(here::here("R/spatial_regions.R"))
}
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- May 15, 2026 - Initial draft for ecRxiv
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
The Eurasian lynx (_Lynx lynx_) is the only wild cat species in Norway.
According to [Rovdata](https://rovdata.no/Gaupe.aspx), it lives mainly in forest areas over large parts of the country, and the densest populations are in Midt-Norway and eastern Norway.
The species is also widespread in Europe and Asia and is listed as endangered (sterkt truet) on the Norwegian Red List (2021).
Rovdata is responsible for monitoring lynx in Norway.
The indicator is part of the Norwegian Nature Index and is described on [Naturindeksen](https://naturindeks.no/Indicators/gaupe).

2. **Methods**:
Indicator values for lynx abundance are taken from the Naturindeks database (via the `NIcalc` API in this workflow).
Population monitoring and status reporting for lynx are coordinated by [Rovdata](https://rovdata.no/Gaupe.aspx), which publishes national and regional information based on the national monitoring programme for large carnivores.
For example, the 2025 status report gives a national population estimate expressed as a range and relates this to counts of family groups of lynx [@rovdata_gaupe_status_2025].

3. **Interpretation**:
An indicator value of 1 means that the lynx population in the reporting unit is at least as large as expected under the reference condition used in Naturindeksen.
Values decreasing toward zero indicate that the population index is further below that reference.

4. **Results and development needs**:
The indicator is used in Naturindeksen.


## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: Gaupe (_Lynx lynx_) er det eneste ville kattedyret i Norge. 
Ifølge [Rovdata](https://rovdata.no/Gaupe.aspx) lever arten hovedsakelig i skogsområder over store deler av landet, og de tetteste bestandene er i Midt-Norge og på Østlandet.
Arten er også utbredt over store deler av Europa og Asia, og er vurdert som sterkt truet på Norsk rødliste (2021), som oppgitt på samme side.
Rovdata har ansvar for overvåking av gaupe i Norge.
Indikatoren inngår i Naturindeksen og er beskrevet på [Naturindeksen](https://naturindeks.no/Indicators/gaupe).

2. **Metoder**: Indikatorverdier for gaupebestand hentes fra Naturindeks-databasen (via `NIcalc`-API i denne arbeidsflyten).
Bestandsovervåking og statusrapportering for gaupe koordineres av [Rovdata](https://rovdata.no/Gaupe.aspx).
Status for 2025 oppgis blant annet som et intervall for antall gauper, knyttet til antall familiegrupper [@rovdata_gaupe_status_2025].

3. **Tolkning**: En indikatorverdi på 1 betyr at gaupebestanden i rapporteringsenheten er minst like stor som forventet under referansetilstanden i Naturindeksen.
Verdier som nærmer seg null, viser at bestandsindekset ligger lenger under referansen.

4. **Resultater og utviklingsbehov**: Indikatoren brukes i Naturindeksen.


## 2. About the underlying data

For Naturindeks reporting, prepared indicator objects are obtained from the Naturindeks database.
Lynx population monitoring in Norway is described and reported by [Rovdata](https://rovdata.no/Gaupe.aspx), under the national monitoring programme for large carnivores.

### 2.1 Spatial and temporal resolution and extent

Indicator values are available for the assessment years in Naturindeksen: 1990, 2000, 2010, 2014, 2019, and 2024.

To produce indicator estimates for these years, input is combined over a short window (typically two to three years), as specified in the Naturindeks indicator protocol.

Spatially, data are organised by the eight large-carnivore management regions (_forvaltningsregioner for rovvilt_, regions 1–8).
Rovdata publishes status by these regions, including national targets for annual family groups of lynx per region [@rovdata_gaupe_status_2025].
In Naturindeksen, indicator and reference values are stored at management-region scale (IC units), then aggregated to five geographic regions and national level for reporting.

### 2.2 Sampling design

[Rovdata](https://rovdata.no/Gaupe.aspx) is responsible for monitoring lynx in Norway and for maintaining an overview of how many lynx there are in the country.
The status pages report population status using family groups of lynx and national/regional targets set by the Storting [@rovdata_gaupe_status_2025].


### 2.3 Original units

The unit of the raw data is the number of family groups of lynx?

### 2.4 Instructions for citing, using and accessing data

Population monitoring and status information: cite [Rovdata](https://rovdata.no/Gaupe.aspx) and the relevant status pages [@rovdata_gaupe_status_2025].
Methodological documentation for the indicator: [Naturindeksen – gaupe](https://naturindeks.no/Indicators/gaupe).

<!-- ### 2.5 Additional comments about the dataset -->

<!--# Optional. Text here -->

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

B3 – Functional state characteristics.

<!-- 
### 3.1.1 Other condition typologies -->

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->


### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

The indicator may covary with other large-carnivore indicators that use the same regional structure (wolf, wolverine, brown bear) where distributions overlap.

### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->


### 3.5 Spatial aggregation

Following the metadata pathway Aggregate (population model, large-carnivore management region) – [Scale – Truncate] – Aggregate (national level): values are handled per forvaltningsregion for rovvilt in Naturindeksen, then aggregated to five geographic regions and national level with area weights (see Section 8).
Each unit is normalised against _X~0~_ and _X~100~_ for that region (linear scaling with truncation to [0, 1]).

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition for lynx in Naturindeksen is a boreal/montane forest ecosystem with minimal human impact and a climate similar to the reference period used in the Norwegian Nature Index framework [@nybo_fagsystem_2017].

### 4.2 Reference levels

_X~100~_ is set per large-carnivore management region (IC unit in Naturindeksen), using suitable area and density within each _rovviltregion_ and expert assessment: it represents the expected lynx population (or equivalent index) under reference conditions in that unit.

_X~0~_ is defined as no lynx in the unit (functional absence).


#### 4.2.1 Spatial resolution and validity

Reference levels and indicator values apply at the scale of forvaltningsregioner for rovvilt (IC units).
Aggregated results for the five Naturindeks geographic regions and for Norway are summaries of those units with area weights.

## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->


### 5.2 Reported uncertainties
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->



## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

<!--# Import and prepare the first dataset for analysis. Change the header from Dataset A to the name of the actual dataset. -->

Run the download chunk once when you have API credentials. Credentials are read from `R/_pw.R` (not committed to git). The export is saved under `data/` (also gitignored) so the document can be rendered without calling the API on every build.

```{r lyab-import-api}
#| eval: false
#| code-summary: "Download lynx data from Naturindeks API"
source(here::here("R", "_pw.R"))

lyabImport <- NIcalc::importDatasetApi(
  username = myUser,
  password = myPwd,
  eco = NULL,
  indic = c("Gaupe"),
  year = c("1990", "2000", "2010", "2014", "2019", "2024")
)

saveRDS(lyabImport, here::here("data/lyabImport.rds"))
```

```{r lyab-load-import}
#| include: false
#| cache: false
#| code-summary: "Load saved Naturindeks import"

lyabImport <- readRDS(here::here("data/lyabImport.rds"))
```


## 8. Spatial units

The primary reporting units are forvaltningsregioner for rovvilt (large-carnivore management regions 5–8), which function as ecosystem-relevant IC units for lynx.
Aggregated results for the five Naturindeks geographic regions (NI units: east, south, west, central, north) and for mainland Norway are built from those management-region values with area weights.
Indicator values should be read at management-region scale for detail, and at geographic-region or national scale for overview.

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

Not relevant, as this is the work of a single author. I would, however, like to acknowledge the work of those that created the indicator on _Naturindeks_ [@nater_naturindeks_2025].

::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
