---
title: "Multi-layered forest as share of productive forest area"
subtitle: |
  [NO_MLFA_001]
indicatorID: 
  - NO_MLFA_001
indicatorName: 
  - Multi-layered forest as share of productive forest area
folderName: NO_MLFA_001
country: Norway
continent: Europe
ECT: B2 - Structural State Characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2024 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.002" # Beta-versions are numbers incrementally with values below 0. Operational versions can be given the version numbers above 0. Use this terms similar as for software versioning,
VersionComment: National reference scaling (X0 = 0%, X100 = 30%); scaled regional/national bootstrap
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_MLFA_001
# Leave the url as is, but replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Scale - Transform - Aggregate (area-weighted mean of scaled county values) - Truncate
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Ken Olav Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute for Nature Research
AuthorList: Framstad, E., Storaunet, K. O., Snoksrud, O. and Halbritter, A. H.
date: May 13, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
bibliography: references.bib
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```


<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```
::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 08 Nov. 2025 - Original PR
- 13 May 2026 - ecRxiv template workflow; MLFA denominator includes NA as ensiktet; national scaling (X0 = 0%, X100 = 30%), truncation, county merge for split counties; scaled regional/national bootstrap

:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
This indicator represents share of productive forest area with more than two vertical tree layers.
In forests shaped by natural disturbances and succession, a substantial part of the forest will be multi-layered.
Only where large stand-replacing disturbances occur at a certain intervals may one tree layer be characteristic.
Modern even-aged forestry practices will result in one tree layer for much of the managed forest.
Multi-layered forests will generally have more varied habitats and resources for species than a one-layered forest.
This indicator has been proposed as a voluntary indicator in EU.

2. **Methods**:
Data for this indicator come from the National Forest Inventory (NFI), where the number of tree layers within the 250 m^2^ NFI plot has been recorded for productive forest by current methods since the mid-1990s [@viken2021-feltinstruks].

3. **Interpretation**:
The non-scaled indicator is the area-based ratio *andel flersjiktet / (enskiktet + tosjiktet + NA + flersjiktet)*.
For reporting on a 0–1 condition scale, each county is scaled with national reference levels X0 = 0% and X100 = 30% (provisional); regional and national scaled values are area-weighted means of scaled county values.

4. **Results and development needs**:
We report scaled indicator values (0–1) with bootstrap uncertainty at county, regional, and national levels.
The upper reference level may be revised when expert-based reference values are agreed.
The indicator should not be used at finer spatial resolution than allows reliable estimates, probably regions.


## 2. About the underlying data

The number of tree layers is recorded for each 250 m^2^ NFI inventory plot [@viken2021-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the share of the productive forest area with more than two tree layers, for included NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI. 

### 2.1 Spatial and temporal resolution and extent

Data are available from the mid-1990s, when current recording methods were established. Annual estimates are possible due to representative sampling every year.
For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010.
The data are representative for all productive forest land within the selected region. Spatial resolution should be limited to the resolution allowing for reliable estimates, probably counties.

### 2.2 Sampling design

All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.

### 2.3 Original units

The indicator is specified as the share of productive forest area (%) with more than two vertical tree layers (`best_form` = 3), relative to productive forest where the layer classification is ensiktet, tosjiktet, NA (enskiktet), or flersjiktet.


### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2025-10-20, between NIBIO and the specified users.
Further use of these data requires a separate data user agreement with NIBIO.


### 2.5 Additional comments about the dataset

None

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The vertical tree layers represent a structural attribute of forests.
Hence, this indicator represents a B2 Biotic structural characteristic.

### 3.1.1 Other condition typologies

None


### 3.2 Ecosystem condition characteristic

The vertical tree layers represent a structural attribute of forests. The indicator is characterised as Functionally important species and biophysical structures.

### 3.3 Collinearities with other indicators

There may be some positive correlations with indicators for old forest and dead wood.


### 3.4 Impact factors

Natural disturbances and succession at small scale or with moderate severity will generally result in multi-layered forests.
Severe, stand-replacing disturbances may result in one-layered forests, but such disturbances are expected to be rare.
Modern even-aged forestry practices will result in one tree layer for much of the managed forest.

### 3.5 Spatial aggregation

The indicator should not be used at finer spatial resolution than allows reliable estimates, probably regions. 

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems, where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances. 

### 4.2 Reference levels

For the analyses below we use national reference levels applied at county level: X0 = 0% (no multi-layered forest within the eligible denominator) and X100 = 30% multi-layered share within that denominator (provisional, pending expert review).
The lower reference value (X0) is set to 0%.
For lack of better knowledge, the relationship between indicator values and ecosystem condition is assumed to be linear for values between X100 and X0.
County values are scaled to 0–1 with `scale_indicator`; values above X100 are truncated to 1.
Regional and national scaled point estimates are area-weighted means of scaled county values (not a direct ratio from all plots at once).


#### 4.2.1 Spatial resolution and validity

The reference value will be given as an overall aggregated value for the share of multi-layered productive forest for larger areas such as counties.
Although indicator values from NFI plots may technically be aggregated at finer spatial units than counties, scaling indicator values at such finer spatial levels may result in biases.

## 5. Uncertainties

### 5.1 General uncertainties

Uncertainties relate to the variability in the frequency of forest layers among NFI plots and assessment errors for tree layers within NFI plots.
This uncertainty is quantified by bootstrapping in sections 9.3–9.4; the results report the resulting uncertainty for scaled values.

There will also be uncertainty related to the setting of the upper reference value and to the shape of the relationship between indicator values and ecosystem condition.
These uncertainties will not be quantified.

### 5.2 Reported uncertainties

The results report only scaled values, where the uncertainty reflects the bootstrap workflows in sections 9.3–9.4.

For unscaled values in each reporting period, plots are resampled with replacement within each geographic unit (county, region, or nation).
Each replicate recomputes the same ratio from the resampled plots (total multi-layered `au_areal` divided by total eligible `au_areal`).
The median of that distribution is stored as `indicator_value` in the county/region/national non-scaled uncertainty objects and used when transforming summaries for scaled counties.

For scaled county uncertainty, the summaries from that non-scaled bootstrap (mean, median, quartiles, and confidence interval bounds) are passed through the same national linear scaling function as the point estimates (section 9.4); this is an approximation when many replicates exceed X100, because truncation is not applied inside each replicate before summarising the scaled distribution.

For scaled regional and national uncertainty, a second bootstrap resamples plots, recomputes county non-scaled ratios in each replicate, scales each county with X0 and X100, then takes the area-weighted mean of scaled county values within each region or for Norway; the distribution of those aggregated values is summarised with mean, median, standard error, confidence interval, and quartiles.
The scaled tables in section 10 use the bootstrap median as the reported point estimate where a bootstrap distribution is shown.

## 6. References

::: {#refs}
:::

## 7. Import and prepare datasets

R package versions for this indicator are managed with `renv`.
The project lockfile `renv.lock` records the package versions used to render the document.
After cloning the repository, run `renv::restore()` in the project root before rendering.

### 7.1 Dataset Multi-layered forest

The number of tree layers is recorded for each 250 m^2^ NFI inventory plot [@viken2021-feltinstruks]. There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of the share of the productive forest area with more than two tree layers, for included NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.

```{r load-data}

# Load main data file
lsk_data_raw <- read_csv(here("data/LSK_data_naturregnskap.csv"), 
                     show_col_types = FALSE,
                     locale = locale(encoding = "UTF-8")) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values are reported for counties, regions, and nationally.
Where counties are split into parts in the NFI export, `normalize_county_parts()` merges those labels to a single county name before aggregation (same approach as other forest indicators using this dataset).
Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, probably counties (i.e., de facto Ecosystem Assets).

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.
We retain only productive forest plots (`arealtype` = 1), and then create the derived variables used in subsequent aggregation steps.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

Next we filter the raw dataset to include only productive forest plots, standardise the geography fields, and compute plot-level weights for the MLFA ratio.
The indicator is based on the NFI variable `best_form`, where a value of `"3"` indicates multi-layered forest (more than two tree layers; *flersjiktet*).
Missing `best_form` corresponds to height classes 1 and 2 and is treated as single-layer forest (*enskiktet*).
Those plots stay in the dataset and count toward the denominator together with plots coded explicitly as *enskiktet* or *tosjiktet* (`best_form` `"1"` or `"2"`).
The reported share is *andel flersjiktet divided by (enskiktet + tosjiktet + NA + flersjiktet)*: the total productive forest area represented by multi-layered plots, divided by the total area of all plots that belong to one of those four classes (with NA counted as *enskiktet*).
Plots with any other `best_form` code are left out of both the numerator and the denominator.
Each plot contributes its sample area `au_areal` (hectares) to those totals.

```{r create-derived-variables}
#| cache: false

lsk_data <- lsk_data_raw |>
  dplyr::filter(arealtype == 1) |>
  dplyr::mutate(
    best_form_chr = as.character(best_form),
    best_form_eligible = is.na(best_form) |
      best_form_chr %in% c("1", "2", "3"),
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    fylke_name = normalize_county_parts(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),
    indicator_continuous = dplyr::if_else(
      best_form_eligible & best_form_chr %in% "3",
      1,
      0
    ),
    indicator_continuous_weighted = indicator_continuous * au_areal,
    mlfa_num_weighted = dplyr::if_else(
      best_form_eligible,
      as.numeric(best_form_chr %in% "3") * au_areal,
      0
    ),
    mlfa_den_weighted = dplyr::if_else(
      best_form_eligible,
      au_areal,
      0
    )
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    fylke_name,
    fylke_name_clean,
    best_form,
    indicator_continuous,
    indicator_continuous_weighted,
    mlfa_num_weighted,
    mlfa_den_weighted
  )

```

### 9.2 Calculate non-scaled area-weighted indicator values

With the indicator-specific values we can produce area-weighted summaries for each reporting period at county, region, and national scales.

```{r calculate-mlfa-all-periods}

mlfa_results <- purrr::imap(
  reporting_periods,
  ~ calculate_mlfa_period_indicators(lsk_data, .y, .x)
) |>
  purrr::compact()

mlfa_results_county_all_periods <- purrr::map_dfr(mlfa_results, "county")
mlfa_results_region_all_periods <- purrr::map_dfr(mlfa_results, "region")
mlfa_results_national_all_periods <- purrr::map_dfr(mlfa_results, "national")

```

### 9.3 Bootstrap uncertainty estimation for non-scaled indicator values

Uncertainty in the non-scaled indicator estimates is quantified using bootstrapping.
For each 5-year period and for each spatial unit (county, region, or Norway as a whole), we draw many bootstrap samples of plots with replacement from the plots in that unit.
In every replicate we recalculate the indicator as *total multi-layered sample area divided by total eligible sample area* (the same definition as in section 9.1), then summarise the resulting distribution (mean, standard error, 95% interval, quartiles).
The bootstrap summaries from this step feed scaled county uncertainty in section 9.4.

```{r bootstrap-mlfa-all-periods}

mlfa_bootstrap <- mlfa_results |>
  purrr::imap(
    ~ list(
      period = .y,
      data = .x$data,
      county = .x$county,
      region = .x$region,
      national = .x$national,
      national_boot = bootstrap_mlfa_indicator(.x$data, n_bootstrap = 1000),
      county_boot = purrr::map(
        split(.x$data, .x$data$fylke_name),
        ~ bootstrap_mlfa_indicator(.x, n_bootstrap = 1000)
      ),
      region_boot = purrr::map(
        split(.x$data, .x$data$region_code),
        ~ bootstrap_mlfa_indicator(.x, n_bootstrap = 1000)
      )
    )
  )

mlfa_results_county_non_scaled_uncertainty <- purrr::map_dfr(
  mlfa_bootstrap,
  ~ {
    county_summary <- purrr::imap_dfr(
      .x$county_boot,
      function(boot, county_name) {
        dplyr::tibble(
          fylke_name = county_name,
          mean = boot$mean,
          indicator_value = boot$median,
          se = boot$se,
          ci_lower = boot$ci_lower,
          ci_upper = boot$ci_upper,
          q1 = boot$q1,
          median = boot$median,
          q3 = boot$q3
        )
      }
    )

    .x$county |>
      dplyr::select(fylke_name, period, total_area, n_plots) |>
      dplyr::left_join(county_summary, by = "fylke_name")
  }
)

mlfa_results_region_non_scaled_uncertainty <- purrr::map_dfr(
  mlfa_bootstrap,
  ~ {
    region_summary <- purrr::imap_dfr(
      .x$region_boot,
      function(boot, region_code) {
        dplyr::tibble(
          region_code = region_code,
          mean = boot$mean,
          indicator_value = boot$median,
          se = boot$se,
          ci_lower = boot$ci_lower,
          ci_upper = boot$ci_upper,
          q1 = boot$q1,
          median = boot$median,
          q3 = boot$q3
        )
      }
    )

    .x$region |>
      dplyr::select(region_code, period, total_area, n_plots) |>
      dplyr::left_join(region_summary, by = "region_code")
  }
)

mlfa_results_national_non_scaled_uncertainty <- purrr::map_dfr(
  mlfa_bootstrap,
  ~ .x$national |>
    dplyr::mutate(
      mean = .x$national_boot$mean,
      indicator_value = .x$national_boot$median,
      se = .x$national_boot$se,
      ci_lower = .x$national_boot$ci_lower,
      ci_upper = .x$national_boot$ci_upper,
      q1 = .x$national_boot$q1,
      median = .x$national_boot$median,
      q3 = .x$national_boot$q3
    )
  )

```

### 9.4 Scale indicator values and aggregate to regional and national levels

For each county, the non-scaled point estimate is scaled with national reference levels X0 = 0% and X100 = 30%.
Regional and national scaled point estimates are area-weighted means of scaled county values, weighted by each county's represented area (`total_area` from the county aggregation).

For scaled county uncertainty, the non-scaled bootstrap summaries from section 9.3 are transformed with the same national linear scaling function as for the point estimates.
For scaled regional and national uncertainty, each bootstrap replicate resamples plots, calculates county non-scaled MLFA ratios, scales each county with X0 and X100, then aggregates scaled county values within each region or for Norway before the bootstrap distribution is summarised (`bootstrap_region_scaled_agg_mlfa`, `bootstrap_national_scaled_agg_mlfa` in `R/functions.R`).

The `scale_indicator` function truncates scaled values above 1 (i.e. non-scaled shares above 30% become 1 on the condition scale).

```{r scale-and-aggregate-mlfa}

# National reference levels (provisional): X0 = 0%, X100 = 30%
x0 <- 0
x100 <- 0.30

# Scale county point estimates
county_scaled_all_periods <- mlfa_results_county_all_periods |>
  dplyr::mutate(
    scaled_value = scale_indicator(indicator_value, x0, x100)
  )

# Scaled uncertainty at county level: apply national scaling to non-scaled bootstrap summaries
county_scaled_with_uncertainty <- mlfa_results_county_non_scaled_uncertainty |>
  dplyr::left_join(
    county_scaled_all_periods |>
      dplyr::select(period, fylke_name, scaled_value),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    scaled_mean = scale_indicator(mean, x0, x100),
    scaled_median = scale_indicator(indicator_value, x0, x100),
    scaled_ci_lower = scale_indicator(ci_lower, x0, x100),
    scaled_ci_upper = scale_indicator(ci_upper, x0, x100),
    scaled_q1 = scale_indicator(q1, x0, x100),
    scaled_q3 = scale_indicator(q3, x0, x100)
  )

# Non-scaled regional and national aggregates (from plot-level regions, for truncation reporting)
region_non_scaled_agg_all_periods <- mlfa_results_region_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

national_non_scaled_agg_all_periods <- mlfa_results_national_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    indicator_value = sum(indicator_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  )

# Scaled regional and national point estimates: area-weighted mean of scaled *county* values
region_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- county_scaled_all_periods |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    scaled_value = sum(scaled_value * total_area, na.rm = TRUE) / sum(total_area, na.rm = TRUE),
    total_area = sum(total_area, na.rm = TRUE),
    n_plots = sum(n_plots, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National")

# Truncation: non-scaled ratio exceeds X100 (30%)
truncation_summary <- dplyr::bind_rows(
  mlfa_results_county_all_periods |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      exceeds_threshold, original_value_pct
    ),
  region_non_scaled_agg_all_periods |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_name = assign_region_name(region_code),
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      exceeds_threshold, original_value_pct
    ),
  national_non_scaled_agg_all_periods |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = indicator_value > x100,
      original_value_pct = round(indicator_value * 100, 2),
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      exceeds_threshold, original_value_pct
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_30pct = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_30pct = round(100 * n_exceeding_30pct / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_pct))

```

Non-scaled values above 30% (X100) are truncated to 1.0 after scaling.
The next chunk summarises how often that occurred by period and level, and lists affected units where applicable.

```{r report-truncation-mlfa}
#| echo: false

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding 30%` = n_exceeding_30pct,
    `% exceeding 30%` = pct_exceeding_30pct
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))

if (nrow(truncated_values) > 0) {
  knitr::kable(
    truncated_values |>
      dplyr::mutate(
        location = dplyr::case_when(
          level == "County" ~ fylke_name,
          level == "Region" ~ region_name,
          level == "National" ~ "National",
          TRUE ~ NA_character_
        )
      ) |>
      dplyr::select(period, level, location, original_value_pct) |>
      dplyr::rename(
        Period = period,
        Level = level,
        Location = location,
        `Original value (%)` = original_value_pct
      )
  ) |>
    kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
} else {
  cat("\nNo non-scaled values exceeded 30% (X100) before truncation.\n")
}

```

```{r bootstrap-scaled-uncertainty-mlfa}

region_scaled_uncertainty <- purrr::imap_dfr(
  mlfa_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    if (nrow(period_data) == 0) {
      return(NULL)
    }

    bootstrap_region_scaled_agg_mlfa(
      period_data,
      x0 = x0,
      x100 = x100,
      n_bootstrap = 1000
    ) |>
      dplyr::rename(
        boot_mean = mean,
        boot_median = median
      ) |>
      dplyr::mutate(period = period_name)
  }
)

national_scaled_uncertainty <- purrr::imap_dfr(
  mlfa_results,
  function(period_result, period_name) {
    period_data <- period_result$data
    if (nrow(period_data) == 0) {
      return(NULL)
    }

    boot_result <- bootstrap_national_scaled_agg_mlfa(
      period_data,
      x0 = x0,
      x100 = x100,
      n_bootstrap = 1000
    )

    dplyr::tibble(
      period = period_name,
      region = "National",
      boot_mean = boot_result$mean,
      boot_median = boot_result$median,
      se = boot_result$se,
      ci_lower = boot_result$ci_lower,
      ci_upper = boot_result$ci_upper,
      q1 = boot_result$q1,
      q3 = boot_result$q3
    )
  }
)

region_scaled_agg_with_uncertainty <- region_scaled_agg_all_periods |>
  dplyr::left_join(region_scaled_uncertainty, by = c("period", "region_code")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

national_scaled_agg_with_uncertainty <- national_scaled_agg_all_periods |>
  dplyr::left_join(national_scaled_uncertainty, by = c("period", "region")) |>
  dplyr::mutate(
    scaled_mean = boot_mean,
    scaled_median = boot_median
  )

```

## 10. Results

Section 9.4 documents truncation (non-scaled county ratios above 30% before scaling) and the bootstrap workflows; those outputs are not repeated here as separate non-scaled tables.

The tables below report scaled indicator values on a 0–1 scale, where 0 corresponds to X0 = 0% and 1 corresponds to X100 = 30% multi-layered share within the eligible denominator.

For counties, scaled uncertainty applies the national scaling function to the non-scaled bootstrap summaries from section 9.3.
For regions and Norway, scaled uncertainty uses the second bootstrap in section 9.4 (resample plots → county non-scaled ratios → scale counties → area-weighted mean of scaled counties).
The reported point estimate in these tables is the bootstrap median where a bootstrap distribution is shown; standard errors and intervals follow the same replicate distribution.

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-mlfa-county-scaled}
#| tbl-cap: 'County-level MLFA values by reporting period, scaled to 0–1 using national X0 = 0% and X100 = 30%. County uncertainty applies the same scaling function to the non-scaled bootstrap summaries (section 9.3). Columns include the scaled point estimate, bootstrap summaries, SE, 95% CI, and quartiles.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code)
  ) |>
  dplyr::select(
    period, fylke_name, region_name, scaled_value, scaled_mean, scaled_median, se,
    scaled_ci_lower, scaled_ci_upper, scaled_q1, scaled_q3, total_area, n_plots
  ) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    scaled_ci_lower = round(scaled_ci_lower, 3),
    scaled_ci_upper = round(scaled_ci_upper, 3),
    scaled_q1 = round(scaled_q1, 3),
    scaled_q3 = round(scaled_q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    County = fylke_name,
    Region = region_name,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = scaled_ci_lower,
    `CI Upper` = scaled_ci_upper,
    `Q1` = scaled_q1,
    `Q3` = scaled_q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  DT::datatable(
    rownames = FALSE,
    options = list(pageLength = 10, scrollX = TRUE, lengthMenu = c(10, 25, 50, 100))
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-mlfa-region-scaled}
#| tbl-cap: 'Regional-level scaled MLFA values by reporting period. Point estimates are area-weighted means of scaled county values. Uncertainty uses the scale-then-aggregate bootstrap (section 9.4). The reported point estimate is the bootstrap median.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::select(
    period, region_name, scaled_value, scaled_mean, scaled_median, se,
    ci_lower, ci_upper, q1, q3, total_area, n_plots
  ) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean %||% NA_real_, 3),
    scaled_median = round(scaled_median %||% NA_real_, 3),
    se = round(se %||% NA_real_, 3),
    ci_lower = round(ci_lower %||% NA_real_, 3),
    ci_upper = round(ci_upper %||% NA_real_, 3),
    q1 = round(q1 %||% NA_real_, 3),
    q3 = round(q3 %||% NA_real_, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    Region = region_name,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-mlfa-national-scaled}
#| tbl-cap: 'National-level scaled MLFA values by reporting period. Point estimates are area-weighted means of scaled county values. Uncertainty uses the scale-then-aggregate bootstrap (section 9.4). The reported point estimate is the bootstrap median.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::select(period, region, scaled_value, scaled_mean, scaled_median, se, ci_lower, ci_upper, q1, q3, total_area, n_plots) |>
  dplyr::mutate(
    scaled_value = round(scaled_value, 3),
    scaled_mean = round(scaled_mean, 3),
    scaled_median = round(scaled_median, 3),
    se = round(se, 3),
    ci_lower = round(ci_lower, 3),
    ci_upper = round(ci_upper, 3),
    q1 = round(q1, 3),
    q3 = round(q3, 3),
    total_area = round(total_area, 0)
  ) |>
  dplyr::rename(
    Period = period,
    `Scaled value` = scaled_value,
    `Bootstrapped (mean)` = scaled_mean,
    `Bootstrapped (median)` = scaled_median,
    `Standard Error` = se,
    `CI Lower` = ci_lower,
    `CI Upper` = ci_upper,
    `Q1` = q1,
    `Q3` = q3,
    `Total Area (ha)` = total_area,
    `Number of Plots` = n_plots
  ) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```


## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [EF, KOS]
* Methodology: [EF, KOS]
* Data curation: [OS]
* Formal analysis: [AHH]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [AHH, OS]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
