---
title: "Ptarmigan"
subtitle: |
  [NO_PTAR_001]
indicatorID: 
  - NO_PTAR_001
indicatorName: 
  - Ptarmigan
verbatimName: 
  - Fjellrype # the indicator name that us used in the native language
folderName: NO_PTAR_001
country: Norway
continent: Europe
ECT: B1 - Compositional state characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T6 Polar-alpine (cryogenic) biome
Ecosystem: 
  - NA
verbatimEcosystem: Fjell
yearAdded: 2026
yearLastUpdate: 2026
status: complete
Version: "1.0.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First version on ecRxiv, but this indicator is already pubslised and described in the Norwegian Nature Index
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_PTAR_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (population model, county level) - [Scale - Trunkate] - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Anders Kolstad
    email: anders.kolstad@nina.no
    affiliations:
      - id: myID
        name: The Norwegian Institute for Nature Research (NINA)
AuthorList: Kolstad, A.
date: April 16, 2026
data_availability: silver
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
csl: ambio.csl
---

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull
library(NIcalc)
library(gghighlight)
knitr::opts_chunk$set(echo = TRUE)
#install.packages('logrittr', repos = 'https://guillaumepressiat.r-universe.dev')
library(logrittr)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```

:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 16 Apr. 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
<!-- What does the metric represent? Why is this relevant for describing ecosystem condition in this ecosystem?
What are the main anthropogenic impact factors? -->
Rock ptarmigan (_Lagopus muta_) is a herbivourous alpine bird that is found throughout the Norwegian mountain range where it lives year round. It responds to anthropogenic disturbance, including climate change, through their effects on both survival and reporoduction. We also have a relatively good amount of data and knowledge about this species and the different populations. This combines to make ptarmigan a suitable indicator for ecosystem condition in Norwegian mountains. The species is also a popular game bird.

2. **Methods**:
<!-- What kind of data is used? How is the data customized (modified, estimated, integrated) to fit its purpose as an indicator? -->
The ptarmigan population size is estimated from a statistical model, using hunting statistics as the main data source, and with the assumption that 10% of the population is shot each year. This value is then normalised based on a reference level, which is an expert-assessed population estimate based on known breeding density and the amount of suitable habitat. The indicator is developed in the work with the Norwegian Nature Index [@nater_naturindeks_2025]. The methods are are described in full on the [Nature index web site](https://www.naturindeks.no/Indicators/fjellrype). 

3. **Interpretation**:
<!-- Add a very short description of how the indicator values should be interpreted. Here's an example for a bumblebee indicator: An indicator value of 1 signifies an intact bumblebee community. Values decreasing toward zero reflect negative ecological change, specifically the decline of species expected to be present. This metric does not account for potentially positive changes, such as increases in species' abundance or the arrival of new species. Tip: Add text on the line directly following the subheader, and don't introduce blank lines. This way the whole paragraph renders as a block quote. -->
An indicator value of 1 means that the ptamigan is as numerous, or more, than what we expect under the reference condition.
Values decreasing toward zero indicates that the average county-level populaton size is decreasing.

4. **Results and development needs**:
<!-- Briefly, what does the indicator say about the ecosystem condition. What is the current status of  the metric (can it be used or is it still in development), and what immediate tasks remain to improve the indicator or to make the indicator operational -->
The indicator value for 2024 for Norway as a whole is 0.35	(95% CI = 0.15 - 0.52), which indicates a relatively severe deviation from the reference condition. The development over time is showing a negative trend. 

## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: Rype (_Lagopus muta_) er en hønsefugl som finnes over hele den norske fjellkjeden, hvor den lever hele året. Arten er følsom på menneskeskapte forstyrrelser, inkludert klimaendringer, gjennom deres effekter på både overlevelse og reproduksjon. Vi har relativt gode mengder data og kunnskap om denne arten og de ulike bestandene. Dette gjør rypa til en egnet indikator for økologisk tilstand i norske fjellområder. Arten er også en populær jaktfugl og reguleres delvsis gjennom kvotesettingen.

2. **Metoder**: Bestandsstørrelsen av rype estimeres ved hjelp av en statistisk modell, der jaktstatistikk er hoveddatakilden. Det brukes en antakelse om at 10 % av bestanden skytes hvert år. Denne verdien normaliseres deretter basert på en referanseverdi, som er et ekspertvurdert bestandsestimat basert på kjent hekketetthet og mengden egnet habitat. Indikatoren er utviklet i arbeidet med Naturindeksen [@nater_naturindeks_2025]. Metodene er beskrevet i detalj på [Naturindeksens nettsider](https://www.naturindeks.no/Indicators/fjellrype).

3. **Tolkning**: En indikatorverdi på 1 betyr at rypebestanden er like stor eller større enn det vi forventer under referansetilstanden. Verdier som nærmer seg null indikerer at den gjennomsnittlige bestandsstørrelsen på fylkesnivå er avtagende.

4. **Resultater og utviklingsbehov**: Indikatorverdien for 2024 for Norge er 0,35 (95 % KI = 0,15–0,52), noe som indikerer et relativt stort avvik fra referansetilstanden. Utviklingen over tid viser en negativ trend.


## 2. About the underlying data

<!--# Give a general introduction the datasets you have used, and provide more specific details in the sub-chapters.  -->
The data comes from hunting statistics (harvest bag data) published by SSB (Statistics Norway). The procedure to download the data is described [here](https://naturindeks.nina.no/docs/24dda0ba-5855-4131-9168-d874980ec6aa.pdf). 

### 2.1 Spatial and temporal resolution and extent

<!--# Describe the temporal and spatial resolution and extent of the data used -->
Indicator values are available for the different assessment years in  _Naturindeksen_: 1990, 2020, 2010, 2011, 2012, 2013, 2014, 2019 and 2024. To produce indicator estimates for these years, one uses the average over 2-3 year. For the year 2024, data from two hunting seasons were included (2022/23-2023/24).

The indicator is calculated for all counties in Norway. Here we report the values for five geographic regions. 

### 2.2 Sampling design

<!-- How was the data sampled? Describe potential sampling biases -->

The data originates from hunters when they report their catches, something they are required to do. The sampling/hunting effort is not known, and assumed to be stable. This is a rather big assumption.  

### 2.3 Original units

<!--# What are the original units for the most relevant  variables in the data-->
The original units of the raw data is the number of individuals. The data are assigned to spatial units, so this could be turned into measures of density.

### 2.4 Instructions for citing, using and accessing data

<!--# Is the data openly available? If not, how can one access it? What are the key references to the datasets?   -->
The underlying data (hunting statistics) are available via SSB, but not in a prepared format (see [here](https://naturindeks.nina.no/docs/24dda0ba-5855-4131-9168-d874980ec6aa.pdf)).

Indicator point estimates and quantiles, for different geographic units, are freely available [@nater_ni2025_2025].


<!-- ### 2.5 Additional comments about the dataset -->

<!--# Optional. Text here -->

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

<!--# Describe the rationale for assigning the indicator to the ECT class. See https://oneecosystem.pensoft.net/article/58218/
This doesnt need to be very long. Maybe just a single sentence. -->
B1 - Compositional state characteristics
<!-- 
### 3.1.1 Other condition typologies -->

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->

The ecosystem characteristic represented in this indicator is an ecosystem where species abundances and composition is similar to that in the reference condition. 

### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

The indicator might be correlated to other indicators, especially other indicators within ECT class B1, such as the species that prey on ptarmigan (arctic fox, golden eagle), or other species that have similar habitats or way of live (such as willow grouse).

### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->
Ptarmigan is sensitive to several antropogenic pressures, especially climate change and hunting, but also habitat destruction and disturbance by human activities. 

### 3.5 Spatial aggregation

<!-- Describe in word the same aggregation pathway presented in the top yaml (the metadata). If you have scaled against the reference value, it is especially important to know if you did any avergaing or summations etc. beforehand. An example text for a rather simple case could be: "The variable is scaled at the initial polygon level and the point estimate is a area-weighted arithmetic mean of all polygins. There is no transformation, and truncation is not needed as the indicator becomes naturally bound between 0 and 1".  -->

The hunting statistics are aggregated to county level using a population model. There the variable value is normalised by scaling and truncating against X~100~ and X~0~. The indicator values are then resampled to five geographic regions, and then to Norway, with a sampling weight proportional to the mountain area in each region or area. 


## 4. Reference condition and levels

### 4.1 Reference condition

<!--# Define the reference condition (or refer to where it is defined). Note the distinction between reference condition and reference levels 10.3897/oneeco.5.e58216  -->
The reference condition is a moutain ecosystem with minimal human impact and a climate similar to the period 1961-1990 [@nybo_fagsystem_2017].

### 4.2 Reference levels

<!--# 

If relevant (i.e. if you have normalised a variable), describe the reference levels used to normalise the variable. 

Use the terminology where X~0~ refers to the reference level (the variable value) denoting the worst possible condition; X~100~denotes the optimum or best possible condition; and X~*n*~, where in is between 0 and 100, denotes any other anchoring points linking the variable scale to the indicator scale (e.g. the threshold value between good and bad condition X~60^). 

Why was the current option chosen and how were the reference levels quantified? If the reference values are calculated as part of the analyses further down, please repeat the main information here.

 -->

X~100~ is defined based on the predicted number of ptarmigan individuals based on a statistical model. This value varies between counties.

X~0~ is defined as a complete absence of ptarmigan.

X~60~ is not specified and defaults to 60% of the X~100~ value. 

#### 4.2.1 Spatial resolution and validity

<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->
 The reference values vary with counties, and normalisation of variable values should be done after aggregating those first to the same county scale. 


## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->
There are obvious limitations to using havest bag data, rather than more standardised monitoring data. This is elaborated in the [indicator documentation on Naturindeksen](https://naturindeks.nina.no/docs/24dda0ba-5855-4131-9168-d874980ec6aa.pdf). However, we feel the hunting data still has considerable value and that it is indeed suitable for this purpose of estimating population sizes. 

### 5.2 Reported uncertainties
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->

The reported uncertainties reflect both the uncertainty in the reference levels (something which is accommodated in the naturindeks workflow), the uncertainty in the county estimates, and when spatially aggregating, also the spatial variation between counties.  

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

### 7.1 Naturindeks

<!--# Import and prepare the first dataset for analysis. Change the header from Dataset A to the name of the actual dataset. -->

In the Naturindeks database I can get indicator values directly, like this:

```{r getData}
#| eval: false
#| code-fold: true
#| code-summary: 'Import data from Naturindeks'


#I get the indicator values, cleaned, normalised and aggregated, from the NIcalc package and the NI database (@fig-hist). 
#The variable values and reference values are not imported. 
#I use ´NIcalc::calculateIndex()´ to create an index from a single indicator. 
#This forces NIcalc to output 1000 draws regardless of whether the indicator uses
#a custom distribution or if the uncertainty is reported by just the point estimate,
#the quartiles and the distribution family (as in this case).

# password is hidden for now
source(here::here("R", "_pw.R"))

# Import data set
ptarImport <- NIcalc::importDatasetApi(username = myUser,password = myPwd,
                        eco = NULL, indic = c("Fjellrype"),
                        year = c("2010","2014","2019", "2024"))
# Assemble
ptarInput <- NIcalc::assembleNiObject(inputData = ptarImport,
                        predefNIunits = c(allArea = T, parts = T, counties = F),
                        indexType = "thematic")
# Calculate index for amphibians 
ptarIndex <- NIcalc::calculateIndex(
  x = ptarInput, 
  nsim = 1000,
  fids = FALSE, # use fidelity to the ecosystem in the weigthing 
  tgroups = FALSE, # Use weighting based on trophic groups
  keys = "ignore", # What do do with key indicators 
  w = 0, # Controls how much influence key indicators get relative to others. Only used in one specific mode (keys = "specialWeight")
  awbs = TRUE, # Use area-weighted BSunits. BS units are municipalities or counties, and the area is the total area, not the mountain area.
  awBSunit = "Fjell")

# I checked the outputs and NIunitWeights corresponds to the area of mountains in each municipality.
# I cannot change the mountain area used in the weighting. This is not ideal since we now have a new moutain definition,
# but I'ts probably not adding a big systematic bias either.

#saveRDS(ptarIndex, here::here("data/ptarIndex_data.RDS"))
```

However, we will use a slightly different approach where we get the variable values and the reference values from the NI database.
The reference values have uncertainties attached to them, which we will ignore here.
The variable values have uncertainties as well, in the form of probability function. 

```{r}
# password is hidden for now
source(here::here("R", "_pw.R"))

# Import data set
ptarImport <- NIcalc::importDatasetApi(
  username = myUser,
  password = myPwd,
  eco = NULL,
  indic = c("Fjellrype"),
  year = c("2010", "2014", "2019", "2024")
)

# Assemble
ptarAssemble <- NIcalc::assembleNiObject(
  inputData = ptarImport,
  predefNIunits = c(allArea = T, parts = T, counties = F),
  indexType = "thematic"
)
```


Nå simulerer jeg *n* verdier fra sannsynelighetsfunksjonen til indikatoren. 

First I get the reference values


```{r refs}
x100 <- ptarImport$referenceValues$referenceValues |>
  select(ICunitId, ICunitName, x100 = expectedValue)
x100
```

These are unique to each county (ICunit)

Then we get the variable values as resamples values from a probability distribution.
Jeg bruker ´tradObs´ der det er en vanlig fordelingsfamilie, og customObs der customDist!= NA. Når distributionFamilyName == NA så er customDistribution != NA.

```{r vars} 
myYears <- as.character(c(2010, 2014, 2019, 2024))
obstype <- NULL

for (i in 1:length(myYears)) {
  obs <- ptarAssemble$indicatorValues[[i]]$distributionFamilyName

  obs[!is.na(obs)] <- "tradObs"
  obs[is.na(obs)] <- "customObs"

  obstype[[i]] <- obs
}

allVariableValues <- tibble(ICunitId = NULL, value = NULL, year = NULL)

for (i in 1:length(myYears)) {
  #print(i)
  myMat <- NIcalc::sampleObsMat(
    ICunitId = ptarAssemble$indicatorValues[[i]]$ICunitId,
    value = ptarAssemble$indicatorValues[[i]]$expectedValue,
    distrib = ptarAssemble$indicatorValues[[i]]$distributionFamilyName,
    mu = ptarAssemble$indicatorValues[[i]]$distParameter1,
    sig = ptarAssemble$indicatorValues[[i]]$distParameter2,
    customDistribution = ptarAssemble$indicatorValues[[i]]$customDistribution,
    obsType = obstype[[i]],
    nsim = 1000
  )
  ICunitIds <- row.names(myMat)
  
  myMat <-  myMat |>
    as_tibble() |>
    add_column(ICunitId = ICunitIds, .before = 1) |>
    pivot_longer(cols = -any_of("ICunitId")) |>
    select(-name) |>
    add_column(year = myYears[i])
  
  allVariableValues <- allVariableValues |> 
    bind_rows(myMat)
  #assign(paste0("myMat", myYears[i]), myMat)
}
```

This is the point in the spatial aggregation where we should normalise against the reference values.
So let's add inn the X100 values.

```{r}
allVariableValues <- allVariableValues |>
  left_join(x100 |> mutate(ICunitId = as.character(ICunitId)), by = join_by(ICunitId))
```

Then we can normalise also the scatter plot with variables against indicator values.

```{r}
library(ecTools)

```


Then we can aggregate by a weighted resampling from counties to geographic regions, and then to Norway, making sure sample from the the inferential confidence intervals and not the descriptive ones we have now.

Let's first join all the data into one very long dataframe, before we add in the region names and the mountain area for each county.

```{r}

```

```{r getRegNames}

# Radene i denne matrisen er ICunitIDs (fylker). Jeg kan knytte de til navnene til NIunits.
link <- as.data.frame(ptarAssemble$NIunits) |>
  select(-wholeArea) %>=%
  rownames_to_column("BCunit") %>=%
  pivot_longer(cols = -any_of("BCunit")) %>=%
  filter(value > 0) %>=%
  select(-value) %>=%
  rename(region = name)

anyDuplicated(link$ICunit)


link2 <- as.data.frame(ptarAssemble$ICunits) %>=%
  rename(ICunitID = Fjellrype) |>
  rownames_to_column("BCunit")

link <- link %>=%
  left_join(link2, by = join_by(BCunit)) |>
  distinct(ICunitID, region)


#nord <- link$ICunitID[link$region == "N"]
#midt <- link$ICunitID[link$region == "C"]
#vest <- link$ICunitID[link$region == "W"]
#sor <-  link$ICunitID[link$region == "S"]
#ost <-  link$ICunitID[link$region == "E"]


```



```{r}

# Here are the variable values estimates per region
regions <- c("N", "C", "W", "S", "E")

varVal <- data.frame(
  reg = rep(regions, each = length(myYears)),
  year = rep(myYears, length(regions)),
  low = NA,
  med = NA,
  upp = NA
)

for (i in regions) {
  for (n in myYears) {
    tempMat <- get(paste0("myMat", n))
    tempMat <- as.data.frame(tempMat)
    myMatrX <- x100

    if (i == "N") {
      tempMat <- tempMat[row.names(tempMat) %in% nord, ]
    }
    if (i == "C") {
      tempMat <- tempMat[row.names(tempMat) %in% midt, ]
    }
    if (i == "W") {
      tempMat <- tempMat[row.names(tempMat) %in% vest, ]
    }
    if (i == "S") {
      tempMat <- tempMat[row.names(tempMat) %in% sor]
    }
    if (i == "E") {
      tempMat <- tempMat[row.names(tempMat) %in% ost, ]
    }

    #   if(i == "N") x100 <- x100[row.names(x100) %in% nord,]
    #   if(i == "C") x100 <- x100[row.names(x100) %in% midt,]
    #   if(i == "W") x100 <- x100[row.names(x100) %in% vest,]
    #   if(i == "S") x100 <- x100[row.names(x100) %in% sor,]
    #   if(i == "E") x100 <- x100[row.names(x100) %in% ost,]

    varVal[varVal$reg == i & varVal$year == n, 3:5] <-
      quantile(colSums(tempMat), c(0.025, .5, .975))
  }
}
```





```{r readDat}
#| echo: false
ptarIndex <- readRDS(here::here("data/ptarIndex_data.RDS"))
```

```{r fig-hist}
#| code-summary: 'Simple histogram'
#| fig-cap: 'Histogram of some aggregated indicator values for Norway.'
hist(ptarIndex$wholeArea$'2024'$index )
# The point estimate (median) and the 95% CI math those on the Naturindeks website pretty close. 
```



## 8. Spatial units

<!--# Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpreted at. Potential spatial delineation data (maksing layers etc.) should be introduced under chapert 7. We recommend using the SEEA EA terminology of Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). -->

We treat the Norwegian counties as BSunits, and the five geographic regions in Norway (east, south, west, central, north) as the Ecosystem assets. The Ecosystem accounting Area is the alpine areas within the whole of mainland Norway.

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

The data imported from Naturindeks contains the indicator values.
We have not imported the variable or reference values, and therefore 
do not plot the variable values against the indicator values.

The indicator uses scaling and truncation, but not any transformation (i.e. there's a linear relationship between the variable values and the indicator values within the zero to 1 range). 

Now we prepare the data for export (@tbl-exp). We do thus using a generic function.

```{r NIprepare_export}
#| code-summary: 'Function'
create_output <- function(dat) {
  regions <- names(dat)

  out <- list()

  for (region in regions) {
    years <- names(ptarIndex[[region]])

    for (year in years) {
      col_name <- paste(region, year, sep = "_")

      # Optional renaming: wholeArea becomes All
      col_name <- gsub("^wholeArea_", "All_", col_name)

      out[[col_name]] <- dat[[region]][[as.character(year)]][["index"]]
    }
  }

  as_tibble(out)
}

```


```{r export}
#| eval: false
#| #| code-summary: 'Export data'
ptarExport <- create_output(dat = ptarIndex)
write.csv(ptarExport, here::here("data/results_NO_PTAR_001_v1.0.0.csv"), row.names = F)
```

```{r tbl-exp}
#| code-summary: 'Print dataframe'
#| tbl-cap: 'Table showing the first 6 (out of 1000) rows from the exported indicator values file.'
ptarExport <- read.csv(here::here("data/results_NO_PTAR_001_v1.0.0.csv"))
head(ptarExport)
```

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

```{r tbl-results}
#| warning: false
#| message: false
#| tbl-cap: 'Indicator values and uncertainty estimates for NO_PTAR_001.'
ptarResults <- ptarExport |>
  pivot_longer(cols = everything(), names_to = "name", values_to = "value") |>
  separate(col = name, into = c("Region", "Year"), sep = "_") |>
  group_by(Region, Year) |>
  summarise(
    median = round(median(value), 2),
    p_025 = round(quantile(value, probs = 0.0025), 2),
    p_25 = round(quantile(value, probs = 0.25), 2),
    p_75 = round(quantile(value, probs = 0.75), 2),
    p_975 = round(quantile(value, probs = 0.975), 2),
  ) |>
  add_column(
    ECT = "B1",
    Indicator = "Ptarmigan",
    Ecosystem = "Mountains",
    ID = "NO_PTAR_001"
  )

ptarResults |>
  unite(col = "CI_95", c(p_025, p_975), sep = " - ") |>
  unite(col = "IQR", c(p_25, p_75), sep = " - ") |>
  select(
    ECT,
    Indicator,
    ID,
    "Accounting area" = Region,
    Ecosystem,
    "Closing year" = Year,
    "Indicator value" = median,
    "95% CI" = CI_95,
    "50% CI" = IQR
  ) |>
  kable()
```


```{r fig-trend}
#| fig-cap: 'Trend in indicator values over time for NO_PTAR_001. The indicator values are for the national scale (mountains on mainland Norway), with regional trend lines shown as unlabeled grey lines. Errorbars are 95% confidence intervals.'
#| warning: false
ptarResults |>
  ggplot(aes(x = as.numeric(Year), y = median, colour = Region)) +
  geom_line(linetype = 2) +
  gghighlight::gghighlight(Region == "All", use_direct_label =F) +
  geom_point(size = 10) +
  gghighlight::gghighlight(Region == "All", unhighlighted_params = list(size = 2), use_direct_label =F) +
  geom_errorbar(aes(ymin = p_025, ymax = p_975), width = 0.2) +
  theme_bw(base_size = 15) +
  labs(x = "Year", y = "Indicator value") +
  guides(colour = "none") +
  scale_x_continuous(breaks = c(2010, 2014, 2019, 2024))
```

## 11. Export file

The indicator values are exported and can be found under `data/results_NO_PTAR_001_v1.0.0.csv`.

## 12. Author contributions

Not relevant, as this is the work of a single author. I would, however, like to acknowledge the work of those that created the indicator on _Naturindeks_ [@nater_naturindeks_2025].



::: {.callout-tip collapse="true"}
## Session Info

This workflow uses `renv`. Run `renv::restore()` to recreate the working environment. 
 
:::