---
title: "Rowan, aspen, and goat willow"
subtitle: |
  NO_ROSE_001  
indicatorID: 
  - NO_ROSE_001
indicatorName: 
  - Rowan, aspen, and goat willow
verbatimName: 
  - Rogn, osp og selje
folderName: NO_ROSE_001
country: Norway
continent: Europe
ECT: B2 Structural state characteristics
Realm: 
  - Terrestrial (T)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
  - T2.2 Deciduous temperate forests
verbatimEcosystem: 
  - T2.1.2 Boreal/subalpin skog
  - T2.2.2 Temperert edellauvskog
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "000.001" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: Initial version # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_ROSE_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (NFI plots to county/sub-county) - [Scale - Truncate] - Aggregate (geographic region) - Aggregate (national level)
Precision: Approximate 95% bootstrap interval (2.5th–97.5th percentile of 1000 bootstrap draws) as inferential uncertainty around the point estimate (median). At county or sub-county level, bootstrapping is applied directly to NFI plot data; at regional and national level, uncertainty is propagated from scaled county bootstrap draws.
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud Helen Halbritter
    affiliations:
      - name: University of Bergen
  - name: Ulrika Jansson Asplund
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Erik Framstad
    email: erik.framstad@nina.no
    affiliations:
      - name: Norwegian Institute for Nature Research
  - name: Oliver Snoksrud
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
  - name: Ken Olaf Storaunet
    affiliations:
      - name: Norwegian Institute of Bioeconomy Research
AuthorList: Halbritter, A. H., Jansson Asplund, U., Framstad, E., Snoksrud, O., and Storaunet, K. O.
date: May 7, 2026
data_availability: bronze
code_reproducibility: gold
open_science_badge: silver
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
if (file.exists("renv/activate.R")) {
  source("renv/activate.R")
} else if (file.exists("../renv/activate.R")) {
  source("../renv/activate.R")
}

library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(janitor)
library(stringi)
library(DT)
library(gt)
library(ecTools)
library(MetBrewer)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
source(here("R/functions.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 07 May. 2026 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

1. **Background and rationale**:
Rowan (*Sorbus aucuparia*), aspen (*Populus tremula*) and goat willow (*Salix caprea*) (collectively abbreviated ROS) are boreal tree species which represent important substrate, habitat or food sources for many fungi, lichens, bryophytes, and animals.
Under natural forest conditions these tree species favour disturbances like floods, fires and storms with ensuing succession to provide openings in the forest and avoid competition from dense spruce forest.
The aggregated volume per hectare of ROS species is considered to be positively associated with ecosystem condition.
Data for this indicator come from the National Forest Inventory (NFI), where the volume of ROS species for trees >10 cm diameter at breast-height has been recorded for each NFI plot since 1994. The recorded mean volume of ROS species is compared to reference values for regions, based on expert assessments. 

2. **Methods**:
For each NFI plot, the combined volume of rowan, aspen and goat willow is taken from the NFIvariable `ROS_10_PRHA` (m³ ha⁻¹).
Results are calculated for sequential 5-year periods from 2010 and aggregated with area weights to county or sub-county, regional and national units.
The value for each county or sub-county is scaled against its specific reference level before regional and national values are calculated as area-weighted means of scaled county values.

3. **Interpretation**:
An indicator value of 1 implies that the ROS volume is at least as high as expected in the reference condition.
Values decreasing toward zero reflect lower combined volume of rowan, aspen and goat willow relative to the reference.

4. **Results and development needs**:
For the current reporting period (2020-2024) the national indicator value is 0.45 (95% CI: 0.41 - 0.49). The regional indicator values range from 0.34 - 0.67, with the East having the lowest value and the South having the highest value.
The indicator values increase over time nationally and for all regions.
There are no specific development needs for the indicator measure.
However more thorough analyses and review of the assessment of reference values and the scaling function is needed. 


## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**:
Rogn (*Sorbus aucuparia*), osp (*Populus tremula*) og selje (*Salix caprea*) (samlet forkortet ROS) er boreale treslag som representerer viktige substrat, habitat eller matkilder for mange sopper, lav, moser og dyr.
I skog med naturlig dynamikk er disse treslagene avhengige av forstyrrelser som flom, brann og stormer, med påfølgende suksesjon, for å skape åpninger i skogen og unngå konkurranse fra tette granbestand. 
Samlet volum per hektar av ROS-arter anses å være positivt knyttet til økosystemtilstanden. Data for denne indikatoren kommer fra Landsskogtakseringen (LSK), der volumet av ROS-arter for trær >10 cm diameter i brysthøyde har blitt registrert for hver LSK-prøveflate siden 1994.
Det registrerte gjennomsnittlige volumet av ROS-arter blir sammenlignet med referanseverdier for regioner, basert på ekspertvurderinger. 

2. **Metoder**: 
For hver LSK-prøveflate tas samlet volum av rogn, osp og selje fra variabelen ROS_10_PRHA (m³ ha⁻¹).
Resultatene beregnes for påfølgende 5-årsperioder fra 2010 og aggregeres med arealvekter til fylkes/delfylkesnivå, regioner og nasjonalt.
Verdien for hvert fylke/delfylke skaleres mot sin spesifikke referanseverdi før regionale og nasjonale verdier beregnes som arealveide gjennomsnitt av skalerte fylkesverdier. 

3. **Tolkning**:
En indikatorverdi på 1 betyr at ROS-volumet er minst like høyt som forventet under referansetilstanden.
Verdier som går mot null, viser lavere samlet volum av rogn, osp og selje sammenlignet med referansen.  

4. **Resultater og utviklingsbehov**: 
For nåværende rapporteringsperiode (2020-2024) er den nasjonale indikatorverdien 0.45 (95% CI: 0.41 - 0.49). Regionale indikatorverdier varierer mellom 0.34 - 0.67, med Øst som har lavest verdi og Sør som har høyest verdi.
Indikatorverdien øker over tid både nasjonalt og for alle regioner.
Det er ingen spesifikke utviklingsbehov, bortsett fra en mulig gjennomgang av referanseverdier og skaleringsfunksjonen.


## 2. About the underlying data

The combined volume of rowan, aspen and goat willow is recorded for each Norwegian National Forest Inventory (NFI) plot [@viken2025-feltinstruks].
There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county.
The indicator value is an unbiased area-weighted estimate of combined rowan-aspen-goat willow volume (m³ ha⁻¹) for NFI plots in any selected region.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year.
For this indicator data from each full inventory cycle of 5 years are used, starting in 2010, meaning that results reported for 2024 are based on data from the years 2020-2024.
The data are available upon request from NFI.

### 2.1 Spatial and temporal resolution and extent

Data are available from 2010, when NFI had extended the grid to all forest land. Annual estimates are possible due to representative sampling every year. For this indicator all data from each full inventory cycle of 5 years are used, starting in 2010. Spatial resolution should be limited to the resolution applied to reference values, that is counties or sub-county units for scaling and regions for aggregated reporting.

### 2.2 Sampling design

There are about 12 000 NFI plots in forest ecosystems, distributed on a regular grid, with varying grid density for forest below/above the coniferous forest line and in Finnmark county [@viken2025-feltinstruks].
In each NFI plot, all individual trees with a diameter ≥5 cm at breast height are determined to species and measured for several variables, including diameter at breast height.
From the data, the volume of individual trees are calculated.
For this indicator, we use the combined volume per hectare of trees of rowan, aspen, and goat willow with a diameter at breast height of at least 10 cm.
All NFI plots are sampled over a 5-year period, with a representative sampling scheme every year, allowing unbiased estimates from data for each year. 

### 2.3 Original units

The indicator is specified as combined standing volume per hectare (m³ ha⁻¹) for rowan, aspen and goat willow on the NFI plot (`ROS_10_PRHA`).

### 2.4 Instructions for citing, using and accessing data

The data used for this indicator have been made available by the Norwegian National Forest Inventory under a specific data use agreement of 2026-05-29, between NIBIO and the specified users.
Further use of the data requires a separate data user agreement with NIBIO

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

The ROS species represent important structural components in boreal forests and deviate in various properties from the more common coniferous spruce and pine.
Hence, the indicator is classified as a B2 Structural state characteristic.


### 3.2 Ecosystem condition characteristic 

### 3.3 Collinearities with other indicators

No direct correlation with other indicators is expected. 

### 3.4 Impact factors

Natural disturbances such as fires, floods, or windthrows are important to maintain site conditions for recruitment of rowan, aspen, and goat willows (ROS).
Human impacts with negative effects on ROS species are linked to forestry activities, especially clear-cutting, dense planting, other forest management, and fire prevention, although clearcuts may also open the forest for increased recruitment of ROS species.
Human management resulting in high abundances of livestock or wild ungulates will negatively affect recruitment of ROS species.
Human management of ground conditions, other tree species and animal browsing may also directly facilitate recruitment and growth of ROS species. See @solberg2011_ros. 

### 3.5 Spatial aggregation

The spatial aggregation pathway is Aggregate (NFI plots to county/sub-county) – [Scale – Truncate] – Aggregate (geographic region) – Aggregate (national level).
For each NFI plot we create `indicator_continuous_weighted = ros_10_prha * au_areal`.
Non-scaled indicator values for each reporting period are calculated separately for each geographic unit as area-weighted arithmetic means of all NFI plots within that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
This is done for counties or sub-counties, regions, and at national level.
Scaling and truncation are applied only after the non-scaled county or sub-county aggregates have been calculated: each county or sub-county is scaled against its own lower reference level _X~0~_ (0 m³ ha⁻¹) and upper reference level _X~100~_ using a linear scaling function, and values above 1 are truncated to 1 at that step.
Regional and national scaled point estimates are area-weighted means of the already scaled county or sub-county values, using each county or sub-county's represented area as weight.
Non-scaled regional and national point estimates are still calculated directly from all plots within each region or the whole country.

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition is considered to be intact forest ecosystems where effects of human activities are clearly subordinate to those of natural ecosystem processes and disturbances.  

### 4.2 Reference levels

The lower reference level _X~0~_ is set to 0 m³ ha⁻¹ and the upper reference level _X~100~_ is set separately for each county or sub-county.
X0 denotes the worst possible condition and X100 denotes the best possible condition on the original volume scale.
Between _X~0~_ and _X~100~_, indicator values are scaled linearly to a 0-1 indicator scale.
Scaled values above 1 are truncated to 1 when volume exceeds the county-specific X100.
The county and sub-county _X~100~_ values used in the analyses are listed in @tbl-rose-reference-values. _X~0~_ is 0 m³ ha⁻¹ for all units.

#### 4.2.1 Spatial resolution and validity

Reference levels are defined for regional groupings of counties or sub-counties.
Scaling is therefore applied to area-weighted non-scaled indicator values that have first been aggregated to the matching county or sub-county unit.
Regional and national reporting uses area-weighted aggregation of these scaled county or sub-county values.
The indicator should not be interpreted at finer spatial resolution than the county or sub-county reference levels allow.

## 5. Uncertainties
### 5.1 General uncertainties
Reference levels are held fixed in this workflow (their own uncertainties are not propagated).
Regional and national values are area-weighted summaries of counties, so they inherit both model uncertainty at county level and any mismatch between county boundaries and ecological variation within regions. 

### 5.2 Reported uncertainties

The uncertainties reported in @sec-results are inferential and refer to uncertainty around the point estimates, not to the full spatial variation in rowan-aspen-goat willow volume among NFI plots.
For each reporting period, non-scaled indicator values are bootstrapped by resampling NFI plots with replacement within the relevant geographic unit and recalculating the area-weighted mean.
This is done separately for counties or sub-county units, regions, and the national level.
For scaled county or sub-county values, the bootstrap mean, median, standard error, confidence interval, and quartiles from the non-scaled bootstrap are transformed with the same sub-county-specific linear scaling function used for the point estimate.
For scaled regional and national values, uncertainty is propagated from the scaled county or sub-county bootstrap draws (section 9.4).
The resulting Monte Carlo distributions are summarised with median and quantiles for the results tables in section 10.

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->
::: {#refs}
:::

## 7. Import and prepare datasets

The combined volume of rowan, aspen and goat willow is quantified for each NFI plot from `ROS_10_PRHA`.
For this indicator data from each full 5-year inventory cycle, starting in 2010, are used.

```{r load-data}
#| code-summary: "Load data"

lsk_data_file <- here("data/LSK_data_okologisk_tilstand_2026-06-03.csv")

lsk_data_raw <- read_csv(
  lsk_data_file,
  show_col_types = FALSE,
  locale = locale(encoding = "UTF-8")
) |>
  clean_names()

```

## 8. Spatial units

Aggregated indicator values should not be applied for spatial units finer than those that allow reliable estimates, e.g., regions.

## 9. Analyses

### 9.1 Prepare data and create derived variables

This section prepares the NFI plot data for calculating the indicator.

First we define the 5-year reporting periods that match the temporal structure of the indicator.

```{r reporting-periods}
#| code-summary: "Define reporting periods"

# Define multiple 5-year reporting periods starting from 2010
reporting_periods <- list(
  "2010-2014" = 2010:2014,
  "2015-2019" = 2015:2019,
  "2020-2024" = 2020:2024
)

```

Next we filter the raw dataset to include both productive and unproductive forest plots (`arealtype` 1 and 12) and standardise the county names.
Area-weighted values are created using `indicator_continuous_weighted = ros_10_prha * au_areal`.

```{r create-derived-variables}
#| cache: false
#| code-summary: "Create derived variables"

lsk_data <- lsk_data_raw |>
  filter(
    # productive and unproductive forest plots
    arealtype %in% c(1, 12)
  ) |>
  mutate(
    fylke_name = region_fylke |>
      stringr::str_remove("^\\d+-\\d+\\s+") |>
      stringr::str_remove("^\\d+\\s*") |>
      stringr::str_squish(),
    county_part = assign_bbca_county_part(fylke_name),
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(
          stringr::str_replace_all(
            normalize_county_parts(fylke_name),
            "[0-9.-]",
            " "
          )
        )
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(fylke_name_clean),
    region_name = assign_region_name(region_code),

    indicator_continuous_weighted = ros_10_prha * au_areal
  )

```

Finally, we retain only the columns required for subsequent aggregation and reporting.

```{r select-downstream-fields}
#| code-summary: "Select downstream columns"

lsk_data <- lsk_data |>
  dplyr::select(
    sesong,
    au_areal,
    arealtype,
    region_code,
    region_name,
    county_part,
    fylke_name,
    fylke_name_clean,
    ros_10_prha,
    indicator_continuous_weighted
  )

```


### 9.2 Calculate non-scaled area-weighted indicator values

For each reporting period, we calculate non-scaled indicator values separately for counties or sub-counties, regions, and the whole country.
Within each geographic unit, the point estimate is the area-weighted arithmetic mean of all NFI plots in that unit: sum of `indicator_continuous_weighted` divided by sum of `au_areal`.
County or sub-county estimates use the plot-level `fylke_name` field; regional estimates use `region_code`; the national estimate uses all plots in the period.

```{r calculate-rose-all-periods}
#| code-summary: "Calculate non-scaled values"

 rose_results <- purrr::imap(
   reporting_periods,
   ~ calculate_period_indicators(lsk_data, .y, .x)
 ) |>
   purrr::compact()
 
 rose_results_county_all_periods   <- purrr::map_dfr(rose_results, "county")
 rose_results_region_all_periods   <- purrr::map_dfr(rose_results, "region")
 rose_results_national_all_periods <- purrr::map_dfr(rose_results, "national")

```

### 9.3 Bootstrap draws per county

For each county and period, NFI plots are resampled with replacement 1000 times.
Each replicate computes the area-weighted mean non-scaled indicator value, giving a distribution of 1000 draws per county.

```{r county-draws-long}
#| code-summary: "Bootstrap draws per county (1000 per county per period)"
#| cache: true
#| cache.extra: !expr file.mtime(here::here("R/functions.R"))

county_draws_long <- purrr::imap_dfr(rose_results, function(period_result, period_name) {
  county_meta <- period_result$county |>
    dplyr::select(fylke_name, county_part, total_area, region_code)

  purrr::imap_dfr(
    split(period_result$data, period_result$data$fylke_name),
    function(county_data, county_name) {
      boot <- bootstrap_indicator(county_data, n_bootstrap = 1000)
      dplyr::tibble(
        period           = period_name,
        fylke_name       = county_name,
        sim_id           = seq_along(boot$draws),
        non_scaled_value = boot$draws
      )
    }
  ) |>
    dplyr::left_join(county_meta, by = "fylke_name")
})

```

The table below (@tbl-rose-county-nonscaled) shows the direct area-weighted mean rowan-aspen-goat willow volume per county or sub-county for all reporting periods, with approximate 95% bootstrap confidence intervals (2.5th–97.5th percentile of 1000 draws). Use the search box and page controls to browse rows.

```{r tbl-rose-county-nonscaled}
#| tbl-cap: "Non-scaled rowan-aspen-goat willow volume (m³ ha⁻¹) by county or sub-county for all reporting periods. Point estimate is the direct area-weighted mean from section 9.2; 95% CI is derived from 1000 bootstrap draws."
#| echo: false

county_draws_long |>
  dplyr::group_by(period, fylke_name, county_part, region_code, total_area) |>
  dplyr::summarise(
    ci_low  = round(quantile(non_scaled_value, 0.025), 2),
    ci_high = round(quantile(non_scaled_value, 0.975), 2),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    rose_results_county_all_periods |>
      dplyr::select(period, fylke_name, n_plots, indicator_value),
    by = c("period", "fylke_name")
  ) |>
  dplyr::mutate(
    region_name = assign_region_name(region_code),
    mean_m3_ha = round(indicator_value, 2),
    `95% CI` = paste0(ci_low, " – ", ci_high)
  ) |>
  dplyr::arrange(period, region_code, fylke_name) |>
  dplyr::select(
    Period = period,
    County = fylke_name,
    `Sub-county` = county_part,
    Region = region_name,
    `Mean (m³ ha⁻¹)` = mean_m3_ha,
    `95% CI`,
    `Total area (ha)` = total_area,
    `Number of plots` = n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Mean (m³ ha⁻¹)`,
    decimals = 2
  ) |>
  gt::fmt_number(
    columns = c(`Total area (ha)`, `Number of plots`),
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 9.4 Scale indicator values and aggregate to regional and national levels {#sec-rose-reference-values}

County or sub-county reference levels are listed in @tbl-rose-reference-values.
_X~0~_ is 0 m³ ha⁻¹ for all units; _X~100~_ is the rowan-aspen-goat willow volume at best condition for each county or sub-county accounting unit.

```{r rose-reference-values}
#| echo: false
#| code-summary: "Reference values by county or sub-county"

rose_reference_values <- dplyr::tibble(
  county_part = c(
    "Øs", "OA", "HeS", "HeN", "OpS", "OpN", "BuS", "BuV", "Ve",
    "TeS", "TeN", "AA", "VA", "Ro", "Ho", "SF", "MR", "ST", "NT",
    "NoS", "NoN", "Tr", "Fi"
  ),
  x100_m3_ha = c(
    9.0, 11.5, 7.0, 3.0, 6.0, 4.5, 7.0, 3.5, 15.0,
    8.5, 4.5, 5.5, 7.5, 7.5, 7.5, 7.0, 9.0, 4.5, 5.0,
    5.6, 4.0, 4.5, 2.0
  )
) |>
  dplyr::mutate(x100 = x100_m3_ha)

x0 <- 0
county_reference <- rose_reference_values |>
  dplyr::select(county_part, x100)

```

```{r tbl-rose-reference-values}
#| echo: false
#| tbl-cap: "County or sub-county reference values for scaling. _X~0~_ is 0 m³ ha⁻¹ for all units."

rose_reference_values |>
  dplyr::select(`Sub-county` = county_part, x100_m3_ha) |>
  gt::gt() |>
  gt::cols_label(
    x100_m3_ha = gt::html("<i>X</i><sub>100</sub> (m³ ha⁻¹)")
  ) |>
  gt::cols_align(align = "left", columns = `Sub-county`) |>
  gt::cols_align(align = "right", columns = x100_m3_ha) |>
  gt::fmt_number(
    columns = x100_m3_ha,
    decimals = 1
  ) |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::tab_options(
    table.align = "left",
    table.width = gt::px(260)
  )

```

Each county's 1000 non-scaled draws from section 9.3 are scaled with `ecTools::ec_normalise()` (linear, X0 = 0, county-specific X100).
@fig-rose-norm shows the mapping for each sub-county.

```{r county-draws-scaled}
#| code-summary: "Scale county draws; derive county uncertainty and feed ec_upscale"

county_draws_scaled <- county_draws_long |>
  dplyr::inner_join(county_reference, by = "county_part") |>
  dplyr::mutate(
    scaled_value = ec_normalise(
      non_scaled_value,
      x0 = 0,
      x100 = x100,
      fun = "linear"
    ),
    nation = "National"
  )

county_scaled_with_uncertainty <- county_draws_scaled |>
  dplyr::group_by(period, fylke_name, county_part, total_area) |>
  dplyr::summarise(
    median = median(scaled_value),
    p_025  = as.numeric(quantile(scaled_value, 0.025)),
    p_25   = as.numeric(quantile(scaled_value, 0.25)),
    p_75   = as.numeric(quantile(scaled_value, 0.75)),
    p_975  = as.numeric(quantile(scaled_value, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    rose_results_county_all_periods |> dplyr::select(period, fylke_name, n_plots),
    by = c("period", "fylke_name")
  )

```

```{r fig-rose-norm}
#| fig-cap: "Linear normalisation of rowan-aspen-goat willow volume (m³ ha⁻¹) to indicator values (0–1). Each panel shows one sub-county accounting unit, with its own X100 reference value (upper right corner of the scaling line). The x-axis is free-scaled per panel to reflect the range of observed values in that sub-county. Points are a random sample of 100 bootstrap draws (out of 1000) pooled across all reporting periods, illustrating the spread of resampled area-weighted means around the direct area-weighted mean from section 9.2. The diagonal line shows the linear scaling function between X0 = 0 m³ ha⁻¹ (lower reference, worst condition) and the sub-county-specific X100 (upper reference, reference condition); values above X100 are truncated to 1."
#| code-summary: "Normalisation diagnostic figure"
#| cache: false
#| fig-height: 8
#| fig-width: 8

norm_plot_data <- county_draws_scaled |>
  dplyr::group_by(county_part) |>
  dplyr::slice_sample(n = 100) |>
  dplyr::ungroup()

rose_norm_plot <- ec_norm_plot(norm_plot_data$non_scaled_value, norm_plot_data$scaled_value) +
  ggplot2::facet_wrap(~ norm_plot_data$county_part, scales = "free_x") +
  ggplot2::labs(
    x = "Rowan-aspen-goat willow volume (m³ ha⁻¹)",
    y = "Scaled indicator value"
  ) +
  ggplot2::theme(axis.text = ggplot2::element_text(size = 9))
rose_norm_plot$layers[[1]]$aes_params$size <- 3

# save plot for later, and supress message
suppressMessages(
  ggsave(
    filename = here::here("img/rose_norm_plot.png"),
    plot = rose_norm_plot,
    scale = 1.5
  )
)

rose_norm_plot
```

The scaled draws are then passed to `ecTools::ec_upscale()`, which propagates uncertainty through the spatial hierarchy.
For each of 1000 Monte Carlo iterations, one draw is sampled per county and the area-weighted mean gives one regional value; a second `ec_upscale()` call repeats this from regions to national level.

```{r rose-ec-upscale-regional}
#| code-summary: "ec_upscale: county to region"
#| eval: false

rose_mc_regional <- ecTools::ec_upscale(
  data        = county_draws_scaled,
  variable    = scaled_value,
  weight      = total_area,
  start_units = fylke_name,
  end_units   = region_code,
  year        = period,
  end_units_name = "region_code",
  n = 1000
)
saveRDS(rose_mc_regional, here::here("data/rose_mc_regional.rds"))
```

```{r rose-load-mc-regional}
#| include: false
rose_mc_regional <- readRDS(here::here("data/rose_mc_regional.rds"))
```

```{r rose-ec-upscale-national}
#| code-summary: "ec_upscale: region to national"
#| eval: false

region_area_weights <- county_draws_scaled |>
  dplyr::distinct(period, fylke_name, total_area, region_code) |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(region_area = sum(total_area), .groups = "drop")

rose_mc_regional_for_national <- rose_mc_regional |>
  dplyr::left_join(region_area_weights, by = c("period", "region_code")) |>
  dplyr::mutate(nation = "National")

rose_mc_national <- ecTools::ec_upscale(
  data        = rose_mc_regional_for_national,
  variable    = sampled_mean,
  weight      = region_area,
  start_units = region_code,
  end_units   = nation,
  year        = period,
  end_units_name = "nation",
  n = 1000
)
saveRDS(rose_mc_national, here::here("data/rose_mc_national.rds"))
```

```{r rose-load-mc-national}
#| include: false
rose_mc_national <- readRDS(here::here("data/rose_mc_national.rds"))
```

```{r rose-mc-combined}
#| code-summary: "Combine regional and national MC draws; build summary tables"

rose_mc_combined <- dplyr::bind_rows(
  rose_mc_regional |> dplyr::select(period, region = region_code, sampled_mean),
  rose_mc_national |> dplyr::select(period, region = nation,      sampled_mean)
)

county_scaled_all_periods <- rose_results_county_all_periods |>
  dplyr::left_join(county_reference, by = "county_part") |>
  dplyr::filter(!is.na(x100)) |>
  dplyr::mutate(
    scaled_value = ec_normalise(indicator_value, x0 = x0, x100 = x100, fun = "linear")
  )

region_scaled_agg_all_periods <- rose_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region_name = assign_region_name(region_code))

national_scaled_agg_all_periods <- rose_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(scaled_value = median(sampled_mean), .groups = "drop") |>
  dplyr::mutate(region = "National")

region_scaled_agg_with_uncertainty <- rose_mc_regional |>
  dplyr::group_by(period, region_code) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = assign_region_name(region_code)) |>
  dplyr::left_join(
    rose_results_region_all_periods |> dplyr::select(period, region_code, total_area, n_plots),
    by = c("period", "region_code")
  )

national_scaled_agg_with_uncertainty <- rose_mc_national |>
  dplyr::group_by(period) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_25   = as.numeric(quantile(sampled_mean, 0.25)),
    p_75   = as.numeric(quantile(sampled_mean, 0.75)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region = "National") |>
  dplyr::left_join(
    rose_results_national_all_periods |> dplyr::select(period, total_area, n_plots),
    by = "period"
  )

truncation_summary <- dplyr::bind_rows(
  county_scaled_all_periods |>
    dplyr::left_join(
      rose_reference_values |> dplyr::select(county_part, reference_value_m3_ha = x100_m3_ha),
      by = "county_part"
    ) |>
    dplyr::mutate(
      level = "County",
      exceeds_threshold = indicator_value > x100,
      original_value_m3_ha = round(indicator_value, 2),
      region_code = NA_character_,
      region_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    ),
  region_scaled_agg_all_periods |>
    dplyr::left_join(
      rose_results_region_all_periods |> dplyr::select(period, region_code, indicator_value),
      by = c("period", "region_code")
    ) |>
    dplyr::mutate(
      level = "Region",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    ),
  national_scaled_agg_all_periods |>
    dplyr::left_join(
      rose_results_national_all_periods |> dplyr::select(period, indicator_value),
      by = "period"
    ) |>
    dplyr::mutate(
      level = "National",
      exceeds_threshold = scaled_value >= 1,
      original_value_m3_ha = round(indicator_value, 2),
      reference_value_m3_ha = NA_real_,
      region_code = NA_character_,
      region_name = "National",
      fylke_name = NA_character_
    ) |>
    dplyr::select(
      period, level, region_code, region_name, fylke_name,
      reference_value_m3_ha, exceeds_threshold, original_value_m3_ha
    )
)

truncation_counts <- truncation_summary |>
  dplyr::group_by(period, level) |>
  dplyr::summarise(
    n_total = dplyr::n(),
    n_exceeding_reference = sum(exceeds_threshold, na.rm = TRUE),
    pct_exceeding_reference = round(100 * n_exceeding_reference / n_total, 1),
    .groups = "drop"
  )

truncated_values <- truncation_summary |>
  dplyr::filter(exceeds_threshold) |>
  dplyr::arrange(period, level, dplyr::desc(original_value_m3_ha))

```

At county or sub-county level, truncation is evaluated by comparing the non-scaled point estimate to the matching X100.
At regional and national level, the tables report cases where the `ec_upscale()` median equals 1.
The following tables summarise how often truncation occurred in each period and list the affected units.

```{r report-truncation}
#| code-summary: "Report truncation"

truncation_counts |>
  dplyr::rename(
    Period = period,
    Level = level,
    `Total` = n_total,
    `Exceeding reference` = n_exceeding_reference,
    `% Exceeding reference` = pct_exceeding_reference
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `% Exceeding reference`,
    decimals = 1
  )

if (nrow(truncated_values) > 0) {
  truncated_values |>
    dplyr::mutate(
      location = dplyr::case_when(
        level == "County" ~ fylke_name,
        level == "Region" ~ region_name,
        level == "National" ~ "National",
        TRUE ~ NA_character_
      )
    ) |>
    dplyr::select(
      period,
      level,
      location,
      reference_value_m3_ha,
      original_value_m3_ha
    ) |>
    dplyr::rename(
      Period = period,
      Level = level,
      Location = location,
      `Reference (m3/ha)` = reference_value_m3_ha,
      `Original Value (m3/ha)` = original_value_m3_ha
    ) |>
    gt::gt() |>
    gt::tab_style(
      style = gt::cell_text(weight = "bold"),
      locations = gt::cells_column_labels()
    ) |>
    gt::fmt_number(
      columns = c(`Reference (m3/ha)`, `Original Value (m3/ha)`),
      decimals = 1
    )
} else {
  cat("\nNo values exceeded their reference and required truncation.\n")
}

```

The time series below (@fig-rose-timeSeries) summarises regional and national values.

```{r fig-rose-timeSeries}
#| fig-cap: "Indicator values for rowan-aspen-goat willow volume by region and reporting period. Points show the MC median; error bars are 95% credible intervals from the Monte Carlo propagation. National values use filled circles and a thicker line."
#| code-summary: "Time series of indicator values by region and period"
#| cache: false
#| warning: false
#| message: false

rose_region_labels <- c(
  "1" = "Øst", "2" = "Sør", "3" = "Vest",
  "4" = "Midt", "5" = "Nord", "National" = "National"
)
rose_region_levels <- c("National", "Midt", "Nord", "Øst", "Sør", "Vest")
rose_region_cols <- setNames(
  MetBrewer::met.brewer("Archambault", n = length(rose_region_levels)),
  rose_region_levels
)

rose_plot_data <- rose_mc_combined |>
  dplyr::mutate(
    region_name = dplyr::recode(region, !!!rose_region_labels),
    series_type = dplyr::if_else(region == "National", "National", "Region")
  ) |>
  dplyr::group_by(period, region_name, series_type) |>
  dplyr::summarise(
    median = median(sampled_mean),
    p_025  = as.numeric(quantile(sampled_mean, 0.025)),
    p_975  = as.numeric(quantile(sampled_mean, 0.975)),
    .groups = "drop"
  ) |>
  dplyr::mutate(region_name = factor(region_name, levels = rose_region_levels))

ggplot2::ggplot(rose_plot_data,
                ggplot2::aes(x = period, y = median, colour = region_name, group = region_name)) +
  ggplot2::geom_line(
    ggplot2::aes(linewidth = series_type),
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::geom_pointrange(
    ggplot2::aes(ymin = p_025, ymax = p_975, shape = series_type),
    size = 0.8, linewidth = 0.45,
    position = ggplot2::position_dodge(width = 0.18)
  ) +
  ggplot2::scale_colour_manual(values = rose_region_cols, name = NULL) +
  ggplot2::scale_shape_manual(values = c(National = 19, Region = 17), name = NULL) +
  ggplot2::scale_linewidth_manual(values = c(National = 1, Region = 0.55), guide = "none") +
  ggplot2::scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position    = "bottom",
    legend.box         = "horizontal",
    panel.grid.minor   = ggplot2::element_blank(),
    panel.grid.major.x = ggplot2::element_blank()
  ) +
  ggplot2::guides(
    colour = ggplot2::guide_legend(nrow = 2,
                                   override.aes = list(size = 0.7, linewidth = 0.45)),
    shape  = ggplot2::guide_legend(override.aes = list(size = 0.8, linewidth = 0.45))
  ) +
  ggplot2::labs(y = "Indicator value", x = "")
```

## 10. Results {#sec-results}

The following tables present scaled indicator results at county or sub-county, regional, and national scales for each reporting period.
Tables report the median as the indicator value, with an approximate 95% interval (2.5% and 97.5% quantiles) and an approximate interquartile range (25% and 75% quantiles).

### 10.1 County-level results (scaled with uncertainty)

```{r tbl-county-results-table}
#| tbl-cap: 'County or sub-county indicator values for rowan-aspen-goat willow volume by reporting period. Values are scaled to 0–1 using county-specific X0 and X100 via ec_normalise(). Uncertainty is derived from 1000 bootstrap draws scaled with the same function (section 9.3–9.4). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the scaled draw distribution.'
#| echo: false

county_scaled_with_uncertainty |>
  dplyr::mutate(
    fylke_name_clean = stringi::stri_trans_general(
      stringr::str_squish(
        tolower(stringr::str_replace_all(.data$fylke_name, "[0-9.-]", " "))
      ),
      "Latin-ASCII"
    ),
    region_code = assign_region_code(.data$fylke_name_clean),
    region_name = assign_region_name(.data$region_code),
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code, .data$fylke_name) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Rowan-aspen-goat willow volume",
    ID = "NO_ROSE_001",
    `Accounting area` = .data$fylke_name,
    Region = .data$region_name,
    `Sub-county` = .data$county_part,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  gt::gt() |>
  gt::tab_style(
    style = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  ) |>
  gt::fmt_number(
    columns = `Number of plots`,
    decimals = 0,
    use_seps = TRUE
  ) |>
  gt::opt_interactive(
    use_search = TRUE,
    use_pagination = TRUE,
    use_page_size_select = TRUE,
    page_size_default = 12,
    page_size_values = c(12, 24, 66)
  )
```

### 10.2 Regional results (scaled with uncertainty)

```{r tbl-region-results-table}
#| tbl-cap: 'Regional indicator values for rowan-aspen-goat willow volume by reporting period. Scaled values and uncertainty are derived from the ecTools::ec_upscale() Monte Carlo propagation (1000 draws per region per period). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

region_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period, .data$region_code) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Rowan-aspen-goat willow volume",
    ID = "NO_ROSE_001",
    `Accounting area` = .data$region_name,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

### 10.3 National results (scaled with uncertainty)

```{r tbl-national-results-table}
#| tbl-cap: 'National indicator values for rowan-aspen-goat willow volume by reporting period. Scaled values and uncertainty are derived from the ecTools::ec_upscale() Monte Carlo propagation (1000 draws). The median is reported as the indicator value, with approximate 95% and interquartile intervals from the MC distribution.'
#| echo: false

national_scaled_agg_with_uncertainty |>
  dplyr::mutate(
    median = round(.data$median, 2),
    p_025 = round(.data$p_025, 2),
    p_25 = round(.data$p_25, 2),
    p_75 = round(.data$p_75, 2),
    p_975 = round(.data$p_975, 2)
  ) |>
  dplyr::arrange(.data$period) |>
  tidyr::unite("CI_95", c(p_025, p_975), sep = " - ") |>
  tidyr::unite("IQR", c(p_25, p_75), sep = " - ") |>
  dplyr::transmute(
    ECT = "B2",
    Indicator = "Rowan-aspen-goat willow volume",
    ID = "NO_ROSE_001",
    `Accounting area` = .data$region,
    Ecosystem = "Boreal/montane forests",
    `Reporting period` = .data$period,
    `Indicator value` = .data$median,
    `95% CI` = .data$CI_95,
    `50% CI` = .data$IQR,
    `Number of plots` = .data$n_plots
  ) |>
  knitr::kable() |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## 11. Export file

Regional and national Monte Carlo draws are exported in one table (`rose_mc_combined`: `year`, `region`, `sampled_mean`; `region` contains the five NI region codes and `"National"`).

```{r export}
#| eval: false
#| code-summary: "Export regional and national MC-propagated indicator draws"

write.csv(
  rose_mc_combined,
  here::here("data/results_NO_ROSE_001.csv"),
  row.names = FALSE
)
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [EF, KOS, UJA]
* Methodology: [EF, KOS, UJA]
* Data curation: [OS]
* Formal analysis: [AHH]
* Validation: [AHH, KOS]
* Visualization: [AHH]
* Writing – original draft: [All]
* Writing – review & editing: [All]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
