---
title: "Wolf abundance"
subtitle: |
  [NO_WOAB_001]
indicatorID: 
  - NO_WOAB_001
indicatorName: 
  - Wolf abundance [Ulvebestand]
verbatimName: 
  - Ulvebestand
folderName: NO_WOAB_001
country: Norway
continent: Europe
ECT: B3 Functional state characteristics
Realm: 
  - Terrestrial (T) # e.g. Freshwater (F)
Biome: 
  - T2 Temperate-boreal forests and woodlands biome
Ecosystem: 
  - T2.1 Boreal/montane forests
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First version on ecRxiv; indicator described in the Norwegian Nature Index
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_WOAB_001
Normalised: Yes
SpatialAggregationPathway: Aggregate (population model, large-carnivore management region) - [Scale - Truncate] - Aggregate (national level)
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: Aud H. Halbritter
    email: aud.halbritter@uib.no
    affiliations:
      - id: myID
        name: University of Bergen
AuthorList: Halbritter, A. H.
date: May 15, 2026
data_availability: gold
code_reproducibility: gold
open_science_badge: gold
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE # Change to FALSE
bibliography: references.bib
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull
library(NIcalc)

# This workflow uses `renv`. Please run `renv::restore()` in order to reproduce the working environment and run the analysis.
# renv::restore()

knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
if (file.exists(here::here("R/spatial_regions.R"))) {
  source(here::here("R/spatial_regions.R"))
}
```



<!--# The following parts are autogenerated. Do not edit. -->

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- May 15, 2026 - Initial draft for ecRxiv
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--# 
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow. 
 -->

1. **Background and rationale**:
The grey wolf (_Canis lupus_) is the largest member of the dog family and a social species; most wolves live in pairs or packs that defend territories.
In Norway the species is found mainly in southeastern parts of the country along the border with Sweden, although individual wolves can roam over long distances and may in principle occur anywhere on the mainland.
The species also occurs in large parts of Europe, Asia, and North America.
It is listed as critically endangered (kritisk truet) on the Norwegian Red List (2021).
Rovdata is responsible for monitoring wolves in Norway.
The indicator is part of the Norwegian Nature Index and is described on [Naturindeksen](https://naturindeks.no/Indicators/ulv).

2. **Methods**:
The wolf population size is estimated from ....
This value is then normalised based on a reference level, which is an expert-assessed population estimate based on known breeding density and the amount of suitable habitat.
The indicator is developed in the work with the Norwegian Nature Index [@nater_naturindeks_2025]. The methods are are described in full on the [Nature index web site](https://www.naturindeks.no/Indicators/ulv). 


3. **Interpretation**:
An indicator value of 1 means that the wolf population in the reporting unit is at least as large as expected under the reference condition used in Naturindeksen.
Values decreasing toward zero indicate that the population index is further below that reference.

4. **Results and development needs**:
The indicator is used in Naturindeksen.


## 1.2 Native language summary - Norwegian

1. **Bakgrunn og begrunnelse**: Ulv (_Canis lupus_) er det største medlemmet av hundefamilien og en sosial art; de fleste lever i par eller flokker som hevder territorier.
I Norge finnes arten hovedsakelig i sørøstlige deler av landet ved grensen mot Sverige, men enkeltdyr kan streife langt og kan i prinsippet dukke opp hvor som helst på fastlandet.
Arten finnes også i store deler av Europa, Asia og Nord-Amerika, og er vurdert som kritisk truet på Norsk rødliste (2021).
Rovdata har ansvar for overvåking av ulv i Norge.
Indikatoren inngår i Naturindeksen og er beskrevet på [Naturindeksen](https://naturindeks.no/Indicators/ulv).

2. **Metoder**: ...
Denne verdien normaliseres deretter basert på en referanseverdi, som er et ekspertvurdert bestandsestimat basert på kjent hekketetthet og mengden egnet habitat. Indikatoren er utviklet i arbeidet med Naturindeksen [@nater_naturindeks_2025]. Metodene er beskrevet i detalj på [Naturindeksens nettsider](https://www.naturindeks.no/Indicators/ulv).

3. **Tolkning**: En indikatorverdi på 1 betyr at ulvebestanden i rapporteringsenheten er minst like stor som forventet under referansetilstanden i Naturindeksen.
Verdier som nærmer seg null, viser at bestandsindekset ligger lenger under referansen.

4. **Resultater og utviklingsbehov**: Indikatoren brukes i Naturindeksen.


## 2. About the underlying data

The data comes from ...
The procedure to download the data is described [here](https://naturindeks.nina.no/docs/24dda0ba-5855-4131-9168-d874980ec6aa.pdf). 

### 2.1 Spatial and temporal resolution and extent

Indicator values are available for the assessment years in Naturindeksen: 1990, 2000, 2010, 2014, 2019, and 2024.

To produce indicator estimates for these years, input is combined over a short window (typically two to three years), as specified in the Naturindeks indicator protocol.

Spatially, data are organised by the eight large-carnivore management regions (_forvaltningsregioner for rovvilt_, regions 1–8).
In Naturindeksen, indicator and reference values are stored at management-region scale (IC units), then aggregated to five geographic regions and national level for reporting, using the same spatial framework as for brown bear and lynx in this repository.

### 2.2 Sampling design

[Rovdata](https://rovdata.no/Ulv.aspx) is responsible for monitoring wolves in Norway and for maintaining an overview of how many wolves there are in the country.
Populations are estimated from annual winter monitoring, including counts of wolves with wholly Norwegian residency and wolves with residency on both sides of the border with Sweden, and reporting of wolf litters in Norwegian and border territories relative to national targets [@rovdata_ulv_status_2025].
Further detail on field methods (for example registration of tracks and signs) is linked from [Rovdata](https://rovdata.no/Ulv.aspx).

### 2.3 Original units

At the monitoring stage, Rovdata reports population status using numbers of wolves (including ranges for a given winter) and documentation of litters in territories [@rovdata_ulv_status_2025].

### 2.4 Instructions for citing, using and accessing data

Population monitoring and status information: cite [Rovdata](https://rovdata.no/Ulv.aspx) and the relevant status pages [@rovdata_ulv_status_2025].
Methodological documentation for the indicator: [Naturindeksen – ulv](https://naturindeks.no/Indicators/ulv).

<!-- ### 2.5 Additional comments about the dataset -->

<!--# Optional. Text here -->

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

B3 – Functional state characteristics.

<!-- 
### 3.1.1 Other condition typologies -->

<!--# Optional: Add text about other standards or typologies, such as alternative national condition typologies, and how the indicator relates to these -->


### 3.2 Ecosystem condition characteristic

<!--# 

Describe the ecosystem condition characteristic represented in the indicator. See 10.3897/oneeco.6.e58218 for information on what these characteristics might be.
For example, and indicator called 'Trenching in mires' could be made to represent an ecosystem characteristic 'Intact hydrology'. The term 'characteristic' is used similar to the term 'criteria' in Multiple Criteria Decision Making.  

-->


### 3.3 Collinearities with other indicators

<!--# Describe known collinearities with other metrics (indicators or variables) that could become problematic if they were also included in the same Ecosystem Condition Assessment as the indicator described here. -->

The indicator may covary with other large-carnivore indicators that use the same regional structure (brown bear, lynx, wolverine) where distributions overlap.

### 3.4 Impact factors

<!--# Describe the main natural and anthropogenic factors that affecst the metric -->


### 3.5 Spatial aggregation

<!-- Describe in word the same aggregation pathway presented in the top yaml (the metadata). If you have scaled against the reference value, it is especially important to know if you did any avergaing or summations etc. beforehand. An example text for a rather simple case could be: "The variable is scaled at the initial polygon level and the point estimate is a area-weighted arithmetic mean of all polygins. There is no transformation, and truncation is not needed as the indicator becomes naturally bound between 0 and 1".  -->

Following the metadata pathway Aggregate (population model, large-carnivore management region) – [Scale – Truncate] – Aggregate (national level): values are handled per forvaltningsregion for rovvilt in Naturindeksen, then aggregated to five geographic regions and national level with area weights (see Section 8).
Each unit is normalised against _X~0~_ and _X~100~_ for that region (linear scaling with truncation to [0, 1]).

## 4. Reference condition and levels

### 4.1 Reference condition

The reference condition for wolf in Naturindeksen is a boreal/montane forest ecosystem with minimal human impact and a climate similar to the reference period used in the Norwegian Nature Index framework [@nybo_fagsystem_2017].

### 4.2 Reference levels

_X~100~_ is set per large-carnivore management region (IC unit in Naturindeksen), using suitable area and density within each _rovviltregion_ and expert assessment: it represents the expected wolf population (or equivalent index) under reference conditions in that unit.

_X~0~_ is defined as no wolves in the unit (functional absence).


#### 4.2.1 Spatial resolution and validity

<!--# 

Describe the spatial resolution of the reference levels. E.g. is it defined as a fixed value for all areas, or does it vary. Also, at what spatial scale are the reference levels valid? For example, if the reference levels have a regional resolution (varies between regions), it might mean that it is only valid and correct to use for normalising local variable values that are first aggregated to regional scale. However, sometimes the reference levels are insensitive to this and can be used to scale variables at the local (e.g. plot) scale. 

 -->

Reference levels and indicator values apply at the scale of forvaltningsregioner for rovvilt (IC units).
Aggregated results for the five Naturindeks geographic regions and for Norway are summaries of those units with area weights.

## 5. Uncertainties
### 5.1 General uncertainties
<!--# Describe the main reasons why you might not trust 100% the final value to represent the given ecosystem characteristic for the give ecosystem type. What aspects of this condition metric are most sensitive to subjective decitions? Is the spatial sampling sufficient? Does the choice of modeling framework affect the values? Is the metric relevant do describe ecosystem condition for the give ecosystem type?  -->


### 5.2 Reported uncertainties
<!-- Describe the nature of the errors you (hopefully) provide alongside the point estimate for the condition metric. It is especially important to clarify if the errors are inferential (referring to the uncertainty of the point estimate) or descriptive (referreing to the data itself). Here' an example text: "What we report as uncertainties in @tbl-finalTable are inferential and refer to the uncertainty around the point estimates. It does not directly describe variation in the data itself. 
However, the variation in the mean is defined by the spatial variation only, and quartiles are obtained by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples".   -->



## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

Run the download chunk once when you have API credentials. Credentials are read from `R/_pw.R` (not committed to git). The export is saved under `data/` (also gitignored).

```{r woab-import-api}
#| eval: false
#| code-summary: "Download wolf data from Naturindeks API"
source(here::here("R", "_pw.R"))

woabImport <- NIcalc::importDatasetApi(
  username = myUser,
  password = myPwd,
  eco = NULL,
  indic = c("Ulv"),
  year = c("1990", "2000", "2010", "2014", "2019", "2024")
)

saveRDS(woabImport, here::here("data/woabImport.rds"))
```

```{r woab-load-import}
#| eval: false
#| include: false
#| cache: false
#| code-summary: "Load saved Naturindeks import (enable when data/woabImport.rds exists)"

woabImport <- readRDS(here::here("data/woabImport.rds"))
```

## 8. Spatial units

<!--# Describe the spatial units that you rely on in your analyses. Highlight the spatial units (the resolution) that the indicator values should be interpreted at. Potential spatial delineation data (maksing layers etc.) should be introduced under chapert 7. We recommend using the SEEA EA terminology of Basic Spatial Units (BSU), Ecosystem Asses (EA) and Ecosystem Accounting Area (EAA). -->

Primary reporting units are forvaltningsregioner for rovvilt (regions 1–8).
Aggregated reporting uses the five Naturindeks geographic regions (east, south, west, central, north) and national level.

```{r woab-rovvilt-ni-mapping}
#| eval: false
#| echo: true
#| cache: false
#| code-summary: "Rovvilt region -> Naturindeks region lookup (enable when data are imported)"

rovvilt_ni_mapping()
```

## 9. Analyses

<!--# 

Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->

## 10. Results

<!--# 

Repeat the final results here. Typically this is a map or table of indicator values.

This is typically where people will harvest data from, so make sure to include all relevant output here, but don't clutter this section with too much output either.

-->

## 11. Export file

<!--# 

Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
```


## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structur below, and just delete the rows you dont use.  -->

Not relevant, as this is the work of a single author. I would, however, like to acknowledge the work of those that created the indicator on _Naturindeks_ [@nater_naturindeks_2025].

::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
