---
  title: "Vegetation indicator for pH"
subtitle: |
  [NO_FUPH_001] (open lowland) 
indicatorID: 
  - [NO_FUPH_005]
indicatorName: 
  - Vegetation indicator for pH
verbatimName: 
  - Vegetasjonsindikator for pH # the indicator name that us used in the native language
folderName: NO_FUPH_005 # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
ECT: A1 - Physical State characteristics
Realm: # Please copy and paste Realms, Biomes and Ecostsem names from here: https://github.com/NINAnor/ecRxiv/wiki#ecosystem-typology 
  - Terrestrial (T)
Biome: 
  - T3 Shrublands and shrubby woodlands biome
  - T4 Savannas and grasslands biome
Ecosystem: 
  - T4.5 Temperate grasslands
  - T3.4 Rocky pavements
  - T3.2 Seas dry temp shrublands
verbatimEcosystem:
  - åpen lavland
yearAdded: 2026
yearLastUpdate: 2026
status: complete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: This indicator has been calculated before under different names and codes. Most rencently NO_FUNC_004 version 000.001 on ecRxiv 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/NO_FUMO_004
Normalised: Yes # Yes or no depending on whether the metric has been normliased or not.
SpatialAggregationPathway: Scale/Truncate - Aggregate (zero-one inflated beta-regression) 
Precision: 95% confidence intervals for the true value of the spatially aggregated indicator values derived from the beta-regression standard errors. # What kind of quantitative unceratinties are reported alongside the point estimates. Can be either inferrential (i.e. bootstrapped confidence intervals or standard errors) or descriptive (variation, standard deviation or percentiles from the distribution of the raw indicator values).
format: 
  html:
  embed-resources: true
code-fold: true
toc: true
toc-title: Contents
toc-depth: 3
smooth-scroll: true
execute: 
  cache: false
author:
  - name: Francesca Jaroszynska              # Enter name
email: francesca.jaroszynska@nina.no  # Enter email
affiliations:
  - id: myID
name: Norwegian institute for nature research # Enter affiliations
AuthorList: Jaroszynska, F.  
date: September 14, 2026
data_availability: silver
code_reproducibility: silver
open_science_badge: silver
callout-icon: false
lightbox: true
code-links:
  - text: Add a review
icon: github
href: https://github.com/NINAnor/ecRxiv
hide: FALSE 
bibliography: references.bib
codeReviewers:
  - name: Aud Halbritter
email: aud.halbritter@uib.no
affiliations:
  - id: reviewerInst
name: University of Bergen
reviewType: code review
github: 
---
  
  <!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->
  
  <!--# Load all you dependencies here -->
  
  ```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(sf)
library(ecTools)
library(glmmTMB)
library(MASS)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull
library(arrow)
library(gt)
knitr::opts_chunk$set(echo = TRUE)
```

```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```



<!--# The following parts are autogenerated. Do not edit. -->
  
  ```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
  
  
  > **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`


::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class", 
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>   
  kable()

```

:::
  
  ::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
Presentation of moisture indicator for ANO, GRUK and ASO datasets

:::
  
  <hr />
  
  <!--# Document you work below.  -->
  
  ## 1. Summary
  
  1. **Background and rationale**: This indicator represents plant community responses to Moisture. It represents the lowest score of deviations from an optimal state either to higher or to lower Moisture levels. 

2. **Methods**: The data originate from field sampled plant community data are used from three monitoring programmes: 1. the Norwegian nature monitoring program ANO (Tingstad et al. 2019), 2. GRUK, an open calcareous grasslands monitoring programme (Evju & Vang 2026), and finally 3. the ASO monitoring programme (Bär et al. 2025). We combine these with plant functional trait values from Tyler et al. (2021) and plant strategies towards stressors from Grime (1974). Indicator values are based on calculating community-weighted means (Diekmann 2003, Töpper et al. 2018) or cumulative abundance (see Framstad et al. 2022) of functional trait values for plant community data. Deviations in such an indicator beyond a certain range of indicator values indicate a directional shift in the plant community. As reference communities we use ‘generalized species lists’ underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020, see Evju et al. 2023 for naturally open lowland ecosystems). For each major ecosystem type a separate set of indicators, related to the main drivers of change today, is computed. The lowest scoring functional plant community indicator is then reported as the ‘Functional plant community index’.

3. **Interpretation**:
  An indicator value of 1 indicates no discernible difference in the functional plant composition due to Moisture impacts. 
Values decreasing toward zero reflect a shift in species composition and community trait values away from the reference state, either as a response to increased or decreased Moisture.

4. **Results and development needs**:
  
  The functional plant community indicator for moisture and its 95% CI is slightly below 0.6 indicating that the functional plant community composition deviates more than accepted under good ecological condition. The result is mainly driven by deviations towards high moisture values, but deviations towards low moisture values also occur (ca. 3/4 upper deviations and 1/4 lower deviations).

To improve this indicator in the future, the perhaps most pressing issue it to regionalise the generalised species list, as currently they have a bias towards southern and eastern Norway. The indicator values also need validation against independent datasets.

Access to further years of data for the ASO monitoring scheme would strengthen its temporal and spatial representation, and thus the indicator's robustness.




## 2. About the underlying data {#sec-about}
We use three types of data for building functional plant community indicators for ecological condition:
  
  - as observation data  we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019), the ASO monitoring scheme (Bär et al. 2025), and the GRUK monitoring scheme (Evju and Vang 2026).
  - as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (Halvorsen et al. 2020)
- Swedish functional plant indicator data from Tyler et al. (2021) and indicator values for plant strategies from Grime (1974)

For details see point 7 below.

### 2.1 Spatial and temporal resolution and extent
The ANO monitoring data available now represent one cycle of sampling from 2019-2024 comprising ca. 1000 sites across Norway.
ANO has no defined end date. 2025 is the first year of repeated sampling for about 20% of the approximately 1000 ANO sites for all of mainland Norway (i.e. the Ecosystem Accounting Area). Annual sampling is spatially representative, in principle allowing to compute area-representative estimates for single years but also multiple-yr-reporting periods. The indicator at hand is calculated based on the ANO data from 2020-2024.

The GRUK monitoring data follows a similar five-year sampling cycle from 2020-2024. Monitoring sites were randomly selected across the Oslo fjord basin, resulting in ~80 sites surveyed annually. The sites comprise open, shallow-soil calcareous grassland in three sub-areas - the inner, middle and outer Oslo fjord region.

The ASO monitoring dataset is spatially representative of Norway, with each year's sampling pool representative at a national and regional scale. ASO follows a 5-year monitoring cycle, though only data from 2022 are represented here.

### 2.2 Sampling design
At every ANO site, the plant community is recorded (presence and percent cover of every vascular plant species) in the central 1 sqm plot of each of systematically arranged 18 monitoring points.

Within each GRUK site, one or several polygons matching the NiN mapping units T2-C-7 and T2-C-8 were delineated, and 5, 10, or 15 0.5 x 0.5 m permanent plots established, randomly selected from a 10 x 10 m grid within the polygons. All vascular plant species were recorded, including their percentage cover (visual estimate). 

ASO is structured using a nested design consisting of ASO areas (10x10 km), ASO plots (500x500 m), and semi-natural grasslands. Within each semi-natural grassland up to three transects were established, the length of each chosen to reflect the size of the grassland. 

### 2.3 Original units
Abundance data for vascular plants in ANO is given as % cover within a 1 m2 sample quadrat, and within a 0.5 x 0.5 m sample quadrat for GRUK. This is transformed to relative abundance data per quadrat. Each ANO and GRUK quadrat is classified to its respective ecosystem type according to ‘Nature in Norway’ (NiN; <https://artsdatabanken.no/NiN>). Species diversity in ASO was recorded using cover and frequency classes (cf. modified measurement scale A7 for NiN2). Environmental indicator values from Tyler et al. (2021) are given on an ordinal scale from 1 (lowest affinity to the respective variable) to n (highest affinity to the respective variable), the number of levels (n) varying between environmental indicators.




### 2.4 Instructions for citing, using and accessing data
The ANO data are openly available at https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054. See @sec-about for detail on the other datasets.  

### 2.5 Additional comments about the dataset
The ANO program is run by the Norwegian Environment Agency. On their web sites you may also find some additional summary data. https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/
  
## 3. Indicator properties
  
### 3.1 Ecosystem Condition Typology Class (ECT)
The functional plant community indicator for Moisture reflects the availability of nutrients and soil chemical properties and thus represents the ECT class 'A1 Physical characteristics'.

### 3.2 Ecosystem condition characteristic
The functional plant community indicator for Moisture reflects the availability of water and thus represents the characteristic 'Abiotic factors'.

### 3.3 Collinearities with other indicators
Different functional plant community indicators for different environmental variables (light, moisture, ...) might show a certain degree of co-linearity with one another as they are derived from the very same plant community data.
Functional plant community indicators can potentially exhibit co-linearities with other vegetation-oriented indicators, such as NDVI, though this depends greatly on which functional plant community indicator ends up as the leading one (i.e. the one with the lowest scaled score).


### 3.4 Impact factors
The plant community composition in open lowland ecosystems is subject to changes with climate warming and precipitation changes as well as pollution and disturbance from human infrastructure.

### 3.5 Spatial aggregation
The variable is scaled at the initial survey-point level and the indicator estimate is an arithmetic mean of all survey-points. There is no transformation, but indicator values are truncated at 1.

## 4. Reference condition and levels

### 4.1 Reference condition
In Norway, the reference condition for natural ecosystems has been one of largely intact ecosystems without large-scale human modifications (Nybø and Evju 2017). For semi-natural ecosystems, human influence is allowed, and indeed a prerequisite, but then in the sense of appropriate, stable and traditional land-use practices. In terms of Moisture, the reference condition is defined by a reference community.

### 4.2 Reference levels
Reference and limit values (scaling values) for functional plant community indicators are derived as the 0.025, 0.5, and 0.975 quantiles (lower limit values, reference value, upper limit value, respectively) from distributions of community weighted means of the respective plant indicators originating from bootstrapped generalized species lists.
Bootstraps were set up with resampling of the original generalized species lists 1000 times. In every sample a random 1/3 of the original generalized species list was sampled (as ANO points on average contain roughly a third of the potential species richness in the generalized species lists -> local variation in species composition thus is large and needs to be reflected in the bootstraps), with the exception of dominant species (abundance value of 5 or higher in the original scale of the generalized species lists), which were retained in every sample.
All reference distributions, and thus all scaling values are unique to NiN mapping units at a scale of 1:5000.

#### 4.2.1 Spatial resolution and validity
The same reference levels apply to all types of ecosystems and all regions.

## 5. Uncertainties
### 5.1 General uncertainties
Uncertainties relate partly to sampling and observer variability (and possible observer bias) in the recording of the plant cover data. Presumably, there is also uncertainty related to the scoring of indicator values for species, and this will probably vary among species (cf. Tyler et al. 2021). These potential sources uncertainty can only be assumed and not quantified.
Spatial variability as a source of uncertainty can be quantified. For every mean or median indicator value (after scaling) for every region (or any other delimited area of interest) we can calculate its corresponding standard error or confidence interval quantiles as a measure of spatial uncertainty for a geographical area.
There is some uncertainty related to the ecosystem categorization at the field sites (see Ullerud et al. 2018 and Naas et al. 2023), but since ANO monitoring does not include spatial delineation of ecosystem polygons this problem is considered smaller than in spatial ecosystem mapping which these references mainly relate to.
Other uncertainties connected to the indicator arise from community weighted means calculated from plant communities of different species richness in the observation data; since no uncertainty measure for these community-weighted means is calculated and propagated, the actual uncertainty of functional community signals for the single community observations is unknown.

There is some uncertainty about the appropriateness of using (Halvorsen et al. 2020) as a reference dataset and a reference plant community for different nature types, and especially for different geographical and environenmetal regions of Norway. The dataset has a known bias towards south and eastern Norway, and may not reflect the natural plant communities equally well i the western and northern parts. 

### 5.2 Reported uncertainties
For this indicator, we report 95% confidence intervals as inferential uncertainty based on a statistical model (zero-one inflated beta-regression).

## 6. References
Artsdatabanken (2023, 11. august). Fremmede arter i Norge - med økologisk risiko 2023. http://www.artsdatabanken.no/lister/fremmedartslista/2023.

Bär, Johansen, Vesterbukt, Svalheim, Daugstad, Hinderaker, Thorvaldsen, Albertsen, Carlsen, Grenne, Maurset, Øien, Heimstad, Lyngstad, Hasvik, Ombler, Wehn, Moen, Blütecher, Leh, Davidsen, Paulsen, Vassvik, Sundsdal, Velle, Solstad, Bele, Solbu, Winther, Henriksen, Nesheim-Hauge, Høydal (2025). Overvåking av semi-naturlig eng (ASO). Version 1.8. Norwegian Institute of Bioeconomy Research. Sampling event dataset https://doi.org/10.15468/gq6wa5 accessed via GBIF.org on 2026-09-11.

Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

Evju, M., Olsen, S.L., Prestø, T., Vange, V., Bratli, H. & Töpper, J. 2023. Ecological condition of naturally open areas below the forest line. Background, proposed indicators, and knowledge needs. NINA Report 2341. Norwegian Institute for Nature Research.

Evju M, & Vang R 2026. Overvåking av åpen grunnlendt kalkmark i Oslofjordområdet. Version 1.20. Norwegian Institute for Nature Research. Occurrence dataset. https://ipt.nina.no/resource?r=gruk_2020&v=1.20

Framstad, E., Kolstad, A. L., Nybø, S., Töpper, J. & Vandvik, V. 2022. The condition of forest and mountain ecosystems in Norway. Assessment by the IBECA method. NINA Report 2100. Norwegian Institute for Nature Research.

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E. (2020). Towards a systematics of ecodiversity: The EcoSyst framework. Global Ecology and Biogeography, 29(11), 1887-1906. doi:10.1111/geb.13164

Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik. V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2020. Setting reference levels and limits for good ecological condition in terrestrial ecosystems. Insights from a case study based on the IBECA approach. Ecological Indicators 116: 106492.

Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2021. An index-based assessment of ecological condition and its links to international frameworks. Ecological Indicators 124: 107252.

Naas, A.E., Halvorsen, R., Horvath, P., Wollan, A.K., Bratli, H., Brynildsrud, K. et al. (2023) What explains inconsistencies in field-based ecosystem mapping? Applied Vegetation Science, 26, e12715. Available from: https://doi.org/10.1111/avsc.12715

Nybø, S. & Evju, M. (red.) 2017. Fagsystem for fastsetting av god økologisk tilstand. Forslag fra et ekspertråd, https://regjeringen.no/no/dokument/rapportar-og-planar/id438817/.

Tingstad, L., Evju, M., Sickel, H., & Töpper, J. 2019. Utvikling av nasjonal arealrepresentativ naturovervåking (ANO). Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag. NINA Rapport 1642.

Töpper, J., Velle, L.G. & Vandvik, V. 2018. Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition). NINA Report 1529b. Norwegian Institute for Nature Research.

Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923

Ullerud HA, Bryn A, Halvorsen R, Hemsing LØ. Consistency in land-cover mapping: Influence of field workers, spatial scale and classification system. Appl Veg Sci. 2018; 21: 278–288. https://doi.org/10.1111/avsc.12368


## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

  
  ### 7.1 Dataset A
  Import ANO testing data:
  First, download from  Miljødirektoratet (downloaded spring 2026).
```{r, eval = F, echo = F}
# load packages
library(tidyverse)
library(pwr)
library(lme4)
library(simr)
library(readxl)
library(formattable)
library(sf)
library(tmap)
library(WorldFlora)
library(zen4R)
library(tidylog)
library(rgbif)
library(here)
library(ecTools)
library(glmmTMB)


# download 
url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", mode="wb") 
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb.zip", exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb")
```

...then upload from local storage
The geodatabase contains several layers; 'ANO_SurveyPoint' contains site and point data including alien plant species cover.
```{r loadData}
#Sys.setlocale("LC_ALL", "no_NB.utf8") #works with æøå or use "Norwegian"
ano_species <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb", layer="ANO_Art", quiet = T)
ano_geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/ANO/naturovervaking_eksport.gdb", layer="ANO_SurveyPoint", quiet = T)
#head(ANO.sp)
#head(ANO.geo)

```

  ### 7.2 Dataset B
  WorldFlora backbone
```{r, eval = F, echo = F}
# download WFO plant list from zenodo
#download_zenodo("https://doi.org/10.5281/zenodo.20782718", files = "_DwC_backbone_R.zip", path = "/Users/francesca.jaroszynsk/OneDrive - NINA/nina_projects/ANO/lowlands/ecRxiv_veg_indicators/data")

wfo_backbone <- read_delim("C:/Users/francesca.jaroszynsk/OneDrive - NINA/nina_projects/ANO/lowlands/ecRxiv_veg_indicators/data/_DwC_backbone_R/classification.csv", delim = "\t")
```

  ### 7.3. Dataset C
Import Reference data from NiN:
```{r}
load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/reference from NiN/Eco_State.RData")
 # str(Eco_State)

# for GRUK
natopen_NiN_ref <- read_rds("C:/Users/francesca.jaroszynsk/OneDrive - NINA/nina_projects/ANO/lowlands/ecRxiv_veg_indicators/data/NiN/natopen_NiN_ref.RDS")
natopen_NiN_species <- read_rds("C:/Users/francesca.jaroszynsk/OneDrive - NINA/nina_projects/ANO/lowlands/ecRxiv_veg_indicators/data/NiN/natopen_NiN_ref_spInfo.RDS")

```

The generalized species lists underlying the ecosystem categorization in NiN (Halvorsen et al. 2020) represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the ‘expected combination of frequency and cover’ of occurrence

1 = < 1/32
2 = 1/32 - 1/8
3 = 1/8 - 3/8
4 = 3/8 - 4/5
5 = 3/8 - 4/5 + dominance
6 = > 4/5

For the purpose of this project, these steps are simplified to maximum expected combination of frequency and cover, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.



  ### 7.4. Dataset D
Import plant indicator values from Tyler et al. (2021):
```{r}
ind_tyler <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/ind.Tyler.RDS")
```

The Swedish plant indicator values dataset published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we use indicator data for moisture and Moisture as these are thought to be subject to potential change due to ongoing pressures in the respective ecosystems (see details above under 3.4 'Impact factors').

  ### 7.5 Dataset E
Import GRUK data
```{r, eval = F, echo = F}
# read in data
GRUK_species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUK_alle_artsdata_2020-24.xlsx", sheet="Arter i ruter")
GRUK_ruter <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUK_alle_artsdata_2020-24.xlsx", sheet="Ruter")
GRUK_sirkler <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUK_alle_artsdata_2020-24.xlsx", sheet="Sirkler")
GRUK_polygoner <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUK_alle_artsdata_2020-24.xlsx", sheet="Polygoner")

```

  ### 7.6 Dataset F
  Import ASO data

```{r, eval = F, echo = F}
# ASO dataset

# GBIF query
#aso_ds <- dataset_search(doi = "10.15468/gq6wa5")
#aso_ds$data |>
#  select(title, datasetKey, publishingOrganizationTitle)
## extract dataset key
#aso_key <- aso_ds$data$datasetKey[1]
## Request GBIF download
#occ <- occ_search(datasetKey = aso_key, limit = 20000)
#ASO_species <- occ$data

# no abundance data. Falling back onto the available abundance data from 2022 only

ASO_species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/ASO/Semi-naturlig_eng_S123_2022.xlsx", sheet = "transektregistreringer_4")
ASO_points <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/ASO/Semi-naturlig_eng_S123_2022.xlsx", sheet = "surveyPoint_0")
```


## 8. Spatial units
The spatial units for functional plant community indicators are governed by the ANO data, which consist of 1000 randomly chosen sites in Norway. Each site is a 500 x 500 m grid cell with 18 monitoring points of 250sqm and a central 1 x 1 m vegetation plot. The basic unit for which observational community data exist, and for which every functional plant community indicator is computed, is this central 1 x 1 m vegetation plot. Aggregation of these basic spatial units to higher ones - like sites, municipalities, counties, regions, or the national level - should be done with appropriate consideration of the spatial structure of the data, i.e. imbalances and clustering (highly unequal number of points of an ecosystem type between sites).

## 9. Analyses

<!--# 
  
  Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed. 

Code folding is activated, meaning the code will be hidden by default in the html (one can click to expand it).

Caching is also activated (from the top YAML), meaning that rendering to html will be quicker the second time you do it. This will create a folder inside you project folder (called INDICATORID_cache). Sometimes caching created problems because some operations are not rerun when they should be rerun. Try deleting the cash folder and try again.

-->
  
  #### Data handling
  - Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)


```{r spatial_data, warnings = FALSE}
## Read Norway / regions, join to open lowlands points

nor <- st_read(paste0(here(), "/data/geo/outlineOfNorway_EPSG25833.shp"), quiet = TRUE) |>
  st_as_sf() |>
  st_transform(crs = st_crs(ano_lowlands))  # originally ano_lowlands

reg <- st_read(paste0(here(), "/data/geo/regions.shp"), quiet = TRUE) |>
  st_as_sf() |>
  st_transform(crs = st_crs(ano_lowlands))

# update region names
reg$region <- c(
  "Northern_Norway",
  "Central_Norway",
  "Eastern_Norway",
  "Western_Norway",
  "Southern_Norway"
)

regnor <- st_intersection(reg, nor)
```

```{r species_dictionary}
species_dict_pattern <- tibble::tibble(
  pattern = c(
    "Aconitum lycoctonum",
    "Aconitum septentrionale",
    "Agrostis hyemalis",
    "Anagallis minima",
    "Antennaria lapponica",
    "Antennaria porsildii",
    "Arabis wahlenbergii",
    "Arctous alpinus",
    "Aristavena setacea",
    "Atriplex lapponica",  
    "Atriplex praecox",    
    "Artemisia rupestris",
    "Betula tortuosa",
    "Blysmopsis rufa",
    "Cardamine dentata",
    "Cardamine nymanii",
    "Carex adelostoma",
    "Carex concolor",
    "Carex leersii",
    "Carex mackenziei",
    "Carex myosuroides",
    "Carex paupercula",
    "Carex simpliciuscula",
    "Carex viridula",
    "Carex xvacillans",
    "Chamaepericlymenum suecicum",
    "Chamerion angustifolium",
    "Cherleria biflora",
    "Cicerbita alpina",
    "Cirsium acaulon",
    "Clinopodium acinos",
    "Cornus suecia",
    "Cotoneaster scandinavicus",
    "Cotoneaster symondsii",
    "Cyanus montanus",
    "Dactylorhiza viridis",
    "Diphasiastrum alpinum",
    "Diphasiastrum complanatum",
    "Dryopteris affinis",
    "Empetrum hermaphroditum",
    "Elymus alaskanus",
    "Elymus repens",
    "Epilobium angustifolium",
    "Erysimum virgatum",
    "Euphrasia frigida",
    "Festuca prolifera",
    "Festuca trachyphylla",
    "Galium album",
    "Galium elongatum",
    "Hedlundia meinichii",
    "Helictotrichon pratense",
    "Helictotrichon pubescens",
    "Hieracium alpina",
    "Hieracium alpinum",
    "Hieracium hieracium",
    "Hieracium hieracioides",
    "Hieracium murorum",
    "Hieracium oreadea",
    "Hieracium prenanthoidea",
    "Hieracium pilosella",
    "Hieracium vulgata",
    "Hieracium vulgatum",
    "Huperzia appressa",
    "Huperzia arctica",
    "Hylotelephium maximum",
    "Juncus arcuatus",
    "Kali turgida",
    "Leontodon autumnalis",
    "Lepidotheca suaveolens",
    "Listera cordata",
    "Listera ovata",
    "Loiseleuria procumbens",
    "Luzula pallidula",
    "Lychnis alpina",
    "Matricaria discoidea",
    "Minuartia rubella",
    "Minuartia stricta",
    "Mycelis muralis",
    "Omalotheca norvegica",
    "Omalotheca supina",
    "Omalotheca sylvatica",
    "Ononis arvensis",
    "Oreopteris limbosperma",
    "Oxycoccus microcarpus",
    "Oxycoccus palustris",
    "Papaver dahlianum",   
    "Papaver lapponicum",  
    "Pentanema salicinum",
    "Phalaris minor",
    "Pilosella peletariana",
    "Pinus unicinata",
    "Poa alpigena",
    "Poa angustifolia",
    "Poa humilis",
    "Polygonum boreale",   
    "Polygonum excelsius", 
    "Polygonum norvegicum",
    "Potentilla anserina",
    "Potentilla arenosa",
    "Potentilla tabernaemontani",
    "Puccinellia retroflexa",
    "Pyrola grandiflora",
    "Pyrola norvegica",
    "Rosa dumalis",
    "Rosa vosagica",
    "Roegneria canina",
    "Rubus fruticosus",
    "Rumex alpestris",
    "Rumex graminifolius",
    "Salicornia pojarkovae",
    "Schedonorus elatior",
    "Sedum maximum",
    "Sedum telephium",
    "Selinum carvifolium",
    "Sorbus hybrida",
    "Spergula marina",
    "Spergularia salina",
    "Stellaria uliginosa",
    "Syringa emodi",
    "Taraxacum crocea",
    "Taraxacum croceum",
    "Thinopyrum junceiforme",
    "Trientalis europaea",
    "Trifolium pallidum",
    "Veratrum lobelianum"
  ),
  replacement = c(
    "Aconitum septentrionale",
    "Aconitum lycoctonum",
    "Agrostis scabra",
    "Lysimachia minima",
    "Antennaria alpina",
    "Antennaria alpina",
    "Arabis hirsuta",
    "Arctous alpina",
    "Deschampsia setacea",
    "Atriplex longipes",
    "Atriplex longipes",
    "Artemisia norvegica",
    "Betula pubescens",
    "Blysmus rufus",
    "Cardamine pratensis",
    "Cardamine pratensis",
    "Carex buxbaumii",
    "Carex aquatilis",
    "Carex echinata",
    "Carex norvegica",
    "Kobresia myosuroides",
    "Carex magellanica",
    "Kobresia simpliciuscula",
    "Carex flava",
    "Carex vacillans",
    "Cornus suecia",
    "Chamaenerion angustifolium",
    "Minuartia biflora",
    "Lactuca alpina",
    "Cirsium acaule",
    "Acinos arvensis",
    "Cornus suecica",
    "Cotoneaster integerrimus",
    "Cotoneaster simonsii",
    "Centaurea montana",
    "Coeloglossum viride",
    "Lycopodium alpinum",
    "Lycopodium complanatum",
    "Dryopteris filix-mas",
    "Empetrum nigrum",
    "Elymus kronokensis",
    "Elytrigia repens",
    "Chamaenerion angustifolium",
    "Erysimum strictum",
    "Euphrasia wettsteinii",
    "Festuca rubra",
    "Festuca brevipila",
    "Galium mollugo",
    "Galium palustre",
    "Sorbus aucuparia",
    "Avenula pratensis",
    "Avenula pubescens",
    "Hieracium Alpina",
    "Hieracium Alpina",
    "Hieracium Hieracium",
    "Hieracium umbellatum",
    "Hieracium Vulgata",
    "Hieracium Oreadea",
    "Hieracium Prenanthoidea",
    "Pilosella officinarum",
    "Hieracium Vulgata",
    "Hieracium umbellatum",
    "Huperzia selago",
    "Huperzia selago",
    "Hylotelephium telephium",
    "Luzula arcuata",
    "Kali turgidum",
    "Scorzoneroides autumnalis",
    "Matricaria discoidea",
    "Neottia cordata",
    "Neottia ovata",
    "Kalmia procumbens",
    "Luzula pallescens",
    "Viscaria alpina",
    "Matricaria discoidea",
    "Sabulina rubella",
    "Sabulina stricta",
    "Lactuca muralis",
    "Gnaphalium norvegicum",
    "Gnaphalium supinum",
    "Gnaphalium sylvaticum",
    "Ononis spinosa",
    "Thelypteris limbosperma",
    "Vaccinium microcarpum",
    "Vaccinium oxycoccos",
    "Papaver radicatum",
    "Papaver radicatum",
    "Inula salicina",
    "Phalaris arundinacea",
    "Pilosella officinarum",
    "Pinus mugo",
    "Poa pratensis",
    "Poa pratensis",
    "Poa pratensis",
    "Polygonum oxyspermum",
    "Argentina anserina",
    "Potentilla nivea",
    "Potentilla verna",
    "Puccinellia capillaris",
    "Pyrola rotundifolia",
    "Pyrola rotundifolia",
    "Rosa vosagiaca",
    "Rosa vosagiaca",
    "Elymus caninus",
    "Rubus plicatus",
    "Rumex acetosa",
    "Rumex acetosella",
    "Salicornia procumbens",
    "Schedonorus arundinaceus",
    "Hylotelephium telephium",
    "Hylotelephium telephium",
    "Selinum carvifolia",
    "Hedlundia hybrida",
    "Spergularia marina",
    "Spergularia marina",
    "Stellaria alsine",
    "Syringa vulgaris",
    "Taraxacum officinale",
    "Taraxacum officinale",
    "Elytrigia juncea",
    "Lysimachia europaea",
    "Trifolium pratense",
    "Veratrum album"
  )
)

clean_species_from_dictionary <- function(x, dict = species_dict_pattern) {
  x_clean <- x
  for (k in seq_len(nrow(dict))) {
    pat  <- dict$pattern[k]
    repl <- dict$replacement[k]
    x_clean[grepl(pat, x_clean)] <- repl
  }
  x_clean
}
```

```{r dataHandling_Tyler, warning=F} 
#, message=F, results='hide'

# 2.0 Plant indicator data

ind_tyler <- ind_tyler |>
  rename(scientific_name = Scientific_name) |> 
  mutate(scientific_name_original = scientific_name,
         scientific_name = str_replace_all(scientific_name, "ssp.", "subsp."),         # correct subspecies labelling
         scientific_name = str_replace_all(scientific_name, "\u00EB", "e"),
         scientific_name = str_remove_all(scientific_name, "agg."),                    # remove aggregates
         scientific_name = str_replace_all(scientific_name, " x ", " \u00D7 ")) |>     # correct hybrids labelling
  filter(!is.na(scientific_name), !scientific_name == "") |> 
  distinct(scientific_name_original, scientific_name, Moisture, Nitrogen) |>           # select indicator here
  tibble()

ind_tyler <- ind_tyler |> 
  mutate(scientific_name = clean_species_with_patterns(species, species_dict_pattern))

# remove certain species
ind_tyler <- ind_tyler |>  
  filter( !(scientific_name_original %in% list("Ammophila arenaria x Calamagrostis epigejos",
                                               "Anemone nemorosa x ranunculoides",
                                               "Armeria maritima ssp. elongata",
                                               "Asplenium trichomanes ssp. quadrivalens",
                                               "Calystegia sepium ssp. spectabilis",
                                               "Campanula glomerata 'Superba'",
                                               "Dactylorhiza maculata ssp. fuchsii",
                                               "Erigeron acris ssp. droebachensis",
                                               "Erigeron acris ssp. politus",
                                               "Erysimum cheiranthoides L. ssp. alatum",
                                               "Euphrasia nemorosa x stricta var. brevipila",
                                               "Galium mollugo x verum",
                                               "Geum rivale x urbanum",
                                               "Hylotelephium telephium (ssp. maximum)",
                                               "Juncus alpinoarticulatus ssp. rariflorus",
                                               "Lamiastrum galeobdolon ssp. argentatum",
                                               "Lathyrus latifolius ssp. heterophyllus",
                                               "Medicago sativa ssp. falcata",
                                               "Medicago sativa ssp. x varia",
                                               "Monotropa hypopitys ssp. hypophegea",
                                               "Ononis spinosa ssp. hircina",
                                               "Ononis spinosa ssp. procurrens",
                                               "Pilosella aurantiaca ssp. decolorans",
                                               "Pilosella aurantiaca ssp. dimorpha",
                                               "Pilosella cymosa ssp. gotlandica",
                                               "Pilosella cymosa ssp. praealta",
                                               "Pilosella officinarum ssp. peleteranum",
                                               "Poa x jemtlandica (Almq.) K. Richt.",
                                               "Poa x herjedalica Harry Sm.",
                                               "Ranunculus peltatus ssp. baudotii",
                                               "Sagittaria natans x sagittifolia",
                                               "Salix repens ssp. rosmarinifolia",
                                               "Stellaria nemorum L. ssp. montana",
                                               "Trichophorum cespitosum ssp. germanicum")
))

# prepare dataset for WFO matching
tyler_prepared_wfo <- WFO.prepare(ind_tyler$scientific_name)


# reconnect subspecies with corresponding species from authorship column
tyler_prepared <- tyler_prepared_wfo |>
  tibble() |> 
  mutate(
    # first word of Authorship
    subspecies = if_else(!is.na(Authorship), word(Authorship, 1, 1), ""),
    clean_string = word(spec.name, 1, 2),
    clean_string = if_else(
      str_detect(spec.name, " x$") |                                # ending with " x" OR
      str_detect(spec.name, "\u00D7$") |                            # ending with "×" OR
      str_detect(spec.name, "^\\s*[x\u00D7]\\b"),                   # starting with x/× hybrid marker
      paste(spec.name, subspecies),
      clean_string
    ),
    clean_string = if_else(                                         # remove 'sect.'
      str_detect(spec.name, "sect."),
      paste(word(clean_string, 1, 1), word(spec.full, 3, 3)),
      clean_string
    )
    ) |>
  # unique values in spec.full and clean_string only
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone (slow)
tyler_sp_matched <- WFO.match(spec.data = tyler_prepared$clean_string,
                             WFO.data = wfo_backbone,
                             Fuzzy = 0.15,
                             Fuzzy.max = 50,
                             Fuzzy.one = FALSE)



# finalise accepted name column according to latest taxonomical nomenclature
tyler_sp_clean <- tyler_sp_matched |>
  # make copy of the original species string
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
tyler_sp_clean |> filter(clean_string != accepted_name) #|> view()
tyler_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()


# bind new species names onto indicator dataset
tyler_species_clean <- left_join(tyler_prepared, tyler_sp_clean, by = "clean_string") |> 
  full_join(ind_tyler, by = join_by(spec.full == scientific_name)) |> 
  # filter out sect. species and subspecies
  #filter(!grepl("subsp.", scientific_name_original)) |> 
  distinct()

# correct incorrect corrections. haha
tyler_species_clean <- tyler_species_clean |> 
  mutate(
  flag_species_revert =
    case_when(
      accepted_name == "Rosa vinodora" ~ "reverted",
      #accepted_name == "Salix mollissima" ~ "reverted",
      accepted_name == "Rosa canina subsp. glauca" ~ "edited",
      clean_string == "Heracleum 'kungsholm'" ~ "reverted",
      clean_string == "Iris germanica" ~ "edited",
      clean_string == "Hedlundia atrata" ~ "reverted",
      clean_string == "Hedlundia faohraei" ~ "reverted",
      grepl("Hieracium", clean_string) & grepl("sect.", spec.full) ~ "edited",
      is.na(accepted_name) ~ "added",                                                          # fill missing accepted_name with clean_string
      TRUE ~ ""
      
    ),
  accepted_name = case_when(
    accepted_name = "Clinopodium acinos" ~ "Acinos arvensis",
    accepted_name == "Rosa vinodora" ~ "Rosa inodora",
    clean_string == "Heracleum 'kungsholm'" ~ clean_string,
    #accepted_name == "Salix mollissima" ~ clean_string,
    clean_string == "Iris germanica" ~ clean_string,
    clean_string == "Hedlundia atrata" ~ clean_string,
    clean_string == "Hedlundia faohraei" ~ clean_string,
    grepl("Hieracium", clean_string) & grepl("sect.", spec.full) ~ paste(word(spec.full, 1, 1), word(spec.full, 3, 3)),
    is.na(accepted_name) ~ clean_string,
    TRUE ~ accepted_name
  ))

# check name changes
tyler_species_clean |> filter(clean_string != accepted_name)


tyler_species_clean |> filter(is.na(accepted_name))


tyler_indicators <- tyler_species_clean |> 
  select(species = accepted_name,
         Moisture) |> 
  filter(!is.na(Moisture))

# remove unused dataframes
#rm(ind_tyler, tyler_prepared, tyler_prepared_wfo, tyler_sp_clean, tyler_sp_matched, tyler_species_clean)


# cache the indicator data here
#saveRDS(tyler_indicators, paste0(here::here(),"/data/cache/tyler_indicators.RDS"))

```

```{r dataHandling_GRUK, warning = FALSE}

# 2.1 GRUK data

## 2.1.1 GRUK species data handling
GRUK_species <- GRUK_species |> 
  janitor::clean_names() |> 
  rename(scientific_name = latinsk_navn,
         art_dekning = dekning_percent)


# fix species names
GRUK_species <- GRUK_species |>
  mutate(scientific_name_original = scientific_name,
         scientific_name = str_replace_all(scientific_name, "ssp.", "subsp."),         # correct subspecies labelling
         scientific_name = str_replace_all(scientific_name, "\u00EB", "e"),
         scientific_name = str_remove_all(scientific_name, "agg."),                    # remove aggregates
         scientific_name = str_replace_all(scientific_name, " x ", " \u00D7 ")) |>     # correct hybrids labelling
  filter(!is.na(scientific_name), !scientific_name == "") |> 
  tibble()

# update species names
GRUK_species <- GRUK_species |> 
  mutate(scientific_name = clean_species_with_patterns(species, species_dict_pattern))

GRUK_prepared_wfo <- WFO.prepare(GRUK_species$scientific_name)

GRUK_prepared <- GRUK_prepared_wfo |>
  tibble() |> 
  mutate(
    # first word of Authorship
    subspecies = if_else(!is.na(Authorship), word(Authorship, 1, 1), ""),
    clean_string = word(spec.name, 1, 2),
    clean_string = if_else(
      str_detect(spec.name, " x$") |                                # ending with " x" OR
      str_detect(spec.name, "\u00D7$") |                            # ending with "×" OR
      str_detect(spec.name, "^\\s*[x\u00D7]\\b"),                   # starting with x/× hybrid marker
      paste(spec.name, subspecies),
      clean_string
    ),
    clean_string = if_else(                                         # remove 'sect.'
      str_detect(spec.name, "sect."),
      paste(word(clean_string, 1, 1), word(spec.full, 3, 3)),
      clean_string
    )
    ) |>
  # unique values in spec.full and clean_string only
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone
GRUK_sp_matched <- WFO.match(spec.data = GRUK_prepared$clean_string,
                             WFO.data = wfo_backbone,
                             Fuzzy = 0.15,
                             Fuzzy.max = 50,
                             Fuzzy.one = FALSE)


# create accepted name column according to latest taxonomical nomenclature
GRUK_sp_clean <- GRUK_sp_matched |>
  # First, store the original string clearly
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # 1) Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # 2) Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # 3) Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
GRUK_sp_clean |> filter(clean_string != accepted_name)
GRUK_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name)


# correct incorrect corrections
GRUK_sp_clean <- GRUK_sp_clean |> 
  mutate(
    flag_species_revert =
      case_when(
        clean_string == "Hieracium vulgata" ~ "added",
        clean_string == "Hieracium murorum" ~ "edited",
        clean_string == "Potentilla anserina" ~ "added",
        TRUE ~ ""
        
      ),
    accepted_name = case_when(
        clean_string == "Hieracium vulgata" ~ "Hieracium Vulgata",
        clean_string == "Hieracium murorum" ~ "Hieracium Vulgata",
        clean_string == "Potentilla anserina" ~ "Argentina anserina",
      TRUE ~ accepted_name
    ))



# bind new species names onto original dataset
GRUK_species_clean <- full_join(GRUK_prepared, GRUK_sp_clean, by = "clean_string") |> 
  full_join(GRUK_species, by = join_by(spec.full == scientific_name)) |> 
  mutate(accepted_name = case_when(
    is.na(accepted_name)& !grepl("Hieracium", spec.full) ~ str_to_sentence(spec.full),
    is.na(accepted_name)& grepl("Hieracium", spec.full) ~ spec.full,
    TRUE ~ accepted_name
  )) |> 
  distinct() |> 
  filter(!is.na(accepted_name))


# check for original species with no matched accepted name.
GRUK_species_clean |> 
  filter(is.na(accepted_name)) |> 
  tibble()
# none.


#GRUK_species_clean |> group_by(accepted_name, global_id) |> mutate(n = n_distinct(art_dekning)) |>  filter(n > 1)

# merge species data with indicators

GRUK_species_ind <- GRUK_species_clean |>
  select(parent_global_id, polygon_id, rute_id, creation_date, species = accepted_name, art_dekning) |> 
  left_join(tyler_indicators,
            by = join_by("species")) |> 
  tibble()


summary(GRUK_species_ind)

# checking which species didn't find a match
unique(GRUK_species_ind[is.na(GRUK_species_ind$Moisture),'species'])# |> view()




#########################################################
## 2.1.2 GRUK ruter data handling
names(GRUK_ruter)

GRUK_ruter <- GRUK_ruter |> 
  janitor::clean_names() |>
  select(
    global_id,
    polygon_id,
    rute_id,
    rute_id_loknr,
    dekning_karplanter_feltsjikt = dekning_percent_av_karplanter_i_feltsjikt,
    dekning_moser =  dekning_percent_av_moser,
    dekning_lav = dekning_percent_av_lav,
    dekning_stro = dekning_percent_av_stro,
    dekning_bar_jord_grus_stein_berg = dekning_percent_av_bar_jord_grus_stein_berg,
    precision,
    UTM33_E_ne = utm33_e_ne, 
    UTM33_N_ne = utm33_n_ne,
    UTM33_E_sw = utm33_e_sw,
    UTM33_N_sw = utm33_n_sw,
    areal_m2,
    registeringsdato
  ) |> 
  mutate(year = year(registeringsdato)) |> 
  select(-registeringsdato)


# make coordinates numeric
GRUK_ruter <- GRUK_ruter |> 
  mutate(UTM33_E_ne = as.numeric(UTM33_E_ne),
         UTM33_N_ne = as.numeric(UTM33_N_ne),
         UTM33_E_sw = as.numeric(UTM33_E_sw),
         UTM33_N_sw = as.numeric(UTM33_N_sw) )

# calculate central coordinates for each plot
GRUK_ruter <- GRUK_ruter |> 
  mutate(UTM33_N = (UTM33_N_ne + UTM33_N_sw)/2,
         UTM33_E = (UTM33_E_ne + UTM33_E_sw)/2)

# some of the calculations throw NA's because there's only one set of coordinates, coalesce that set into the calculation column 
GRUK_ruter <- GRUK_ruter |> 
  mutate (UTM33_N = coalesce(UTM33_N,UTM33_N_ne),
          UTM33_E = coalesce(UTM33_E,UTM33_E_ne),
          UTM33_N = coalesce(UTM33_N,UTM33_N_sw),
          UTM33_E = coalesce(UTM33_E,UTM33_E_sw) 
          )


## 2.1.3 GRUK sirkler data handling
## merge information on mapping units and condition variables from GRUK.sirkler into GRUK.ruter
names(GRUK_ruter)
names(GRUK_sirkler)

GRUK_sirkler <- GRUK_sirkler |> 
  janitor::clean_names() |> 
  select(
    global_id,
    kartleggingsenhet_1_5000,
    spor_etter_slitasje_og_slitasjebetinget_erosjon = spor_etter_slitasje_og_slitasjebetinget_erosjon_percent,
    dekning_nakent_berg= dekning_percent_av_nakent_berg,
    total_dekning_vedplanter_i_feltsjikt = total_dekning_percent_av_vedplanter_i_feltsjikt,
    dekning_busker_busksjikt = dekning_percent_av_busker_i_busksjikt,
    dekning_tresjikt = dekning_percent_av_tresjikt,
    dekning_problemarter = dekning_percent_av_problemarter,
    total_dekning_fremmede_arter = total_dekning_percent_av_fremmede_arter,
    registeringsdato
  ) |> 
  mutate(year = year(registeringsdato)) |> 
  select(-registeringsdato)


GRUK_variables <- GRUK_ruter |> 
  full_join(GRUK_sirkler, by = c("global_id", "year")) |>           # join plots with circles
  filter(!is.na(UTM33_N))                                           # filter away plots with no coordinates

summary(GRUK_variables)



## merge information on condition and quality from GRUK.polygoner into GRUK.variables
# transform GRUK.variables into spatial object
GRUK_variables <- st_as_sf(GRUK_variables, coords = c("UTM33_E","UTM33_N"), remove = FALSE, crs = 25833)


## 2.1.4 GRUK polygoner data handling
# transform GRUK.polygoner into spatial object
GRUK_polygoner <- st_as_sf(GRUK_polygoner, wkt = "WKT" ,remove=F, crs = 25833) |> 
  janitor::clean_names() |> 
  rename(
    nin_id = ni_nid,
    year = ar,
    nin_kartleggingsenheter = ni_n_kartleggingsenheter,
    nin_beskrivelsesvariabler = ni_n_beskrivelsesvariabler
  )





tm_shape(GRUK_polygoner) +
  tm_graticules() +
  tm_polygons("polygon_id") +
  tm_shape(GRUK_variables) +
  tm_dots("rute_id")

# run a spatial join to get columns from GRUK.polygoner into GRUK.variables
#GRUK.variables <- st_join(GRUK.variables,GRUK.polygoner[,c(3:4,9,15,18,20,22,24,60)])
#names(GRUK.variables)[1:33]<-c("GlobalID","PolygonID.x","RuteID","RuteID_loknr",
#                               "Dekning_karplanter_feltsjikt","Dekning_moser","Dekning_lav","Dekning_strø",
#                               "Dekning_bar_substrat","Precision","UTM33_E_ne","UTM33_N_ne",
#                               "UTM33_E_sw","UTM33_N_sw","areal(m2)","UTM33_N","UTM33_E","Kartleggingsenhet_1til5000",
#                               "erosjon_prosent","Dekning_nakentberg",
#                               "Totaldekning_vedplanter_feltsjikt","Dekning_busker_busksjikt","Dekning_tresjikt",
#                               "Dekning_problemarter","Totaldekning_fremmedearter","LokalitetID","PolygonID.y",
#                               "Kartleggingsdato","Lokalitetskvalitet","Kommune","Tilstand","Naturmangfold","NiNKartleggingsenheter")

# check how good the spatial join worked
#cbind(GRUK.variables$PolygonID.x,GRUK.variables$PolygonID.y)
#GRUK.variables[7,]
#GRUK.polygoner[GRUK.polygoner$PolygonID=="46-2",]
# some points could not be matched to polygons -> merge by PolygonID instead, drop geometry of GRUK.variables first
GRUK_variables <- GRUK_variables |> 
  st_drop_geometry()

names(GRUK_variables)
names(GRUK_polygoner)


GRUK_variables <- GRUK_variables |> 
  left_join(GRUK_polygoner, by = c("year", "polygon_id")) # there are 94 rows only found in GRUK_polygoner


summary(GRUK_variables) 
summary(as.factor(GRUK_variables$tilstand)) # 5 plots with NAs


# edit the column names
GRUK_variables <- GRUK_variables |> 
  rename(erosion_percent = spor_etter_slitasje_og_slitasjebetinget_erosjon)



## adding information on ecosystem and condition variables to species+indicator data
names(GRUK_species_ind)
names(GRUK_variables)

GRUK_all <- GRUK_species_ind |> 
  left_join(GRUK_variables |>  select(-polygon_id, -rute_id), by = join_by(parent_global_id == global_id))



# fixing variable types
GRUK_all <- GRUK_all |> 
  mutate(across(
    c(species, kartleggingsenhet_1_5000, lokalitetskvalitet, kommune, tilstand, naturmangfold, nin_kartleggingsenheter),
    as.factor
  ),
  across(
    c(areal_m2, dekning_nakent_berg, dekning_problemarter),
    as.numeric
  )) |> 
# trimming away the points without information on NiN, species or cover  
    filter(!is.na(species), !is.na(art_dekning))

summary(GRUK_all)


#rm(GRUK_polygoner, GRUK_prepared, GRUK_prepared_wfo, GRUK_ruter, GRUK_sirkler, GRUK_sp_clean, GRUK_sp_matched, GRUK_species, GRUK_species_clean, GRUK_species_ind, GRUK_variables)

#saveRDS(GRUK_all, paste0(here::here(),"/data/cache/GRUK_all.RDS"))
#saveRDS(GRUK_species_ind, paste0(here::here(),"/data/cache/GRUK_species_ind.RDS"))



```

```{r dataHandling_ASO, warning = FALSE}
# 2.2 ASO
# 2.2.1 ASO points data

ASO_points <- st_as_sf(x = ASO_points, 
                    coords = c("x", "y"),
                    crs = "+proj=longlat +datum=WGS84 +ellps=WGS84")

ASO_points <- ASO_points |> 
  janitor::clean_names()

# Rename using base R
nms <- names(ASO_points)

nms[nms == "dominerende_kartleggingsenhet_1_5000_t32"] <- "nin_grunntype"
nms[nms == "aktuell_bruksintensitet_7jb_ba"]           <- "bruksintensitet"
nms[nms == "beitetrykk_7jb_bt"]                        <- "beitetrykk"
nms[nms == "slatteintensitet_7jb_si"]                  <- "slatteintensitet"
nms[nms == "spor_etter_ferdsel_med_tunge_kjoretoy_m_dir_prtk"] <-
  "tungekjoretoy"
nms[nms == "spor_etter_slitasje_og_slitasjebetinget_erosjon_m_dir_prse"] <-
  "slitasje"

names(ASO_points) <- nms

## fix NiN-variables
# remove variable code in the data
ASO_points <- ASO_points |>
  mutate(
    bruksintensitet = bruksintensitet |>
      str_remove("^7JB-BA_") |>
      na_if("X") |>
      as.numeric(),
    
    beitetrykk = beitetrykk |>
      str_remove("^7JB-BT_") |>
      na_if("X") |>
      as.numeric(),
    
    slatteintensitet = slatteintensitet |>  # 4 NAs
      str_remove("^7JB-SI_") |>
      na_if("X") |>
      as.numeric(),
    
    tungekjoretoy = tungekjoretoy |>
      str_remove("^MDirPRTK_") |>
      na_if("X") |>
      as.numeric(),
    
    slitasje = slitasje |>
      str_remove("^MDirPRSE_") |>
      na_if("X") |>
      as.numeric()
  )



# 2.2.2 species data

## fixing variable names and issues in ASO.sp
head(as.data.frame(ASO_species))

ASO_species <- rename(ASO_species, art_dekning = Dekning)

# fix species names
ASO_species <- ASO_species |>
  mutate(scientific_name_original = scientific_name) |> 
  separate(
    col  = Navn,
    into = c("norsk_navn", "scientific_name"),
    sep  = "_",
    extra = "merge",   # keep any additional _ in the "after" part
    fill  = "right"    # if no _, "after" becomes NA
  ) |> 
  mutate(
    scientific_name = scientific_name  |> 
      str_replace_all("_", " ")  |> 
      str_to_sentence(),
    
    scientific_name = if_else(
      is.na(word(scientific_name, 2)),        # only one word (no species epithet)
      word(scientific_name, 1),              # just genus
      word(scientific_name, 1, 2)            # genus + species
    ) %>%
      str_replace_all("-", " ")  |> 
      str_squish()
  )

ASO_species <- ASO_species  |>  
  mutate(scientific_name = clean_species_with_patterns(species, species_dict_pattern))


# filter out NAs for WFO cleaning
ASO_species_prepared <- filter(ASO_species, !is.na(scientific_name))


ASO_prepared_wfo <- WFO.prepare(ASO_species_prepared$scientific_name)

ASO_prepared <- ASO_prepared_wfo |>
  mutate(
    spec.name = case_when(
      !is.na(Authorship) & Authorship != "" ~ paste0(spec.name, "-", Authorship),
      TRUE ~ spec.name) |>
      str_squish() |> 
      str_to_sentence()
  ) |>
  rename(clean_string = spec.name) |>
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone
ASO_sp_matched <- WFO.match(spec.data = ASO_prepared$clean_string,
                             WFO.data = wfo_backbone,
                             Fuzzy = 0.15,
                             Fuzzy.max = 50,
                             Fuzzy.one = FALSE)


# create accepted name column according to latest taxonomical nomenclature
ASO_sp_clean <- ASO_sp_matched |>
  # First, store the original string clearly
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # 1) Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # 2) Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # 3) Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
ASO_sp_clean |> filter(clean_string != accepted_name)
ASO_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name)


# correct incorrect corrections. haha
ASO_sp_clean <- ASO_sp_clean |> 
  mutate(
    flag_species_revert =
      case_when(
        clean_string == "Hieracium vulgata" ~ "added",
        clean_string == "Taraxacum crocea" ~ "added",
        clean_string == "Taraxacum hamata" ~ "added",
        TRUE ~ ""
        
      ),
    accepted_name = case_when(
      clean_string == "Hieracium vulgata" ~ "Hieracium Vulgata",
      clean_string == "Taraxacum crocea" ~ clean_string,
      clean_string == "Taraxacum hamata" ~ clean_string,
      TRUE ~ accepted_name
    )) |> 
  # add "sp." back onto genus-level identifications
  mutate(accepted_name = case_when(
    is.na(word(accepted_name, 2)) ~ paste(accepted_name, "sp."),
    TRUE ~ accepted_name
  ))



# bind new species names onto original dataset
ASO_species_clean <- left_join(ASO_prepared, ASO_sp_clean, by = "clean_string") |> 
  full_join(ASO_species, by = join_by(spec.full == scientific_name)) |> 
  distinct() |> 
  # add on Hieracium group info lost in Norwegian name
  mutate(
    accepted_name = case_when(
      norsk_navn == "skogsvevegruppa" ~ paste("Hieracium", str_sub(karplantenavn, 19, 27))  |> str_squish(),
      norsk_navn == "beitesvevegruppa" ~ paste("Hieracium", str_sub(karplantenavn, 20, 26))  |> str_squish(),
      norsk_navn == "skjermsvevegruppa" ~ paste("Hieracium", str_sub(karplantenavn, 21, 32))  |> str_squish(),
      TRUE ~ accepted_name
    )
  )


# check for original species with no matched accepted name.
ASO_species_clean |> 
  filter(is.na(accepted_name)) |> 
  tibble()



# 2.2.3 merge with indicator data

## merge species data with indicators
ASO_species_ind <- ASO_species_clean |>
  select(species = accepted_name, art_dekning, ParentGlobalID) |> 
  filter(!is.na(art_dekning)) |> 
  left_join(tyler_indicators |> 
              select(species, Moisture)) |> 
  tibble()



# checking which species didn't find a match
unique(ASO_species_ind[is.na(ASO_species_ind$Moisture),'species'])



## adding information on ecosystem and condition variables to species data
ASO_all <- ASO_species_ind |> 
  left_join(ASO_points |>  
              select(global_id, omradenummer_flatenummer, eng_id, aso_id, nin_grunntype, annen_dominerende_kartleggingsenhet), 
            by = join_by(ParentGlobalID == global_id))        # check why there are 276 points with no species data?



# fixing variable types
ASO_all <- ASO_all |> 
  mutate(across(
    c(species, nin_grunntype, omradenummer_flatenummer, eng_id, aso_id),
    as.factor
  )) |> 
  # trimming away the points without information on NiN, species or cover  
  filter(!is.na(species), !is.na(art_dekning), !is.na(nin_grunntype))

summary(ASO_all)

 #rm(ASO_species, ASO_prepared, ASO_sp_matched, ASO_points, ASO_sp_clean, ASO_species_clean, ASO_species_prepared, ASO_prepared_wfo)

#saveRDS(ASO_all, paste0(here::here(),"/data/cache/ASO_all.RDS"))
#saveRDS(ASO_species_ind, paste0(here::here(),"/data/cache/ASO_species_ind.RDS"))



```

```{r dataHandling_ANO, warning = FALSE}
#### 2.3 ANO monitoring data

# 2.3.1 extract species data
ano_species <- ano_species |> 
  mutate(scientific_name = str_replace_all(art_navn, "_", " "),
         scientific_name_original = scientific_name,
         scientific_name = str_replace_all(scientific_name, "ssp.", "subsp."),         # correct subspecies labelling
         scientific_name = str_replace_all(scientific_name, "\u00EB", "e"),
         scientific_name = str_remove_all(scientific_name, "agg."),                    # remove aggregates
         scientific_name = str_replace_all(scientific_name, " x ", " \u00D7 "),     # correct hybrids labelling
         scientific_name = str_to_sentence(scientific_name),
         scientific_name = if_else(
           is.na(word(scientific_name, 2)),        # only one word (no species epithet)
           word(scientific_name, 1),              # just genus
           word(scientific_name, 1, 2)            # genus + species
           ) |> 
           #str_replace_all("-", " ")  |> 
           str_squish()
         ) |> 
  filter(!is.na(scientific_name), !scientific_name == "") |> 
  tibble()


# recode species names for WFO matching
ano_species <- ano_species |> 
  mutate(scientific_name = clean_species_with_patterns(species, species_dict_pattern))

ano_prepared_wfo <- ano_species |> 
  distinct(scientific_name, scientific_name_original)

# prepare dataset for WFO matching
ano_prepared_wfo <- WFO.prepare(ano_prepared_wfo$scientific_name)

# reconnect subspecies with corresponding species from authorship column
ano_prepared <- ano_prepared_wfo |>
  tibble() |> 
  mutate(
    spec.name = spec.name |>
      str_replace_all("\\bsect\\b\\.?", " ") |>
      str_squish()
  ) |>
  rename(clean_string = spec.name) |> 
  # unique values in spec.full and clean_string only
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone (slow)
ano_sp_matched <- WFO.match(spec.data = ano_prepared$clean_string,
                              WFO.data = wfo_backbone,
                              Fuzzy = 0.15,
                              Fuzzy.max = 50,
                              Fuzzy.one = FALSE)


# finalise accepted name column according to latest taxonomical nomenclature
ano_sp_clean <- ano_sp_matched |>
  # make copy of the original species string
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # 1. Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # 2. Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # 3. Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
ano_sp_clean |> filter(clean_string != accepted_name) #|> view()
ano_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()


# correct incorrect corrections
ano_sp_clean <- ano_sp_clean |> 
  mutate(
    flag_species_revert =
      case_when(
        grepl("Hieracium", clean_string) ~ "edited",
        clean_string == "Poa jemtlandica" ~ "edited",
        TRUE ~ ""
        
      ),
    accepted_name = case_when(
        grepl("Hieracium", clean_string) ~ clean_string,
        accepted_name == "Poa jemtlandica" ~ "Poa alpina",
      TRUE ~ accepted_name
    ))

# check name changes
ano_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()


ano_species_clean <- left_join(ano_prepared, ano_sp_clean, by = "clean_string") |> 
  full_join(ano_species, by = join_by(spec.full == scientific_name)) |> 
  # filter out sect. species and subspecies
  #filter(!grepl("subsp.", scientific_name_original)) |> 
  distinct() |> 
  filter(!is.na(accepted_name))

ano_species_clean |> filter(is.na(accepted_name))


## merge species data with indicators
ANO_species_ind <- ano_species_clean |>                                 # fix species match here
  select(species = accepted_name, art_dekning, parentglobalid) |> 
  full_join(tyler_indicators |> 
              select(species, Moisture)) |> 
  tibble()



# checking which species didn't find a match
unique(ANO_species_ind[is.na(ANO_species_ind$Moisture),'species']) # |> view()



# fixing variable types
ANO_species_ind <- ANO_species_ind |> 
  mutate(across(
    c(species),
    as.factor
  )) |> 
  # trimming away the points without information on NiN, species or cover  
  filter(!is.na(species), !is.na(art_dekning))

summary(ANO_species_ind)



## fix NiN information
ano_geo <- ano_geo |>
  mutate(
    # 3 first characters of kartleggingsenhet_1m2, no hyphen
    hovedtype_rute = substr(kartleggingsenhet_1m2, 1, 3),
    hovedtype_rute = gsub("-", "", hovedtype_rute),
    
    # recode to hovedøkosystem
    hovedoekosystem_rute = recode(
      hovedtype_rute,
      "T4"  = "Forest", "T30" = "Forest",
      "T3"  = "Mountain", "T7"  = "Mountain", "T14" = "Mountain", "T22" = "Mountain",
      "V1"  = "Wetland", "V2"  = "Wetland", "V3"  = "Wetland", "V4"  = "Wetland", "V5"  = "Wetland", "V6"  = "Wetland", 
      "V7"  = "Wetland", "V8"  = "Wetland",
      "T31" = "Seminat", "T32" = "Seminat", "T33" = "Seminat", "T34" = "Seminat", "V9"  = "Seminat", "V10" = "Seminat",
      "T2"  = "Natopen", "T8"  = "Natopen", "T11" = "Natopen", "T12" = "Natopen", "T13" = "Natopen", "T15" = "Natopen", 
      "T16" = "Natopen", "T18" = "Natopen", "T21" = "Natopen", "T24" = "Natopen", "T29" = "Natopen",
      "T41" = "Seminat", # DOUBLE CHECK THIS!
      "T45" = "Seminat"  # DOUBLE CHECK THIS!
    )
  )

# 2. Fix NiN variable names
ano_geo <- ano_geo |> 
  rename(
  groeftingsintensitet = bv_7jb_ba,
  bruksintensitet      = bv_7jb_bt,
  beitetrykk           = bv_7jb_si,
  slatteintensitet     = bv_7tk,
  tungekjoretoy        = bv_7se,
  slitasje             = forekomst_ntyp
)

## 3. Clean NiN codes (remove prefixes, X -> NA, to numeric)

ano_geo <- ano_geo |>
  mutate(
    groeftingsintensitet = groeftingsintensitet |>
      str_remove("^7GR-GI_") |>
      na_if("X") |>
      as.numeric(),
    
    bruksintensitet = bruksintensitet |>
      str_remove("^7JB-BA_") |>
      na_if("X") |>
      as.numeric(),
    
    beitetrykk = beitetrykk |>
      str_remove("^7JB-BT_") |>
      na_if("X") |>
      as.numeric(),
    
    slatteintensitet = slatteintensitet |>
      str_remove("^7JB-SI_") |>
      na_if("X") |>
      as.numeric(),
    
    tungekjoretoy = tungekjoretoy |>
      str_remove("^7TK_") |>
      na_if("X") |>
      as.numeric(),
    
    slitasje = slitasje |>
      str_remove("^7SE_") |>
      na_if("X") |>
      as.numeric()
  )

## 4. Filter lowland plots points
ano_lowlands <- ano_geo# |>
  #filter(hovedoekosystem_rute %in% c("Natopen", "Seminat", "T45", "T41"))


ano_lowlands <- st_join(
  ano_lowlands,
  regnor,
  left = TRUE,
  join = st_nearest_feature
)


ANO_species_ind <- ANO_species_ind |> 
  left_join(ano_lowlands |> select("globalid","ano_flate_id","ano_punkt_id","ssb_id","aar", "hovedtype_rute","kartleggingsenhet_1m2",
                             #"bruksintensitet","beitetrykk","slatteintensitet", "tungekjoretoy","slitasje",
                             "vedplanter_total_dekning","busker_dekning","tresjikt_dekning","roesslyng_dekning"), 
            by = join_by(parentglobalid == globalid)) |> 
  filter(!is.na(hovedtype_rute))


#saveRDS(ano_lowlands, paste0(here::here(),"/data/cache/ano_lowlands.RDS"))
#saveRDS(ANO_species_ind, paste0(here::here(),"/data/cache/ANO_species_ind.RDS"))


```

```{r dataHandling_NiN, warning = FALSE}

# 2.4.1 data handling NiN GRUK
head(natopen_NiN_ref)
head(natopen_NiN_species)

colnames(natopen_NiN_ref)[1] <- "species"
head(natopen_NiN_ref)

natopen_NiN_ref <- merge(natopen_NiN_ref, natopen_NiN_species[,c(1,4)], by.x="species", by.y="ScientificName", all.x=T)
unique(natopen_NiN_ref[is.na(natopen_NiN_ref$Phylum),'species']) # Pucinella does not exist in ind.dat, so we don't care
# we're only interested in vascular plants and ferns, which we have indicators on
unique(natopen_NiN_ref$Phylum)
natopen_NiN_ref <- natopen_NiN_ref %>%
  filter(Phylum %in% c("Magnoliophyta","Pteridophyta"))
unique(natopen_NiN_ref$Phylum)



natopen_NiN_ref$species
# only genus and species name
natopen_NiN_ref$sp.orig <- natopen_NiN_ref$species
natopen_NiN_ref$species <- word(natopen_NiN_ref$species, 1,2)
natopen_NiN_ref <- natopen_NiN_ref[!is.na(natopen_NiN_ref$species),]

natopen_NiN_ref <- natopen_NiN_ref |> 
    mutate(species = str_to_sentence(species)
    # if needed later: filter out groups, e.g. trees
    # |> filter(spgr != "a1a")
  ) |>
  select(-sp.orig) |> 
  pivot_longer(
    cols = -c(species, Phylum),     # columns to pivot
    names_to   = "nin_id",                                   # new column for former column names
    values_to  = "cover"                                     # new column for values
  ) |> 
  filter(cover != 0)                                         # filter out 0 cover 


natopen_NiN_ref <- natopen_NiN_ref %>%
  mutate(species = clean_species_with_patterns(species, species_dict_pattern)) |> 
  group_by(species, nin_id) |> 
  summarise(cover = max(cover),           # summing cover of species which have ended up appearing several times per nature type
    .groups = "drop"
  )

natop_nin_prepared_wfo <- natopen_NiN_ref |> distinct(species)

# prepare dataset for WFO matching
natop_nin_prepared_wfo <- WFO.prepare(natop_nin_prepared_wfo$species)


# reconnect subspecies with corresponding species from authorship column
natop_nin_prepared <- natop_nin_prepared_wfo |>
  rename(clean_string = spec.name) |> 
  tibble() |> 
  mutate(
    clean_string = clean_string |>
      str_replace_all("\\bsect\\b\\.?", " ") |>
      str_squish()
  ) |> 
  # unique values in spec.full and clean_string only
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone
natop_nin_sp_matched <- WFO.match(spec.data = natop_nin_prepared$clean_string,
                            WFO.data = wfo_backbone,
                            Fuzzy = 0.15,
                            Fuzzy.max = 50,
                            Fuzzy.one = FALSE)




# finalise accepted name column according to latest taxonomical nomenclature
natop_nin_sp_clean <- natop_nin_sp_matched |>
  # make copy of the original species string
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # 1. Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # 2. Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # 3. Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
natop_nin_sp_clean |> filter(clean_string != accepted_name) #|> view()
natop_nin_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()


# bind new species names onto indicator dataset
natop_nin_species_clean <- left_join(natop_nin_prepared, natop_nin_sp_clean, by = "clean_string") |> 
  full_join(natopen_NiN_ref, by = join_by(spec.full == species)) |> 
  # filter out sect. species and subspecies
  #filter(!grepl("subsp.", scientific_name_original)) |> 
  distinct()

natop_nin_species_clean |> filter(is.na(accepted_name))


# Merge with indicator values
natop_nin_sp_ind <- natop_nin_species_clean |>
  select(nin_id, species = accepted_name, cover) |> 
  left_join(tyler_indicators |> 
              select(species, Moisture),
            by = join_by(species)) |> 
  tibble() |>
  filter(grepl("T2-C-7", nin_id) | grepl("T2-C-8", nin_id)) |> 
  mutate(nin_id = str_remove(nin_id, "_BN")) |> 
      group_by(species, nin_id, Moisture) |> 
  summarise(cover = max(cover),           # summing cover of species which have ended up appearing several times per nature type
    .groups = "drop"
  )



# double check species matching
natop_nin_sp_ind |> filter(is.na(species))  
natop_nin_sp_ind |> filter(is.na(Moisture)) |>  distinct(species)




# 2.4.2 reference data - data handling ANO

### Inspect Eco_State structure 

str(Eco_State)

# species list, env data, abundance data (same as your checks)
Eco_State$Concept_Data$Species$Species_List$species
t(Eco_State$Concept_Data$Env$Env_Data)
t(Eco_State$Concept_Data$Species$Species_Data)

# Transpose & prepare species abundance data

NiN_sp <- Eco_State$Concept_Data$Species$Species_Data |>
  t() |> 
  as_tibble()

NiN_sp <- NiN_sp |>
  mutate(
    species_original = as_factor(as.vector(Eco_State$Concept_Data$Species$Species_List$species)),
    species_group = as_factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code)),
    # only genus + species
    species = word(species_original, 1, 2),
    species = str_to_sentence(species)
    # if needed later: filter out groups, e.g. trees
    # |> filter(spgr != "a1a")
  ) |>
  select(-species_original) |> 
  group_by(species, species_group) |> 
  summarise(
    across(
      where(is.numeric),  # all numeric columns (e.g. abundances)
      ~ sum(.x, na.rm = TRUE)
    ),
    .groups = "drop"
  ) |> 
  pivot_longer(
    cols = -c(species, species_group),     # columns to pivot
    names_to   = "nin_id",                                   # new column for former column names
    values_to  = "cover"                                     # new column for values
  ) |> 
  filter(!species_group %in% c("a2m", "a2lb", "a2l")) |>     # filter out moss and lichen species 
  filter(cover != 0)                                         # filter out 0 cover 



### Environment data 

NiN_env <- Eco_State$Concept_Data$Env$Env_Data

NiN_sp <- NiN_sp |> 
  left_join(NiN_env, 
                    by = join_by(nin_id == ID)) |> 
  filter(Nature_Type %in% c("Semi_Natural ", "Coastal_Heath")) |>   # filter for the desired nature type
  mutate(nin_code = case_match(
      nin_id,
      "CH01" ~ "T34-C1",
      "CH02" ~ "T34-C2a",
      "CH03" ~ "T34-C2b",
      "CH04" ~ "T34-C2c",
      "CH06" ~ "T34-C3",
      "CH07" ~ "T34-C4a",
      "CH08" ~ "T34-C4b",
      "CH09" ~ "T34-C4c",
      #"CH10" ~ "T34-C4d",
      #"CH11" ~ "T34-C5a",
      "CH12" ~ "T34-C5b",
      #"CH13" ~ "T34-C5c",
      #"CH14" ~ "T34-C6a",
      "CH15" ~ "T34-C6b",
      #"CH16" ~ "T34-C6c",
      "SN08" ~ "T32-C1C2",
      "SN09" ~ "T32-C3C4",
      "SN10" ~ "T32-C5C20a",
      "SN11" ~ "T32-C7C8",
      "SN12" ~ "T32-C5C20b",
      "SN13" ~ "T32-C9a",
      "SN14" ~ "T32-C9b",
      "SN15" ~ "T32-C15",
      "SN16" ~ "T32-C21C6a",
      "SN17" ~ "T32-C21C6b",
      "SN18" ~ "T32-C10a",
      "SN19" ~ "T32-C10b",
      "SN20" ~ "T32-C16",
      "SN21" ~ "T41a",
      "SN22" ~ "T41b",
      "SN23" ~ "T45-C1C2",
      "SN24" ~ "T45-C3",
      "SN25" ~ "V10-C1C2",
      "SN26" ~ "V10-C3",
      .default = NA_character_
    )
  ) |>
  filter(!is.na(nin_code))


NiN_sp <- NiN_sp |> 
  mutate(species = recode(
    species,
    "Aconitum lycoctonum"          = "Aconitum septentrionale",
    "Anagallis arvensis"           = "Lysimachia arvensis",
    "Anagallis minima"             = "Lysimachia minima",
    "Arctous alpinus"              = "Arctous alpina",
    "Betula tortuosa"              = "Betula pubescens",
    "Blysmopsis rufa"              = "Blysmus rufus",
    "Chamerion angustifolium"      = "Chamaenerion angustifolium",
    "Cardamine nymanii"            = "Cardamine pratensis",
    "Carex adelostoma"             = "Carex buxbaumii",
    "Carex leersii"                = "Carex echinata",
    "Carex paupercula"             = "Carex magellanica",
    "Carex simpliciuscula"         = "Kobresia simpliciuscula",
    "Carex _vacillans"             = "Carex vacillans",
    "Carex viridula"               = "Carex flava",
    "Chamaepericlymenum suecicum"  = "Cornus suecia",
    "Cornus suecia"                = "Cornus suecica",
    "Cicerbita alpina"             = "Lactuca alpina",
    "Dactylorhiza fuchsii"         = "Dactylorhiza maculata",
    "Dactylorhiza sphagnicola"     = "Dactylorhiza majalis",
    "Diphasiastrum alpinum"        = "Lycopodium alpinum",
    "Diphasiastrum complanatum"    = "Lycopodium complanatum",
    "Elymus alaskanus"             = "Elymus kronokensis",
    "Empetrum hermaphroditum"      = "Empetrum nigrum",
    "Erigeron acer"                = "Erigeron acris",
    "Erigeron eriocephalus"        = "Erigeron uniflorus",
    "Festuca altissima"            = "Drymochloa sylvatica",
    "Festuca prolifera"            = "Festuca rubra",
    "Galium album"                 = "Galium mollugo",
    "Galium elongatum"             = "Galium palustre",
    "Glaux maritima"               = "Lysimachia maritima",
    "Helictotrichon pratense"      = "Avenula pratensis",
    "Helictotrichon pubescens"     = "Avenula pubescens",
    "Hieracium alpina"             = "Hieracium Alpina",
    "Hieracium alpinum"            = "Hieracium Alpina",
    "Hieracium aurantiacum"        = "Pilosella aurantiaca",
    "Hieracium dovrense"           = "Hieracium Alpestria",
    "Hieracium hieracium"          = "Hieracium Hieracium",
    "Hieracium hieracioides"       = "Hieracium umbellatum",
    "Hieracium lactucella"         = "Pilosella lactucella",
    "Hieracium murorum"            = "Hieracium Vulgata",
    "Hieracium oreadea"            = "Hieracium Oreadea",
    "Hieracium prenanthoidea"      = "Hieracium Prenanthoidea",
    "Hieracium vulgata"            = "Hieracium Vulgata",
    "Hieracium pilosella"          = "Pilosella officinarum",
    "Hieracium vulgatum"           = "Hieracium umbellatum",
    "Hierochloã« alpina"           = "Hierochloë alpina",
    "Hierochloã« hirta"            = "Hierochloë hirta",
    "Hierochlo\x95 hirta"          = "Hierochloë hirta",
    "Hierochloã« odorata"          = "Hierochloë odorata",
    "Huperzia appressa"            = "Huperzia selago",
    "Hylotelephium maximum"        = "Hylotelephium telephium",
    "Lappula myosotis"             = "Lappula squarrosa",
    "Lepidotheca suaveolens"       = "Matricaria discoidea",
    "Listera cordata"              = "Neottia cordata",
    "Listera ovata"                = "Neottia ovata",
    "Leontodon autumnalis"         = "Scorzoneroides autumnalis",
    "Loiseleuria procumbens"       = "Kalmia procumbens",
    "Logfia arvensis"              = "Filago arvensis",
    "Mentha _verticillata"         = "Mentha verticillata",
    "Minuartia rubella"            = "Sabulina rubella",
    "Minuartia stricta"            = "Sabulina stricta",
    "Mycelis muralis"              = "Lactuca muralis",
    "Omalotheca supina"            = "Gnaphalium supinum",
    "Omalotheca norvegica"         = "Gnaphalium norvegicum",
    "Omalotheca sylvatica"         = "Gnaphalium sylvaticum",
    "Ononis arvensis"              = "Ononis spinosa",
    "Oreopteris limbosperma"       = "Thelypteris limbosperma",
    "Oxycoccus microcarpus"        = "Vaccinium microcarpum",
    "Oxycoccus palustris"          = "Vaccinium oxycoccos",
    "Phalaris minor"               = "Phalaris arundinacea",
    "Phalaroides arundinacea"      = "Phalaris arundinacea",
    "Pinus unicinata"              = "Pinus mugo",
    "Platanthera montana"          = "Platanthera chlorantha",
    "Poa alpigena"                 = "Poa pratensis",
    "Poa angustifolia"             = "Poa pratensis",
    "Poa laxa"                     = "Poa flexuosa",
    "Poa _herjedalica"             = "Poa herjedalica",
    "Poa _jemtlandica"             = "Poa alpina",
    "Poa lindebergii"              = "Poa arctica",
    "Potentilla anserina"          = "Argentina anserina",
    "Pyrola grandiflora"           = "Pyrola rotundifolia",
    "Rhamnus catharticus"          = "Rhamnus cathartica",
    "Rumex alpestris"              = "Rumex acetosa",
    "Salix _fragilis"              = "Salix fragilis",
    "Saxifraga _opdalensis"        = "Saxifraga opdalensis",
    "Sorbus hybrida"               = "Hedlundia hybrida",
    "Spergularia salina"           = "Spergularia marina",
    "Syringa emodi"                = "Syringa vulgaris",
    "Taraxacum crocea"             = "Taraxacum officinale",
    "Taraxacum croceum"            = "Taraxacum officinale",
    "Taraxacum erythrospermum"     = "Taraxacum officinale",
    "Taraxacum hamatum"            = "Taraxacum officinale",
    "Trientalis europaea"         = "Lysimachia europaea",
    "Trifolium pallidum"          = "Trifolium pratense",
    "Vicia orobus"                = "Vicia cassubica"
  ),
  species = str_replace(species, "Hierochlo.? hirta", "Hierochloë hirta"),
  species = str_replace_all(species, "spp.", "sp.")
    # and similar for other bad names
  ) |> 
    group_by(species, species_group, nin_id, Nature_Type, Sub_Type, nin_code) |> 
  summarise(cover = max(cover),           # summing cover of species which have ended up appearing several times per nature type
    .groups = "drop"
  )


nin_prepared_wfo <- NiN_sp |> 
  distinct(species)

# prepare dataset for WFO matching
nin_prepared_wfo <- WFO.prepare(nin_prepared_wfo$species)


# reconnect subspecies with corresponding species from authorship column
nin_prepared <- nin_prepared_wfo |>
  rename(clean_string = spec.name) |> 
  tibble() |> 
  mutate(
    clean_string = clean_string |>
      str_replace_all("\\bsect\\b\\.?", " ") |>
      str_squish()
  ) |> 
  # unique values in spec.full and clean_string only
  distinct(spec.full, clean_string)



# standardise names to the WFO backbone
nin_sp_matched <- WFO.match(spec.data = nin_prepared$clean_string,
                            WFO.data = wfo_backbone,
                            Fuzzy = 0.15,
                            Fuzzy.max = 50,
                            Fuzzy.one = FALSE)




# finalise accepted name column according to latest taxonomical nomenclature
nin_sp_clean <- nin_sp_matched |>
  # make copy of the original species string
  rename(clean_string = spec.name.ORIG) |>
  group_by(clean_string) |>
  summarise(
    # 1. Flag multiple scientificName suggestions
    flag_multiple_suggestions = n_distinct(scientificName) > 1,
    
    # 2. Candidate accepted_name from Old.name when possible
    accepted_name = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ 
        # take one Old.name where New.accepted == TRUE and Old.name non-empty
        Old.name[New.accepted == TRUE & Old.name != ""][1],
      TRUE ~ 
        # otherwise fall back to (one) scientificName
        scientificName[1]
    ),
    
    # 3. Where did accepted_name come from?
    accepted_from = case_when(
      any(New.accepted == TRUE & Old.name != "") ~ "Old.name",
      TRUE ~ "scientificName"
    ),
    .groups = "drop"
  )


# check the species that change name where many options were available
nin_sp_clean |> filter(clean_string != accepted_name) #|> view()
nin_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()



# correct incorrect corrections. haha
nin_sp_clean <- nin_sp_clean |> 
  mutate(
    flag_species_revert =
      case_when(
        grepl("Hieracium", clean_string) ~ "edited",
        TRUE ~ ""
        
      ),
    accepted_name = case_when(
      grepl("Hieracium", clean_string) ~ clean_string,
      TRUE ~ accepted_name
    ))

# check name changes
nin_sp_clean |> filter(flag_multiple_suggestions == TRUE, clean_string != accepted_name) #|> view()



# bind new species names onto indicator dataset
nin_species_clean <- left_join(nin_prepared, nin_sp_clean, by = "clean_string") |> 
  full_join(NiN_sp, by = join_by(spec.full == species)) |> 
  # filter out sect. species and subspecies
  #filter(!grepl("subsp.", scientific_name_original)) |> 
  distinct()

nin_species_clean |> filter(is.na(accepted_name))




# Merge with indicator values
nin_sp_ind <- nin_species_clean |>
  select(Nature_Type, Sub_Type, nin_id, nin_code, species = accepted_name, species_group, cover) |> 
  left_join(tyler_indicators |> 
              select(species, Moisture),
            by = join_by(species)) |> 
  tibble()


# double check species matching
nin_sp_ind |> filter(is.na(species))  
nin_sp_ind |> filter(is.na(Moisture)) |>  distinct(species)

nin_sp_ind



# merge nin datasets
nin_sp_ind <- nin_sp_ind |> 
  full_join(natop_nin_sp_ind |> rename(nin_code = nin_id))
  


# create cover scale data frame
coverscale <- data.frame(
  orig = 0:6,
  cov  = c(0, 1/32, 1/8, 3/8, 0.6, 4/5, 1)
)


# prepare for bootstrapping

abun_wide <- nin_sp_ind |>
  filter(cover != 0) |> 
  mutate(cover = coverscale$cov[match(cover, coverscale$orig)]) |> 
  select(
    sp = species,
    nin_code,
    cover
  ) |> 
  pivot_wider(
    names_from  = nin_code,
    values_from = cover,
    values_fill = 0
  )

# Create wide indicator matrix
ind_wide <- nin_sp_ind  |> 
  select(
    sp = species,
    Moisture
    )  |> 
  group_by(sp) |> 
  summarise(
    Moisture = first(Moisture),
    .groups  = "drop"
  )


# Extract sp, abun, ind for indBoot.freq()
sp    <- abun_wide$sp
abun  <- abun_wide[, -1]  # drop 'sp' column; remaining columns are NiN types
ind   <- ind_wide[, -1]   # drop 'sp'; remaining columns are indicators



rm(wfo_backbone)

```

This leaves us with the monitoring data including plant indicators (ano_lowlands) and the reference data including plant indicators (nin_sp_ind):
  
  <!-- Print head of data set with horizontal scrolling -->
  ```{r ano-sp-tab, echo=F}
head(ANO.sp.ind) |>
  kable("html",caption = "ANO species occurrence data set with attached plant trait data.") |> 
  kable_styling("striped") |> scroll_box(width = "100%")
```

```{r wetland_mountain_forest_seminatural-ref-data-tab, echo=FALSE}
head(nin_sp_ind) |> 
  kable("html", caption = "Reference data set for semi-natural ecosystems") |> 
  kable_styling("striped") |> scroll_box(width = "100%")

```

<br />
  
  For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
In order to get distributions of these metrics rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

<br />
  
  ##### Bootstrap functions for frequency abundance and cumulative density
  - function to calculate community weighted means (indBoot.freq() ) or cumulative density (indBoot.HeatOverhang() ) of selected indicator values (ind)
- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
- with a given number of iterations (iter),
- with species given a certain minimum abundance occurring in all bootstraps (obl), and
- with a given re-sampling ratio of the original species list (rat)
- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)

```{r bootstrapping}

indBoot.freq <- function(sp, abun, ind, iter, obl, rat = 2/3,var.abun = F) {
  
  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
  colnames(ind.b) <- colnames(abun)
  ind.b <- as.data.frame(ind.b)  
  
  ind <- as.data.frame(ind)
  ind.list <- as.list(1:length(colnames(ind)))
  names(ind.list) <- colnames(ind)
  
  for (k in 1:length(colnames(ind)) ) {
    ind.list[[k]] <- ind.b }
  
  for (j in 1:length(colnames(abun)) ) {
    
    dat <- cbind(sp,abun[,j],ind)
    dat <- dat[dat[,2]>0,]            # only species that are present in the ecosystem
    dat <- dat[!is.na(dat[,3]),]      # only species that have indicator values
    
    for (i in 1:iter) {
      
      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
      dat.b <- rbind(dat[dat[,2] >= obl,],
                     dat[match(speciesSample,dat$sp),]
      )
      
      if (var.abun==T) {
        for (m in 1:nrow(coverscale[-1,]) ) {
          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
        }
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
      }
      
      for (k in 1:length(colnames(ind))) {
        
        if ( nrow(dat.b)>2 ) {
          
          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
          ind.list[[k]][i,j] <- ind.b
          
        } else {ind.list[[k]][i,j] <- NA}
        
      }
      
      #      print(paste(i,"",j)) 
    }
    
  }
  return(ind.list)
}
```

```{r}
#colnames(NiN.forest.cov)
# 1st column is the species
# 2nd-27th column is the abundances of sp in different ecosystem types
# 28th-29th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 0.8 or more (dominant species) must be included in each sample
# each sample re-samples 1/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
```

Running the bootstraps:
```{r boot2, eval = F}


lowland_ref_cov <- indBoot.freq(
  sp      = sp,
  abun    = abun,
  ind     = ind,
  iter    = 1000,
  obl     = 0.8,
  rat     = 1/2,
  var.abun = FALSE
)



# fixing NaNs
for (i in 1:length(lowland_ref_cov) ) {
  for (j in 1:ncol(lowland_ref_cov[[i]]) ) {
    v <- lowland_ref_cov[[i]][,j]
    v[is.nan(v)] <- NA
    lowland_ref_cov[[i]][,j] <- v
  }
}



#saveRDS(lowland_ref_cov, paste0(here::here(),"/data/cache/lowland_ref_cov.RDS"))


```

```{r, include = F}
# Data from cache
lowland_ref_cov <- readRDS(paste0(here::here(), "/data/cache/lowland_ref_cov.RDS"))
```

```{r}
head(lowland_ref_cov[[1]]) |>
  kable("html", caption = "Table showing the first 6 rows of the bootstrapped data set.") |> kable_styling("striped") |> scroll_box(width = "100%")
```

This results in an R-list for forest ecosystem, with a data frame for every selected plant indicator and each data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.

Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:
  
  - Median = reference values
- 0.025 and 0.975 quantiles = lower and upper limit values
- min and max of the respective indicators scale = min/max values



```{r scalingValues, attr.output='style="max-height: 300px;"', results='hide'}

# every NiN-type is represented by one 'generalisert artsliste'
# some NiN-types are represented by two such species lists
# in some cases two NiN-types are represented by the same species list

# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types), i.e. excluding types with a- and b-suffix
names(lowland_ref_cov[["Moisture"]])


# 2 indicator-value indicators: Moisture, "Moisture"
indEll.n=2
# creating a table to hold:
# for every indicator: the 0.5 quantile (median), 0.025 quantile and  0.975 quantile for each NiN-type (3 metrics -> 3 columns per indicator), for Heat_requirement in alpine/arctic systems we do 0.05 quantile and  0.95 quantile and ignore the former as it is a one-sided indicator
# for every nature type (nrows)
# coercing the values into the table

myQuantiles <- c(0.025, 0.5, 0.975)

# 1) Moisture bootstrap matrix
moist_mat  <- lowland_ref_cov[["Moisture"]]
all_names  <- colnames(moist_mat)

# 2) Strip a/b/c/d suffixes from column names to get base NiN IDs
base_names <- all_names %>%
  str_remove("[abcd]$")

# 3) Map: each column index -> base NiN ID
nin_map <- tibble(
  col_index = seq_along(all_names),
  col_name  = all_names,
  base_nin  = base_names
)

# 4) Count how many columns per base NiN
nin_counts <- nin_map %>%
  count(base_nin, name = "n_cols")

# Singles: NiN types represented by exactly one column
nin_single <- nin_counts %>%
  filter(n_cols == 1) %>%
  pull(base_nin)

# Multis: NiN types represented by multiple columns
nin_multi <- nin_counts %>%
  filter(n_cols > 1) %>%
  pull(base_nin)

# 5) Quantiles for single NiN types (one column each)
tab_single <- map_dfr(nin_single, function(base_id) {
  # find the column index that has this base NiN
  idx <- nin_map %>%
    filter(base_nin == base_id) %>%
    pull(col_index)
  
  vals <- as.matrix(moist_mat[, idx])
  q    <- quantile(vals, probs = myQuantiles, na.rm = TRUE)
  
  tibble(
    NiN              = base_id,
    Moisture_025     = q[1],
    Moisture_50      = q[2],
    Moisture_975     = q[3],
    Moisture_025_dup = q[1],
    Moisture_50_dup  = q[2],
    Moisture_975_dup = q[3]
  )
})

# 6) Quantiles for multi NiN types (several columns per type)
tab_multi <- map_dfr(nin_multi, function(base_id) {
  # find all columns whose stripped base_nin equals this base_id
  idx <- nin_map %>%
    filter(base_nin == base_id) %>%
    pull(col_index)
  
  vals <- as.matrix(moist_mat[, idx])
  q    <- quantile(vals, probs = myQuantiles, na.rm = TRUE)
  
  tibble(
    NiN              = base_id,
    Moisture_025     = q[1],
    Moisture_50      = q[2],
    Moisture_975     = q[3],
    Moisture_025_dup = q[1],
    Moisture_50_dup  = q[2],
    Moisture_975_dup = q[3]
  )
})

# 7) Combine singles and multis into final table
tab <- bind_rows(tab_single, tab_multi)

tab

# take first nin type where multiple are listed
tab <- tab |> mutate(NiN = str_replace(NiN, "(.*?C[^C]*?)C.*", "\\1"))


# making it a proper data frame
dim(tab)
round(tab[,2:7],digits=2)

summary(tab)
tab <- tab |> 
  mutate(NiN = if_else(
  !NiN %in% c("T2-C-7", "T2-C-8"), 
  gsub("C", "C-", NiN), 
  NiN
))
  
tab


# creating final objects holding the reference and limit values for all indicators

# ref object for indicators
lowland_ref_cov_val <- data.frame(grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
                                                    county=rep('all',(nrow(tab)*2*indEll.n)),
                                                    region=rep('all',(nrow(tab)*2*indEll.n)),
                                                    Ind=c(rep(c('Moist1','Moist2'),nrow(tab)),
                                                          rep(c('Moisture1','Moisture2'),nrow(tab))
                                                          
                                                    ),
                                                    Rv=c(rep(tab$Moisture_50,each=2),
                                                         rep(tab$Moisture_50_dup,each=2)
                                                    ),
                                                    Gv=c(y.Moist,y.Moisture),
                                                    maxmin=c(rep(c(1,12),nrow(tab))  # 9 levels of Moisture
                                                    )
)

lowland_ref_cov_val <- lowland_ref_cov_val |> 
  tibble() |> 
  mutate(grunn = as.factor(grunn),
         Ind = as.factor(Ind))

summary(lowland_ref_cov_val)

#saveRDS(lowland_ref_cov_val, paste0(here::here(),"/data/cache/lowland_ref_cov_val.RDS"))


```

```{r}
lowland_ref_cov_val |>
  kable("html", caption = "Reference values (Rv and threshold values (Gv) for each indicator and nature type combination") |> kable_styling("striped") |> scroll_box(width = "100%", height = "300px")
```


```{r merging datasets}
# ANO
ano_lowlands <- ano_lowlands |>
  select(globalid, ano_flate_id, ano_punkt_id, ssb_id, aar, hovedoekosystem_250m2, hovedoekosystem_rute, hovedtype_rute, hovedtype_250m2, hovedtype_1m2, kartleggingsenhet_1m2, kartleggingsenhet_250m2) |> 
  mutate(
    X = st_coordinates(ano_lowlands)[, 1],
    Y = st_coordinates(ano_lowlands)[, 2],
    kartleggingsenhet_1m2 = word(kartleggingsenhet_1m2, 1, 1)
  ) |>
  st_drop_geometry() |> 
  mutate(globalid_chr = as.character(globalid)) %>%
  semi_join(
    ANO_species_ind %>% mutate(parent_chr = as.character(parentglobalid)),
    by = c("globalid_chr" = "parent_chr")
  ) |> 
  filter(hovedtype_rute %in% c("T32", "T34", "T45", "V10", "T41")) |> 
    rename(kartleggingsenhet = kartleggingsenhet_1m2) |> 
  mutate(scale = "1m2")


# GRUK
gruk_lowlands <- GRUK_all |>
  rename(globalid = parent_global_id) |> 
  mutate(globalid_chr = as.character(globalid)) |> 
  select(-c(species, art_dekning, Moisture, polygon_id, rute_id, precision, UTM33_E_ne, UTM33_N_ne, UTM33_E_sw, UTM33_N_sw)) |> 
  distinct() |> 
  rename(
    X = UTM33_E,
    Y = UTM33_N, 
    aar = year) |> 
st_drop_geometry() |> 
  semi_join(
    GRUK_species_ind  |>  mutate(parent_chr = as.character(parent_global_id)) |>  st_drop_geometry(),
    by = c("globalid_chr" = "parent_chr")
  ) |> 
  mutate(hovedtype_rute = str_sub(kartleggingsenhet_1_5000, 1, 2)) |> 
  select(globalid, globalid_chr, hovedtype_rute, aar, X, Y, kartleggingsenhet_1_5000, nin_kartleggingsenheter, hovedokosystem) |> 
  rename(kartleggingsenhet = kartleggingsenhet_1_5000) |> 
  mutate(scale = "1_5000")

# ASO

ASO_lowlands <- ASO_all |>
  rename(globalid = ParentGlobalID) |> 
  mutate(globalid_chr = as.character(globalid)) |> 
  select(-c(species, art_dekning, Moisture)) |> 
  distinct() |> 
  st_as_sf(crs = "+proj=longlat +datum=WGS84 +ellps=WGS84") |>
  st_transform(crs = 25833) |> 
mutate(
  X = st_coordinates(st_sfc(geometry))[, 1],
  Y = st_coordinates(st_sfc(geometry))[, 2]) |> 
  st_drop_geometry() |> 
semi_join(
    ASO_species_ind  |>  mutate(parent_chr = as.character(ParentGlobalID)) |>  st_drop_geometry(),
    by = c("globalid_chr" = "parent_chr")
  ) |> 
  mutate(hovedtype_rute = str_sub(nin_grunntype, 1, 3)) |> 
  select(globalid, globalid_chr, aso_id, hovedtype_rute, X, Y, nin_grunntype) |> 
  rename(kartleggingsenhet = nin_grunntype)


combined_lowlands <- bind_rows(ano_lowlands, gruk_lowlands, ASO_lowlands, .id = "dataset") |> 
  mutate(dataset = case_match(
    dataset, 
    "1" ~ "ANO",
    "2" ~ "GRUK",
    "3" ~ "ASO"
  )) |> 
  # harmonise projections
  st_as_sf(coords = c("X", "Y"), remove = FALSE, crs = 25833) |> 
  # join to Norway regions
  st_join(
    regnor,
  left = TRUE,
  join = st_nearest_feature
) |> 
  st_drop_geometry()



# combine species data

GRUK_species_ind <- GRUK_species_ind |> 
  filter(!is.na(species), !is.na(art_dekning), !is.na(Moisture))

ASO_species_ind <- ASO_species_ind |> 
  filter(!is.na(species), !is.na(art_dekning), !is.na(Moisture))


combined_species <-  bind_rows(ANO_species_ind |> select(-c(busker_dekning, tresjikt_dekning, roesslyng_dekning, ssb_id, vedplanter_total_dekning, kartleggingsenhet_1m2)), 
                               GRUK_species_ind |> rename(parentglobalid = parent_global_id),
                               ASO_species_ind |> rename(parentglobalid = ParentGlobalID                       ),
                               .id = "dataset") |> 
  mutate(dataset = case_match(
    dataset, 
    "1" ~ "ANO",
    "2" ~ "GRUK",
    "3" ~ "ASO"
  ))
```

Once test data and the scaling values from the reference data are in place, we can calculate CWMs of the selected indicator for the ANO community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type. An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

For scaling ANO-data against the reference we use the normalisation function from package ecTools.

```{r}
#| warning: false

#### 1. calculating scaled and non-truncated values for the indicators based on the datasets ####

ref_grunn <- unique(sub("\\-.*", "", as.character(lowland_ref_cov_val$grunn)))

for (i in seq_len(nrow(combined_lowlands))) {
  tryCatch({
    
    ds       <- combined_lowlands$dataset[i]         # "ANO" or "GRUK"
    hoved_i  <- as.character(combined_lowlands$hovedtype_rute[i])
    global_i <- as.character(combined_lowlands$globalid_chr[i])
    nin_i    <- as.character(combined_lowlands$kartleggingsenhet[i])
    
    # 1) hovedtype must exist in ref
    cond_hoved <- !is.na(hoved_i) && hoved_i %in% ref_grunn
    if (!isTRUE(cond_hoved)) next
    
    # 2) species data for this row and dataset
    dat <- combined_species |>
      dplyr::filter(
        dataset == ds,
        parentglobalid == global_i
      ) |>
      dplyr::select(art_dekning, Moisture)
    
    if (nrow(dat) == 0) next
    
    combined_lowlands$richness[i] <- nrow(dat)
    dat <- dat[!is.na(dat$Moisture), ]
    if (nrow(dat) == 0) next
    
    # 3) CWM Moisture
    val <- sum(dat$art_dekning * dat$Moisture, na.rm = TRUE) /
      sum(dat$art_dekning, na.rm = TRUE)
    
    ## Moisture1
    idx1 <- lowland_ref_cov_val$Ind == "Moisture1" &
      lowland_ref_cov_val$grunn == nin_i
    idx1[is.na(idx1)] <- FALSE
    
    if (sum(idx1) == 1) {
      ref1    <- as.numeric(lowland_ref_cov_val$Rv[idx1][[1]])
      lim1    <- as.numeric(lowland_ref_cov_val$Gv[idx1][[1]])
      maxmin1 <- as.numeric(lowland_ref_cov_val$maxmin[idx1][[1]])
      
      if (!any(is.na(c(ref1, lim1, maxmin1)))) {
        combined_lowlands$Moisture1[i] <- ec_normalise(
          variable = val,
          x0   = maxmin1,
          x60  = lim1,
          x100 = ref1,
          fun  = "linear"
        )
      }
    }
    
    ## Moisture2
    idx2 <- lowland_ref_cov_val$Ind == "Moisture2" &
      lowland_ref_cov_val$grunn == nin_i
    idx2[is.na(idx2)] <- FALSE
    
    if (sum(idx2) == 1) {
      ref2    <- as.numeric(lowland_ref_cov_val$Rv[idx2][[1]])
      lim2    <- as.numeric(lowland_ref_cov_val$Gv[idx2][[1]])
      maxmin2 <- as.numeric(lowland_ref_cov_val$maxmin[idx2][[1]])
      
      if (!any(is.na(c(ref2, lim2, maxmin2)))) {
        combined_lowlands$Moisture2[i] <- ec_normalise(
          variable = val,
          x0   = maxmin2,
          x60  = lim2,
          x100 = ref2,
          fun  = "linear"
        )
      }
    }
    
  }, error = function(e) {
    cat("ERROR at row", i, ":", conditionMessage(e), "\n")
  })
}


combined_lowlands <- combined_lowlands |>
  dplyr::mutate(
    moisture  = Moisture1 * Moisture2,
    Moisture1 = if_else(Moisture1 == 1, NA_real_, Moisture1),
    Moisture2 = if_else(Moisture2 == 1, NA_real_, Moisture2)
  )



```

```{r}
head(ano_lowlands) |>
  kable("html", caption = "ANO plots for ecosystems with scaled indicator values in the far-right columns.") |> 
  kable_styling("striped") |> scroll_box(width = "100%")
```


We can visualise the normalisation with a generic figure (@fig-norm).

```{r fig-norm}
#| fig-cap: 'A figure illustrating the normalisation process.The placement of the x60 values are arbitrary in this figure and does not give a representative picture of the relationship between the variable and indicator values.'

var <- seq(0, 100, by = 0.5)
var2 <- seq(101, 200, by = 0.5)

ind <- ec_normalise(var, x0 = 0, x100 = 100, x60 = 50)
ind2 <- ec_normalise(var2, x100 = 101, x0 = 200, x60 = 150)

ec_norm_plot(
  variable = c(var, var2),
  indicator = c(ind, ind2)
) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200),
    labels = c("lowest possible\nvalue", "2.5%", "50%", "97.5%", "highest possible\nvalue"),
    expand = expansion(add = 30)
  ) +
  labs(x = "Variable values as percentiles of \nbootstrapped reference values")
```

```{r}
#| eval: false
#| include: false

ec_norm_plot(
  variable = c(var, var2),
  indicator = c(ind, ind2)
) +
  scale_x_continuous(
    breaks = c(0, 50, 100, 150, 200),
    labels = c("lowest possible\nvalue", "2.5%", "50%", "97.5%", "highest possible\nvalue"),
    expand = expansion(add = 30)
  ) +
  labs(x = "Variabelverdier som persentiler ift. \nbootstrappede referanse verdier",
       y = "Indikatorverdi")

ggsave(here::here("img/norm_plot.png"))
```

#### Calculating national and regional indices

We can calculate national and region-wise indicator values using a mixed-effects, zero-one inflated beta-regression analysis. Since the indicator values are on a scale from 0-1 the data per definition dont follow a Gaussian distribution, hence the zero-one inflated beta-regression. Since the underlying monitoring data are spatially aggregated (several points per site), not all points are independent, hence the mixed-effects model.

The uncertainty around the national and region indicator values is derived from the same statistical model.

The national and regional estimates are computed separately.

```{r tmbs}
#| warning: false
## Fit ordered beta regression null-model with a random effect
# Norway

Moisture_no_ano <- glmmTMB(
  moisture ~ 1 + (1|ano_flate_id),
  data = combined_lowlands |> filter(dataset == "ANO"),
  family = ordbeta
)

Moisture1_no_ano <- glmmTMB(
  Moisture1 ~ 1 + (1|ano_flate_id),
  data = combined_lowlands |> filter(dataset == "ANO", !is.na(Moisture1)),
  family = ordbeta
)

Moisture2_no_ano <- glmmTMB(
  Moisture2 ~ 1 + (1|ano_flate_id),
  data = combined_lowlands |> filter(dataset == "ANO", !is.na(Moisture2)),
  family = ordbeta
)

summary(Moisture_no_ano)
summary(Moisture1_no_ano)
summary(Moisture2_no_ano)

# regions
Moisture_re_ano <- glmmTMB(
  moisture ~ 0 + region + (1|ano_flate_id),
  data = combined_lowlands |> filter(dataset == "ANO"),
  family = ordbeta
)

Moisture1_re_ano <- glmmTMB(
  Moisture1 ~ 0 + region + (1|ano_flate_id),
  data = combined_lowlands[!is.na(combined_lowlands$Moisture1),],
  family = ordbeta
)

Moisture2_re_ano <- glmmTMB(
  Moisture2 ~ 0 + region + (1|ano_flate_id),
  data = combined_lowlands[!is.na(combined_lowlands$Moisture2),],
  family = ordbeta
)

summary(Moisture_re_ano)
summary(Moisture1_re_ano)
summary(Moisture2_re_ano)
```


## 10. Results

The model estimates and standard errors are on a logit scale, but we can calculate indicator values and confidence intervals (CI) by transforming back from logit to values between 0-1
```{r}

NO_FUMO_005_table <- tibble(
  Region = c(
    "Norway",
    "Central Norway",
    "Eastern Norway",
    "Northern Norway",
    "Southern Norway",
    "Western Norway"
  ),
  indicator_value = c(
    plogis(summary(Moisture_no_ano)$coef$cond[, 1]),
    plogis(summary(Moisture_re_ano)$coef$cond[, 1])
  ),
  CI_upper = c(
    plogis(
      summary(Moisture_no_ano)$coef$cond[, 1] +
        1.96 * summary(Moisture_no_ano)$coef$cond[, 2]
    ),
    plogis(
      summary(Moisture_re_ano)$coef$cond[, 1] +
        1.96 * summary(Moisture_re_ano)$coef$cond[, 2]
    )
  ),
  CI_lower = as.numeric(
    plogis(
      summary(Moisture_no_ano)$coef$cond[, 1] -
        1.96 * summary(Moisture_no_ano)$coef$cond[, 2]
    ),
    plogis(
      summary(Moisture_re_ano)$coef$cond[, 1] -
        1.96 * summary(Moisture_re_ano)$coef$cond[, 2]
    )
  ),
  upper_indicator_value = c(
    plogis(summary(Moisture2_no_ano)$coef$cond[, 1]),
    plogis(summary(Moisture2_re_ano)$coef$cond[, 1])
  ),
  n_upper_indicator_value = c(
    summary(Moisture2_no_ano)$nobs,
    as.vector(table(Moisture2_re_ano$frame$region))
  ),
  lower_indicator_value = c(
    plogis(summary(Moisture1_no_ano)$coef$cond[, 1]),
    plogis(summary(Moisture1_re_ano)$coef$cond[, 1])
  ),
  n_lower_indicator_value = c(
    summary(Moisture1_no_ano)$nobs,
    as.vector(table(Moisture1_re_ano$frame$region))
  )
)
NO_FUMO_005_table
```

To express the results as a distribution, we need to simulate mean estimates based on the models intercept and associated standard error.
```{r}

## Norway
# Fixed-effect intercept and SE
b0 <- fixef(Moisture_no_ano)$cond["(Intercept)"]
se_b0 <- sqrt(vcov(Moisture_no_ano)$cond["(Intercept)", "(Intercept)"])
# Dispersion parameter (For glmmTMB this is usually on the response/positive scale via sigma() )
phi <- sigma(Moisture_no_ano)


# draw estimates and backtransform from logit-scale
eta_draws <- rnorm(1000, mean = b0, sd = se_b0)
mu_draws  <- plogis(eta_draws)


## regions
b_re <- fixef(Moisture_re_ano)$cond
vc_re <- vcov(Moisture_re_ano)$cond
phi <- sigma(Moisture_re_ano)

eta_draws_reg <- MASS::mvrnorm(
  n = 1000,
  mu = b_re,
  Sigma = vc_re
)

mu_draws_reg <- plogis(eta_draws_reg)


# combining Moisture_no_sim and Moisture_re_sim
mu <- mu_draws_reg |>
  as_tibble() |>
  add_column(Norway = mu_draws) |>
  pivot_longer(everything(), names_to = "part") |>
  mutate(part = case_when(
    part == "regionCentral_Norway" ~ "C",
    part == "regionEastern_Norway" ~ "E",
    part == "regionNorthern_Norway" ~ "N",
    part == "regionWestern_Norway" ~ "W",
    part == "regionSouthern_Norway" ~ "S",
    .default = part
  )) #|>
  #add_column(year = "2024")
```


```{r tbl-results}
#| tbl-cap: 'Indicator values and 95% confidence intervals for NO_FUMO_005, calculated for Norwegian open lowland ecosystems.'
mu |>
  group_by(part) |>
  summarise(
    indicator = mean(value),
    low = round(quantile(value, probs = 0.025), 2),
    high = round(quantile(value, probs = .975), 2)
  ) |>
  unite(col = "95% CI", c(low, high), sep = " - ") |>
  gt::gt()
```

## 11. Export file

<!--# 
  
  Optional: Display the code (dont execute it) or the workflow for exporting the indicator values to file. Ideally the indicator values are exported as a georeferenced shape or raster file with indicators values, reference values and errors. You can also chose to export the raw (un-normalised or unscaled variable) as a separate product. You should not save large sptaial output data on GitHub. You can use eval=FALSE to avoid code from being executed (example below - delete if not relevant) 

-->

```{r export}
#| eval: false
arrow::write_parquet(mu,"data/results_NO_FUMO_004.parquet")
```

Demonstrating how to read back the data:

```{r readingBack}
arrow::read_parquet(here::here("data/results_NO_FUMO_004.parquet"))
```

## 12. Author contributions



**CRediT statement**

* Conceptualization: [JT]
* Methodology: [JT]
* Data curation: [JT, ALK]
* Formal analysis: [JT]
* Investigation: [JT]




**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [JT].

**Data processing**
Raw data were curated and cleaned by [JT and AK].

**Statistical analysis**
All analyses were conducted by [JT] , with verification and minor edits by [AK].

**Visualization**
Figures were produced by [AK].

**Reproducibility**
All code was reviewed and validated by [AK].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This workflow uses renv. Run `renv::restore()` to get started.

:::