---
title: "Enter species or nature type name here"
subtitle: "[INDICATORID]" 
indicatorID: INDICATORID
indicatorName: NAME OF SPECIES OR NATUR TYPE # Short name for the species or nature type. Add more by adding another line starting '-'
verbatimName: 
  - VERBATIM_INDICATOR_NAME # the indicator/species/natur type name that us used in the native language
folderName: XX_YYYY_ZZZ # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/'). 
# If it's a single indicator being reported, the folder name should be the same as the indicatorID. 
# If there are multiple ID's then the folder name could look something like XX_YYY_ZZZ-ÅÅÅ.
country: Norway
continent: Europe
Realm: # Please copy and paste Realms, Biomes and Ecostsem names from here: https://github.com/NINAnor/ecRxiv/wiki#ecosystem-typology 
  - write realm here # e.g. Freshwater (F)
Biome: 
  - write biome here # e.g. F1 Rivers and streams biome
Ecosystem: 
  - write ecosystem here # e.g. F1.3 Freeze-thaw streams
verbatimEcosystem: # add name of the ecosystem or naturetype according to any relevant local or national typology. E.g. fjell, skog, ... 
yearAdded: 2026 # the year when the indicator was first submitted to ecRxiv
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: Comment here # Describe briefly how this version differs from earlier versions 
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/XX_YYYY_ZZZ
# To be filled out by ecRxiv during the typesetting, when they will replace XX_YYYY_ZZ with the name of the root folder for the current indicator(s). It should be the same as the indicatorID, but if there are more indicator IDs presented in the same folder, the folder name might be different, e.g. XX_YYYY_AAA-BBB 
Precision: # Optional. Describe the precision metric you have used. Precision may be reported either as variation among observed values (e.g. standard deviation, variance, range, interquartile range) or as uncertainty around an estimated value (e.g. standard error, confidence interval, credible interval). 
writtenLanguage: Norwegian
format: 
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute: 
  cache: true
author:
  - name: John Doe              # Enter name
    email: john.doe@example.no  # Enter email
    affiliations:
      - id: myID
        name: A Fictive Institute for Nature Research # Enter affiliations
  - name: First Last              #  Enter subsequent authors like this, or remove if not relevant
    affiliations:
      - ref: myID               # To reuse affiliations reference the id like this
AuthorList: Doe, J. and Last, F.  
date: January 1, 1901 # Enter date
data_availability: none
code_reproducibility: none
open_science_badge: none
callout-icon: false
lightbox: true
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: TRUE # Change to FALSE
bibliography: references.bib
codeReviewers:
  - name: Jane Smith
    email: jane.smith@example.no
    affiliations:
      - id: reviewerInst
        name: Example Institute
    reviewType: code review
    github: janesmith
  - name: John Doe
    affiliations:
      - ref: myID
    reviewType: reproducibility check
    github: johndoe
---

<!--# This is a template for how to document analyses in a specific Norwegian monitoring project. Try to keep to the headers in the template. This is because it easier to read if all the similar ecRxiv submissions are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting, it is recommended to fill in as much as you can in the top yaml part of the file, where all the metadata goes. This will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
library(DT)
library(readxl)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")
  
  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
# The following inserts a badge specifying the GPL GNU version 3 license. ecRxiv recommends using this license, but you can decide to use another. Contact ecRxiv for help in insert a badge for a different license
license_badge()
```


::: 
 

> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID: ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added", 
    "Last update",
    "Version",
    "Version comment",
    "Precision")) |>   
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 01 Jan. 1901 - Original PR
:::


<hr />

<!--# Document you work below.  -->

## Oppsummering
<!-- <standardisert oppsummering, lik den som leveres i hovedrapport> -->

Skriv tekst her, og så kan du henvise til figurer slik (@fig-innledning), og til referanser slik [@nater_ni2025_2025].

## English summary


## 1. Innledning
<!-- Om arten/naturtypen: Organismegruppe, Status rødlista, Utbredelse, Trusler. Kort om dette delprosjektet (hvilke tiltak ser det på effekten av, hvor. -->

```{r fig-innledning}
#| fig-cap: 'Figurtekst skrives her.'
#| out-width: '70%'
#| code-summary: 'Kode for å vise bilde.'
# pictures are added to the img folder
pth <- here::here("img/MY_PICTURE.png")
knitr::include_graphics(pth)
```


## 2. Formål med delprosjektet

### 2.1 Målsetning med tiltakene

<!-- Hva er målsetningen med tiltakene som gjennomføres.
Hvilke indikatorer skal tiltakene endre. -->


### 2.2 Formål med overvåking

<!-- Hva skal overvåkingen avdekke. -->

### 2.3 Forventninger

<!-- Hva forventer vi å observere av endringer i indikatorene – på kort og lang sikt. -->


## 3. Overvåkingslokalitetene

### 3.1 Beskrivelse av lokalitetene

<!-- Hvor, kart, plassering av lokalitetene i Norge.
Status og utfordringer/påvirkninger. -->


### 3.2 Tiltak

<!-- Beskrivelse av tiltak, hva, når, hvem utfører. Legg inn bilde. -->


## 4. Datainnsamling

<!-- Beskrivelse av datainnsamling. Avgrensning av lokalitet, utlegging av utvalgsenheter, antall, metodikk for innsamling av data fra utvalgsenheter. Fyll inn tabellen dummyData_effektovervaaking.xlsx. Gi den gjerne ett nytt navn, og bruk koden under for å vise tabellen.-->



Referer til tabellen under (@tbl-datainnsamling).

```{r tbl-datainnsamling}
#| tbl-cap: 'Skriv tabelltekst her.'
#| code-summary: 'Kode for å vise tabell.'

pth2 <- here::here("data/dummyTable_effektovervaaking.xlsx")
tbl <- read_excel(pth2)
DT::datatable(tbl)

```

## 5. Dataanalyser

<!-- Kort beskrivelse: hvilke analyser er gjennomført, for å undersøke endringer i indikatorer ift. påvirkningsfaktorer og tiltak, jf. forventninger (under formål). -->

## 6. Resultater

<!-- Strukturen på denne vil variere, avhengig av art/naturtype. Det kan være lokalitet for lokalitet (eks. GRUK), eller indikator for indikator. Figurer, tabeller som i tidligere år, med oppdaterte resultater. -->


## 7. Oppsummering

### 7.1 Hva viser årets resultater

<!-- Tiltakenes effekt per nå, endringer siden i fjor, usikkerhet. -->


### 7.2 Anbefaling for videreføring

<!-- Hva er kunnskapsbehovet, om tiltaket (hvordan gjennomføres) og om effekten (variasjon mellom år, mellom lokaliteter.

Anbefalinger for videreføring av delprosjektet – endringer i metodikk? Utvidelse? Endret frekvens for datainnsamling? Avslutning? -->



## 8. References

::: {#refs}
:::


## 9. Author contributions

<!-- Fill in CReDIT and MeRIT statements to clarify the contributions from individual authors. For MeRIT, see https://www.nature.com/articles/s41467-023-37039-1. For CRedit, see https://credit.niso.org/. You can start with the template structure below, and just delete the rows you dont use.  -->

**CRediT statement**

* Conceptualization: [Initials]
* Methodology: [Initials]
* Data curation: [Initials]
* Formal analysis: [Initials]
* Investigation: [Initials]
* Software: [Initials]
* Validation: [Initials]
* Visualization: [Initials]
* Resources: [Initials]
* Writing – original draft: [Initials]
* Writing – review & editing: [Initials]
* Supervision: [Initials]
* Project administration: [Initials]
* Funding acquisition: [Initials]


**MeRIT attribution**

This part supplements the CRediT statement with more granularity for method section.

**Study design**
The study was designed by [AB, CD].

**Data collection / fieldwork**
Field data were collected by [AB, EF] between [years]. Sampling protocol was developed by [CD].

**Laboratory procedures**
DNA extraction was performed by [GH]. Sequencing was conducted by [IJ].

**Data processing**
Raw data were curated and cleaned by [EF]. Code for preprocessing was written by [KL].

**Statistical analysis**
All analyses were conducted by [CD] using [software], with verification by [AB].

**Visualization**
Figures were produced by [KL], with input from [CD].

**Reproducibility**
All code was reviewed and validated by [AB, CD].



::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.  

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
